### 2023-05-08 scripts/Barneby_Lives.Rmd
init

---
title: "Barneby Lives!"
author: "steppe"
date: "9/13/2021"
output: html_document
---
The purpose of this script is to assist Botanical Collectors in accessioning their collections.

It is our lived experience that many herbaria now request digital copies of collection information in there own format. This script aims to alleviate the investment in time this requires, and in doing so allow the collector more time to do less onerous tasks; such as collection and study.

While in theory all of the tasks in this script could be automated, doing so would be an unconscionable action against local taxonomic authorities. While several database options for updating collections (and spell checking nomen) which are current and well curated (e.g. Kew Plants of the World) exist, we feel many proposals and combinations require interpretation by those in the regions attempting to understand the material at hand.

This script will perform the following tasks:

**Geographic:**

**1)** Convert coordinates to a Decimal Degrees format, if needed (e.g. from DMS 72*17.55 -> 72.29861). 
**2)** Retrieve appropriate political place names. (i.e. State, County).
**3)** Retrieve accurate elevation values of coordinates.
**4)** Retrieve Public Land Ownership (e.g. United States Forest Service, Bureau of Land Management).
**5)** Project coordinates from one system to another, if needed (e.g. from an UTM zone to WGS84).
**6)** Retrieve Cadastral data from the PLSS for Township, Range, Section (TRS)

**Temporal:**

**1)** Provide a 'written' date in the style of choosing. (e.g. May 12th 2021 OR 12th May 2021), for the purpose of labelling.

**Taxonomic:**

**1)** Provide a 'spell check' for family. 
**2)** Provide a 'spell check' for genus. 
**3)** Provide a 'spell check' for epithet.
**4)** Provide a 'spell check' for authorities. 
**5)** In addition you may check for synonymy at each of these taxonomic levels, and accept or reject suggestions.

**Directions:**

I preface this with the anecdote that a good deal of collectors and curators consider these data non-essential now. Regardless, if you are so inclined to gain access to a free google maps account these services are provided.

**1)** Write directions, from a town of notable size or repute, to the closet parking spot to the collection locality.

Note that these directions may not reflect the actual route traveled. *Caveat emptor.*

We would like to acknowledge the following developers of the libraries, and databases, which this script utilizes. As is everything under the sun, nothing here is new and none of this is novel, it is merely the regurgitated contents from the fingertips and intellect of others.

Of final note, this script is named for the late and great Rupert Charles Barneby, and by no small means is a dig at certain attributes of the fictional 'Hayduke'.

Files required to use this script:

- PAD-US database for retrieving land ownership information if needed (*note does not include identities of Private owners*). 
- World Flora Online static copy of the database.

In order to not bog people's computers down with opening and storing many large files we make use of downloading items into R repeatedly.

```{r Libraries, message=FALSE, warning=FALSE}
#install.packages("WorldFlora", "elevatr")
library(tidyverse)
library(sf)
#library(parzer)
library(tigris)
#library(WorldFlora)
library(elevatr)
```

Please load your data here. *If you want to create a copy of your herbarium ledger and move the copy to a folder containing this script that would be a path of quite low resistance.*

```{r}
path_f <- "L:/Pollen_Reference_Library/Herbarium_spreadsheets-ACTIVE"
files <- list.files(path_f)
files <- paste(path_f, sep = "/", files) 
df <- readxl::read_excel(files[5], sheet = 2) %>%  
  mutate(across(.cols = everything(),
                .fns = ~ str_replace(string = ., pattern =  '[°*]', ""))) %>% 
  mutate(collection_number = as.numeric(collection_number)) %>%
  mutate(tissue_col = as.numeric(tissue_col)) %>% 
  mutate(county = paste(county, "Co.", sep = " ")) %>% 
  mutate(across(starts_with('elevation'), function(x) as.numeric(as.character(x)))) %>% 
  mutate(across(c('latitude', 'longitude'), function(x) as.numeric(x)))

rm(path_f, files)
```

# Geographic

**Convert coordinates from Degrees, Minutes, Seconds to Decimal Degrees.**
```{r Convert coords from DMS to DD, eval = F}

df <- df %>% 
  mutate(Longitude.dd = parzer::parse(Longitude),
          Latitude.dd = parzer::parse(Latitude))

```

**Create a Spatial (Simple Features) Object**

```{r}
df <- st_as_sf(df, coords = c("longitude (raw)", "latitude"), remove = F, crs = 4269)
```

**Retrieve state of collections.**

```{r Retrieve State of Collection, message = F, wanrning = F}
states.sf <- tigris::states(cb = FALSE, resolution = "5m", class = "sf") %>% 
  dplyr::select(state = NAME)

df <- df %>% dplyr::select(-state, -county)
df <- st_intersection(df, states.sf)

rm(states.sf)
```

**Retrieve counties of collections**
```{r Retrieve county of collection}
states_vec <- df %>% ungroup() %>% distinct(state) %>% pull(state) %>% unique()

counties.sf <- tigris::counties(state = states_vec, cb = FALSE, resolution = "5m", class = "sf") %>% 
  dplyr::select(NAME) %>% rename(county = NAME)

df <- st_intersection(df, counties.sf)

rm(counties.sf, states_vec)
```

**Retrieve elevation of site**
```{r retrieve sites elevation}
df <- df %>% dplyr::select(-elevation_ft,  -elevation_m, -elevation_ft_final,)

prj_dd <- "EPSG:4269"

df_points <- df %>% dplyr::select('longitude..raw.', 'latitude')
df_elev_epqs <- get_elev_point(df_points, prj = prj_dd, src = "epqs")
df_elev_epqs <- df_elev_epqs %>% 
  rename(elevation_m = elevation) %>% 
  dplyr::select(elevation_m) %>% 
  mutate(elevation_ft = elevation_m *  3.28084) %>% 
  mutate(across(c('elevation_m','elevation_ft'), round, 0)) %>% 
  st_drop_geometry()
# st_geometry(df_elev_epqs) <- NULL # just in case last operation does not work. 
df <- cbind(df, df_elev_epqs)

rm(prj_dd, df_elev_epqs, df_points)
```

**Retrieve Ownership of site**

# Background 
This section will require you to download a slightly hefty file in order to determine ownership of a site.

```{r}
knitr::include_graphics("files/spatial/doi-unified-regions.jpg", dpi = NA)
```
The image above may help you determine which files are necessary for your work. You may know download these two ways.

First off you may manually download them to the 'spatial' subfolder in Barneby Lives.

Or you may use the following code...

```{r}
install.packages("rvest")
page <- rvest::html("https://www.sciencebase.gov/catalog/item/6026ed89d34eb1203113955e")

page %>% 
  html_nodes("a") %>% 
  html_attr("href")
```

# Practice

```{r}
ownership.f <- as.list(list.files("files/spatial/PADUS", pattern = "^PADUS2_1_Region"))
ownership.sf <- lapply(owernship.f, function(x) read_sf(x))

rm(ownership.f, ownership.sf)
```

**Retrieve Township, Range, Section information from the PLSS**
Bear in mind these are *usually* only present in Western States.

I thought I had a way of patching this through a REST API once upon a time... Well if you are interested in this follow along... Refer to the file "Barneby_Lives/files/spatial/PLSS-CadNSDI-Data-Set-Availability.pdf" to find a proper reference source to download a CAD from. Usually they come direct from BLM (AK, MT, UT are a few of the exceptions)

```{r}
plss.f <-as.list(list.files("files/spatial/PLSS"))
plss.sf <- lapply(owernship.f, function(x) read_sf(x))

rm(plss.f, plss.sf)
```

# Temporal

Simple stuff here, let's get a nice date format for making labels. I do not like nor accept that a format such as 07.07.2021 is acceptable for an museum label, is this month 7 or day 7 ? oh wait... Either way, let's make it unambiguous.

```{r Date formats}

df <- df %>% 
  mutate(date_digital = case_when(

  ))
```

# Taxonomic

**Spelling**

We will largely use the backbone from the World Flora online as a dictionary to search against using the WorldFlora package. 
```{r}

library("WorldFlora")
data("WFO.example")
test <- WorldFlora::WFO.match(spec.data = df, 
   WFO.file = "L:/Barneby_Lives/files/taxonomic/WFO_Backbone/classification.txt",
   spec.name = 'binomial',
   #Genus = ,
   #Species = ,
   #Infraspecific.rank = ,
   #Infraspecifc = , 
   #Authorship = ,
   Fuzzy = 0.1)

spec.test <- data.frame(spec.name=c("Faidherbia albida", "Acacia albida",
    "Omalanthus populneus", "Pygeum afric"))

WFO.match(spec.data=spec.test, WFO.data=WFO.example, counter=1, verbose=TRUE)

```

**Taxonomy** 
We will switch over to Kews Plants of the World to see if a conserved name or some other alternative opinion exists for our taxa. 
```{r}
taxize::pow_synonyms()
```

**Family**
These things change. 
```{r}
taxize::classification("", db = "pow")
```

### 2023-05-09 scripts/crop2boundary.Rmd
creating tiles for making virtual rasters

---
title: "crop"
author: "steppe"
date: "`r Sys.Date()`"
output: html_document
---

```{r library}
library(sf)
library(terra)
library(tidyverse)
```

```{r define domain of data}

bound <- data.frame(
  y = c(30, 30, 50, 50, 30), 
  x = c(-85, -125, -125, -85, -85)
) %>% 
  st_as_sf(coords = c('x', 'y'), crs = 4326) %>% 
  st_bbox() %>% 
  st_as_sfc()

bb_vals <- c(-125, -85, 30, 50)
w_illi <- ext(bb_vals)

tile_cells <- st_make_grid(
  bound,
  what = "polygons", square = T, flat_topped = F, crs = 4326,
  n = c(
  abs(round( (bb_vals[2] - bb_vals[1]) / 5, 0)), # rows
    abs(round( (bb_vals[3] - bb_vals[4]) / 5, 0))) #cols
) %>% 
  st_as_sf() %>% 
  rename(geometry = x) %>% 
  mutate(x = st_coordinates(st_centroid(tile_cells))[,1],
         y = st_coordinates(st_centroid(tile_cells))[,2], 
         .before = geometry, 
         across(c('x', 'y'), round, 1),
         cellname = paste0('n', abs(y), 'w', abs(x))
         )

tile_cells

```

```{r import data which defines extent of analysis}

geom_p <- '../geodata/geomorphons/'
geo_f <- paste0(geom_p, list.files(path = geom_p, recursive = T))

geomorphons <- vrt(geo_f)
w_illi <- ext(c(-125, -85, 30, 50))

geomorphons <- crop(geomorphons, w_illi)
plot(geomorphons)

makeTiles()

(125-85) / 5 # 8 tiles x
(50-30) / 5 # 4 tiles y

```

```{r import states to crop to}
counties <- st_read('../geodata/USCounties')
st <- tigris::states() %>% select(STATEFP, STUSPS, STATE = NAME) %>% 
  filter(STUSPS %in% c('AZ', 'AK', 'CA', 'CO', 'ID', 'KS', 'NE', 'ND', 'NM', 'NV', 
                       'MN', 'MT', 'SD', 'TX', 'OK', 'OR', 'UT', 'WA', 'WY'))
```

```{r}
ggplot() + 
  geom_sf(data = st, aes(fill = STATEFP), color = NA)
```

### 2023-05-09 scripts/crop2boundary.Rmd
format for subset tiles for vrts

@@
 ---
+
+Many of these datasets are very large we are going to crop them to the area of analysis, and save the subsets.
 
@@
   rename(geometry = x) %>% 
-  mutate(x = st_coordinates(st_centroid(tile_cells))[,1],
-         y = st_coordinates(st_centroid(tile_cells))[,2], 
+  mutate(x = st_coordinates(st_centroid(.))[,1],
+         y = st_coordinates(st_centroid(.))[,2], 
          .before = geometry, 
@@
 
-tile_cells
+object.size(tile_cells)
 
@@
 
-```{r import data which defines extent of analysis}
+```{r geomorphons}
 
-geom_p <- '../geodata/geomorphons/'
+tile_cellsV <- vect(tile_cells)
+
+geom_p <- '../geodata_raw/geomorphons/'
 geo_f <- paste0(geom_p, list.files(path = geom_p, recursive = T))
@@
 geomorphons <- vrt(geo_f)
-w_illi <- ext(c(-125, -85, 30, 50))
+geomorphons <- crop(geomorphons, st_union(tile_cells))
 
-geomorphons <- crop(geomorphons, w_illi)
-plot(geomorphons)
+fname <- paste0('../geodata/geomorphons/', tile_cells$cellname, "_.tif")
+makeTiles(geomorphons, tile_cellsV,filename = fname, na.rm = F)
 
-makeTiles()
-
-(125-85) / 5 # 8 tiles x
-(50-30) / 5 # 4 tiles y
-
+rm(geomorphons)
 ```

### 2023-05-09 scripts/crop2boundary.Rmd
cropped rasters

@@
 library(tidyverse)
+source('functions.R')
 ```
@@
 
-object.size(tile_cells)
-
+tile_cellsV <- vect(tile_cells)
 ```
 
-```{r geomorphons}
+```{r subset physical variables to domain, eval = F}
 
-tile_cellsV <- vect(tile_cells)
+mason(dirin = '../geodata_raw/geomorphons/',
+      dirout = '../geodata/geomorphons/',
+      grid = tile_cellsV, fnames = 'cellname')
 
-geom_p <- '../geodata_raw/geomorphons/'
-geo_f <- paste0(geom_p, list.files(path = geom_p, recursive = T))
+mason(dirin = '../geodata_raw/aspect/',
+      dirout = '../geodata/aspect/',
+      grid = tile_cellsV, fnames = 'cellname')
 
-geomorphons <- vrt(geo_f)
-geomorphons <- crop(geomorphons, st_union(tile_cells))
+mason(dirin = '../geodata_raw/slope/',
+      dirout = '../geodata/slope/',
+      grid = tile_cellsV, fnames = 'cellname')
 
-fname <- paste0('../geodata/geomorphons/', tile_cells$cellname, "_.tif")
-makeTiles(geomorphons, tile_cellsV,filename = fname, na.rm = F)
-
-rm(geomorphons)
 ```

### 2023-05-09 scripts/functions.R

@@ -1,0 +1,30 @@
+#' a wrapper around terra::makeTiles for setting the domain of a project
+#' 
+#' @param dirin the input directory, including path, with the dataset
+#' @param dirout the output directory, including path, for the subset dataset to go
+#' @param grid spatvector defining the tiles, should contain a column containing grid cell names
+#' @param fnames name of column containing filenames in the grid
+mason <- function(dirin, dirout, grid, fnames, ...){
+
+  # identify the paths to the rasters
+  dirin <- sub('/$', '', dirin)
+  paths2rast <- file.path(dirin, list.files(path = paste0(dirin, '/'), recursive = T, ...))
+
+  # create and crop the virtual raster
+  virtualRast <- terra::vrt(paths2rast)
+ # virtualRast_sub <- terra::crop(virtualRast, terra::ext(grid))
+
+  # set file names and write out product
+  columns <- terra::values(grid)
+  cellname <- columns[,fnames]
+  dirout <- sub('/$', '', dirout)
+  
+  fname <- file.path(dirout, paste0(cellname, ".tif"))
+  terra::makeTiles(virtualRast_sub, grid, filename = fname, na.rm = F)
+  
+  return(fname)
+  rm(virtualRast_sub)
+  gc()
+
+}
+

### 2023-05-09 scripts/crop2boundary.Rmd
battling plss

@@
 
+rm(mason)
 ```
 
-```{r import states to crop to}
-counties <- st_read('../geodata/USCounties')
-st <- tigris::states() %>% select(STATEFP, STUSPS, STATE = NAME) %>% 
-  filter(STUSPS %in% c('AZ', 'AK', 'CA', 'CO', 'ID', 'KS', 'NE', 'ND', 'NM', 'NV', 
-                       'MN', 'MT', 'SD', 'TX', 'OK', 'OR', 'UT', 'WA', 'WY'))
+```{r crop political boundary data to extent, eval = F}
+counties <- st_read('../geodata_raw/USCounties/tl_2020_us_county.shp', quiet = T) %>% 
+  st_transform(st_crs(tile_cells)) %>% 
+  select(STATEFP, COUNTY = NAME)
+
+states <- tigris::states() %>% 
+  select(STATEFP, STATE = NAME, STUSPS) %>% 
+  st_drop_geometry()
+
+counties <- counties[st_intersects(counties, st_union(tile_cells)) %>% lengths > 0, ]
+counties <- left_join(counties, states, by = "STATEFP") %>% 
+  select(-STATEFP) %>% 
+  arrange(STATE)
+
+states <- counties %>%
+  distinct(STUSPS) %>% 
+  pull(STUSPS)
+
+st_write(counties, dsn = file.path('../geodata/political', 'political.shp'))
+
+rm(counties)
 ```
 
-```{r}
-ggplot() + 
-  geom_sf(data = st, aes(fill = STATEFP), color = NA)
+```{r crop mountain names database to extent, eval = F}
+
+mtns <- st_read('../geodata_raw/globalMountains/GMBA_Inventory_v2.0_standard_basic.shp', 
+        quiet = T) %>% 
+  st_make_valid() %>% 
+  select(MapName) %>% 
+  st_crop(., st_union(tile_cells))
+
+st_write(mtns, dsn = file.path('../geodata/mountains', 'mountains.shp'), quiet = T)
+
+rm(mtns)
 ```
+
+```{r crop geographic place name database to extent, eval = F}
+
+files <- paste0('../geodata_raw/GNIS/Text/DomesticNames_', states, '.txt')
+cols <- c('feature_name', 'prim_lat_dec', 'prim_long_dec')
+
+places <- lapply(files, read.csv, sep = "|") %>% 
+  map(., ~ .x |> select(all_of(cols))) %>% 
+  data.table::rbindlist() %>% 
+  st_as_sf(coords = c('prim_long_dec', 'prim_lat_dec'), crs = 4269) %>% 
+  st_transform(4326)
+
+places <- places[st_intersects(places, st_union(tile_cells)) %>% lengths > 0, ]
+
+st_write(places, dsn = file.path('../geodata/places', 'places.shp'), quiet = T)
+
+rm(places, cols, files)
+```
+
+```{r crop padus to domain, eval = F}
+
+padus <- st_read(dsn = '../geodata_raw/PADUS3/PAD_US3_0.gdb', 
+                 layer = 'PADUS3_0Fee') %>% 
+  filter(State_Nm %in% states) %>% 
+  select(Mang_Name, Unit_Nm) %>% 
+  st_cast('MULTIPOLYGON')
+
+tile_cells <- st_transform(tile_cells, st_crs(padus))
+padus <- padus[st_intersects(padus, st_union(tile_cells)) %>% lengths > 0, ]
+padus <- st_transform(padus, crs = 4326)
+
+st_write(padus, dsn = file.path('../geodata/pad', 'pad.shp'), quiet = T)
+rm(padus)
+```
+
+```{r geological map, eval = F}
+geological <- st_read(
+  '../geodata_raw/geologicalMap/USGS_StateGeologicMapCompilation_ver1.1.gdb', 
+          layer = 'SGMC_Geology', quiet = T) %>% 
+  select(GENERALIZED_LITH, UNIT_NAME)
+
+tile_cells <- st_transform(tile_cells, st_crs(geological))
+geological <- geological[st_intersects(geological, st_union(tile_cells)) %>% lengths > 0, ]
+geological <- st_crop(geological, st_union(tile_cells))
+geological <- st_transform(geological, 4326)
+
+st_write(geological, dsn = file.path('../geodata/geology', 'geology.shp'), quiet = T)
+
+tile_cells <- st_transform(tile_cells, crs = 4326)
+rm(geological)
+```
+
+```{r allotments, eval = F}
+
+blm <- st_read(
+  '../geodata_raw/BLMAllotments/BLM_Natl_Grazing_Allotment_Polygons.shp') %>% 
+  select(ALLOT_NAME)
+usfs <- st_read(
+  '../geodata_raw/USFSAllotment/S_USA.Allotment.shp', quiet = T
+) %>% 
+  select(ALLOT_NAME = ALLOTMENT_)
+
+allotments <- bind_rows(blm, usfs) %>% 
+  st_make_valid()
+tile_cells <- st_transform(tile_cells, st_crs(allotments))
+allotments <- allotments[st_intersects(allotments, st_union(tile_cells)) %>% lengths > 0, ]
+allotments <- st_crop(allotments, st_union(tile_cells))
+allotments <- st_transform(allotments, 4326)
+
+st_write(allotments, dsn = file.path('../geodata/allotments', 'allotments.shp'), quiet = T)
+
+rm(blm, usfs, allotments)
+```
+
+```{r public land survey system only for west the front range}
+
+bound_nm <- data.frame(
+  y = c(31, 31, 50, 50, 31), 
+  x = c(-105, -125, -125, -105, -103)
+) %>% 
+  st_as_sf(coords = c('x', 'y'), crs = 4326) %>% 
+  st_bbox() %>% 
+  st_as_sfc()
+
+township <- st_read(
+  '../geodata_raw/nationalPLSS/ilmocplss.gdb', layer = 'PLSSTownship', quiet = T
+) %>% 
+  st_cast('POLYGON') %>% 
+  select(TWNSHPLAB)
+
+bound_nm <- st_transform(bound_nm, st_crs(township))
+township <- township[st_intersects(township, bound_nm) %>% lengths > 0, ]
+
+section <- st_read(
+  '../geodata_raw/nationalPLSS/ilmocplss.gdb', layer = 'PLSSFirstDivision', quiet = T
+) %>% 
+  st_cast('POLYGON') %>% 
+  select(FRSTDIVLAB)
+
+section <- section[st_intersects(section, bound_nm) %>% lengths > 0, ]
+
+trs <- st_join(township, section, .predicate = st_overlaps) %>% 
+  mutate(TRS = paste0(TWNSHPLAB, ' S', FRSTDIVLAB)) %>% 
+  select(TRS)
+
+plot(trs)
+
+before <- object.size(trs)
+trs <- st_simplify(trs)
+after <- object.size(trs)
+after
+
+st_write(trs, 
+         dsn = file.path('../geodata/plss', 'plss.shp'), quiet = T)
+
+rm(section, township, trs)
+```

### 2023-05-09 scripts/crop2boundary.Rmd
names added

@@
       dirout = '../geodata/slope/',
+      grid = tile_cellsV, fnames = 'cellname')
+
+mason(dirin = '../geodata_raw/elevation/',
+      dirout = '../geodata/elevation/',
       grid = tile_cellsV, fnames = 'cellname')
@@
 
-```{r public land survey system only for west the front range}
+```{r public land survey system only for land west of CO border, eval = F}
 
@@
   y = c(31, 31, 50, 50, 31), 
-  x = c(-105, -125, -125, -105, -103)
+  x = c(-95, -125, -125, -95, -95)
 ) %>% 
@@
   st_cast('POLYGON') %>% 
-  select(TWNSHPLAB)
+  select(TWNSHPLAB, PLSSID)
 
@@
 township <- township[st_intersects(township, bound_nm) %>% lengths > 0, ]
+township <- st_drop_geometry(township)
 
@@
   st_cast('POLYGON') %>% 
-  select(FRSTDIVLAB)
+  select(FRSTDIVLAB, PLSSID, FRSTDIVID)
 
@@
 
-trs <- st_join(township, section, .predicate = st_overlaps) %>% 
-  mutate(TRS = paste0(TWNSHPLAB, ' S', FRSTDIVLAB)) %>% 
+trs <- inner_join(section, township, 'PLSSID') %>% 
+  mutate(TRS = paste(TWNSHPLAB, FRSTDIVLAB)) %>% 
   select(TRS)
-
-plot(trs)
 
@@
 after <- object.size(trs)
-after
+object.size(trs)
 
@@
 
-rm(section, township, trs)
+rm(section, township, trs, bound_nm, before, after)
 ```

### 2023-05-09 scripts/functions.R

@@ -12,7 +12,7 @@ mason <- function(dirin, dirout, grid, fnames, ...){
 
   # create and crop the virtual raster
   virtualRast <- terra::vrt(paths2rast)
- # virtualRast_sub <- terra::crop(virtualRast, terra::ext(grid))
+  virtualRast_sub <- terra::crop(virtualRast, terra::ext(grid))
 
   # set file names and write out product
   columns <- terra::values(grid)
@@ -22,7 +22,6 @@ mason <- function(dirin, dirout, grid, fnames, ...){
   fname <- file.path(dirout, paste0(cellname, ".tif"))
   terra::makeTiles(virtualRast_sub, grid, filename = fname, na.rm = F)
   
-  return(fname)
   rm(virtualRast_sub)
   gc()
 

### 2023-05-09 scripts/taxonomy_setup.Rmd
exploring taxonomy process

---
title: "Barneby Lives!"
author: "steppe"
date: "9/13/2021"
output: html_document
---

```{r}
library(tidyverse)
library(kewr)
```

stages...
verify the spelling of the genus (parse these from the POWO webshot)
verify the spelling of the epithet (parse these from the POWO webshot)

if not verified
fuzzy match genus... (somehow get a local list.....)
fuzzy match epithet...

find authorship...

```{r}
re <- search_powo("Ivesia lycopodioides")

# some considerations....

# If the species cannot be found, RESULTS will be NULL 
search_powo("Ivesia lycopodioides var. lycopodioides")
# if results are NULL, can we strip from subsp./var./ssp. until the end to query only species?

# in the situation where there is a synonym of the group RESULTS will be a list of length two. 
re_syn <- search_powo("Ivesia lycopodioides")
# results will be a list of length 2

# when the appropriate name is searched RESULTS will be a list of length one...
re_perf <- search_powo("Potentilla lycopodioides")

re1 <- re[['results']]

```

```{r}

data("WFO.example")

WFO.remember()

test <- WorldFlora::WFO.match(
  spec.data = WFO.example, 
   spec.name = 'binomial',
   Fuzzy = 0.1)

family_spellings <- vascular.families %>% 
  dplyr::pull(Family)

pec.test <- data.frame(spec.name=c("Faidherbia albida", "Acacia albida",
"Faidherbia albidum", "Faidherbia albidus",
"Faidherbia albiida",
"Prunus africanus", "Prunos africanea",
"Prunus afrocaneus", "Prunus afrocaneos"))

match1 <- WFO.match.fuzzyjoin(spec.data=spec.test, WFO.data=WFO.example,
fuzzydist.max = 6)
```

### 2023-05-10 scripts/taxonomy_setup.Rmd
function to retrieve found results from powo

@@
 library(kewr)
+library(WorldFlora)
+WFO.remember()
 ```
@@
 ```{r}
-re <- search_powo("Ivesia lycopodioides")
-
+x <- search_powo("Ivesia lycopodioides")
 # some considerations....
@@
 # If the species cannot be found, RESULTS will be NULL 
-search_powo("Ivesia lycopodioides var. lycopodioides")
+re_none <- search_powo("Ivesia lycopodioides var. lycopodioides")
+
+try_again <-
+
+is.null(re_none[["results"]]){
+
+  try_again
+
+} else {final <- result_grabber(query_results)}
+
 # if results are NULL, can we strip from subsp./var./ssp. until the end to query only species?
@@
 
-re1 <- re[['results']]
+# once everything has gone correctly - i.e. we have found the species, we come to
+# this stage of the process where we extract the values.
+
+result_grabber <- function(x){
+
+  # subset the appropriate data frame, there is one if clean data were entered, 
+  # and two if a synonym was entered.
+  results <- x[['results']]
+  resultT <- data.frame(results[which(lapply(results, length) == 8)])
+
+  # the main variables which all results should have. 
+  family = resultT$family
+  authority = resultT$author
+  full_name = resultT$name
+  split_name <- unlist(str_split(full_name, pattern = " "))
+  genus = split_name[1]
+  epithet = split_name[2]
+
+  # the infra species information if relevant. 
+  if( length(split_name) > 2){
+     infraspecies = split_name[length(split_name)]
+  } else {infraspecies = NA}
+
+  if( length(split_name) > 2){
+     infrarank = str_extract(full_name, ' var\\. | spp\\. | subsp\\. ') |>
+       str_trim(side = 'both')
+  } else {infrarank = NA}
+
+  taxonomic_info <- data.frame(
+    family, name_authority = paste(full_name, authority), 
+    full_name, genus, epithet, infrarank, infraspecies, authority
+  )
+
+  return(taxonomic_info)
+}
+
+result_grabber(re_syn)
 
@@
 ```{r}
-
-data("WFO.example")
-
-WFO.remember()
 

### 2023-05-10 scripts/taxonomy_setup.Rmd
subset POWO results lists by whether accepted name is TRUE

--- 2023-05-09 scripts/taxonomy_setup.Rmd
@@
 library(kewr)
+library(WorldFlora)
+WFO.remember()
+source('functions.R')
 ```
@@
 ```{r}
-re <- search_powo("Ivesia lycopodioides")
 
-# some considerations....
+dropped_infra <- search_powo("Ivesia lycopodioides var. lycopodioides")
+valid_binomial <- search_powo("Ivesia lycopodioides")
+kept_infra <- search_powo("Potentilla lycopodioides var. megalopetala")
+what_do_w_authority_autonyms <- search_powo("Potentilla lycopodioides var. lycopodioides")
 
-# If the species cannot be found, RESULTS will be NULL 
-search_powo("Ivesia lycopodioides var. lycopodioides")
-# if results are NULL, can we strip from subsp./var./ssp. until the end to query only species?
+powo_searcher("Potentilla lycopodioides var. megalopetala")
+powo_searcher("Potentilla lycopodioides var. lycopodioides")
 
-# in the situation where there is a synonym of the group RESULTS will be a list of length two. 
-re_syn <- search_powo("Ivesia lycopodioides")
-# results will be a list of length 2
-
-# when the appropriate name is searched RESULTS will be a list of length one...
-re_perf <- search_powo("Potentilla lycopodioides")
-
-re1 <- re[['results']]
+lapply(c('Poa annua', 'Linnaea borealis'), powo_searcher)
 

### 2023-05-10 scripts/functions.R

@@ -27,3 +27,136 @@ mason <- function(dirin, dirout, grid, fnames, ...){
 
 }
 
+
+#' try to download from Kew again, with just the binomial
+#' 
+#' several infraspecies fail from POWO, retry the query with just the base name
+#' @param x output from the first step of 'powo_searcher'
+try_again <- function(x) {
+  q <- x[['query']]
+  only_binomial <- unlist(stringr::str_split(q, pattern = " "))
+  only_binomial <- paste(only_binomial[1], only_binomial[2])
+  
+  results <- search_powo(only_binomial)
+  return(results)
+}
+
+#' collect the results of a successful powo search
+#' 
+#' @param x a successful powo search query
+result_grabber <- function(x) {
+  # subset the appropriate data frame, there is one if clean data were entered,
+  # and two if a synonym was entered.
+  results <- x[['results']]
+  
+  clean <- which(unlist(lapply(results, `[`, 'accepted')) == T)
+  resultT <- data.frame(results[clean])
+  
+  # the main variables which all results should have.
+  family <- resultT$family
+  authority <- resultT$author
+  full_name <- resultT$name
+  split_name <- unlist(stringr::str_split(full_name, pattern = " "))
+  genus <- split_name[1]
+  epithet <- split_name[2]
+  
+  # the infra species information if relevant.
+  if (length(split_name) > 2) {
+    infraspecies <- split_name[length(split_name)]
+  } else {
+    infraspecies <- NA
+  }
+  
+  if (length(split_name) > 2) {
+    infrarank <-
+      stringr::str_extract(full_name, ' var\\. | spp\\. | subsp\\. ') |>
+      stringr::str_trim(side = 'both')
+  } else {
+    infrarank <- NA
+  }
+  
+  if (is.null(authority)){
+    
+    res <- autonym(genus, epithet, infrarank, infraspecies)
+    name_authority <- res[[1]]
+    authority <- res[[2]]
+    
+  } else{name_authority = paste(full_name, authority)}
+  
+  taxonomic_info <- data.frame(
+    family,
+    'name_authority' = name_authority,
+    full_name,
+    genus,
+    epithet,
+    infrarank,
+    infraspecies,
+    authority
+  )
+  
+  return(taxonomic_info)
+}
+
+#' query plants of the world online for taxonomic information
+#' 
+#' a wrapper for kewr::search_powo
+#' @param x a vector of species names to submit, these should have clean spelling
[…]

### 2023-05-10 scripts/taxonomy_setup.Rmd
spell check function

@@
 
+# the first letter of the genus must be correct, 
+# the first three letters of the species must be correct
+
+```{r}
+
+cla <- data.table::fread('../taxonomic_data/WFO_Backbone/classification.txt')
+genLKPtab <- unique(cla, by = 'genus')[taxonomicStatus == 'Accepted','genus'] %>% 
+  arrange(genus) %>% 
+  mutate(Grp = str_extract(genus, '[A-Z][a-z]{1}'))%>% 
+  rename(strings = genus)
+
+speLKPtab <- unique(cla, by = 'specificEpithet')[
+  taxonomicStatus == 'Accepted','specificEpithet'] %>% 
+  arrange(specificEpithet) %>% 
+  mutate(Grp = str_extract(specificEpithet, '[a-z]{3}')) %>% 
+  rename(strings = specificEpithet)
+
+spell_check('Poa annua') # works
+spell_check('Poar annua') # works
+spell_check('Poa annur') # works
+spell_check('Poar annur') # works
+spell_check('Pou annua') # works
+
+```
+
 stages...
@@
 ```{r}
-
-dropped_infra <- search_powo("Ivesia lycopodioides var. lycopodioides")
-valid_binomial <- search_powo("Ivesia lycopodioides")
-kept_infra <- search_powo("Potentilla lycopodioides var. megalopetala")
-what_do_w_authority_autonyms <- search_powo("Potentilla lycopodioides var. lycopodioides")
-
-powo_searcher("Potentilla lycopodioides var. megalopetala")
-powo_searcher("Potentilla lycopodioides var. lycopodioides")
-
-lapply(c('Poa annua', 'Linnaea borealis'), powo_searcher)
-
+lapply(c('Cryptantha glomerata', 'Sphenosciadium capitellatum'), powo_searcher)
 ```
-
-```{r}
-
-test <- WorldFlora::WFO.match(
-  spec.data = WFO.example, 
-   spec.name = 'binomial',
-   Fuzzy = 0.1)
-
-family_spellings <- vascular.families %>% 
-  dplyr::pull(Family)
-
-pec.test <- data.frame(spec.name=c("Faidherbia albida", "Acacia albida",
-"Faidherbia albidum", "Faidherbia albidus",
-"Faidherbia albiida",
-"Prunus africanus", "Prunos africanea",
-"Prunus afrocaneus", "Prunus afrocaneos"))
-
-match1 <- WFO.match.fuzzyjoin(spec.data=spec.test, WFO.data=WFO.example,
-fuzzydist.max = 6)
-```

### 2023-05-10 scripts/functions.R

@@ -159,4 +159,56 @@ autonym <- function(genus, epithet, infrarank, infraspecies){
   return(list(name_authority, authority))
 }
 
-
+#' check that genera and specific epithets are spelled (almost) correctly
+#' 
+#' @param x a vector of species names
+spell_check <- function(x) {
+  
+  pieces <- unlist(stringr::str_split(x, pattern = " "))
+  genus <- pieces[1]
+  species <- pieces[2]
+  binom <- paste(genus, species)
+  
+  if (any(grep(x = cla$scientificName, pattern = binom, fixed = T))) {
+    return(binom)
+  } else{ # try and determine which piece is incorrect. 
+    
+    #subset datasets really quickly
+    genus2char <- stringr::str_extract(genus, '[A-Z][a-z]{1}')
+    species3char <- stringr::str_extract(species, '[a-z]{3}')
+    gen_strings <-
+      dplyr::filter(genLKPtab, Grp == genus2char) |> dplyr::pull(strings)
+    spe_strings <-
+      dplyr::filter(speLKPtab, Grp == species3char) |> dplyr::pull(strings)
+    
+    # check to see if both genus and species are clean
+    if (any(grep(x = gen_strings, pattern = paste0('^', genus, '$')))) {
+      clean_genus_Tag <- genus
+      return(clean_genus_Tag)
+    } else {
+      possible_genus_Tag <- gen_strings[which.min(adist(genus, gen_strings))]
+    }
+    
+    # is species clean
+    if (any(grep(x = spe_strings, pattern = paste0('^', species, '$')))) {
+      clean_species_Tag <- species
+    } else {
+      possible_species_Tag <- spe_strings[which.min(adist(species, spe_strings))]
+    }
+    
+    if (exists('clean_genus_Tag') & exists ('clean_species_Tag'))
+    {
+      return('Both the genus and species appear to be spelled correctly, but are not in my lookup table')
+    } else{
+      combos <- ls()[grep(ls(), pattern = 'Tag')]
+      search_q <- combos[c(grep(combos, pattern = 'genus'), grep(combos, pattern = 'species'))]
+      search_nom <- paste(unlist(mget(search_q)), collapse = " ")
+      
+      if(any(grep(x = cla$scientificName, pattern = search_nom, fixed = T))){
+        return(search_nom)
+      } else{
+        return(paste0 (search_nom, 'We cannot figure this one out'))
+      }
+    }
+  }
+}

### 2023-05-11 scripts/taxonomy_setup.Rmd
spepecies spell check support for infraspecies, and formatted output results

@@
 library(WorldFlora)
-WFO.remember()
+# WFO.remember()
 source('functions.R')
@@
 
-```{r}
+```{r prepare materials for look up tables, eval = F}
 
-cla <- data.table::fread('../taxonomic_data/WFO_Backbone/classification.txt')
+cla <- data.table::fread('../taxonomic_data/WFO_Backbone/classification.txt', na.strings = "")
+
+epiLKPtab <- cla[, c('scientificName', 'specificEpithet', 'infraspecificEpithet', 'verbatimTaxonRank')]
+epiLKPtab <- filter(epiLKPtab, is.na(verbatimTaxonRank) | verbatimTaxonRank != "f.")
+
 genLKPtab <- unique(cla, by = 'genus')[taxonomicStatus == 'Accepted','genus'] %>% 
@@
   rename(strings = genus)
-
 speLKPtab <- unique(cla, by = 'specificEpithet')[
@@
 
-spell_check('Poa annua') # works
-spell_check('Poar annua') # works
-spell_check('Poa annur') # works
-spell_check('Poar annur') # works
-spell_check('Pou annua') # works
+spell_check('Penstemon scariosus var. garrettii') # works
+spell_check('Penstemen scariosus var. garrettii') # works
+spell_check('Penstemen scariosius var garratti') # works
+spell_check('Penstemen scariosius ssp garratti') # works
+spell_check('Penstemen scariosus') # works
 
 ```
-
-stages...
-verify the spelling of the genus (parse these from the POWO webshot)
-verify the spelling of the epithet (parse these from the POWO webshot)
-
-if not verified
-fuzzy match genus... (somehow get a local list.....)
-fuzzy match epithet...
-
-find authorship...
 

### 2023-05-11 scripts/functions.R

@@ -165,49 +165,81 @@ autonym <- function(genus, epithet, infrarank, infraspecies){
 spell_check <- function(x) {
   
   pieces <- unlist(stringr::str_split(x, pattern = " "))
-  genus <- pieces[1]
-  species <- pieces[2]
+  genus <- pieces[1] ; species <- pieces[2]
   binom <- paste(genus, species)
   
-  if (any(grep(x = cla$scientificName, pattern = binom, fixed = T))) {
-    return(binom)
-  } else{ # try and determine which piece is incorrect. 
+  # infraspecies should be found without much hassle due to their length
+  if(length(pieces) == 4){
+    infras <- na.omit(epiLKPtab)
+    full_name <- paste(genus, species, 
+                       stringr::str_replace(pieces[3], 'ssp\\.|ssp', 'subsp.'), pieces[4])
     
-    #subset datasets really quickly
-    genus2char <- stringr::str_extract(genus, '[A-Z][a-z]{1}')
-    species3char <- stringr::str_extract(species, '[a-z]{3}')
-    gen_strings <-
-      dplyr::filter(genLKPtab, Grp == genus2char) |> dplyr::pull(strings)
-    spe_strings <-
-      dplyr::filter(speLKPtab, Grp == species3char) |> dplyr::pull(strings)
-    
-    # check to see if both genus and species are clean
-    if (any(grep(x = gen_strings, pattern = paste0('^', genus, '$')))) {
-      clean_genus_Tag <- genus
-      return(clean_genus_Tag)
+    if (any(grep( x = infras$scientificName, pattern = full_name, fixed = T))) {
+      return(data.frame(Query = x, Result = full_name, Match = 'exact'))
     } else {
-      possible_genus_Tag <- gen_strings[which.min(adist(genus, gen_strings))]
-    }
-    
-    # is species clean
-    if (any(grep(x = spe_strings, pattern = paste0('^', species, '$')))) {
-      clean_species_Tag <- species
-    } else {
-      possible_species_Tag <- spe_strings[which.min(adist(species, spe_strings))]
-    }
-    
-    if (exists('clean_genus_Tag') & exists ('clean_species_Tag'))
-    {
-      return('Both the genus and species appear to be spelled correctly, but are not in my lookup table')
-    } else{
-      combos <- ls()[grep(ls(), pattern = 'Tag')]
[…]

### 2023-05-11 scripts/functions.R
powo searcher returns intiial query with results

@@ -132,10 +132,10 @@ powo_searcher <- function(x) {
   
   # this is the end of the process, return empty results without error, or real results
   if (exists('taxonomic_info')) {
-    return(taxonomic_info)
+    return(cbind(query = x, taxonomic_info))
   } else {
     taxonomic_info <- result_grabber(results_to_process)
-    return(taxonomic_info)
+    return(cbind(query = x, taxonomic_info))
   }
 }
 

### 2023-05-11 scripts/taxonomy_setup.Rmd
powo searcher returns intiial query with results
map maker script and content order for labels

@@
 
-# the first letter of the genus must be correct, 
+# the first two letters of the genus must be correct, 
 # the first three letters of the species must be correct
@@
 
-cla <- data.table::fread('../taxonomic_data/WFO_Backbone/classification.txt', na.strings = "")
+cla <- data.table::fread('../taxonomic_data_raw/WFO_Backbone/classification.txt', na.strings = "")
 
-epiLKPtab <- cla[, c('scientificName', 'specificEpithet', 'infraspecificEpithet', 'verbatimTaxonRank')]
-epiLKPtab <- filter(epiLKPtab, is.na(verbatimTaxonRank) | verbatimTaxonRank != "f.")
+sppLKPtab <- cla[, c('scientificName', 'specificEpithet', 'infraspecificEpithet', 'verbatimTaxonRank')]
+sppLKPtab <- filter(sppLKPtab, is.na(verbatimTaxonRank) | verbatimTaxonRank != "f.")
 
@@
   rename(strings = genus)
-speLKPtab <- unique(cla, by = 'specificEpithet')[
+epiLKPtab <- unique(cla, by = 'specificEpithet')[
   taxonomicStatus == 'Accepted','specificEpithet'] %>% 
@@
 
-spell_check('Penstemon scariosus var. garrettii') # works
-spell_check('Penstemen scariosus var. garrettii') # works
-spell_check('Penstemen scariosius var garratti') # works
-spell_check('Penstemen scariosius ssp garratti') # works
-spell_check('Penstemen scariosus') # works
+write.csv(sppLKPtab, '../taxonomic_data/species_lookup_table.csv', row.names = F)
+write.csv(epiLKPtab, '../taxonomic_data/epithet_lookup_table.csv', row.names = F)
+write.csv(genLKPtab, '../taxonomic_data/genus_lookup_table.csv', row.names = F)
 
@@
 ```
+
+```{r}
+library(sf)
+
+dummy_pts <- data.frame(
+  collectionNo = paste0(rep('RCB', 20), seq(1:20)),
+  x = runif(20, min = -120, max = -85), 
+  y = runif(20, min = 35, max = 48)
+) %>% sf::st_as_sf(coords = c('x', 'y'), crs = 4269)
+
+states_sf <- st_read('../geodata/political/political.shp', quiet = T)
+
+map_maker <- function(x, states, path, collection_col){
+
+  if(st_crs(x) == sf::st_crs(states)) {
+    pts } else {pts <- sf::st_transform(x, sf::st_crs(states))}
+
+  ints <- sf::st_intersection(pts, states)
+
+  focal_state <- dplyr::filter(states, STATE == ints$STATE)
+
+  p <- ggplot() + 
+    geom_sf(data = focal_state, fill = NA, color = 'grey15') + 
+    geom_sf(data = pts, size = 0.5) +
+    theme_void()
+
+  fname <- file.path(path, 'maps', paste0('map_', sf::st_drop_geometry(x[1,'collectionNo']), '.png'))
+  ggsave(filename =  fname, plot = p, device = 'png', dpi = 300, width = 1, height = 1, units = 'in', 
+         bg = 'transparent')
+}
+
+dummy_pts_l <- split(dummy_pts, dummy_pts$collectionNo)
+
+out <- lapply(dummy_pts_l, map_maker, states_sf, '../results', collection_col = 'collectionNo')
+
+ggplot() + 
+    geom_sf(data = states_sf, fill = NA, color = 'grey15') + 
+    geom_sf(data = dummy_pts, size = 0.5)
+
+```

### 2023-05-11 scripts/functions.R

@@ -244,3 +244,30 @@ spell_check <- function(x) {
     }
   }
 }
+
+
+#' make a quick county dot map to display the location of the collection
+#' 
+#' @param x an sf dataframe of coordinates to make maps for, requires collection number and spatial attributes
+#' @param path a directory to store the map images in before merging
+#' @param collection_col column specify the collection number or other UNIQUE id for the collection
+
+map_maker <- function(x, states, path, collection_col){
+  
+  if(st_crs(x) == sf::st_crs(states)) {
+    pts } else {pts <- sf::st_transform(x, sf::st_crs(states))}
+  
+  ints <- sf::st_intersection(pts, states)
+  
+  focal_state <- dplyr::filter(states, STATE == ints$STATE)
+  
+  p <- ggplot() + 
+    geom_sf(data = focal_state, fill = NA, color = 'grey15') + 
+    geom_sf(data = pts, size = 0.5) +
+    theme_void() 
+  
+  fname <- file.path(path, 'maps', paste0('map_', sf::st_drop_geometry(x[1,'collectionNo']), '.png'))
+  ggsave(filename =  fname, plot = p, device = 'png', dpi = 300, width = 1, height = 1, units = 'in', 
+         bg = 'transparent')
+}
+

### 2023-05-11 scripts/crop2boundary.Rmd
added function for political information

@@
 ```
+
+```{r modify names of variables to be in proper formats, eval = F}
+
+sf::st_read('../geodata/political/political.shp', quiet = T) %>% 
+  rename(County = COUNTY, State = STATE) %>% 
+  st_write(., '../geodata/political/political.shp', quiet = T, append = F)
+
+sf::st_read('../geodata/allotments/allotments.shp', quiet = T) %>% 
+  dplyr::mutate(Allotment = str_to_title(ALLOT_NAME)) %>% 
+  dplyr::select(-ALLOT_NAME) %>% 
+  st_write(., '../geodata/allotments/allotments.shp', quiet = T, append = F)
+
+sf::st_read('../geodata/plss/plss.shp', quiet = T) %>% 
+  rename(trs = TRS) %>% 
+  st_write(., '../geodata/plss/plss.shp', quiet = T, append = F)
+
+sf::st_read('../geodata/pad/pad.shp', quiet = T) %>% 
+  sf::st_make_valid() %>% 
+  dplyr::filter(sf::st_is_valid(.)) %>% 
+  mutate(Unit_Nm = str_replace(Unit_Nm, "National Forest", "NF"),
+         Unit_Nm = str_replace(Unit_Nm, "Field Office", "FO"),
+         Unit_Nm = str_replace(Unit_Nm, "National Wildlife Refuge", "NWR")) %>% 
+  st_write(., '../geodata/pad/pad.shp', quiet = T, append = F)
+
+```

### 2023-05-11 scripts/project_template.Rmd
added function for political information

---
title: "Barneby Lives usage example"
author: "steppe"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
dummy_pts <- data.frame(
  collectionNo = paste0(rep('RCB', 50), seq(1:50)),
  day = c(sample(1:31, size = 50, replace = T)),
  month = c(sample(5:10, size = 50, replace = T)),
  year = 2023,
  x = runif(50, min = -120, max = -85), 
  y = runif(50, min = 35, max = 48)
) %>% sf::st_as_sf(coords = c(x = 'x', y = 'y'), crs = 4326)

```

```{r Acquire administrative geodata}

political <- sf::st_read('../geodata/political/political.shp', quiet = T)
allotment <- sf::st_read('../geodata/allotments/allotments.shp', quiet = T)
plss <- sf::st_read('../geodata/plss/plss.shp', quiet = T)
ownership <- sf::st_read('../geodata/pad/pad.shp', quiet = T)

x <- sf::st_join(x, political) 
x <- sf::st_join(x, allotment) 
x <- sf::st_join(x, ownership)

x_plss <- sf::st_transform(x, sf::st_crs(plss))
x_plss <- sf::st_join(x_plss, plss)

x_vars <- dplyr::left_join(x, x_plss)

places <- sf::st_read('../geodata/places/places.shp', quiet = T)

head(dummy_ownership)

out <- political_grabber(dummy_pts)
out <- place_grabber(dummy_pts)

# if within 0.5 km of place, use
# if dist > 0.5 km direct

```

### 2023-05-11 scripts/functions.R

@@ -271,3 +271,93 @@ map_maker <- function(x, states, path, collection_col){
          bg = 'transparent')
 }
 
+
+#' gather political site information
+#' 
+#' this function grabs information on the state, county, and township of collections
+#' @param x an sf data frame of collection points
+#' @param y a column which unambiguously identifies each collection
+political_grabber <- function(x) {
+  
+  political <- sf::st_read('../geodata/political/political.shp', quiet = T)
+  allotment <- sf::st_read('../geodata/allotments/allotments.shp', quiet = T)
+  plss <- sf::st_read('../geodata/plss/plss.shp', quiet = T)
+  ownership <- sf::st_read('../geodata/pad/pad.shp', quiet = T)
+  
+  # write attributes to data set
+  
+  x <- sf::st_join(x, political)
+  x <- sf::st_join(x, allotment)
+  x <- sf::st_join(x, ownership)
+  
+  x_plss <- sf::st_transform(x, sf::st_crs(plss))
+  x_plss <- sf::st_join(x_plss, plss) %>% 
+    sf::st_drop_geometry()
+  
+  x_vars <- dplyr::left_join(x, x_plss) %>% 
+    relocate(any_of(c('State', 'County', 'Mang_Name', 'Unit_Nm', 'trs')),
+             .before = geometry)
+  return(x_vars)
+  
+  rm(political, allotment, plss, ownership)
+}
+
+
+
+#' gather place and site information
+#' 
+#' this function grabs information on the state, county, and township of collections
+#' @param x an sf data frame of collection points
+#' @param y a column which unambiguously identifies each collection
+place_grabber <- function(x) {
+  
+  mountains <- sf::st_read('../geodata/mountains/mountains.shp', quiet = T)
+  places <- sf::st_read('../geodata/places/places.shp', quiet = T)
+  
+  x <- sf::st_join(x, mountains)
+  
+  st_nearest(x, places)
+  x <- sf::st_join(x, places)
+  
+  return(x)
+  
+}
+
+
+
+
+#' gather distance and azimuth from shop to center of town
+#'
+#' Help provide some simple context between the building and the 
+#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
+#' @param places_data an sf/dataframe/dataframe which contains coordinates for the locations in the data
+distAZE <- function(x, places_data){
+  
[…]

### 2023-05-12 scripts/crop2boundary.Rmd
cleaning places

@@
 
+sf::st_read('../geodata/places/places.shp', quiet = T) %>% 
+  mutate(
+    fetr_nm = str_remove(fetr_nm, ' Census Designated Place' ),
+    fetr_nm = str_remove(fetr_nm, 'Township of '), 
+    fetr_nm = str_remove(fetr_nm, ' \\(historical\\)'),
+    fetr_nm = str_remove(fetr_nm, 'Village of '),
+    fetr_nm = str_remove(fetr_nm, 'Town of '), 
+    fetr_nm = str_remove(fetr_nm, 'City of ')
+    ) %>% 
+  filter(!str_detect(fetr_nm,
+                     'Election Precinct| County|Ditch|Unorganized Territory|Canal|Lateral|Division|
+                     River Division|Drain Number|Waste Pond|Commissioner District|Lake Superior|Lake Michigan')) %>% 
+  st_write(., '../geodata/places/places.shp', quiet = T, append = F)
+
 ```
+
+' Lake', ' lk.'
+' Canyon', ' cnyn.'
+' Campground', ' cmpgrnd.'
+' Creek', ' crk.'
+' River', ' rvr.'
+' Reservoir', ' rsvr.'
+' Fork', ' fk.'
+' Number', ' #'
+' Mountain', ' mtn.'
+' Mountains', ' mtns.'
+' Meadow', ' mdw.'
+' Meadows' , ' mdws.'
+' Point', ' pt.'

### 2023-05-12 scripts/project_template.Rmd
cleaning places

@@
 dummy_pts <- data.frame(
-  collectionNo = paste0(rep('RCB', 50), seq(1:50)),
-  day = c(sample(1:31, size = 50, replace = T)),
-  month = c(sample(5:10, size = 50, replace = T)),
+  collectionNo = paste0(rep('RCB', 500), seq(1:500)),
+  day = c(sample(1:31, size = 500, replace = T)),
+  month = c(sample(5:10, size = 500, replace = T)),
   year = 2023,
-  x = runif(50, min = -120, max = -85), 
-  y = runif(50, min = 35, max = 48)
+  x = runif(500, min = -120, max = -85), 
+  y = runif(500, min = 35, max = 48)
 ) %>% sf::st_as_sf(coords = c(x = 'x', y = 'y'), crs = 4326)
@@
 
-political <- sf::st_read('../geodata/political/political.shp', quiet = T)
-allotment <- sf::st_read('../geodata/allotments/allotments.shp', quiet = T)
-plss <- sf::st_read('../geodata/plss/plss.shp', quiet = T)
-ownership <- sf::st_read('../geodata/pad/pad.shp', quiet = T)
-
-x <- sf::st_join(x, political) 
-x <- sf::st_join(x, allotment) 
-x <- sf::st_join(x, ownership)
-
-x_plss <- sf::st_transform(x, sf::st_crs(plss))
-x_plss <- sf::st_join(x_plss, plss)
-
-x_vars <- dplyr::left_join(x, x_plss)
-
 places <- sf::st_read('../geodata/places/places.shp', quiet = T)
-
-head(dummy_ownership)
 

### 2023-05-12 scripts/functions.R

@@ -315,49 +315,24 @@ place_grabber <- function(x) {
   places <- sf::st_read('../geodata/places/places.shp', quiet = T)
   
   x <- sf::st_join(x, mountains)
+
+  nearest_place <- places[sf::st_nearest_feature(dummy_pts, places),]
+  distance <- round(
+    as.numeric(
+      st_distance(x, nearest_place, by_element = T)),
+    -2) / 1000 # distance in kilometers
   
-  st_nearest(x, places)
-  x <- sf::st_join(x, places)
+  azimuth <- nngeo::st_azimuth(
+    nearest_place, 
+    x) |> round(0)
+  
+  places <- data.frame(place = st_drop_geometry(nearest_place), distance, azimuth)
+  
+  x <- dplyr::bind_cols(x, places) %>% 
+    relocate(any_of(c('place', 'distance', 'azimuth')),
+             .before = geometry)
   
   return(x)
   
 }
 
-
-
-
-#' gather distance and azimuth from shop to center of town
-#'
-#' Help provide some simple context between the building and the 
-#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
-#' @param places_data an sf/dataframe/dataframe which contains coordinates for the locations in the data
-distAZE <- function(x, places_data){
-  
-  locality <- sf::st_drop_geometry(x)
-  locality <- locality[1, 'Locality']
-  
-  focal <- places_data[grep(locality, places_data$NAME), ]
-  if(nrow(focal) > 1){
-    union_loc <- sf::st_union(x) |> sf::st_point_on_surface()
-    focal <- focal[sf::st_nearest_feature(union_loc, focal),]
-  }
-  
-  location_from <- sf::st_centroid(focal)
-  
-  location_from <- sf::st_transform(location_from, 5070)
-  x_planar <- sf::st_transform(x, 5070)
-  distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')
-  
-  azy <- nngeo::st_azimuth(
-    location_from, 
-    x_planar
-  )
-  
-  distances <- data.frame(
-    st_drop_geometry(x),
-    Distance = round(as.numeric(distances / 1609.34), 1),
-    Azimuth = round(as.numeric(azy), 0),
-    geometry = x[,'geometry']
-  )
-  
-}

### 2023-05-12 scripts/crop2boundary.Rmd
more abbreviations, fn for naming place

--- 2023-05-11 scripts/crop2boundary.Rmd
@@
   mutate(Unit_Nm = str_replace(Unit_Nm, "National Forest", "NF"),
+         Unit_Nm = str_replace(Unit_Nm, "District Office", "DO"), 
          Unit_Nm = str_replace(Unit_Nm, "Field Office", "FO"), 
-         Unit_Nm = str_replace(Unit_Nm, "National Wildlife Refuge", "NWR")) %>% 
+         Unit_Nm = str_replace(Unit_Nm, "National Park", "NP"),
+         Unit_Nm = str_replace(Unit_Nm, 'National Grassland', 'NG'),
+         Unit_Nm = str_replace(Unit_Nm, "National Wildlife Refuge|National Wildlife Range", "NWR"),
+         Unit_Nm = str_replace(Unit_Nm, 'National Monument', 'NM'),
+         Unit_Nm = str_replace(Unit_Nm, "National Recreation Area", "NRA"), 
+         Unit_Nm = str_replace(Unit_Nm, "State Resource Management Area", "SRMA"), 
+         Unit_Nm = str_replace(Unit_Nm, "State Trust Lands", "STL"), 
+         Unit_Nm = str_replace(Unit_Nm, "Department", "Dept."),
+         Unit_Nm = str_replace(Unit_Nm, "Recreation Area", "Rec. Area"),
+         Unit_Nm = str_replace(Unit_Nm, 'Wildlife Habitat Management Area', "WHMA"), 
+         Unit_Nm = str_replace(Unit_Nm, 'Wildlife Management Area', "WMA")) %>% 
   st_write(., '../geodata/pad/pad.shp', quiet = T, append = F)
 
+sf::st_read('../geodata/places/places.shp', quiet = T) %>% 
+  mutate(
+    fetr_nm = str_remove(fetr_nm, ' Census Designated Place' ),
+    fetr_nm = str_remove(fetr_nm, 'Township of '), 
+    fetr_nm = str_remove(fetr_nm, ' \\(historical\\)'),
+    fetr_nm = str_remove(fetr_nm, 'Village of '),
+    fetr_nm = str_remove(fetr_nm, 'Town of '), 
+    fetr_nm = str_remove(fetr_nm, 'City of ')
+    ) %>% 
+  filter(!str_detect(fetr_nm,
+                     'Election Precinct| County|Ditch|Unorganized Territory|Canal|Lateral|Division|
+                     River Division|Drain Number|Waste Pond|Commissioner District|Lake Superior|Lake Michigan')) %>% 
+  mutate(
+    fetr_nm = str_replace(fetr_nm, ' Canyon', ' cnyn.'),
+    fetr_nm = str_replace(fetr_nm, ' Campground', ' cmpgrnd.'),
+    fetr_nm = str_replace(fetr_nm, ' Creek', ' crk.'),
+    fetr_nm = str_replace(fetr_nm, ' Fork', ' fk.'),
+    fetr_nm = str_replace(fetr_nm, ' Lake', ' lk.'),
+    fetr_nm = str_replace(fetr_nm, ' Meadows', ' mdws.'),
+    fetr_nm = str_replace(fetr_nm, ' Meadow', ' mdw.'),
+    fetr_nm = str_replace(fetr_nm, ' Mountains', ' mtns.'),
+    fetr_nm = str_replace(fetr_nm, ' Mountain', ' mtn.'),
+    fetr_nm = str_replace(fetr_nm, ' Number', ' #'),
+    fetr_nm = str_replace(fetr_nm, ' Point', ' pt.'),
+    fetr_nm = str_replace(fetr_nm, ' Reservoir', ' rsvr.'),
+    fetr_nm = str_replace(fetr_nm, ' River ', ' rvr.')
+  ) %>% 
+  st_write(., '../geodata/places/places.shp', quiet = T, append = F)
+
 ```
+
+' Point', ' pt.'

### 2023-05-12 scripts/project_template.Rmd
more abbreviations, fn for naming place

@@
 ```{r}
+library(tidyverse)
+library(sf)
+source('functions.R')
+```
+
+```{r}
 dummy_pts <- data.frame(
@@
   year = 2023,
-  x = runif(500, min = -120, max = -85), 
+  x = runif(500, min = -120, max = -100), 
   y = runif(500, min = 35, max = 48)
@@
 
+rm(out)
 # if within 0.5 km of place, use

### 2023-05-12 scripts/taxonomy_setup.Rmd
more abbreviations, fn for naming place

@@
 
-map_maker <- function(x, states, path, collection_col){
-
-  if(st_crs(x) == sf::st_crs(states)) {
-    pts } else {pts <- sf::st_transform(x, sf::st_crs(states))}
-
-  ints <- sf::st_intersection(pts, states)
-
-  focal_state <- dplyr::filter(states, STATE == ints$STATE)
-
-  p <- ggplot() + 
-    geom_sf(data = focal_state, fill = NA, color = 'grey15') + 
-    geom_sf(data = pts, size = 0.5) +
-    theme_void()
-
-  fname <- file.path(path, 'maps', paste0('map_', sf::st_drop_geometry(x[1,'collectionNo']), '.png'))
-  ggsave(filename =  fname, plot = p, device = 'png', dpi = 300, width = 1, height = 1, units = 'in', 
-         bg = 'transparent')
-}
-
 dummy_pts_l <- split(dummy_pts, dummy_pts$collectionNo)

### 2023-05-12 scripts/functions.R

@@ -319,20 +319,24 @@ place_grabber <- function(x) {
   nearest_place <- places[sf::st_nearest_feature(dummy_pts, places),]
   distance <- round(
     as.numeric(
-      st_distance(x, nearest_place, by_element = T)),
+      sf::st_distance(x, nearest_place, by_element = T)),
     -2) / 1000 # distance in kilometers
   
   azimuth <- nngeo::st_azimuth(
-    nearest_place, 
-    x) |> round(0)
+    nearest_place, # from 
+    x) |> round(0) # to
   
-  places <- data.frame(place = st_drop_geometry(nearest_place), distance, azimuth)
-  
-  x <- dplyr::bind_cols(x, places) %>% 
-    relocate(any_of(c('place', 'distance', 'azimuth')),
-             .before = geometry)
+  places <- data.frame(
+    'place' = sf::st_drop_geometry(nearest_place), distance, azimuth)  |>
+    rename(place = fetr_nm)
+    
+  x <- dplyr::bind_cols(x, places)|>
+    dplyr::mutate(area_from =
+                    paste0('From ', place, ' ', distance, ' km at ', azimuth, '°.'),
+                  .before = geometry) |>
+    select(-place, -distance, -azimuth) 
   
   return(x)
-  
 }
+
 

### 2023-05-12 scripts/functions.R
prototype geomorpho grabber

@@ -340,3 +340,73 @@ place_grabber <- function(x) {
 }
 
 
+
+
+#' gather physical characteristics of the site
+#' 
+#' this function grabs information on the elevation, azimuth, geomorphon, and geology of the site
+#' @param x an sf data frame of collection points
+#' @param y a column which unambiguously identifies each collection
+physical_grabber <- function(x) {
+  
+  round_df <- function(df, digits) {
+    nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
+    df[,nums] <- round(df[,nums], digits = digits)
+    return(df)
+  }
+  
+  x_spat <- terra::vect(x)
+  geology <- terra::vect('../geodata/geology/geology.shp')
+  x_geo <- terra::extract(geology, x_spat)
+  
+  asp <- '../geodata/aspect/'; slo <-  '../geodata/slope'
+  geo <-  '../geodata/geomorphons'; elev <-  '../geodata/elevation'
+  
+  paths2rast <- file.path(asp, list.files(path = paste0(asp, '/'), recursive = T))
+  aspect <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(slo, list.files(path = paste0(slo, '/'), recursive = T))
+  slope <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(geo, list.files(path = paste0(geo, '/'), recursive = T))
+  geomorphon <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(elev, list.files(path = paste0(elev, '/'), recursive = T))
+  elevation <- terra::vrt(paths2rast)
+
+  asp_val <- terra::extract(aspect, x_spat)
+  names(asp_val) <- c('ID', 'aspect')
+  slo_val <- terra::extract(slope, x_spat)
+
+  geoLKPtab <- data.frame(unit = 1:10, geomorphon = c(
+    'flat', 'peak', 'ridge', 'shoulder', 'spur', 
+    'slope', 'hollow', 'footslope', 'valley', 'pit'
+  ))
+  
+  geo_val <- terra::extract(geomorphon, x_spat)
+  names(geo_val) <- c('ID', 'unit')
+  geo_val <- dplyr::left_join(geo_val, geoLKPtab, by = 'unit') |>
+    dplyr::select(-unit)
+  
+  ele_val <- terra::extract(elevation, x_spat)
+  
+  values <- data.frame(cbind(
[…]

### 2023-05-12 scripts/project_template.Rmd
prototype geomorpho grabber
evaluation with political

@@
 dummy_pts <- data.frame(
-  collectionNo = paste0(rep('RCB', 500), seq(1:500)),
-  day = c(sample(1:31, size = 500, replace = T)),
-  month = c(sample(5:10, size = 500, replace = T)),
+  collectionNo = paste0(rep('RCB', 1000), seq(1:1000)),
+  day = c(sample(1:31, size = 1000, replace = T)),
+  month = c(sample(5:10, size = 1000, replace = T)),
   year = 2023,
-  x = runif(500, min = -120, max = -100), 
-  y = runif(500, min = 35, max = 48)
+  x = runif(1000, min = -120, max = -100), 
+  y = runif(1000, min = 35, max = 48)
 ) %>% sf::st_as_sf(coords = c(x = 'x', y = 'y'), crs = 4326)
@@
 
-out <- political_grabber(dummy_pts)
-out <- place_grabber(dummy_pts)
+out1 <- political_grabber(dummy_pts, 'collectionNo')
+out2 <- place_grabber(dummy_pts)
+out3 <- physical_grabber(dummy_pts)
 
-rm(out)
-# if within 0.5 km of place, use
-# if dist > 0.5 km direct
+out <- out1 %>% 
+  dplyr::mutate(
+    Gen = paste0(
+      Country, ', ', State, ', ', County, ' Co., ', Mang_Name, " ", Unit_Nm, " ", trs), 
+    Gen = stringr::str_replace_all(Gen, "NA", ""), 
+    Gen = stringr::str_replace_all(Gen, "  ", ""))
 

### 2023-05-12 scripts/functions.R

@@ -277,7 +277,9 @@ map_maker <- function(x, states, path, collection_col){
 #' this function grabs information on the state, county, and township of collections
 #' @param x an sf data frame of collection points
 #' @param y a column which unambiguously identifies each collection
-political_grabber <- function(x) {
+political_grabber <- function(x, y) {
+  
+  y_quo <- rlang::enquo(y)
   
   political <- sf::st_read('../geodata/political/political.shp', quiet = T)
   allotment <- sf::st_read('../geodata/allotments/allotments.shp', quiet = T)
@@ -291,18 +293,23 @@ political_grabber <- function(x) {
   x <- sf::st_join(x, ownership)
   
   x_plss <- sf::st_transform(x, sf::st_crs(plss))
-  x_plss <- sf::st_join(x_plss, plss) %>% 
-    sf::st_drop_geometry()
+  x_plss <- sf::st_join(x_plss, plss) |>
+    sf::st_drop_geometry() %>% 
+    dplyr::select(y, trs)
   
-  x_vars <- dplyr::left_join(x, x_plss) %>% 
-    relocate(any_of(c('State', 'County', 'Mang_Name', 'Unit_Nm', 'trs')),
-             .before = geometry)
+  x_vars <- dplyr::left_join(x, x_plss, by = y) |> 
+    dplyr::mutate(Country = 'U.S.A.') |> 
+    dplyr::relocate(any_of(c('Country', 'State', 'County', 'Mang_Name', 'Unit_Nm', 'trs')),
+             .before = geometry) |>
+    dplyr::distinct(.keep_all = T) |> # some points may theoretically fall on an exact border
+    dplyr::group_by( .data[[y]] ) |>
+    dplyr::slice_head(n = 1) |> 
+    dplyr::ungroup()
+  
   return(x_vars)
   
   rm(political, allotment, plss, ownership)
 }
-
-
 
 #' gather place and site information
 #' 
@@ -328,13 +335,13 @@ place_grabber <- function(x) {
   
   places <- data.frame(
     'place' = sf::st_drop_geometry(nearest_place), distance, azimuth)  |>
-    rename(place = fetr_nm)
+    dplyr::rename(place = fetr_nm)
     
   x <- dplyr::bind_cols(x, places)|>
     dplyr::mutate(area_from =
                     paste0('From ', place, ' ', distance, ' km at ', azimuth, '°.'),
                   .before = geometry) |>
[…]

### 2023-05-12 scripts/project_template.Rmd
included string parsing to describe site

@@
 out3 <- physical_grabber(dummy_pts)
-
-out <- out1 %>% 
-  dplyr::mutate(
-    Gen = paste0(
-      Country, ', ', State, ', ', County, ' Co., ', Mang_Name, " ", Unit_Nm, " ", trs), 
-    Gen = stringr::str_replace_all(Gen, "NA", ""), 
-    Gen = stringr::str_replace_all(Gen, "  ", ""))
-
 ```

### 2023-05-12 scripts/functions.R

@@ -301,10 +301,20 @@ political_grabber <- function(x, y) {
     dplyr::mutate(Country = 'U.S.A.') |> 
     dplyr::relocate(any_of(c('Country', 'State', 'County', 'Mang_Name', 'Unit_Nm', 'trs')),
              .before = geometry) |>
-    dplyr::distinct(.keep_all = T) |> # some points may theoretically fall on an exact border
+    dplyr::distinct(.keep_all = T) |> # with large enough sample size some points fall on an exact border
     dplyr::group_by( .data[[y]] ) |>
     dplyr::slice_head(n = 1) |> 
     dplyr::ungroup()
+  
+  x_vars <- x_vars %>% 
+    dplyr::mutate(
+      Gen = paste0(
+        Country, ', ', State, ', ', County, ' Co., ', Mang_Name, " ", Unit_Nm, " ", trs), 
+      Gen = stringr::str_replace_all(Gen, "NA", ""), 
+      Gen = stringr::str_replace_all(Gen, "  ", ""),
+      Gen = stringr::str_trim(Gen),
+      Gen = stringr::str_remove(Gen, ",$"), 
+      .before = geometry)
   
   return(x_vars)
   

### 2023-05-15 scripts/project_template.Rmd
added text

@@
 
-```{r}
+```{r dummy data}
 dummy_pts <- data.frame(
@@
 
+# first convert from american date format (m%d%year%) to ISO 8601 for both collection and determination date
+
+# second from ISO 8601 prepare notation dates for day of collection and date of determination in text format
+
+```{r Date parsing}
+
+#' create herbarium format dates for specimens
+#' 
+#' this function will return up to two additional columns with x. 'coll_date_text' a text format for the date of collection and 'det_date_text" a shorter text format date for identification
+#' @param x an (sf/tibble/) data frame with both the collection date
+#' and determination date
+#' @param coll_date date of collection, expected to be of the format 'DD/MM/YYYY', minor checks for compliance to the format will be carried out before returning an error.
+#' @param det_date date of determination, same format and processes as above. 
+date_parser <- function(x, coll_date, det_date){
+
+    if(missing(det_date)) {
+
+    } else {
+
+    }
+
+}
+
+# FIRST CHECK IF FORMAT IS CORRECT ... 
+IsDate <- function(mydate, date.format = "%m/%d/%y") {
+  tryCatch(!is.na(as.Date(mydate, date.format)),  
+           error = function(err) {FALSE})  
+}
+
+date <- c('12/11/2003','07/09/2023', '13/01/2016')
+IsDate(date)
+
+stringr::str_split(date, pattern = '/', simplify = T)[,2] # capture the month
+
+date_text <- function(x, col2text){
+
+  text_col <- !!enquo(col2text)
+
+  month_tab <- data.frame(Month_Num = sprintf("%02d", 1:12), Month_Abb = month.abb)
+  y <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%Y")
+  d <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%d")
+  m <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%m")
+  m_text <- month_tab[m == month_tab$Month_Num, 'Month_Abb']
+
+ # text <- paste0(d, ' ', m, ', ', y)
+  return(m)
+
+}
+
+fourtet <- data.frame(date, tet = c('abc', 'def', 'ggg'))
+
+t <- date_text(fourtet, 'date')
+
+t %>% 
+  mutate(dtest = date_text(., ))
+
+t == month_tab$Month_Num
+```
+
 ```{r Acquire administrative geodata}

### 2023-05-15 scripts/project_template.Rmd
function to format for label type date

@@
 
-stringr::str_split(date, pattern = '/', simplify = T)[,2] # capture the month
+#' take ISO 8601 date and make it written herbarium label format
+#' 
+#' @param x a data frame with dates
+#' @param y the data frame holding Internationally formatted dates (day, month, year)
+date2text <- function(x, col2text) {
+  x$col2text <- as.Date(x[, col2text], format = "%d/%m/%Y")
+  m <- as.numeric(format(x$col2text, "%m"))
 
-date_text <- function(x, col2text){
+  text <- paste0(format(x$col2text, "%d"), # just grab day of month here
+                 ' ',
+                 month.abb[m], # grab the abbrevation for the month here
+                 ', ',
+                 format(x$col2text, "%Y")) # grab the year here
 
-  text_col <- !!enquo(col2text)
-
-  month_tab <- data.frame(Month_Num = sprintf("%02d", 1:12), Month_Abb = month.abb)
-  y <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%Y")
-  d <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%d")
-  m <- format(as.Date(x[,col2text], format="%d/%m/%Y"), "%m")
-  m_text <- month_tab[m == month_tab$Month_Num, 'Month_Abb']
-
- # text <- paste0(d, ' ', m, ', ', y)
-  return(m)
+  return(text)
 
@@
 
-t <- date_text(fourtet, 'date')
+fourtet %>% 
+  mutate(dtest = date2text(., 'date'))
 
-t %>% 
-  mutate(dtest = date_text(., ))
-
-t == month_tab$Month_Num
 ```

### 2023-05-17 scripts/project_template.Rmd
converting date functions to tidy

--- 2023-05-12 scripts/project_template.Rmd
@@
 
+ first convert from american date format (m%d%year%) to ISO 8601 for both collection and determination date
+
+ second from ISO 8601 prepare notation dates for day of collection and date of determination in text format
+
+```{r Date parsing}
+
+#' create herbarium format dates for specimens
+#' 
+#' this function will return up to two additional columns with x. 'coll_date_text' a text format for the date of collection and 'det_date_text" a shorter text format date for identification
+#' @param x an (sf/tibble/) data frame with both the collection date
+#' and determination date
+#' @param coll_date date of collection, expected to be of the format 'MM/DD/YYYY', minor checks for compliance to the format will be carried out before returning an error.
+#' @param det_date date of determination, same format and processes as above. 
+#' @returns coll_date ISO 8601, det_date ISO 8601, month, day, year, coll_date_text, det_date_text columns for printing onto the label
+date_parser <- function(x, coll_date, det_date){
+
+  coll_date_q <- enquo(coll_date)
+  det_date_q <- enquo(det_date)
+
+  mdy2dmy <- function(x){format(as.Date(x, format = "%m/%d/%Y"), "%d/%m/%Y")}
+
+  x_dmy <- x |> 
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), mdy2dmy,
+           .names = "{col}_dmy"), .before = ncol(.)) |>
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), ~ lubridate::month(x),
+           .names = "{col}_mo"))
+
+  # add in collection dates here
+#    if(missing(det_date)) {
+#      with_date <- dplyr::mutate(x_dmy, coll_date_text = date2text(x_dmy, 'coll_date_dmy'))
+##    } else {
+#     with_date <-  dplyr::mutate(x_dmy, 
+#           coll_date_text = date2text(x_dmy, 'coll_date_dmy'),
+#           det_date_text  = date2text(x_dmy, 'coll_date_dmy'))
+#    }
+
+  return(x_dmy)
+}
+
+```
+
+```{r}
+
+fourtet <- data.frame(col_date = c('12/23/2003','07/09/2023', '1/13/2016'),
+                   #   det_date = c('12/25/2003','07/09/2023', '1/13/2016'),
+                      tet = c('abc', 'def', 'ggg'))
+
+date_parser(fourtet, coll_date = 'col_date')
+
+```
+
 ```{r Acquire administrative geodata}
@@
 out3 <- physical_grabber(dummy_pts)
+
+browseVignettes("ggtree")
 ```

### 2023-05-17 scripts/functions.R

@@ -432,6 +432,22 @@ physical_grabber <- function(x) {
 
 
 
-
+#' take international format date and make it written herbarium label format
+#' 
+#' @param x a data frame with dates
+#' @param y the data frame holding Internationally formatted dates (day, month, year)
+date2text <- function(x, col2text) {
+  x$col2text <- as.Date(x[, col2text], format = "%d/%m/%Y")
+  m <- as.numeric(format(x$col2text, "%m"))
+  
+  text <- paste0(format(x$col2text, "%d"), # just grab day of month here
+                 ' ',
+                 month.abb[m], # grab the abbrevation for the month here
+                 ', ',
+                 format(x$col2text, "%Y")) # grab the year here
+  
+  return(text)
+  
+}
 
 

### 2023-05-18 scripts/project_template.Rmd
reworked dates with lubridate

@@
 
+  names_v <- c('_dmy', '_day', '_mo',  '_yr', '_text')
+  column_names <- c(paste0(coll_date, names_v), paste0(det_date, names_v) )
+
   mdy2dmy <- function(x){format(as.Date(x, format = "%m/%d/%Y"), "%d/%m/%Y")}
@@
    x_dmy <- x |> 
-    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), mdy2dmy,
-           .names = "{col}_dmy"), .before = ncol(.)) |>
-    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), ~ lubridate::month(x),
-           .names = "{col}_mo"))
-
-  # add in collection dates here
-#    if(missing(det_date)) {
-#      with_date <- dplyr::mutate(x_dmy, coll_date_text = date2text(x_dmy, 'coll_date_dmy'))
-##    } else {
-#     with_date <-  dplyr::mutate(x_dmy, 
-#           coll_date_text = date2text(x_dmy, 'coll_date_dmy'),
-#           det_date_text  = date2text(x_dmy, 'coll_date_dmy'))
-#    }
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), lubridate::mdy, .names = "{.col}_dmy")) |>
+    mutate(across(ends_with('_dmy'), ~ lubridate::month(.), .names = "{.col}_mo"),
+           across(ends_with('_dmy'), ~ lubridate::day(.), .names = "{.col}_day"),
+           across(ends_with('_dmy'), ~ lubridate::year(.), .names = "{.col}_yr")) |>
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), date2text, .names = '{.col}_text')) |>
+    rename_with( ~ stringr::str_remove(., '_dmy'), matches("_dmy_.*$")) |>
+    relocate(any_of(column_names), .before = last_col())
 
@@
 fourtet <- data.frame(col_date = c('12/23/2003','07/09/2023', '1/13/2016'),
-                   #   det_date = c('12/25/2003','07/09/2023', '1/13/2016'),
+                      det_date = c('12/25/2003','07/09/2023', '1/13/2016'),
                       tet = c('abc', 'def', 'ggg'))
 
-date_parser(fourtet, coll_date = 'col_date')
+date_parser(fourtet, coll_date = 'col_date', det_date = 'det_date')
 

### 2023-05-18 scripts/functions.R

@@ -435,19 +435,16 @@ physical_grabber <- function(x) {
 #' take international format date and make it written herbarium label format
 #' 
 #' @param x a data frame with dates
-#' @param y the data frame holding Internationally formatted dates (day, month, year)
-date2text <- function(x, col2text) {
-  x$col2text <- as.Date(x[, col2text], format = "%d/%m/%Y")
-  m <- as.numeric(format(x$col2text, "%m"))
+date2text <- function(x) {
   
-  text <- paste0(format(x$col2text, "%d"), # just grab day of month here
+  x1 <- lubridate::mdy(x)
+  
+  text <- paste0(lubridate::day(x1), # just grab day of month here
                  ' ',
-                 month.abb[m], # grab the abbrevation for the month here
+                 lubridate::month(x1, abbr = T, label = T), # grab the abbreviation for the month here
                  ', ',
-                 format(x$col2text, "%Y")) # grab the year here
-  
+                 lubridate::year(x1)) # grab the year here
   return(text)
-  
 }
 
 

### 2023-05-18 scripts/project_template.Rmd
wrapper around parzer fleshed out

@@
   year = 2023,
-  x = runif(1000, min = -120, max = -100), 
-  y = runif(1000, min = 35, max = 48)
-) %>% sf::st_as_sf(coords = c(x = 'x', y = 'y'), crs = 4326)
+  longitude = runif(1000, min = -120, max = -100), 
+  latitude = runif(1000, min = 35, max = 48)
+) %>% sf::st_as_sf(coords = c(x = 'longitude', y = 'latitude'), crs = 4326, remove = F)
 
@@
 
- first convert from american date format (m%d%year%) to ISO 8601 for both collection and determination date
+```{r coordinate parsing if required}
 
- second from ISO 8601 prepare notation dates for day of collection and date of determination in text format
+library(parzer)
 
-```{r Date parsing}
+#' this function parses coordinates from DMS to decimal degrees
+#'
+#'@param x an input data frame to apply operations too
+#'@param lat a name of the column holding the latitude values
+#'@param long a name of the colymn holding the longitude values
+#'@param dms are coordinates in degrees minutes seconds? TRUE for yes, FALSE for decimal degrees
+dms2dd <- function(x, lat, long, dms){
 
-#' create herbarium format dates for specimens
-#' 
-#' this function will return up to two additional columns with x. 'coll_date_text' a text format for the date of collection and 'det_date_text" a shorter text format date for identification
-#' @param x an (sf/tibble/) data frame with both the collection date
-#' and determination date
-#' @param coll_date date of collection, expected to be of the format 'MM/DD/YYYY', minor checks for compliance to the format will be carried out before returning an error.
-#' @param det_date date of determination, same format and processes as above. 
-#' @returns coll_date ISO 8601, det_date ISO 8601, month, day, year, coll_date_text, det_date_text columns for printing onto the label
-date_parser <- function(x, coll_date, det_date){
+  if(missing(lat)){
+    lat = colnames(dummy_pts)[grep('lat', colnames(dummy_pts))] }
+  if(missing(long)){
+    long = colnames(dummy_pts)[grep('long', colnames(dummy_pts))] }
 
-  coll_date_q <- enquo(coll_date)
-  det_date_q <- enquo(det_date)
+  if(missing(dms)){
+    dms = long == parzer::parse_lon(long)
+  }
 
-  names_v <- c('_dmy', '_day', '_mo',  '_yr', '_text')
-  column_names <- c(paste0(coll_date, names_v), paste0(det_date, names_v) )
+  if(dms == T){
+    x$latitude_dd = parzer::parse_lat(lat)
+    x$longitude_dd = parzer::pase_lon(long)
+  } else{
+    colnames(x)[which(names(x) == lat)] <- 'latitude_dd'
+    colnames(x)[which(names(x) == long)] <- 'latitude_dd'
+  }
 
-  mdy2dmy <- function(x){format(as.Date(x, format = "%m/%d/%Y"), "%d/%m/%Y")}
+#  x[,c('latitude_dd', 'longitude_dd')] <- 
+#    sapply(x[c('latitude_dd', 'latitude_dd')], as.numeric)
 
-   x_dmy <- x |> 
-    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), lubridate::mdy, .names = "{.col}_dmy")) |>
-    mutate(across(ends_with('_dmy'), ~ lubridate::month(.), .names = "{.col}_mo"),
-           across(ends_with('_dmy'), ~ lubridate::day(.), .names = "{.col}_day"),
-           across(ends_with('_dmy'), ~ lubridate::year(.), .names = "{.col}_yr")) |>
-    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), date2text, .names = '{.col}_text')) |>
-    rename_with( ~ stringr::str_remove(., '_dmy'), matches("_dmy_.*$")) |>
-    relocate(any_of(column_names), .before = last_col())
+ # if(x$longitude_dd > 0){longitude_dd * -1}; 
+#  if(x$latitude_dd < 0){latitude_dd * -1}
 
-  return(x_dmy)
+  # now overwrite the original DMS values in our exact format
+
+  return(x)
+
+  latitude_dms = paste0(
+    'N ', parzer::pz_degree(x$latitude_dd),
+    '°', round(parzer::pz_minute(x$latitude_dd), 2),
+    "'", round(parzer::pz_second(x$latitude_dd), 2)
+  )
+
+  longitude_dms = paste0(
+    'W ', parzer::pz_degree(x$longitude_dd), 
+    '°', round(parzer::pz_minute(x$longitude_dd), 2), 
+    "'", round(parzer::pz_second(x$longitude_dd), 2)
+  )
+
 }
+
+out <- dms2dd(dummy_pts)
+
+# str detect the lat column
+# str detect the long column
+# ensure input lat is North
+# ensure input long is West
+# str detect the datum column, else assume WGS84
 
@@
 
-```{r}
+```{r date parsing}
 
@@
 
-browseVignettes("ggtree")
 ```

### 2023-05-18 scripts/functions.R

@@ -447,4 +447,32 @@ date2text <- function(x) {
   return(text)
 }
 
+#' create herbarium format dates for specimens
+#' 
+#' this function will return up to two additional columns with x. 'coll_date_text' a text format for the date of collection and 'det_date_text" a shorter text format date for identification
+#' @param x an (sf/tibble/) data frame with both the collection date
+#' and determination date
+#' @param coll_date date of collection, expected to be of the format 'MM/DD/YYYY', minor checks for compliance to the format will be carried out before returning an error.
+#' @param det_date date of determination, same format and processes as above. 
+#' @returns original data frame plus: x_dmy, x_day, x_month, x_year, x_text, det_date_text, columns for both parameters which are supplied as inputs
+date_parser <- function(x, coll_date, det_date){
+  
+  coll_date_q <- enquo(coll_date)
+  det_date_q <- enquo(det_date)
+  
+  names_v <- c('_dmy', '_day', '_mo',  '_yr', '_text')
+  column_names <- c(paste0(coll_date, names_v), paste0(det_date, names_v) )
+  
+  x_dmy <- x |> 
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), lubridate::mdy, .names = "{.col}_dmy")) |>
+    mutate(across(ends_with('_dmy'), ~ lubridate::month(.), .names = "{.col}_mo"),
+           across(ends_with('_dmy'), ~ lubridate::day(.), .names = "{.col}_day"),
+           across(ends_with('_dmy'), ~ lubridate::year(.), .names = "{.col}_yr")) |>
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), date2text, .names = '{.col}_text')) |>
+    rename_with( ~ stringr::str_remove(., '_dmy'), matches("_dmy_.*$")) |>
+    relocate(any_of(column_names), .before = last_col())
+  
+  return(x_dmy)
+}
+
 

### 2023-05-18 scripts/project_template.Rmd
function to parse dms to dd and create dms finished, onto making sf obs

@@
 
-library(parzer)
-
-#' this function parses coordinates from DMS to decimal degrees
-#'
-#'@param x an input data frame to apply operations too
-#'@param lat a name of the column holding the latitude values
-#'@param long a name of the colymn holding the longitude values
-#'@param dms are coordinates in degrees minutes seconds? TRUE for yes, FALSE for decimal degrees
-dms2dd <- function(x, lat, long, dms){
-
-  if(missing(lat)){
-    lat = colnames(dummy_pts)[grep('lat', colnames(dummy_pts))] }
-  if(missing(long)){
-    long = colnames(dummy_pts)[grep('long', colnames(dummy_pts))] }
-
-  if(missing(dms)){
-    dms = long == parzer::parse_lon(long)
-  }
-
-  if(dms == T){
-    x$latitude_dd = parzer::parse_lat(lat)
-    x$longitude_dd = parzer::pase_lon(long)
-  } else{
-    colnames(x)[which(names(x) == lat)] <- 'latitude_dd'
-    colnames(x)[which(names(x) == long)] <- 'latitude_dd'
-  }
-
-#  x[,c('latitude_dd', 'longitude_dd')] <- 
-#    sapply(x[c('latitude_dd', 'latitude_dd')], as.numeric)
-
- # if(x$longitude_dd > 0){longitude_dd * -1}; 
-#  if(x$latitude_dd < 0){latitude_dd * -1}
-
-  # now overwrite the original DMS values in our exact format
-
-  return(x)
-
-  latitude_dms = paste0(
-    'N ', parzer::pz_degree(x$latitude_dd),
-    '°', round(parzer::pz_minute(x$latitude_dd), 2),
-    "'", round(parzer::pz_second(x$latitude_dd), 2)
-  )
-
-  longitude_dms = paste0(
-    'W ', parzer::pz_degree(x$longitude_dd), 
-    '°', round(parzer::pz_minute(x$longitude_dd), 2), 
-    "'", round(parzer::pz_second(x$longitude_dd), 2)
-  )
-
-}
-
  out <- dms2dd(dummy_pts)
 
-# str detect the lat column
-# str detect the long column
-# ensure input lat is North
-# ensure input long is West
-# str detect the datum column, else assume WGS84
-
+ x <- x |>
+    dplyr::relocate(any_of(
+      c('longitude_dms', 'latitude_dms', 'latitude_dd', 'longitude_dd')), 
+    .before = last_col())
 ```

### 2023-05-18 scripts/functions.R

@@ -476,3 +476,53 @@ date_parser <- function(x, coll_date, det_date){
 }
 
 
+
+#' this function parses coordinates from DMS to decimal degrees
+#'
+#'@param x an input data frame to apply operations too
+#'@param lat a name of the column holding the latitude values
+#'@param long a name of the colymn holding the longitude values
+#'@param dms are coordinates in degrees minutes seconds? TRUE for yes, FALSE for decimal degrees
+dms2dd <- function(x, lat, long, dms){
+  
+  # identify columns if they were not supplied
+  if(missing(lat)){
+    lat = colnames(dummy_pts)[grep('lat', colnames(dummy_pts))] }
+  if(missing(long)){
+    long = colnames(dummy_pts)[grep('long', colnames(dummy_pts))] }
+  
+  # test for DMS format if not supplied
+  suppressWarnings(  if(missing(dms)){
+    dms = long == parzer::parse_lon(long)
+  })
+  
+  # convert dms to dd, or rename input columns to dd
+  if(dms == T){
+    x$latitude_dd = parzer::parse_lat(lat)
+    x$longitude_dd = parzer::pase_lon(long)
+  } else{
+    colnames(x)[which(names(x) == lat)] <- 'latitude_dd'
+    colnames(x)[which(names(x) == long)] <- 'longitude_dd'
+  }
+  # ensure the DD signs are appropriate for domain
+  x$latitude_dd <- abs(x$latitude_dd)
+  x$longitude_dd <- abs(x$longitude_dd) * -1
+  
+  # now overwrite the original DMS values in our exact formaty
+  
+  x$latitude_dms = paste0(
+    'N ', parzer::pz_degree(x$latitude_dd),
+    '°', round(parzer::pz_minute(x$latitude_dd), 2),
+    "'", round(parzer::pz_second(x$latitude_dd), 2)
+  )
+  
+  x$longitude_dms = paste0(
+    'W ', parzer::pz_degree(x$longitude_dd), 
+    '°', round(parzer::pz_minute(x$longitude_dd), 2), 
+    "'", round(parzer::pz_second(x$longitude_dd), 2)
+  )
+  
+  return(x)
+  
+}
+

### 2023-05-18 scripts/project_template.Rmd
function to convert datum to sf objects and convert all 2 wgs84

@@
   year = 2023,
+  datum = sample(c('nad83', 'nad27', 'wgs84'), size = 1000, replace = T),
   longitude = runif(1000, min = -120, max = -100), 
   latitude = runif(1000, min = 35, max = 48)
-) %>% sf::st_as_sf(coords = c(x = 'longitude', y = 'latitude'), crs = 4326, remove = F)
+)
 
@@
 
-```{r coordinate parsing if required}
+```{r coordinate parsing and make object spatial in wgs84}
 
 out <- dms2dd(dummy_pts)
+d <- coords2sf(out)
 
- x <- x |>
-    dplyr::relocate(any_of(
-      c('longitude_dms', 'latitude_dms', 'latitude_dd', 'longitude_dd')), 
-    .before = last_col())
 ```

### 2023-05-18 scripts/functions.R

@@ -526,3 +526,42 @@ dms2dd <- function(x, lat, long, dms){
   
 }
 
+
+#' create an sf object row by row to reflect different datum
+#' 
+#' @param x a dataframe which has undergone dms2dd function
+#' @param datum a column, holding the datum values for each observation, works for WGS84, NAD83, NAD27
+coords2sf <- function(x, datum){
+  
+  if(missing(datum)){ # identify datum information
+    datum = colnames(dummy_pts)[grep('datum', colnames(dummy_pts))] }
+  
+  if(length(datum) == 0){x$datum = 'WGS84'} # if column doesn't exist create
+  
+  dat_check <- function(x){if(grepl('nad.*27', ignore.case = T, x = x)) {
+    x = "NAD27"} else if(grepl('nad.*83', ignore.case = T, x = x)) {
+      x = 'NAD83'} else if(grepl('wgs', ignore.case = T, x = x)) {
+        x = "WGS84"
+      } else {x = 'WGS84'}
+  }
+  # now ensure the datum is appropriate for each point
+  x$datum <- sapply(x$datum, dat_check)
+  
+  crs_lkp = data.frame(
+    datum = c('NAD27', 'NAD83', 'WGS84'), 
+    crs = c(4267, 4269, 4326))
+  
+  x <- dplyr::left_join(x, crs_lkp, by = 'datum') 
+  
+  # separate all points by there datum
+  dat_list <- split(x, f = x$datum)
+  dat_list <- purrr::map(dat_list, . %>%  
+                           sf::st_as_sf(., coords = c(x = 'longitude_dd', y = 'latitude_dd'), 
+                                        crs = .$crs[1], remove = F) %>% 
+                           sf::st_transform(4326)) |>
+    dplyr::bind_rows() %>% 
+    dplyr::select(-crs) %>% 
+    dplyr::mutate(datum = 'WGS84', .before = geometry)
+  
+  return(dat_list)
+}

### 2023-05-18 scripts/project_template.Rmd
grab input date columns for relocation

@@
   collectionNo = paste0(rep('RCB', 1000), seq(1:1000)),
-  day = c(sample(1:31, size = 1000, replace = T)),
+  day = c(sample(1:30, size = 1000, replace = T)),
   month = c(sample(5:10, size = 1000, replace = T)),
@@
   latitude = runif(1000, min = 35, max = 48)
-)
+) %>% 
+  mutate(col_date = paste(month, day, year, sep = "/"))
 
@@
 out <- dms2dd(dummy_pts)
-d <- coords2sf(out)
+out <- coords2sf(out)
 
@@
 
-fourtet <- data.frame(col_date = c('12/23/2003','07/09/2023', '1/13/2016'),
-                      det_date = c('12/25/2003','07/09/2023', '1/13/2016'),
-                      tet = c('abc', 'def', 'ggg'))
-
-date_parser(fourtet, coll_date = 'col_date', det_date = 'det_date')
+out <- date_parser(out, coll_date = 'col_date')
 
@@
 
-```{r Acquire administrative geodata}
+```{r Acquire geodata}
 
@@
 ```
+
+```{r gather taxonomic data}
+
+```

### 2023-05-18 scripts/functions.R

@@ -461,7 +461,14 @@ date_parser <- function(x, coll_date, det_date){
   det_date_q <- enquo(det_date)
   
   names_v <- c('_dmy', '_day', '_mo',  '_yr', '_text')
-  column_names <- c(paste0(coll_date, names_v), paste0(det_date, names_v) )
+  if(missing(det_date)){
+    column_names <- c(coll_date, paste0(coll_date, names_v))
+  } else {
+    column_names <- c(coll_date, 
+                      paste0(coll_date, names_v), 
+                        det_date, 
+                        paste0(det_date, names_v))
+  }
   
   x_dmy <- x |> 
     mutate(across(.cols = c(!!coll_date_q, !!det_date_q), lubridate::mdy, .names = "{.col}_dmy")) |>
@@ -470,7 +477,7 @@ date_parser <- function(x, coll_date, det_date){
            across(ends_with('_dmy'), ~ lubridate::year(.), .names = "{.col}_yr")) |>
     mutate(across(.cols = c(!!coll_date_q, !!det_date_q), date2text, .names = '{.col}_text')) |>
     rename_with( ~ stringr::str_remove(., '_dmy'), matches("_dmy_.*$")) |>
-    relocate(any_of(column_names), .before = last_col())
+    relocate(any_of(column_names), .before = geometry)
   
   return(x_dmy)
 }
@@ -521,6 +528,8 @@ dms2dd <- function(x, lat, long, dms){
     '°', round(parzer::pz_minute(x$longitude_dd), 2), 
     "'", round(parzer::pz_second(x$longitude_dd), 2)
   )
+  
+  x$longitude_dms <- gsub('-', "", x$longitude_dms)
   
   return(x)
   

### 2023-05-19 scripts/functions.R
added mountain names to place

@@ -332,8 +332,8 @@ place_grabber <- function(x) {
   places <- sf::st_read('../geodata/places/places.shp', quiet = T)
   
   x <- sf::st_join(x, mountains)
-
-  nearest_place <- places[sf::st_nearest_feature(dummy_pts, places),]
+  
+  nearest_place <- places[sf::st_nearest_feature(x, places),]
   distance <- round(
     as.numeric(
       sf::st_distance(x, nearest_place, by_element = T)),
@@ -348,10 +348,13 @@ place_grabber <- function(x) {
     dplyr::rename(place = fetr_nm)
     
   x <- dplyr::bind_cols(x, places)|>
+    dplyr::mutate(MapName = str_remove(MapName, ' \\(nn\\)')) |>
     dplyr::mutate(area_from =
-                    paste0('From ', place, ' ', distance, ' km at ', azimuth, '°.'),
+                    paste0('From ', place, ' ', distance, ' km at ', azimuth, '°. ', 
+                           MapName, '.'),
                   .before = geometry) |>
-    dplyr::select(-place, -distance, -azimuth) 
+    dplyr::mutate(area_from = str_remove(area_from, ' NA[.]')) |>
+    dplyr::select(-place, -distance, -azimuth, -MapName) 
   
   return(x)
 }
@@ -417,9 +420,11 @@ physical_grabber <- function(x) {
   el_cols <- c('elevation_m', 'elevation_ft')
   values[el_cols] <- lapply(values[el_cols], scales::comma)
   
-  cols <- c('elevation_m', 'elevation_ft', 'aspect', 'slope', 'geomorphon', 'geology') 
+  cols <- c('elevation_m', 'elevation_ft', 'aspect', 'slope', 
+            'geomorphon', 'geology') 
   
   object <- dplyr::bind_cols(x, values) |> 
+    dplyr::select(-ID) |>
     dplyr::relocate(tidyselect::all_of(cols), .before = geometry) |> 
     dplyr::mutate(
       physical_environ = 

### 2023-05-19 scripts/project_template.Rmd
added mountain names to place
fixed hard coded variables added function to identify species lacking taxonomic results

@@
 out <- coords2sf(out)
-
 ```
@@
 out <- date_parser(out, coll_date = 'col_date')
-
 ```
@@
 
-places <- sf::st_read('../geodata/places/places.shp', quiet = T)
-
-out1 <- political_grabber(dummy_pts, 'collectionNo')
-out2 <- place_grabber(dummy_pts)
-out3 <- physical_grabber(dummy_pts)
+out1 <- political_grabber(out, 'collectionNo')
+out2 <- place_grabber(out1)
+out3 <- physical_grabber(out2)
 
@@
 
+t <- taxize::names_list('species')
+t <- c(t, 'agoseris potter')
+
+tout <- spell_check(t)
+tres <- lapply(tout$Result, powo_searcher) %>% 
+  bind_rows()
+
+notFound(tres)
+
+notFound <- function(x){
+
+  library(crayon)
+
+  row_no <- unlist( apply(FUN = grep, X = x, MARGIN = 2, pattern = 'NOT FOUND') )
+  rows <- unique(row_no) # these rows had complications... 
+  not_found <- x[rows, 'query'] # these records were not found.
+
+  recs <- cat(
+  'The record: ' %+%
+  blue$underline$bold(not_found) %+%
+  blue$bold(' (row ') %+% blue$underline$bold(rows) %+% ') did not have all data retrieved.\n')
+
+  message(recs)
+
+}
+
 ```
+
+```{r make herbarium label maps, eval = F}
+
+dummy_pts <- data.frame(
+  collectionNo = paste0(rep('RCB', 20), seq(1:20)),
+  x = runif(20, min = -120, max = -85), 
+  y = runif(20, min = 35, max = 48)
+) %>% sf::st_as_sf(coords = c('x', 'y'), crs = 4269)
+
+states_sf <- st_read('../geodata/political/political.shp', quiet = T)
+
+dummy_pts_l <- split(dummy_pts, dummy_pts$collectionNo)
+out <- lapply(dummy_pts_l, map_maker, states_sf, '../results', collection_col = 'collectionNo')
+
+```

### 2023-05-19 scripts/taxonomy_setup.Rmd
fixed hard coded variables added function to identify species lacking taxonomic results

@@
 ```
-
-```{r}
-library(sf)
-
-dummy_pts <- data.frame(
-  collectionNo = paste0(rep('RCB', 20), seq(1:20)),
-  x = runif(20, min = -120, max = -85), 
-  y = runif(20, min = 35, max = 48)
-) %>% sf::st_as_sf(coords = c('x', 'y'), crs = 4269)
-
-states_sf <- st_read('../geodata/political/political.shp', quiet = T)
-
-dummy_pts_l <- split(dummy_pts, dummy_pts$collectionNo)
-
-out <- lapply(dummy_pts_l, map_maker, states_sf, '../results', collection_col = 'collectionNo')
-
-ggplot() + 
-    geom_sf(data = states_sf, fill = NA, color = 'grey15') + 
-    geom_sf(data = dummy_pts, size = 0.5)
-
-```

### 2023-05-19 scripts/functions.R

@@ -37,7 +37,7 @@ try_again <- function(x) {
   only_binomial <- unlist(stringr::str_split(q, pattern = " "))
   only_binomial <- paste(only_binomial[1], only_binomial[2])
   
-  results <- search_powo(only_binomial)
+  results <- kewr::search_powo(only_binomial)
   return(results)
 }
 
@@ -104,7 +104,7 @@ result_grabber <- function(x) {
 #' notes: results are observed to fail for valid infraspecies on Kew's end, and they seem not
 #' to mention valid infraspecies. 
 powo_searcher <- function(x) {
-  query_results <- search_powo(x)
+  query_results <- kewr::search_powo(x)
   
   if (is.null(query_results[["results"]])) {
     second_try <- try_again(query_results)
@@ -164,11 +164,15 @@ autonym <- function(genus, epithet, infrarank, infraspecies){
 #' @param x a vector of species names
 spell_check <- function(x) {
   
+  sppLKPtab <- read.csv('../taxonomic_data/species_lookup_table.csv')
+  epiLKPtab <- read.csv('../taxonomic_data/epithet_lookup_table.csv')
+  genLKPtab <- read.csv('../taxonomic_data/genus_lookup_table.csv')
+  
   pieces <- unlist(stringr::str_split(x, pattern = " "))
   genus <- pieces[1] ; species <- pieces[2]
   binom <- paste(genus, species)
   
-  # infraspecies should be found without much hassle due to their length
+  # infra species should be found without much hassle due to their length
   if(length(pieces) == 4){
     infras <- na.omit(epiLKPtab)
     full_name <- paste(genus, species, 
@@ -189,8 +193,8 @@ spell_check <- function(x) {
   # species can become difficult due to their short  names, e.g. 'Poa annua'
   } else {
   
-    if (any(grep( x = cla$scientificName,pattern = binom, fixed = T))) {
-      return(data.frame(Query = x, Result = binom, Match = 'exact'))
+    if (any(grep( x = sppLKPtab$scientificName, pattern = binom, fixed = T))) {
+      return(data.frame(Query = x, Result = x, Match = 'exact'))
     } else{
       # try and determine which piece is incorrect.
       
@@ -200,7 +204,7 @@ spell_check <- function(x) {
       gen_strings <-
         dplyr::filter(genLKPtab, Grp == genus2char) |> dplyr::pull(strings)
       spe_strings <-
[…]

### 2023-05-23 scripts/functions.R
adding content to functoins

@@ -4,6 +4,8 @@
 #' @param dirout the output directory, including path, for the subset dataset to go
 #' @param grid spatvector defining the tiles, should contain a column containing grid cell names
 #' @param fnames name of column containing filenames in the grid
+#' @param example
+#' @param export 
 mason <- function(dirin, dirout, grid, fnames, ...){
 
   # identify the paths to the rasters
@@ -32,6 +34,8 @@ mason <- function(dirin, dirout, grid, fnames, ...){
 #' 
 #' several infraspecies fail from POWO, retry the query with just the base name
 #' @param x output from the first step of 'powo_searcher'
+#' @example
+#' @export 
 try_again <- function(x) {
   q <- x[['query']]
   only_binomial <- unlist(stringr::str_split(q, pattern = " "))
@@ -44,6 +48,8 @@ try_again <- function(x) {
 #' collect the results of a successful powo search
 #' 
 #' @param x a successful powo search query
+#' @example
+#' @export 
 result_grabber <- function(x) {
   # subset the appropriate data frame, there is one if clean data were entered,
   # and two if a synonym was entered.
@@ -103,6 +109,8 @@ result_grabber <- function(x) {
 #' @param x a vector of species names to submit, these should have clean spelling
 #' notes: results are observed to fail for valid infraspecies on Kew's end, and they seem not
 #' to mention valid infraspecies. 
+#' @example
+#' @export 
 powo_searcher <- function(x) {
   query_results <- kewr::search_powo(x)
   
@@ -146,6 +154,8 @@ powo_searcher <- function(x) {
 #' @param  epithet derived from result_grabber
 #' @param  infrarank derived from result_grabber
 #' @param  infraspecies derived from result_grabber
+#' @param example
+#' @param export
 autonym <- function(genus, epithet, infrarank, infraspecies){
   
   authority_hunt <- search_powo(paste(genus, epithet))
@@ -162,6 +172,8 @@ autonym <- function(genus, epithet, infrarank, infraspecies){
 #' check that genera and specific epithets are spelled (almost) correctly
 #' 
 #' @param x a vector of species names
+#' @param example
+#' @param export
 spell_check <- function(x) {
   
   sppLKPtab <- read.csv('../taxonomic_data/species_lookup_table.csv')
@@ -255,7 +267,8 @@ spell_check <- function(x) {
 #' @param x an sf dataframe of coordinates to make maps for, requires collection number and spatial attributes
 #' @param path a directory to store the map images in before merging
[…]

### 2023-05-23 scripts/functions.R
added examples to several functions

@@ -4,7 +4,7 @@
 #' @param dirout the output directory, including path, for the subset dataset to go
 #' @param grid spatvector defining the tiles, should contain a column containing grid cell names
 #' @param fnames name of column containing filenames in the grid
-#' @param example
+#' @param example 
 #' @param export 
 mason <- function(dirin, dirout, grid, fnames, ...){
 
@@ -32,10 +32,10 @@ mason <- function(dirin, dirout, grid, fnames, ...){
 
 #' try to download from Kew again, with just the binomial
 #' 
-#' several infraspecies fail from POWO, retry the query with just the base name
+#' several infra species fail from POWO, retry the query with just the base name. This not accessed directly by the user but used inside 'powo_searcher'
 #' @param x output from the first step of 'powo_searcher'
-#' @example
-#' @export 
+#' @example 
+#' @export  
 try_again <- function(x) {
   q <- x[['query']]
   only_binomial <- unlist(stringr::str_split(q, pattern = " "))
@@ -47,6 +47,7 @@ try_again <- function(x) {
 
 #' collect the results of a successful powo search
 #' 
+#' this function is used within 'powo_searcher' to retrieve the results from a successful POWO query. 
 #' @param x a successful powo search query
 #' @example
 #' @export 
@@ -154,8 +155,8 @@ powo_searcher <- function(x) {
 #' @param  epithet derived from result_grabber
 #' @param  infrarank derived from result_grabber
 #' @param  infraspecies derived from result_grabber
-#' @param example
-#' @param export
+#' @param example 
+#' @param export 
 autonym <- function(genus, epithet, infrarank, infraspecies){
   
   authority_hunt <- search_powo(paste(genus, epithet))
@@ -458,12 +459,16 @@ physical_grabber <- function(x) {
   return(object)
 }
 
-
-
-#' take international format date and make it written herbarium label format
+#' take mdy format date and make it written herbarium label format
 #' 
 #' @param x a data frame with dates
-#' @param example
+#' @param example 
+#' first50dates <- paste0(sample(3:9, size = 50, replace = T), '-', 
+#'    sample(1:29, size = 50, replace  = T), '-', 
+#'    rep(2023, times = 50 ))
+#' head(first50dates)
+#' first50dates <- date2text(first50dates)
[…]

### 2023-05-23 scripts/project_template.Rmd
more fn examples

@@
 
-notFound <- function(x){
-
-  library(crayon)
-
-  row_no <- unlist( apply(FUN = grep, X = x, MARGIN = 2, pattern = 'NOT FOUND') )
-  rows <- unique(row_no) # these rows had complications... 
-  not_found <- x[rows, 'query'] # these records were not found.
-
-  recs <- cat(
-  'The record: ' %+%
-  blue$underline$bold(not_found) %+%
-  blue$bold(' (row ') %+% blue$underline$bold(rows) %+% ') did not have all data retrieved.\n')
-
-  message(recs)
-
-}
-
 ```

### 2023-05-23 scripts/functions.R

@@ -34,7 +34,6 @@ mason <- function(dirin, dirout, grid, fnames, ...){
 #' 
 #' several infra species fail from POWO, retry the query with just the base name. This not accessed directly by the user but used inside 'powo_searcher'
 #' @param x output from the first step of 'powo_searcher'
-#' @example 
 #' @export  
 try_again <- function(x) {
   q <- x[['query']]
@@ -49,7 +48,6 @@ try_again <- function(x) {
 #' 
 #' this function is used within 'powo_searcher' to retrieve the results from a successful POWO query. 
 #' @param x a successful powo search query
-#' @example
 #' @export 
 result_grabber <- function(x) {
   # subset the appropriate data frame, there is one if clean data were entered,
@@ -110,7 +108,10 @@ result_grabber <- function(x) {
 #' @param x a vector of species names to submit, these should have clean spelling
 #' notes: results are observed to fail for valid infraspecies on Kew's end, and they seem not
 #' to mention valid infraspecies. 
-#' @example
+#' @example 
+#' pow_results <- lapply(c('Linnaea borealis', 'Astragalus purshii', 'Pinus ponderosa'), powo_searcher) |>
+#'   dplyr::bind_rows()
+#' head(pow_results)
 #' @export 
 powo_searcher <- function(x) {
   query_results <- kewr::search_powo(x)
@@ -150,12 +151,11 @@ powo_searcher <- function(x) {
 
 #' retrieve author results for autonyms
 #' 
-#' this function is interal to result_grabber
+#' this function is internal to result_grabber, it ensures that an author is collected from the KEW API for infra-species which have autonyms. 
 #' @param  genus derived from result_grabber
 #' @param  epithet derived from result_grabber
 #' @param  infrarank derived from result_grabber
 #' @param  infraspecies derived from result_grabber
-#' @param example 
 #' @param export 
 autonym <- function(genus, epithet, infrarank, infraspecies){
   
@@ -168,12 +168,16 @@ autonym <- function(genus, epithet, infrarank, infraspecies){
   
   name_authority <- paste(genus, epithet, authority, infrarank, infraspecies)
   return(list(name_authority, authority))
+  
 }
 
 #' check that genera and specific epithets are spelled (almost) correctly
 #' 
 #' @param x a vector of species names
 #' @param example
+#' names_vec <- c('Astagalus purshii', 'Linnaeus borealius', 'Heliumorus multifora')
[…]

### 2023-05-23 scripts/autonym.R
split out functions

@@ -1,0 +1,21 @@
+#' retrieve author results for autonyms
+#' 
+#' this function is internal to result_grabber, it ensures that an author is collected from the KEW API for infra-species which have autonyms. 
+#' @param  genus derived from result_grabber
+#' @param  epithet derived from result_grabber
+#' @param  infrarank derived from result_grabber
+#' @param  infraspecies derived from result_grabber
+#' @param export 
+autonym <- function(genus, epithet, infrarank, infraspecies){
+  
+  authority_hunt <- search_powo(paste(genus, epithet))
+  
+  results <- authority_hunt[['results']]
+  clean <- which(unlist(lapply(results, `[`, 'accepted')) == T)
+  auth_hunt <- data.frame(results[clean])
+  authority <- auth_hunt$author
+  
+  name_authority <- paste(genus, epithet, authority, infrarank, infraspecies)
+  return(list(name_authority, authority))
+  
+}

### 2023-05-23 scripts/coords2sf.R

@@ -1,0 +1,52 @@
+#' create an sf object row by row to reflect different datum
+#' 
+#' @param x a dataframe which has undergone dms2dd function
+#' @param datum a column, holding the datum values for each observation, works for WGS84, NAD83, NAD27
+#' @param returns an sf/tibble/dataframe of points in the same WGS84 coordinate reference system
+#' @param example mixed_datum <- data.frame(
+#'  datum = (rep(c('nad27', 'NAD83', 'wGs84'), each = 5)), 
+#'  longitude_dd = runif(15, min = -120, max = -100), 
+#'  latitude_dd = runif(15, min = 35, max = 48)
+#'  )
+#' 
+#' wgs84_dat <- coords2sf( mixed_datum )
+#' str(wgs84_dat)
+#' st_crs(wgs84_dat)
+#' @param export 
+coords2sf <- function(x, datum){
+  
+  library(magrittr)
+  
+  if(missing(datum)){ # identify datum information
+    datum = colnames(x)[grep('datum', colnames(x))] }
+  
+  if(length(datum) == 0){x$datum = 'WGS84'} # if column doesn't exist create
+  
+  dat_check <- function(x){if(grepl('nad.*27', ignore.case = T, x = x)) {
+    x = "NAD27"} else if(grepl('nad.*83', ignore.case = T, x = x)) {
+      x = 'NAD83'} else if(grepl('wgs', ignore.case = T, x = x)) {
+        x = "WGS84"
+      } else {x = 'WGS84'}
+  }
+  # now ensure the datum is appropriate for each point
+  x$datum <- sapply(x$datum, dat_check)
+  
+  crs_lkp = data.frame(
+    datum = c('NAD27', 'NAD83', 'WGS84'), 
+    crs = c(4267, 4269, 4326))
+  
+  x <- dplyr::left_join(x, crs_lkp, by = 'datum') 
+  
+  # separate all points by there datum
+  dat_list <- split(x, f = x$datum)
+  dat_list <- purrr::map(dat_list, . |>  
+                           sf::st_as_sf(., coords = c(x = 'longitude_dd', y = 'latitude_dd'), 
+                                        crs = .$crs[1], remove = F) |> 
+                           sf::st_transform(4326)) |>
+    dplyr::bind_rows() %>% 
+    dplyr::select(-crs) %>% 
+    dplyr::mutate(datum = 'WGS84', .before = geometry)
+  
+  return(dat_list)
+}
+

### 2023-05-23 scripts/date2text.R

@@ -1,0 +1,23 @@
+#' take mdy format date and make it written herbarium label format
+#' 
+#' @param x a data frame with dates
+#' @param example 
+#' first50dates <- paste0(sample(3:9, size = 50, replace = T), '-', 
+#'    sample(1:29, size = 50, replace  = T), '-', 
+#'    rep(2023, times = 50 ))
+#'    
+#' head(first50dates)
+#' first50dates <- date2text(first50dates)
+#' head(first50dates)
+#' @param export
+date2text <- function(x) {
+  
+  x1 <- lubridate::mdy(x)
+  
+  text <- paste0(lubridate::day(x1), # just grab day of month here
+                 ' ',
+                 lubridate::month(x1, abbr = T, label = T), # grab the abbreviation for the month here
+                 ', ',
+                 lubridate::year(x1)) # grab the year here
+  return(text)
+}

### 2023-05-23 scripts/date_parser.R

@@ -1,0 +1,37 @@
+#' create herbarium format dates for specimens
+#' 
+#' this function will return up to two additional columns with x. 'coll_date_text' a text format for the date of collection and 'det_date_text" a shorter text format date for identification
+#' @param x an (sf/tibble/) data frame with both the collection date
+#' and determination date
+#' @param coll_date date of collection, expected to be of the format 'MM/DD/YYYY', minor checks for compliance to the format will be carried out before returning an error.
+#' @param det_date date of determination, same format and processes as above. 
+#' @returns original data frame plus: x_dmy, x_day, x_month, x_year, x_text, det_date_text, columns for both parameters which are supplied as inputs
+#' @param example
+#' @param export
+date_parser <- function(x, coll_date, det_date){
+  
+  coll_date_q <- enquo(coll_date)
+  det_date_q <- enquo(det_date)
+  
+  names_v <- c('_dmy', '_day', '_mo',  '_yr', '_text')
+  if(missing(det_date)){
+    column_names <- c(coll_date, paste0(coll_date, names_v))
+  } else {
+    column_names <- c(coll_date, 
+                      paste0(coll_date, names_v), 
+                        det_date, 
+                        paste0(det_date, names_v))
+  }
+  
+  x_dmy <- x |> 
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), lubridate::mdy, .names = "{.col}_dmy")) |>
+    mutate(across(ends_with('_dmy'), ~ lubridate::month(.), .names = "{.col}_mo"),
+           across(ends_with('_dmy'), ~ lubridate::day(.), .names = "{.col}_day"),
+           across(ends_with('_dmy'), ~ lubridate::year(.), .names = "{.col}_yr")) |>
+    mutate(across(.cols = c(!!coll_date_q, !!det_date_q), date2text, .names = '{.col}_text')) |>
+    rename_with( ~ stringr::str_remove(., '_dmy'), matches("_dmy_.*$")) |>
+    relocate(any_of(column_names), .before = geometry)
+  
+  return(x_dmy)
+}
+

### 2023-05-23 scripts/dms2dd.R

@@ -1,0 +1,61 @@
+#' this function parses coordinates from DMS to decimal degrees
+#'
+#' @param x an input data frame to apply operations too
+#' @param lat a name of the column holding the latitude values
+#' @param long a name of the colymn holding the longitude values
+#' @param dms are coordinates in degrees minutes seconds? TRUE for yes, FALSE for decimal degrees
+#' @param returns dataframe(/tibble) with coordinates unambiguously labeled as being in both degress, minutes, seconds (_dms) and decimal degrees (_dd). 
+#' @param example 
+#'  coords <- data.frame(
+#'   longitude_dd = runif(15, min = -120, max = -100), 
+#'   latitude_dd = runif(15, min = 35, max = 48)
+#' )
+#' 
+#' coords_formatted <- dms2dd( coords )
+#' head(coords_formatted)
+#' colnames(coords_formatted)
+#' @param export 
+dms2dd <- function(x, lat, long, dms){
+  
+  # identify columns if they were not supplied
+  if(missing(lat)){
+    lat = colnames(x)[grep('lat', colnames(x))] }
+  if(missing(long)){
+    long = colnames(x)[grep('long', colnames(x))] }
+  
+  # test for DMS format if not supplied
+  suppressWarnings(  if(missing(dms)){
+    dms = long == parzer::parse_lon(long)
+  })
+  
+  # convert dms to dd, or rename input columns to dd
+  if(dms == T){
+    x$latitude_dd = parzer::parse_lat(lat)
+    x$longitude_dd = parzer::pase_lon(long)
+  } else{
+    colnames(x)[which(names(x) == lat)] <- 'latitude_dd'
+    colnames(x)[which(names(x) == long)] <- 'longitude_dd'
+  }
+  # ensure the DD signs are appropriate for domain
+  x$latitude_dd <- abs(x$latitude_dd)
+  x$longitude_dd <- abs(x$longitude_dd) * -1
+  
+  # now overwrite the original DMS values in our exact formaty
+  
+  x$latitude_dms = paste0(
+    'N ', parzer::pz_degree(x$latitude_dd),
+    '°', round(parzer::pz_minute(x$latitude_dd), 2),
+    "'", round(parzer::pz_second(x$latitude_dd), 2)
+  )
+  
+  x$longitude_dms = paste0(
+    'W ', parzer::pz_degree(x$longitude_dd), 
[…]

### 2023-05-23 scripts/functions.R

@@ -271,7 +271,7 @@ spell_check <- function(x) {
 #' @param x an sf dataframe of coordinates to make maps for, requires collection number and spatial attributes
 #' @param path a directory to store the map images in before merging
 #' @param collection_col column specify the collection number or other UNIQUE id for the collection
-#' @param example
+#' @param example 
 #' @param export
 map_maker <- function(x, states, path, collection_col){
   
@@ -298,7 +298,7 @@ map_maker <- function(x, states, path, collection_col){
 #' this function grabs information on the state, county, and township of collections
 #' @param x an sf data frame of collection points
 #' @param y a column which unambiguously identifies each collection
-#' @param example
+#' @param example see the package vignette
 #' @param export
 political_grabber <- function(x, y) {
   
@@ -349,7 +349,7 @@ political_grabber <- function(x, y) {
 #' this function grabs information on the state, county, and township of collections
 #' @param x an sf data frame of collection points
 #' @param y a column which unambiguously identifies each collection
-#' @param example
+#' @param example see the package vignette
 #' @param export
 place_grabber <- function(x) {
   
@@ -392,7 +392,7 @@ place_grabber <- function(x) {
 #' this function grabs information on the elevation, azimuth, geomorphon, and geology of the site
 #' @param x an sf data frame of collection points
 #' @param y a column which unambiguously identifies each collection
-#' @param example
+#' @param example see the package vignette
 #' @param export
 physical_grabber <- function(x) {
   
@@ -469,6 +469,7 @@ physical_grabber <- function(x) {
 #' first50dates <- paste0(sample(3:9, size = 50, replace = T), '-', 
 #'    sample(1:29, size = 50, replace  = T), '-', 
 #'    rep(2023, times = 50 ))
+#'    
 #' head(first50dates)
 #' first50dates <- date2text(first50dates)
 #' head(first50dates)
@@ -535,6 +536,7 @@ date_parser <- function(x, coll_date, det_date){
 #'   longitude_dd = runif(15, min = -120, max = -100), 
 #'   latitude_dd = runif(15, min = 35, max = 48)
 #' )
+#' 
 #' coords_formatted <- dms2dd( coords )
 #' head(coords_formatted)
 #' colnames(coords_formatted)
@@ -635,14 +637,6 @@ coords2sf <- function(x, datum){
   
   return(dat_list)
 }
-
-
-
[…]

### 2023-05-23 scripts/map_maker.R

@@ -1,0 +1,25 @@
+#' make a quick county dot map to display the location of the collection
+#' 
+#' @param x an sf dataframe of coordinates to make maps for, requires collection number and spatial attributes
+#' @param path a directory to store the map images in before merging
+#' @param collection_col column specify the collection number or other UNIQUE id for the collection
+#' @param example 
+#' @param export
+map_maker <- function(x, states, path, collection_col){
+  
+  if(st_crs(x) == sf::st_crs(states)) {
+    pts } else {pts <- sf::st_transform(x, sf::st_crs(states))}
+  
+  ints <- sf::st_intersection(pts, states)
+  
+  focal_state <- dplyr::filter(states, STATE == ints$STATE)
+  
+  p <- ggplot() + 
+    geom_sf(data = focal_state, fill = NA, color = 'grey15') + 
+    geom_sf(data = pts, size = 0.5) +
+    theme_void() 
+  
+  fname <- file.path(path, 'maps', paste0('map_', sf::st_drop_geometry(x[1,'collectionNo']), '.png'))
+  ggsave(filename =  fname, plot = p, device = 'png', dpi = 300, width = 1, height = 1, units = 'in', 
+         bg = 'transparent')
+}

### 2023-05-23 scripts/mason.R

@@ -1,0 +1,30 @@
+#' a wrapper around terra::makeTiles for setting the domain of a project
+#' 
+#' @param dirin the input directory, including path, with the dataset
+#' @param dirout the output directory, including path, for the subset dataset to go
+#' @param grid spatvector defining the tiles, should contain a column containing grid cell names
+#' @param fnames name of column containing filenames in the grid
+#' @param example 
+#' @param export 
+mason <- function(dirin, dirout, grid, fnames, ...){
+
+  # identify the paths to the rasters
+  dirin <- sub('/$', '', dirin)
+  paths2rast <- file.path(dirin, list.files(path = paste0(dirin, '/'), recursive = T, ...))
+
+  # create and crop the virtual raster
+  virtualRast <- terra::vrt(paths2rast)
+  virtualRast_sub <- terra::crop(virtualRast, terra::ext(grid))
+
+  # set file names and write out product
+  columns <- terra::values(grid)
+  cellname <- columns[,fnames]
+  dirout <- sub('/$', '', dirout)
+  
+  fname <- file.path(dirout, paste0(cellname, ".tif"))
+  terra::makeTiles(virtualRast_sub, grid, filename = fname, na.rm = F)
+  
+  rm(virtualRast_sub)
+  gc()
+
+}

### 2023-05-23 scripts/notFound.R

@@ -1,0 +1,27 @@
+#' notify user if an entry had any results not found in POWO
+#' 
+#' simple function to run on 'powo_searcher' results to show species not found which 
+#' @param x output of 'powo_searcher' after binding rows
+#' @param returns messages to consoles indicating search terms, and there status if failed to be found. This desirable because 'powo_searcher' squashes these errors. 
+#' @param example 
+#' names_vec <- taxize::names_list(rank = 'species', size = 10) # 10 random species from taxize, usually 1 or 2 species are not found in Plants of the world online
+#' pow_results <- lapply(names_vec, powo_searcher) |> 
+#'   dplyr::bind_rows() 
+#' pow_results[,1:5] # if there is not a family which is 'NOT FOUND', reshuffle the random species from taxize. 
+#' notFound(pow_results) # little message. 
+#' @param export 
+notFound <- function(x){
+  
+  library(crayon)
+  row_no <- unlist( apply(FUN = grep, X = x, MARGIN = 2, pattern = 'NOT FOUND') )
+  rows <- unique(row_no) # these rows had complications... 
+  not_found <- x[rows, 'query'] # these records were not found. 
+  
+  recs <- cat(
+    'The record: ' %+%
+      blue$underline$bold(not_found) %+%
+      blue$bold(' (row ') %+% blue$underline$bold(rows) %+% ') did not have all data retrieved.\n')
+  
+  message(recs)
+  
+}

### 2023-05-23 scripts/place_grabber.R

@@ -1,0 +1,74 @@
+#' gather physical characteristics of the site
+#' 
+#' this function grabs information on the elevation, azimuth, geomorphon, and geology of the site
+#' @param x an sf data frame of collection points
+#' @param y a column which unambiguously identifies each collection
+#' @param example see the package vignette
+#' @param export
+physical_grabber <- function(x) {
+  
+  round_df <- function(df, digits) {
+    nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
+    df[,nums] <- round(df[,nums], digits = digits)
+    return(df)
+  }
+  
+  x_spat <- terra::vect(x)
+  geology <- terra::vect('../geodata/geology/geology.shp')
+  x_geo <- terra::extract(geology, x_spat)
+  
+  asp <- '../geodata/aspect/'; slo <-  '../geodata/slope'
+  geo <-  '../geodata/geomorphons'; elev <-  '../geodata/elevation'
+  
+  paths2rast <- file.path(asp, list.files(path = paste0(asp, '/'), recursive = T))
+  aspect <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(slo, list.files(path = paste0(slo, '/'), recursive = T))
+  slope <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(geo, list.files(path = paste0(geo, '/'), recursive = T))
+  geomorphon <- terra::vrt(paths2rast)
+  
+  paths2rast <- file.path(elev, list.files(path = paste0(elev, '/'), recursive = T))
+  elevation <- terra::vrt(paths2rast)
+
+  asp_val <- terra::extract(aspect, x_spat)
+  names(asp_val) <- c('ID', 'aspect')
+  slo_val <- terra::extract(slope, x_spat)
+
+  geoLKPtab <- data.frame(unit = 1:10, geomorphon = c(
+    'flat', 'peak', 'ridge', 'shoulder', 'spur', 
+    'slope', 'hollow', 'footslope', 'valley', 'pit'
+  ))
+  
+  geo_val <- terra::extract(geomorphon, x_spat)
+  names(geo_val) <- c('ID', 'unit')
+  geo_val <- dplyr::left_join(geo_val, geoLKPtab, by = 'unit') |>
+    dplyr::select(-unit)
+  
+  ele_val <- terra::extract(elevation, x_spat)
+  
+  values <- data.frame(cbind(
[…]

### 2023-05-23 scripts/political_grabber.R

@@ -1,0 +1,52 @@
+
+
+#' gather political site information
+#' 
+#' this function grabs information on the state, county, and township of collections
+#' @param x an sf data frame of collection points
+#' @param y a column which unambiguously identifies each collection
+#' @param example see the package vignette
+#' @param export
+political_grabber <- function(x, y) {
+  
+  y_quo <- rlang::enquo(y)
+  
+  political <- sf::st_read('../geodata/political/political.shp', quiet = T)
+  allotment <- sf::st_read('../geodata/allotments/allotments.shp', quiet = T)
+  plss <- sf::st_read('../geodata/plss/plss.shp', quiet = T)
+  ownership <- sf::st_read('../geodata/pad/pad.shp', quiet = T)
+  
+  # write attributes to data set
+  
+  x <- sf::st_join(x, political)
+  x <- sf::st_join(x, allotment)
+  x <- sf::st_join(x, ownership)
+  
+  x_plss <- sf::st_transform(x, sf::st_crs(plss))
+  x_plss <- sf::st_join(x_plss, plss) |>
+    sf::st_drop_geometry() %>% 
+    dplyr::select(y, trs)
+  
+  x_vars <- dplyr::left_join(x, x_plss, by = y) |> 
+    dplyr::mutate(Country = 'U.S.A.') |> 
+    dplyr::relocate(any_of(c('Country', 'State', 'County', 'Mang_Name', 'Unit_Nm', 'trs')),
+             .before = geometry) |>
+    dplyr::distinct(.keep_all = T) |> # with large enough sample size some points fall on an exact border
+    dplyr::group_by( .data[[y]] ) |>
+    dplyr::slice_head(n = 1) |> 
+    dplyr::ungroup()
+  
+  x_vars <- x_vars %>% 
+    dplyr::mutate(
+      Gen = paste0(
+        Country, ', ', State, ', ', County, ' Co., ', Mang_Name, " ", Unit_Nm, " ", trs), 
+      Gen = stringr::str_replace_all(Gen, "NA", ""), 
+      Gen = stringr::str_replace_all(Gen, "  ", ""),
+      Gen = stringr::str_trim(Gen),
+      Gen = stringr::str_remove(Gen, ",$"), 
+      .before = geometry)
+  
+  return(x_vars)
+  
+  rm(political, allotment, plss, ownership)
+}

### 2023-05-23 scripts/powo_searcher.R

@@ -1,0 +1,46 @@
+#' query plants of the world online for taxonomic information
+#' 
+#' a wrapper for kewr::search_powo
+#' @param x a vector of species names to submit, these should have clean spelling
+#' notes: results are observed to fail for valid infraspecies on Kew's end, and they seem not
+#' to mention valid infraspecies. 
+#' @example 
+#' pow_results <- lapply(c('Linnaea borealis', 'Astragalus purshii', 'Pinus ponderosa'), powo_searcher) |>
+#'   dplyr::bind_rows()
+#' head(pow_results)
+#' @export 
+powo_searcher <- function(x) {
+  query_results <- kewr::search_powo(x)
+  
+  if (is.null(query_results[["results"]])) {
+    second_try <- try_again(query_results)
+    
+    if (is.null(second_try[["results"]])) {
+      taxonomic_info <- data.frame(
+        family = as.character(),
+        name_authority = as.character(),
+        full_name = as.character(),
+        genus = as.character(),
+        epithet = as.character(),
+        infrarank = as.character(),
+        infraspecies = as.character(),
+        authority = as.character()
+      )
+      taxonomic_info[1,] <- 'NOT FOUND'
+      
+    } else {
+      results_to_process <- second_try
+    }
+    
+  } else {
+    results_to_process <- query_results
+  }
+  
+  # this is the end of the process, return empty results without error, or real results
+  if (exists('taxonomic_info')) {
+    return(cbind(query = x, taxonomic_info))
+  } else {
+    taxonomic_info <- result_grabber(results_to_process)
+    return(cbind(query = x, taxonomic_info))
+  }
+}

### 2023-05-23 scripts/result_grabber.R

@@ -1,0 +1,57 @@
+#' collect the results of a successful powo search
+#' 
+#' this function is used within 'powo_searcher' to retrieve the results from a successful POWO query. 
+#' @param x a successful powo search query
+#' @export 
+result_grabber <- function(x) {
+  # subset the appropriate data frame, there is one if clean data were entered,
+  # and two if a synonym was entered.
+  results <- x[['results']]
+  
+  clean <- which(unlist(lapply(results, `[`, 'accepted')) == T)
+  resultT <- data.frame(results[clean])
+  
+  # the main variables which all results should have.
+  family <- resultT$family
+  authority <- resultT$author
+  full_name <- resultT$name
+  split_name <- unlist(stringr::str_split(full_name, pattern = " "))
+  genus <- split_name[1]
+  epithet <- split_name[2]
+  
+  # the infra species information if relevant.
+  if (length(split_name) > 2) {
+    infraspecies <- split_name[length(split_name)]
+  } else {
+    infraspecies <- NA
+  }
+  
+  if (length(split_name) > 2) {
+    infrarank <-
+      stringr::str_extract(full_name, ' var\\. | spp\\. | subsp\\. ') |>
+      stringr::str_trim(side = 'both')
+  } else {
+    infrarank <- NA
+  }
+  
+  if (is.null(authority)){
+    
+    res <- autonym(genus, epithet, infrarank, infraspecies)
+    name_authority <- res[[1]]
+    authority <- res[[2]]
+    
+  } else{name_authority = paste(full_name, authority)}
+  
+  taxonomic_info <- data.frame(
+    family,
+    'name_authority' = name_authority,
+    full_name,
+    genus,
+    epithet,
+    infrarank,
+    infraspecies,
+    authority
+  )
+  
+  return(taxonomic_info)
+}

### 2023-05-23 scripts/spell_check.R

@@ -1,0 +1,94 @@
+#' check that genera and specific epithets are spelled (almost) correctly
+#' 
+#' @param x a vector of species names
+#' @param example
+#' names_vec <- c('Astagalus purshii', 'Linnaeus borealius', 'Heliumorus multifora')
+#' spelling <- spell_check(names_vec)
+#' spelling 
+#' @param export
+spell_check <- function(x) {
+  
+  sppLKPtab <- read.csv('../taxonomic_data/species_lookup_table.csv')
+  epiLKPtab <- read.csv('../taxonomic_data/epithet_lookup_table.csv')
+  genLKPtab <- read.csv('../taxonomic_data/genus_lookup_table.csv')
+  
+  pieces <- unlist(stringr::str_split(x, pattern = " "))
+  genus <- pieces[1] ; species <- pieces[2]
+  binom <- paste(genus, species)
+  
+  # infra species should be found without much hassle due to their length
+  if(length(pieces) == 4){
+    infras <- na.omit(epiLKPtab)
+    full_name <- paste(genus, species, 
+                       stringr::str_replace(pieces[3], 'ssp\\.|ssp', 'subsp.'), pieces[4])
+    
+    if (any(grep( x = infras$scientificName, pattern = full_name, fixed = T))) {
+      return(data.frame(Query = x, Result = full_name, Match = 'exact'))
+    } else {
+      
+      infras <- na.omit(epiLKPtab)
+      full_name <- paste(genus, species, 
+                         stringr::str_replace(pieces[3], 'ssp\\.|ssp', 'subsp.'), pieces[4])
+      infraspecies_name <-
+        infras[which.min(adist(full_name, infras$scientificName)), 'scientificName'] |> as.character()
+      return(data.frame(Query = x, Result = infraspecies_name, Match = 'fuzzy'))
+    } 
+    
+  # species can become difficult due to their short  names, e.g. 'Poa annua'
+  } else {
+  
+    if (any(grep( x = sppLKPtab$scientificName, pattern = binom, fixed = T))) {
+      return(data.frame(Query = x, Result = x, Match = 'exact'))
+    } else{
+      # try and determine which piece is incorrect.
+      
+      # subset datasets to query each name component separately
+      genus2char <- stringr::str_extract(genus, '[A-Z][a-z]{1}')
+      species3char <- stringr::str_extract(species, '[a-z]{3}')
+      gen_strings <-
+        dplyr::filter(genLKPtab, Grp == genus2char) |> dplyr::pull(strings)
+      spe_strings <-
[…]

### 2023-05-23 scripts/try_again.R

@@ -1,0 +1,13 @@
+#' try to download from Kew again, with just the binomial
+#' 
+#' several infra species fail from POWO, retry the query with just the base name. This not accessed directly by the user but used inside 'powo_searcher'
+#' @param x output from the first step of 'powo_searcher'
+#' @export  
+try_again <- function(x) {
+  q <- x[['query']]
+  only_binomial <- unlist(stringr::str_split(q, pattern = " "))
+  only_binomial <- paste(only_binomial[1], only_binomial[2])
+  
+  results <- kewr::search_powo(only_binomial)
+  return(results)
+}

### 2023-06-04 scripts/valleys_create.Rmd
making empty valleys dataset

---
title: "Creating Named Valleys Dataset"
author: "Payton Lott, Gus Rytting, Reed Benkendorf"
date: "`r Sys.Date()`"
output: html_document
---

The 'Mountains Inventory version 2' dataset generated by Global Mountain Biodiversity Assessment contains a standardized naming schema for all of the worlds mountains. We are interested in developing a similar named schema for many of the valleys of Western North America.

In order to identify drainage areas, 
USGS Watershed Boundary Dataset (WBD) was used as a source of all major watersheds [The National Map](https://apps.nationalmap.gov/downloader/#/).

```{r}
library(tidyverse)
library(sf)
library(terra)
```

```{r mountains}
mtns <- st_read('../geodata/mountains/mountains.shp', quiet = T)
mtn_bbox <- mtns %>% 
  st_transform(4269) %>% 
  st_bbox()

mtns <- st_make_valid(mtns)
```

HU 10 dataset was used as the source for watersheds wherein we would name valleys. It was cropped to the extent of the study area.

```{r load watershed dataset}

pbasin <- '../geodata_raw/watershedBoundaries/WBD_National_GPKG/WBD_National_GPKG.gpkg'
st_layers(pbasin)

hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2')  %>% 
  select(HU2_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu10 <- st_read(dsn = pbasin, layer = 'WBDHU10')  %>% 
  select(HU10_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu10 <- hu10[st_intersects(hu10, st_union(hu2)) %>% lengths > 0, ]

hu10 <- st_transform(hu10, 5070) %>% 
  st_simplify(dTolerance = 75) %>% 
  st_transform(4326) # around 1/9 the size of the original

hu10 <- st_make_valid(hu10)
```

```{r identify watersheds in the great basin area}

gb <- filter(hu2, HU2_NAME == 'Great Basin Region') %>% 
  st_transform(4326)

gb10 <- hu10[st_intersects(hu10, st_union(gb)) %>% lengths > 0, ]
```

```{r overlay with gb mtns}

mtns_gb <- mtns[st_intersects(mtns, st_union(gb10)) %>% lengths > 0, ]
gb10_er <- rmapshaper::ms_erase(gb10, mtns)

ggplot() + 
  geom_sf(data = gb10_er) +
  geom_sf(data = mtns_gb, fill = 'purple', alpha = 0.3)
```

```{r identify western watersheds}
west_pl <- c('California Region', 'Pacific Northwest Region', 
             'Lower Colorado Region', 'Great Basin Region', 'Upper Colorado Region', 
             'Rio Grande Region')
west <- filter(hu2, HU2_NAME %in% west_pl) %>% 
  st_transform(4326)

west10 <- hu10[st_intersects(hu10, st_union(west)) %>% lengths > 0, ]
```

```{r overlay with western mtns}

mtns_west <- mtns[st_intersects(mtns, st_union(west)) %>% lengths > 0, ]
west10 <- st_crop(west10, mtns_west)
west10_er <- rmapshaper::ms_erase(west10, mtns_west)

ggplot() + 
  geom_sf(data = west10) +
  geom_sf(data = mtns_west, fill = 'purple', alpha = 0.3)
```

**First two authors in alphabetical order**

### 2023-06-16 scripts/Barneby_Lives.Rmd
bringing in dististaze from sos-gb

@@
 While in theory all of the tasks in this script could be automated, doing so would be an unconscionable action against local taxonomic authorities.
-While several database options for updating collections (and spell checking nomen) which are current and well curated (e.g.
+While several database options for updating collections (and spell checking names) which are current and well curated (e.g.
 Kew Plants of the World) exist, we feel many proposals and combinations require interpretation by those in the regions attempting to understand the material at hand.
@@
 library(elevatr)
+library(googlesheets4)
 ```
@@
 rm(path_f, files)
+```
+
+An alternative format for loading data is through Google Sheets. This is what we use within our group.
+
+```{r load through google sheets}
+
+library(googlesheets4)
+
+# read in data from the sheet to process
+input <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
+                    sheet = 'Data Entry - Examples') %>% 
+  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))
+
+# determine whether these data have already been processed by the script, using
+# a unique combination of collection name and collection code. 
+processed <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
+                        sheet = 'Processed - Examples') %>% 
+  select(Collection_number, Primary_Collector) %>% 
+  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))
+
+df <- filter(input, ! UNIQUEID %in% processed$UNIQUEID ) %>% 
+  select(-UNIQUEID)
+
+rm(processed, input)
+
+# we will add these data onto our final sheet. 
+sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
+             sheet = 'Processed - Examples', data = df) 
 ```
@@
 ```
+
+**distance from named place**
+
+```{r}
+sites <- data.frame(
+  longitude_dd = runif(15, min = -120, max = -100), 
+  latitude_dd = runif(15, min = 35, max = 48)
+  ) %>% 
+  st_as_sf(coords = c('longitude_dd', 'latitude_dd'))
+
+places <- st_read('../geodata/places/places.shp')
+
+head(places)
+```
+
+```{r}
+
+#' gather distance and azimuth from shop to center of town
+#'
+#' Help provide some simple context between the building and the 
+#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
+#' @param places_data an sf/dataframe/dataframe which contains coordinates for the locations in the data
+distAZE <- function(x, places_data){
+
+  locality <- sf::st_drop_geometry(x)
+  locality <- locality[1, 'Locality']
+
+  focal <- places_data[grep(locality, places_data$ID), ]
+  if(nrow(focal) > 1){
+    union_loc <- sf::st_union(x) |> sf::st_point_on_surface()
+    focal <- focal[sf::st_nearest_feature(union_loc, focal),]
+  }
+
+  location_from <- sf::st_centroid(focal)
+
+  location_from <- sf::st_transform(location_from, 5070)
+  x_planar <- sf::st_transform(x, 5070)
+  distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')
+
+  azy <- nngeo::st_azimuth(
+    location_from, 
+    x_planar
+  )
+
+  distances <- data.frame(
+    st_drop_geometry(x),
+    Distance = round(as.numeric(distances / 1609.34), 1),
+    Azimuth = round(as.numeric(azy), 0),
+    geometry = x[,'geometry']
+  )
+
+}
+
+```

### 2023-06-16 scripts/valleys_create.Rmd
bringing in dististaze from sos-gb

@@
 The 'Mountains Inventory version 2' dataset generated by Global Mountain Biodiversity Assessment contains a standardized naming schema for all of the worlds mountains.
-We are interested in developing a similar named schema for many of the valleys of Western North America.
+We are interested in developing a simpler named schema for many of the valleys and watersheds of Western North America.
 
-In order to identify drainage areas, 
-USGS Watershed Boundary Dataset (WBD) was used as a source of all major watersheds [The National Map](https://apps.nationalmap.gov/downloader/#/).
+USGS Watershed Boundary Dataset (WBD) was used to identify major watersheds [The National Map](https://apps.nationalmap.gov/downloader/#/).
 
@@
   st_bbox()
-
-mtns <- st_make_valid(mtns)
 ```
-
-HU 10 dataset was used as the source for watersheds wherein we would name valleys. It was cropped to the extent of the study area.
 
@@
 
-hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2')  %>% 
+hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2', quiet = T)  %>% 
   select(HU2_NAME = name) %>% 
@@
 
-hu10 <- st_read(dsn = pbasin, layer = 'WBDHU10')  %>% 
-  select(HU10_NAME = name) %>% 
+hu8 <- st_read(dsn = pbasin, layer = 'WBDHU8', quiet = T)  %>% 
+  select(HU8_NAME = name) %>% 
   st_crop(., mtn_bbox)
 
-hu10 <- hu10[st_intersects(hu10, st_union(hu2)) %>% lengths > 0, ]
-
-hu10 <- st_transform(hu10, 5070) %>% 
-  st_simplify(dTolerance = 75) %>% 
-  st_transform(4326) # around 1/9 the size of the original
-
-hu10 <- st_make_valid(hu10)
+hu8 <- st_join(hu2, hu8) 
 ```
 
-```{r identify watersheds in the great basin area}
-
-gb <- filter(hu2, HU2_NAME == 'Great Basin Region') %>% 
-  st_transform(4326)
-
-gb10 <- hu10[st_intersects(hu10, st_union(gb)) %>% lengths > 0, ]
-```
-
-```{r overlay with gb mtns}
-
-mtns_gb <- mtns[st_intersects(mtns, st_union(gb10)) %>% lengths > 0, ]
-gb10_er <- rmapshaper::ms_erase(gb10, mtns)
+```{r}
 
 ggplot() + 
-  geom_sf(data = gb10_er) +
-  geom_sf(data = mtns_gb, fill = 'purple', alpha = 0.3)
+  geom_sf(data = hu8)
+
+gb <- filter(hu2, name == 'Great Basin Region')
 ```
-
-```{r identify western watersheds}
-west_pl <- c('California Region', 'Pacific Northwest Region', 
-             'Lower Colorado Region', 'Great Basin Region', 'Upper Colorado Region', 
-             'Rio Grande Region')
-west <- filter(hu2, HU2_NAME %in% west_pl) %>% 
-  st_transform(4326)
-
-west10 <- hu10[st_intersects(hu10, st_union(west)) %>% lengths > 0, ]
-```
-
-```{r overlay with western mtns}
-
-mtns_west <- mtns[st_intersects(mtns, st_union(west)) %>% lengths > 0, ]
-west10 <- st_crop(west10, mtns_west)
-west10_er <- rmapshaper::ms_erase(west10, mtns_west)
-
-ggplot() + 
-  geom_sf(data = west10) +
-  geom_sf(data = mtns_west, fill = 'purple', alpha = 0.3)
-```
-
-**First two authors in alphabetical order**

### 2023-06-16 scripts/functions.R

@@ -530,7 +530,7 @@ date_parser <- function(x, coll_date, det_date){
 #' @param lat a name of the column holding the latitude values
 #' @param long a name of the colymn holding the longitude values
 #' @param dms are coordinates in degrees minutes seconds? TRUE for yes, FALSE for decimal degrees
-#' @param returns dataframe(/tibble) with coordinates unambiguously labeled as being in both degress, minutes, seconds (_dms) and decimal degrees (_dd). 
+#' @param returns dataframe(/tibble) with coordinates unambiguously labeled as being in both degrees, minutes, seconds (_dms) and decimal degrees (_dd). 
 #' @param example 
 #'  coords <- data.frame(
 #'   longitude_dd = runif(15, min = -120, max = -100), 

### 2023-06-16 scripts/Barneby_Lives.Rmd
distAZE customizef for, but now wrap first step within function

@@
   ) %>% 
-  st_as_sf(coords = c('longitude_dd', 'latitude_dd'))
+  st_as_sf(coords = c('longitude_dd', 'latitude_dd'), crs = 4326)
 
 places <- st_read('../geodata/places/places.shp')
+nf <- st_nearest_feature(sites, places)
 
-head(places)
+sites <- sites %>%
+  mutate(st_drop_geometry(places[nf, 'ID']), .before = geometry)
+
 ```
@@
   locality <- sf::st_drop_geometry(x)
-  locality <- locality[1, 'Locality']
+  locality <- locality[1, 'ID']
 
   focal <- places_data[grep(locality, places_data$ID), ]
-  if(nrow(focal) > 1){
-    union_loc <- sf::st_union(x) |> sf::st_point_on_surface()
-    focal <- focal[sf::st_nearest_feature(union_loc, focal),]
-  }
-
   location_from <- sf::st_centroid(focal)
@@
   distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')
+  place <- data.frame('Place' =  st_drop_geometry(places_data[x$ID, 'fetr_nm']))
 
@@
     st_drop_geometry(x),
-    Distance = round(as.numeric(distances / 1609.34), 1),
+    Distance = round(as.numeric(distances / 1609.34), -1),
     Azimuth = round(as.numeric(azy), 0),
-    geometry = x[,'geometry']
-  )
+    Place = place
+  )  %>% 
+    dplyr::mutate(Site = if_else(
+      Distance < 100, paste('At', fetr_nm, '.'),
+      paste0(Distance, 'm', ' at ', Azimuth, '° from ', fetr_nm, '.')),
+      Site = stringr::str_replace(Site, '\\..$', '.')) %>% 
+    dplyr::select(-any_of(c('Distance', 'Azimuth', 'Place', 'ID', 'fetr_nm')))
+
+  out <- dplyr::bind_cols(x, distances) %>% 
+    dplyr::relocate(Site, .before = geometry) %>% 
+    dplyr::select(-any_of(c('ID')))
 
@@
 
+test <- distAZE(sites, places)
+
 ```

### 2023-06-16 scripts/Barneby_Lives.Rmd
site name writing fn

--- 2023-06-16 scripts/Barneby_Lives.Rmd
@@
   ) %>% 
-  st_as_sf(coords = c('longitude_dd', 'latitude_dd'))
+  st_as_sf(coords = c('longitude_dd', 'latitude_dd'), crs = 4326)
 
-places <- st_read('../geodata/places/places.shp')
-
-head(places)
-```
-
-```{r}
-
-#' gather distance and azimuth from shop to center of town
-#'
-#' Help provide some simple context between the building and the 
-#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
-#' @param places_data an sf/dataframe/dataframe which contains coordinates for the locations in the data
-distAZE <- function(x, places_data){
-
-  locality <- sf::st_drop_geometry(x)
-  locality <- locality[1, 'Locality']
-
-  focal <- places_data[grep(locality, places_data$ID), ]
-  if(nrow(focal) > 1){
-    union_loc <- sf::st_union(x) |> sf::st_point_on_surface()
-    focal <- focal[sf::st_nearest_feature(union_loc, focal),]
-  }
-
-  location_from <- sf::st_centroid(focal)
-
-  location_from <- sf::st_transform(location_from, 5070)
-  x_planar <- sf::st_transform(x, 5070)
-  distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')
-
-  azy <- nngeo::st_azimuth(
-    location_from, 
-    x_planar
-  )
-
-  distances <- data.frame(
-    st_drop_geometry(x),
-    Distance = round(as.numeric(distances / 1609.34), 1),
-    Azimuth = round(as.numeric(azy), 0),
-    geometry = x[,'geometry']
-  )
-
-}
+test <- site_writer(sites)
 

### 2023-06-16 scripts/site_writer.R

@@ -1,0 +1,56 @@
+
+#' gather distance and azimuth from shop to center of town
+#'
+#' Help provide some simple context between the building and the 
+#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
+#' @example
+#' sites <- data.frame(
+#'  longitude_dd = runif(15, min = -120, max = -100), 
+#'  latitude_dd = runif(15, min = 35, max = 48)
+#'  ) %>% 
+#'  st_as_sf(coords = c('longitude_dd', 'latitude_dd'), crs = 4326)
+#'  
+#' head(sites)
+#' distaze_results <- distAZE(sites) # takes some time
+#' head(distaze_results)
+site_writer <- function(x){
+  
+  places_data <- sf::st_read('../geodata/places/places.shp', quiet = T)
+  nf <- sf::st_nearest_feature(sites, places_data) 
+
+  sites <- x %>%
+   dplyr::mutate(st_drop_geometry(places_data[nf, 'ID']), .before = geometry)
+  
+  locality <- sf::st_drop_geometry(sites)
+  locality <- locality[1, 'ID']
+  
+  focal <- places_data[grep(locality, places_data$ID), ]
+  location_from <- sf::st_centroid(focal)
+  
+  location_from <- sf::st_transform(location_from, 5070)
+  x_planar <- sf::st_transform(x, 5070)
+  distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')
+  place <- data.frame('Place' =  st_drop_geometry(places_data[x$ID, 'fetr_nm']))
+  
+  azy <- nngeo::st_azimuth(
+    location_from, 
+    x_planar
+  )
+  
+  distances <- data.frame(
+    st_drop_geometry(x),
+    Distance = round(as.numeric(distances / 1609.34), -1),
+    Azimuth = round(as.numeric(azy), 0),
+    Place = place
+  )  %>% 
+    dplyr::mutate(Site = if_else(
+      Distance < 100, paste0('At ', fetr_nm, '.'),
+      paste0(Distance, 'm', ' at ', Azimuth, '° from ', fetr_nm, '.')),
+      Site = stringr::str_replace(Site, '\\..$', '.')) %>% 
+    dplyr::select(-any_of(c('Distance', 'Azimuth', 'Place', 'ID', 'fetr_nm')))
+  
+  out <- dplyr::bind_cols(x, distances) %>% 
[…]

### 2023-06-18 scripts/Barneby_Lives.Rmd
resolved merge via load library

@@
 #library(parzer)
-library(tigris)
 #library(WorldFlora)

### 2023-06-18 scripts/taxonomy_setup.Rmd
resolved merge via load library

@@
 write.csv(genLKPtab, '../taxonomic_data/genus_lookup_table.csv', row.names = F)
-
 ```

### 2023-06-18 scripts/date2text.R

@@ -1,5 +1,7 @@
 #' take mdy format date and make it written herbarium label format
 #' 
+#' @description This will ensure that dates are formatted in a manner which makes
+#' them readily interpretable to researchers worldwide.
 #' @param x a data frame with dates
 #' @param example 
 #' first50dates <- paste0(sample(3:9, size = 50, replace = T), '-', 

### 2023-06-18 scripts/map_maker.R

@@ -1,10 +1,13 @@
 #' make a quick county dot map to display the location of the collection
 #' 
+#' @description use sf to create a 20th centtury style 'dot map' which features the state
+#' boundary and county lines.
+#' 
 #' @param x an sf dataframe of coordinates to make maps for, requires collection number and spatial attributes
 #' @param path a directory to store the map images in before merging
 #' @param collection_col column specify the collection number or other UNIQUE id for the collection
-#' @param example 
-#' @param export
+#' @example 
+#' @export
 map_maker <- function(x, states, path, collection_col){
   
   if(st_crs(x) == sf::st_crs(states)) {

### 2023-06-18 scripts/mason.R

@@ -1,11 +1,14 @@
 #' a wrapper around terra::makeTiles for setting the domain of a project
 #' 
+#' @description for projects with large spatial domains or using relatively high resolution data, this will help
+#' you make virtual tiles which do not need to be held in memory for the project.
+#' 
 #' @param dirin the input directory, including path, with the dataset
 #' @param dirout the output directory, including path, for the subset dataset to go
 #' @param grid spatvector defining the tiles, should contain a column containing grid cell names
 #' @param fnames name of column containing filenames in the grid
-#' @param example 
-#' @param export 
+#' @example 
+#' @export 
 mason <- function(dirin, dirout, grid, fnames, ...){
 
   # identify the paths to the rasters

### 2023-06-18 scripts/notFound.R

@@ -1,15 +1,15 @@
 #' notify user if an entry had any results not found in POWO
 #' 
-#' simple function to run on 'powo_searcher' results to show species not found which 
+#' @description simple function to run on 'powo_searcher' results to show species not found which 
 #' @param x output of 'powo_searcher' after binding rows
 #' @param returns messages to consoles indicating search terms, and there status if failed to be found. This desirable because 'powo_searcher' squashes these errors. 
-#' @param example 
+#' @example 
 #' names_vec <- taxize::names_list(rank = 'species', size = 10) # 10 random species from taxize, usually 1 or 2 species are not found in Plants of the world online
 #' pow_results <- lapply(names_vec, powo_searcher) |> 
 #'   dplyr::bind_rows() 
 #' pow_results[,1:5] # if there is not a family which is 'NOT FOUND', reshuffle the random species from taxize. 
 #' notFound(pow_results) # little message. 
-#' @param export 
+#' @export 
 notFound <- function(x){
   
   library(crayon)

### 2023-06-18 scripts/physical_grabber.R

@@ -1,10 +1,10 @@
 #' gather physical characteristics of the site
 #' 
-#' this function grabs information on the elevation, azimuth, geomorphon, and geology of the site
+#' @description this function grabs information on the elevation, azimuth, geomorphon, and geology of the site
 #' @param x an sf data frame of collection points
 #' @param y a column which unambiguously identifies each collection
-#' @param example see the package vignette
-#' @param export
+#' @example see the package vignette
+#' @export
 physical_grabber <- function(x) {
   
   round_df <- function(df, digits) {

### 2023-06-18 scripts/spell_check.R

@@ -1,11 +1,12 @@
 #' check that genera and specific epithets are spelled (almost) correctly
 #' 
+#' @description this function attempts to verify the spelling of a user submitted taxonomic name. If necessary it will proceed step-wise by name pieces attempting to place them. 
 #' @param x a vector of species names
-#' @param example
+#' @example
 #' names_vec <- c('Astagalus purshii', 'Linnaeus borealius', 'Heliumorus multifora')
 #' spelling <- spell_check(names_vec)
 #' spelling 
-#' @param export
+#' @export
 spell_check <- function(x) {
   
   sppLKPtab <- read.csv('../taxonomic_data/species_lookup_table.csv')

### 2023-07-06 scripts/running_pipeline.Rmd
refining tutorial

---
title: "running pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{running pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r install from github, eval = F}
devtools::install_github('sagesteppe/BarnebyLives')
```

```{r setup, message=F}
library(tidyverse)
library(BarnebyLives)
```

## load data into R

### load data from a local (on computer) source

Most users may

### load data through google sheets

An alternative format for loading data is through Google Sheets. 
This can facilitate data-sharing and standardization by many collectors in different locations. 
This is the source of data which was used during beta-testing for development.

We will use googlesheets4, which is a part of the tidyverse, but I do not think installs by default with the remainder of the 'verse. 
We will feed in the last part of the URL as the link to the path. 
If this is your first time using googlesheets4, it will walk you through a few steps to log into your google account so you can access data securely. 
After that googlesheets4 is actually really straightforward to use, note that I use the 'drive_auth' function here. 
This function makes non-interactive use easy, it is only used here because the vignette renders as a non-interactive format.

```{r load through google sheets, eF}

library(googlesheets4)
googledrive::drive_auth("reedbenkendorf27@gmail.com")
# read in data from the sheet to process
input <- googlesheets4::read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
                    sheet = 'Data Entry - Examples') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

```

### only process data which have not yet run through the pipeline.

BarnebyLives takes a little bit of time to run, but it also requires the usage of Google Developer credits for

```{r determine which data has been processed, eval = F}

# determine whether these data have already been processed by the script, using
# a unique combination of collection name and collection code. 
read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
                        sheet = 'Processed - Examples') %>% 
  select(Collection_number, Primary_Collector) %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

df <- filter(input, ! UNIQUEID %in% processed$UNIQUEID ) %>% 
  select(-UNIQUEID)

rm(processed, input)
```

## Ensure Coordinates and Dates and formatted appropriately

Date parsing, e.g. convert date into congruent museum formats (month in roman numerals, or spelled out).
This step assumes that you are recording your dates in the American format of (month/day/year), rather than the internationally more common format (day/month/year). We will also parse out the day of month, month of year, and year, which is used in multiple possible downstream software.

```{r Date parsing}
data <- date_parser(input, 'Date_digital')

dplyr::select(data, starts_with('Date')) %>% 
  utils::head()
```

Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD). 
The DMS format is still used in a few applications, such as the Seeds of Success program, we will convert this to the more generally utilized DD format. We will maintain both formats in the data set, so others have them readily available for other uses aside from creating herbarium labels.

```{r Conversion of Degrees Minutes Seconds (DMS) to Decimal Degrees (DD)}
data <- dms2dd(data, dms = F)

dplyr::select(data, starts_with(c('latitude', 'longitude'))) %>% 
  utils::head()
```

For internal usage and data acquisition we will make data set explicitly spatial, rather than just holding coordinates in a column. 
This is also required to write out data in a format readable by Google Earth or another Geographic Information System.

```{r make data spatial}
data <- coords2sf(data)
head(data) # now we can see that it is an sf object
```

## Retrieve Political/Administrative Information

BarnebyLives! will retrieve various components of both political and administrative information. All of these data which include: The *state*, and *county* of collection, and the *township, range, section (TRS)* from the Public Land Survey System (PLSS). 
It will also determine whether a Federal or State Agency manages or owns the land which were collected from, and if so return the name of it. 
If either the United States Forest Service or Bureau of Land Management does manage the land, it will return information on grazing allotments.

Retrieve State, County, Ownership, TRS, and allotment information
```{r Retrieve Political}
data <- political_grabber(data, y = 'UNIQUEID', path = '/hdd/Barneby_Lives-dev/geodata')
```

## Geographic

Retrieve Nearest GNIS place name, and azimuth from it
```{r Retrieve Place and location from it}
data <- site_writer(data, path = '/hdd/Barneby_Lives-dev/geodata')
```

## Site characteristics

Elevation (both meters and feet), Slope, Aspect, Surficial geology, and geomorphons

```{r physical site characteristics}
data <- physical_grabber(data, path = '/hdd/Barneby_Lives-dev/geodata')
```

## Taxonomic

BarnebyLives performs multiple operations related to taxonomy and nomenclature. 
The function **spell_check()**, will attempt to determine whether the supplied binomial scientific name, and family are spelled correctly and offer suggestions if they are not. 
This function requires that the first few letters of the generic name and epithet are spelled correctly. 
It will run a spell check against nearly all published plant name spellings, and search for the most similar spelling within these lists if material is not found. 
It will operate on both pieces of the binomial separately, and can work to identify the species within the context of genus.

```{r check spelling of binomial and family, eval = F}

names <- sf::st_drop_geometry(data) |>
  pull(Full_name)

names_vec <- c('Astragalis purshii', 'Linnaea borealius var. borealis', 'Heliumorus multifora')

lapply(FUN = spell_check, X = names_vec, path = '/hdd/Barneby_Lives-dev/taxonomic_data')

```

A similar check is to determine that the author whom is cited for describing the species, or transferring it to another genus, is spelled using the standard author abbreviations.
This check operates under a similar manner to **spell_check()**.
The list which is consults *is* comprehensive, within a few years, of all taxonomists whom have published plant species.

```{r check spelling of author}
data <- author_check(data, path = '/hdd/Barneby_Lives-dev/taxonomic_data')
```

After BarnebyLives! has verified the spelling of the species it will submit it to Plants of the World Online to determine whether a more current and accepted name has been applied. 
BL! will not automatically over-write your submitted species, and it is important that you interpret the results of query.
Especially as it is possible that **spell_check** has mis-matched your material to another name.

Searches for synonym to species
```{r Search for synonyms to species}

POW_names <- lapply(data$Full_name, powo_searcher) |>
  bind_rows()
```

The final process in the taxonomic step is to ensure that the spellings of any species noted in the *vegetation*, or *associates* fields are accurate. These *will not* then be submitted to Kew, as the fields could quickly overwhelm the API.

```{r check spellings of vegetation and associated species, eval = F}

```

## Directions

directions to a parking spot can be acquired from Google Maps; however this implies the location can be driven to in the first place. 
This will require interactive alterations from the user.

```{r site directions, eval = F}
SoS_gkey = Sys.getenv("Sos_gkey")

directions2site <- directions_grabber(data, api_key = SoS_gkey)
```

## Export collections as 'shapefile' or KML for use in GIS or GoogleEarth

```{r write out data to googlesheets4}

# we will add these data onto our final sheet.  THIS IS THE LAST STEP OF THIS TUTORIAL
sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
             sheet = 'Processed - Examples', data = df)

```

```{r write out collections in a spatial data format}

```

### 2023-07-06 scripts/taxonomy_setup.Rmd
refining tutorial

--- 2023-05-19 scripts/taxonomy_setup.Rmd
@@
 sppLKPtab <- cla[, c('scientificName', 'specificEpithet', 'infraspecificEpithet', 'verbatimTaxonRank')]
-sppLKPtab <- filter(sppLKPtab, is.na(verbatimTaxonRank) | verbatimTaxonRank != "f.")
+sppLKPtab <- filter(sppLKPtab, is.na(verbatimTaxonRank) | verbatimTaxonRank != "f.") %>% 
+  mutate(Grp = str_extract(specificEpithet, '[a-z]{3}'))
 

### 2023-07-06 scripts/running_pipeline.Rmd
Barneby hath risen and full pipe run complete

@@
 
-```{r setup, message=F}
+```{r setup, message = F}
 library(tidyverse)
@@
 
-```{r load through google sheets, eF}
+```{r load through google sheets}
 
@@
 
-names <- sf::st_drop_geometry(data) |>
-  pull(Full_name)
+names <- data$Full_name
+spellchecks <- lapply(FUN = spell_check, X = names, path = '/hdd/Barneby_Lives-dev/taxonomic_data') |>
+  bind_rows() %>% 
+  filter(Match == 'fuzzy')
 
-names_vec <- c('Astragalis purshii', 'Linnaea borealius var. borealis', 'Heliumorus multifora')
+head(spellchecks)
 
-lapply(FUN = spell_check, X = names_vec, path = '/hdd/Barneby_Lives-dev/taxonomic_data')
-
+rm(names)
 ```
@@
   bind_rows()
+
+data <- bind_cols(data, POW_names) %>% 
+  relocate(starts_with('POW_'), .before = geometry)
+
+dplyr::select(data, starts_with('POW')) |>
+  sf::st_drop_geometry() |>
+  utils::head()
 ```
@@
 directions2site <- directions_grabber(data, api_key = SoS_gkey)
+
 ```
@@
 # we will add these data onto our final sheet.  THIS IS THE LAST STEP OF THIS TUTORIAL
-sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
-             sheet = 'Processed - Examples', data = df)
+googledata <- sf::st_drop_geometry(directions2site) %>% 
+  dplyr::select(-UNIQUEID)
 
+write_sheet( ss = '1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
+             sheet = 'Processed - Examples', data = googledata)
+
+?gs4_create
 ```

### 2023-09-02 manuscript/manuscript_APPS.rmd
working on manuscript

--- 
title: BarnebyLives an R package to create herbarium specimen labels
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$,  Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing quality specimens to herbaria is a time intensive task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museums, geospatial, automation, R
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
---

Herbaria house billions of specimens worldwide. They serve as an essential component of all taxonomic, and phylogenetic studies as well as an enormous number of ecological studies. Recent innovations in computing, digitization and data distribution, DNA sequencing, and the development of myriad analytic approaches have brought about a resurgence in museum based study. However, concerns regarding the rate of acquisition of new specimens exist...

The submittal of specimens to herbaria is a time consuming process, especially for younger collectors with limited experience. 
While many young collectors, who are capable of using dichotomous keys to identify their collections, exist they face difficulties navigating several aspects of data collection. 
This scenario results in the delay of deposition of many specimens, and undoubtedly the deposition of many specimens at all.
Problems which face young collectors generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of education on several cartographic and natural history attributes.

The successful generation of an herbarium specimen includes many steps which are easy to take for granted. 
For example, while the acquisition of political information for a collection site appears simple, it is only so if the collector has the adequate resources at their disposal. 
Given the association of boundaries with topographically complex areas (e.g. watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g. ArcMap, which is relatively expensive at 100$ year), or more likely Google Maps by individual site. 
This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.

More evidently difficult tasks involve taxonomy and the rapid rate at which taxonomic names have changed since the publication of many Floras.

# METHODS AND RESULTS

Here we showcase the reduction in errors attributtal to BarnebyLives

\begin{wrapfigure}{l}{0.4\textwidth}
  \centering
    \includegraphics[width=0.4\textwidth]{../graphics/plots/flowchart.png}
  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
\end{wrapfigure}

# CONCLUSIONS

# ACKNOWLEDGEMENTS

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. The package includes three real use-case vignettes (tutorials) on usage. One vignette "setting_up_files" explores setting up a instance for a certain geographic area. Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels.

# ORCID

Jeremie Fant https://orcid.org/0000-0001-9276-1111

# REFERENCES

### 2023-09-02 scripts/running_pipeline.Rmd
working on manuscript

@@
 
-data <- bind_cols(data, POW_names) %>% 
-  relocate(starts_with('POW_'), .before = geometry)
+data <- bind_cols(data, POW_names)
 
@@
   utils::head()
+
+data <- powNAce(data)
+
 ```
@@
 directions2site <- directions_grabber(data, api_key = SoS_gkey)
-
 ```
@@
 
-?gs4_create
 ```

### 2023-09-02 manuscript/flowchart.R

@@ -1,0 +1,87 @@
+library(tidyverse)
+library(DiagrammeR)
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
+
+grViz("
+
+digraph boxes_and_circle {
+  
+  node [shape = oval, fontsize = 28, fontname = arial, style = filled, fontcolor = White]
+
+      subgraph cluster_import{color= 'white'
+        node [color = '#2F0A28']
+        'Raw Spreadsheet'
+      }
+      
+        # Geospatial Operations
+      subgraph cluster_geospatial { style=dashed; color= 'white'; 
+        node [color = '#ffa600']
+  
+          'Raw Spreadsheet' -> 'Clean Coordinates' 
+              [label = '     dms2dd', fontsize = 26, penwidth = 6]
+          'Clean Coordinates' -> 'Geospatial Data' 
+              [label = '     coords2sf', fontsize = 26, penwidth = 6]
+          'Geospatial Data' -> 'Political\nBoundaries' 
+              [label = '     political_grabber', fontsize = 26, penwidth = 6]
+          'Political\nBoundaries' -> 'Site\nCharacteristics'
+              [label = '     physical_grabber', fontsize = 26, penwidth = 6]
+          'Site\nCharacteristics' -> 'Site Name' 
+              [label = '     site_writer', fontsize = 26, penwidth = 6]
+              
+          'Geospatial Data' -> 'Directions'
+              [label = '     directions_grabber', fontsize = 26, penwidth =6, style=dashed]
+          'Directions' -> 'Site Name'  [penwidth = 6, style=dashed]
+      }
+              
+              
+          # Taxonomic Operations
+      subgraph cluster_taxonomy { style=dashed; color= 'white'; 
+        node [color = '#176B87']
+  
+          'Site Name' -> 'Spell Check (Collection)' 
+              [label = '     spell_check', fontsize = 26, penwidth = 6]
+          'Spell Check (Collection)' -> 'Spell Check\n(Authorities)' 
+              [label = '     author_check', fontsize = 26, penwidth = 6]
+          'Spell Check\n(Authorities)' -> 'Spell Check\n(Associated Species)' 
+              [label = '    associates_check', fontsize = 26, penwidth = 6]
+              
+          'Spell Check (Collection)' -> 'Synonym\nCheck' 
+              [label = '    powo_searcher', fontsize = 26, penwidth = 6, style=dashed]
[…]

### 2023-09-03 manuscript/DataSourceTable.Rmd
working on on introduction

---
title: "DataSourcesTable"
author: "steppe"
date: "2023-09-02"
output: html_document
---

```{r echo = F, warning = F, message=F}
library(knitr)
library(kableExtra)
library(gt)
library(tidyverse)
```

```{r Create Table}

local_data <- data.frame(
  Variable = c(
    'County', 'State', 'Ownership', 
    'TRS', 'Place Names', 'Mountains',
    'Elevation', 'Slope', 'Aspect',
    'Geomorphons', 'Surficial Geology', 'Taxonomic Spellings', 
    'Author Abbreviations'
    ),
    Usage = c(
    'Political', 'Political', 'Political',
    'Political', 'Site Name', 'Site Name', 
    'Site Characteristics', 'Site Characteristics', 'Site Characteristics', 
    'Site Characteristics', 'Site Characteristics', 'Spell Checks', 
    'Spell Checks'
    ),
    Source = c(
    'US Census Bureau', 'US Census Bureau', 'US Geological Survey',
    'US Geological Survey', 'US Geological Survey', 'EarthEnv', 
    'Open Topography', 'Open Topography', 'Open Topography',
    'Open Topography', 'World Flora Online', 'US Geological Survey',
    'IPNI'
    ),
  Name = c(
    'Counties', 'States', 'Protected Areas Database', 
    'Public Land Survey System', 'Geographic Names Information System', 'GMBA Mountain Inventory v2',
    'Geomorpho90m - Elevation', 'Geomorpho90m - Slope', 'Geomorpho90m - Aspect',
    'Geomorpho90m - Geomorphons', 'State Geologic Map Compilation', 'World Flora Online',
    'International Plant Names Index'
    ),
  Data_Model = c(
    'Vector', 'Vector', 'Vector',
    'Vector', 'Vector', 'Vector',
    'Raster', 'Raster', 'Raster', 
    'Raster', 'Vector', 'Tabular',
    'Tabular'
    ),
  Subset_Size = c(
    0.073, 0.0, 0.435,
    0.816, 0.081, 0.004,
    4.2, 4.6, 4.1,
    0.455, 0.708, 0.002,
    0.001
  )
)

```

### 2023-09-03 manuscript/manuscript_APPS.rmd
working on on introduction

@@
 
-Herbaria house billions of specimens worldwide.
-They serve as an essential component of all taxonomic, and phylogenetic studies as well as an enormous number of ecological studies.
-Recent innovations in computing, digitization and data distribution, DNA sequencing, and the development of myriad analytic approaches have brought about a resurgence in museum based study.
-However, concerns regarding the rate of acquisition of new specimens exist...
+Worldwide herbaria house roughly 400 million plant specimens (Thiers 2019).
+These specimens were largely collected with the goal of describing the taxonomic diversity of the plant kingdom, and the floristic diversity of the world (Greve et al. 2016). 
+As research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity in the late 20th century, the rate of accessioning new collections to herbaria diminished (Prather 2004, Pyke & Erlich 2010).
+Among other factors, this lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (Funk 2014). 
+Fortunately, renewed interest in collections have brought them back to the forefront of plant sciences ( Ronsted et al. 2020).
+
+Recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have brought about a renaissance in herbarium based studies (Greve et al.
+2016, James et al.
+2019, Brewer et al.
+2019, Ronsted et al.
+2020).
+Current uses of specimen based data extend far beyond their traditional roles in systematics, and studies utilizing collections are regularly carried out to better understand .... of plants (). 
+However, we anticipate that collections will gain their most widespread appreciation and utilization as desires to revitalize the role of natural history in ecology via it's combination with novel approaches, such as remote sensing, meta-barcoding, and landscape studies and emerging in ecology and environmental science (Fevo et al.
+2021).
+
+However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position to share knowledge of collecting and depositing specimens (Daru et al 2017)
 
 The submittal of specimens to herbaria is a time consuming process, especially for younger collectors with limited experience.
+
 While many young collectors, who are capable of using dichotomous keys to identify their collections, exist they face difficulties navigating several aspects of data collection. 
@@
 
+# AUTHOR CONTRIBUTIONS
+The project was conceptualized by R.C.B.
+The program was written by R.C.B Data collection and analysis were performed by R.C.B.
+R.C.B.
+wrote the manuscript with input from all other authors.
+All authors approved the final version of the manuscript.
+
 # ACKNOWLEDGEMENTS

### 2023-09-04 manuscript/manuscript_APPS.rmd
adding sys times

--- 2023-09-02 manuscript/manuscript_APPS.rmd
@@
 author:  |
-    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$,  Jeremie B. Fant$^1$$^,$$^2$
+    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$, Pat Herendeen$^1$$^,$$^2$, Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
@@
 keywords: |
-  Herbarium, collections, natural history museums, geospatial, automation, R
+  Herbarium, collections, natural history museum, geospatial, automation, R, software
 output:
@@
 
-Herbaria house billions of specimens worldwide.
-They serve as an essential component of all taxonomic, and phylogenetic studies as well as an enormous number of ecological studies.
-Recent innovations in computing, digitization and data distribution, DNA sequencing, and the development of myriad analytic approaches have brought about a resurgence in museum based study.
-However, concerns regarding the rate of acquisition of new specimens exist...
+```{r}
+knitr::opts_chunk$set(echo=F, warning = F, message = F)
+```
 
-The submittal of specimens to herbaria is a time consuming process, especially for younger collectors with limited experience. 
-While many young collectors, who are capable of using dichotomous keys to identify their collections, exist they face difficulties navigating several aspects of data collection. 
-This scenario results in the delay of deposition of many specimens, and undoubtedly the deposition of many specimens at all.
-Problems which face young collectors generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of education on several cartographic and natural history attributes.
+Nearly 400 million specimens are housed in herbaria around the world (@thiers2021herbaria).
+These specimens were collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity (@greve2016realising). 
+The rate of accessioning new collections to herbaria diminished in the 20th centry as research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity (@prather2004decline, @pyke2010biological).
+Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion).
+Fortunately, renewed interest in collections have brought herbaria back to the forefront of plant sciences (@ronsted2020integrative).
+
+Recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have brought about a renaissance in herbarium based studies (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
+Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand ....
+of plants ().
+However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, and ...
+(@tosa2021rapid).
+
+However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position where the skills of collecting and processing specimens have declined amongst young persons (Daru et al 2017).
+The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience in the process. 
+While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection.
+This scenario results in not only the delay in the deposition of many specimens, but undoubtedly the deposition of many collections at all.
+Problems which young collectors face generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of formal education on cartography, natural history, and taxonomy at most universities.
 
@@
 
-Here we showcase the reduction in errors attributtal to BarnebyLives
-
 \begin{wrapfigure}{l}{0.4\textwidth}
@@
 
+```{r load libraries}
+library(tidyverse) # data operations
+library(sf) # spatial data
+library(BarnebyLives) # for helping accession collections
+```
+
+```{r import example data and gather some summaries}
+
+test_data <- read.csv('data/test_data.csv', na.strings = "") %>% 
+  drop_na(longitude) %>% 
+  drop_na(latitude) %>% 
+  drop_na(date_digital) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)
+
+n_families <- test_data %>% 
+  st_drop_geometry() %>% 
+  group_by(family) %>% 
+  count() %>% 
+  nrow() # 74 families
+
+n_spp <- test_data %>% 
+  st_drop_geometry() %>% 
+  group_by(binomial) %>% 
+  count() %>% 
+  nrow() # 616 species
+
+n_infraspecies <- test_data %>% 
+  st_drop_geometry() %>% 
+  drop_na(infraspecies) %>% 
+  group_by(binomial, infraspecies) %>% 
+  count() %>% 
+  nrow() # 66 distinct infraspecies
+
+n_sp_authors <- test_data %>% 
+  st_drop_geometry() %>% 
+  drop_na(authority) %>% 
+  count() # 557 groups of authors
+
+n_infra_sp_auths <- test_data %>% 
+  st_drop_geometry() %>% 
+  drop_na(authority1) %>% 
+  group_by(binomial, infraspecies) %>% 
+  count() %>% 
+  nrow() # 22 distinct infra species author groups
+```
+```{r Run pipeline with all steps and benchmarking, eval = F}
+
+system.time({ # if neccessary convert coordinates in degrees minutes second to decimal degrees
+  data <- dms2dd(data, dms = F)
+})
+
+time_coords2sf <- system.time({ # create a spatial (simple features) object
+  data <- coords2sf(data)
+})
+
+time_political_grabber <- system.time({ # grab political information for collection
+  political_grabber(data, y = 'Collection_number', path = p)
+})
+
+time_physical_grabber <- system.time({ # grab sites physical information
+  data <- physical_grabber(data, path = p)
+})
+
+time_site_writer <- system.time({ # write site location notes
+  data <- site_writer(data, path = p)
+})
+
+time_directions_grabber <- system.time({ # write directions to sites
+  SoS_gkey = Sys.getenv("Sos_gkey")
+  data <- directions_grabber(data, api_key = SoS_gkey)
+})
+
+time_spell_check <- system.time({ # ensure appropriate spellings at taxonomic levels
+  data <- spell_check(data, path = p)
+})
+
+time_author_check <- system.time({ # ensure authorities are spelled-noted correctly
+  data <- author_check(data, path = p)
+})
+
+time_powo_searcher <- system.time({ # search for synonyms from plants of the world online
+
+names <- sf::st_drop_geometry(data) %>% 
+  pull(Full_name)
+
+pow_res <- lapply(names,
+      powo_searcher) %>% 
+  bind_rows()
+data <- bind_cols(data, pow_res)
+
+rm(names, pow_res)
+})
+
+time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
+  data <- associate_dropper(data)
+})
+
+time_date_parser <- system.time({ # parse dates into museum formats
+  data <- date_parser(input, coll_date = 'Date_digital', det_date = 'Determined_date')
+})
+
+time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
+  geodata_writer(data, path = 'someplace', 
+               filename = 'Herbarium_Collections_2023',
+               filetype = 'kml')
+})
+
+```
+
+Here we showcase the reduction in errors which are associated with the implementation of BarnebyLives.
+
 # CONCLUSIONS
 
+# AUTHOR CONTRIBUTIONS
+The project was conceptualized by R.C.B.
+The program was written by R.C.B Data collection and analysis were performed by R.C.B.
+R.C.B.
+wrote the manuscript with input from all other authors.
+All authors approved the final version of the manuscript.
+
 # ACKNOWLEDGEMENTS
+The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for the majority of his specimen collection activities.
+Two anonymous peer reviewers who increased the quality of this manuscript are thanked.
+Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Hubert Szczygiel, Caitlin Miller.
 

### 2023-09-04 manuscript/DataSourceTable.R

@@ -1,17 +1,5 @@
----
-title: "DataSourcesTable"
-author: "steppe"
-date: "2023-09-02"
-output: html_document
----
-
-
-```{r echo = F, warning = F, message=F}
 library(gt)
-```
-
-
-```{r Create Table}
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
 
 footnote = 'Counties and States are merged into the same dataset while setting up the package. The value for "County" includes State.'
 
@@ -22,35 +10,35 @@ local_data <- data.frame(
     'Elevation', 'Slope', 'Aspect',
     'Geomorphons', 'Surficial Geology', 'Taxonomic Spellings', 
     'Author Abbreviations'
-    ),
-    Usage = c(
+  ),
+  Usage = c(
     'Political', '', '',
     '', 'Site Name', 'Site Name', 
     'Site Characteristics', '', '', 
     '', '', 'Spell Checks', 
     ''
-    ),
-    Source = c(
+  ),
+  Source = c(
     'US Census Bureau', '', 'US Geological Survey',
     '', '', 'EarthEnv', 
     'Open Topography', '', '',
     '', 'US Geological Survey', 'World Flora Online',
     'IPNI'
-    ),
+  ),
   Name = c(
     'Counties', 'States', 'Protected Areas Database', 
     'Public Land Survey System', 'Geographic Names Information System', 'GMBA Mountain Inventory v2',
     'Geomorpho90m - Elevation', 'Geomorpho90 - Slope', 'Geomorpho90m - Aspect',
     'Geomorpho90m - Geomorphons', 'State Geologic Map Compilation', 'World Flora Online',
     'International Plant Names Index'
-    ),
+  ),
   Data_Model = c(
     'Vector', '', '',
     '', '', '',
     'Raster', '', '', 
     '', 'Vector', 'Text',
     ''
-    ),
+  ),
   Subset_Size = c(
     '0.073', '0.0*', '0.435',
     '0.816', '0.081', '0.004',
@@ -64,13 +52,13 @@ data_sources <- local_data |>
   gt() |>  
   fmt_markdown(columns = everything())|> 
   tab_header(
-      title = 'Data Sources for Package')|> 
+    title = 'Data Sources for Package')|> 
   tab_style(style = cell_text(weight = "bold"),
             locations = cells_column_labels()) |> 
   cols_align(align = "center" ) |> 
   cols_label(
[…]

### 2023-09-04 manuscript/GeographicExtentPlot.R

@@ -1,0 +1,43 @@
+# Geographic extent of tested data
+
+library(sf)
+library(tidyverse)
+library(tigris)
+library(rnaturalearth)
+
+polybb <- st_as_sfc(
+  st_bbox(
+    c(xmin = -85, 
+      xmax = -125, 
+      ymax = 30, 
+      ymin = 50
+      ), 
+    crs = st_crs(4326)
+    )
+  ) 
+
+lakes <- ne_download(category = 'physical', type='lakes', scale = 'small') %>% 
+  st_as_sf() %>% 
+  filter(name_alt == 'Great Lakes') %>% 
+  st_make_valid() 
+
+countries <- ne_countries() %>% 
+  st_as_sf() %>% 
+  filter(name %in% c('United States of America', 'Canada', 'Mexico'))
+
+states <- states() %>% 
+  st_transform(4326) 
+
+test_data <- read.csv('data/test_data.csv') %>% 
+  drop_na(longitude) %>% 
+  drop_na(latitude) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)
+
+ggplot() +
+  geom_sf(data = countries) + 
+  geom_sf(data = states) + 
+  geom_sf(data = lakes, fill = 'blue') + 
+  geom_sf(data = test_data, shape = 1) +
+  geom_sf(data = polybb, fill = NA, linewidth = 2, color = '#2F0A28') + 
+  coord_sf(xlim = c(-85, -125), ylim = c(30, 50))
+

### 2023-09-04 manuscript/manuscript_APPS.rmd
modded  column headers in text file for simple example

@@
 library(tidyverse) # data operations
-library(sf) # spatial data
+# devtools::install_github('sagesteppe/BarnebyLives')
 library(BarnebyLives) # for helping accession collections
@@
 
-test_data <- read.csv('data/test_data.csv', na.strings = "") %>% 
-  drop_na(longitude) %>% 
-  drop_na(latitude) %>% 
-  drop_na(date_digital) %>% 
-  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)
+data <- read.csv('data/test_data.csv', na.strings = "") %>% 
+  drop_na(c('longitude', 'latitude', 'date_digital'))
 
-n_families <- test_data %>% 
-  st_drop_geometry() %>% 
+n_families <- data %>% 
   group_by(family) %>% 
@@
 
-n_spp <- test_data %>% 
-  st_drop_geometry() %>% 
+n_spp <- data %>% 
   group_by(binomial) %>% 
@@
 
-n_infraspecies <- test_data %>% 
-  st_drop_geometry() %>% 
+n_infraspecies <- data %>% 
   drop_na(infraspecies) %>% 
@@
 
-n_sp_authors <- test_data %>% 
-  st_drop_geometry() %>% 
+n_sp_authors <- data %>% 
   drop_na(authority) %>% 
@@
 
-n_infra_sp_auths <- test_data %>% 
-  st_drop_geometry() %>% 
+n_infra_sp_auths <- data %>% 
   drop_na(authority1) %>% 
@@
   nrow() # 22 distinct infra species author groups
+
 ```
+
 ```{r Run pipeline with all steps and benchmarking, eval = F}
 
-system.time({ # if neccessary convert coordinates in degrees minutes second to decimal degrees
+time_dms2dd <- system.time({ # if necessary convert coordinates in degrees minutes second to decimal degrees
   data <- dms2dd(data, dms = F)
+})
+
+time_autofill_checker <- system.time({ # has the spreadsheet software auto-incremented coordinate values?
+  data <- autofill_checker(data)
 })
@@
 
+p2geo <- '/media/steppe/hdd/Barneby_Lives-dev/geodata'
+
 time_political_grabber <- system.time({ # grab political information for collection
-  political_grabber(data, y = 'Collection_number', path = p)
+  data <- political_grabber(data, y = 'Collection_number', path = p2geo)
 })
@@
 time_physical_grabber <- system.time({ # grab sites physical information
-  data <- physical_grabber(data, path = p)
+  data <- physical_grabber(data, path = p2geo)
+}) # not finding slope and aspect
+
+time_site_writer <- system.time({ # write site location notes
+  data <- site_writer(data, path = p2geo)
 })
 
-time_site_writer <- system.time({ # write site location notes
-  data <- site_writer(data, path = p)
-})
+# time_directions_grabber <- system.time({ # write directions to sites
+#   SoS_gkey = Sys.getenv("Sos_gkey")
+#   data <- directions_grabber(data, api_key = SoS_gkey)
+# })
 
-time_directions_grabber <- system.time({ # write directions to sites
-  SoS_gkey = Sys.getenv("Sos_gkey")
-  data <- directions_grabber(data, api_key = SoS_gkey)
-})
+p2tax <- '/media/steppe/hdd/Barneby_Lives-dev/taxonomic_data'
 
 time_spell_check <- system.time({ # ensure appropriate spellings at taxonomic levels
-  data <- spell_check(data, path = p)
+  data <- spell_check(data, path = p2tax)
 })
@@
 time_author_check <- system.time({ # ensure authorities are spelled-noted correctly
-  data <- author_check(data, path = p)
+  data <- author_check(data, path = p2tax)
 })
 
-time_powo_searcher <- system.time({ # search for synonyms from plants of the world online
+# time_powo_searcher <- system.time({ # search for synonyms from plants of the world online
 
-names <- sf::st_drop_geometry(data) %>% 
-  pull(Full_name)
+# names <- sf::st_drop_geometry(data) %>% 
+#   pull(Full_name)
 
-pow_res <- lapply(names,
-      powo_searcher) %>% 
-  bind_rows()
-data <- bind_cols(data, pow_res)
+# pow_res <- lapply(names,
+#       powo_searcher) %>% 
+ #  bind_rows()
+# data <- bind_cols(data, pow_res)
 
-rm(names, pow_res)
-})
+# rm(names, pow_res)
+# })
 

### 2023-09-04 manuscript/flowchart.R

@@ -17,8 +17,10 @@ digraph boxes_and_circle {
       subgraph cluster_geospatial { style=dashed; color= 'white'; 
         node [color = '#ffa600']
   
-          'Raw Spreadsheet' -> 'Clean Coordinates' 
+          'Raw Spreadsheet' -> 'Convert Coordinates\nfrom DMS to DD' 
               [label = '     dms2dd', fontsize = 26, penwidth = 6]
+          'Convert Coordinates\nfrom DMS to DD' -> 'Clean Coordinates' 
+              [label = '     autofill_checker', fontsize = 26, penwidth = 6]
           'Clean Coordinates' -> 'Geospatial Data' 
               [label = '     coords2sf', fontsize = 26, penwidth = 6]
           'Geospatial Data' -> 'Political\nBoundaries' 
@@ -60,14 +62,19 @@ digraph boxes_and_circle {
               [label = '     field_lengths', fontsize = 26, penwidth = 6]
           'Check Field Lengths' -> 'Date Formatting' 
               [label = '     date_parser', fontsize = 26, penwidth = 6]
+
       }
       
       node [color = '#C94277']
-          'Date Formatting' -> 'Processed Spreadsheet' 
+          'Date Formatting' -> 'Processed\nSpreadsheet' 
               [penwidth = 6]
-              
-          'Processed Spreadsheet' -> 'Manual Review' 
+          'Processed\nSpreadsheet' -> 'Manual Review' 
               [penwidth = 6]
+                            
+          'Date Formatting' -> 'Geodata' 
+              [label = '     geodata_writer', fontsize = 26, penwidth = 6, style=dashed]
+          'Geodata' -> 'Manual Review'
+              [penwidth = 6, style = dashed]
       
           # Printed Materials
       subgraph cluster_printing {style=dashed; color='white'; 

### 2023-09-08 manuscript/manuscript_APPS.rmd
working on splitting data from a binomial input column

@@
 author:  |
-    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$, Pat Herendeen$^1$$^,$$^2$, Jeremie B. Fant$^1$$^,$$^2$
+    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
@@
 
-```{r}
+```{r echo = F}
 knitr::opts_chunk$set(echo=F, warning = F, message = F)
@@
 
+\begin{wrapfigure}{l}{0.35\textwidth}
+  \centering
+    \includegraphics[width=0.35\textwidth]{../graphics/plots/flowchart-trim.png}
+  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
+\end{wrapfigure}
+
 # METHODS AND RESULTS
 
-\begin{wrapfigure}{l}{0.4\textwidth}
-  \centering
-    \includegraphics[width=0.4\textwidth]{../graphics/plots/flowchart.png}
-  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
-\end{wrapfigure}
+`r lorem::ipsum(paragraphs = 5)`
 
@@
 data <- read.csv('data/test_data.csv', na.strings = "") %>% 
-  drop_na(c('longitude', 'latitude', 'date_digital'))
+  drop_na(c('Longitude', 'Latitude', 'Date_digital'))
 
 n_families <- data %>% 
-  group_by(family) %>% 
+  group_by(Family) %>% 
   count() %>% 
@@
 n_spp <- data %>% 
-  group_by(binomial) %>% 
+  group_by(Binomial) %>% 
   count() %>% 
@@
 n_infraspecies <- data %>% 
-  drop_na(infraspecies) %>% 
-  group_by(binomial, infraspecies) %>% 
+  drop_na(Infraspecies) %>% 
+  group_by(Binomial, Infraspecies) %>% 
   count() %>% 
@@
 n_sp_authors <- data %>% 
-  drop_na(authority) %>% 
+  drop_na(Binomial_authority) %>% 
   count() # 557 groups of authors
@@
 n_infra_sp_auths <- data %>% 
-  drop_na(authority1) %>% 
-  group_by(binomial, infraspecies) %>% 
+  drop_na(Infraspecific_authority) %>% 
+  group_by(Binomial, Infraspecies) %>% 
   count() %>% 
@@
   data <- physical_grabber(data, path = p2geo)
-}) # not finding slope and aspect
+})
 
@@
 })
-
+lorem
 ```
@@
 
+\begin{wrapfigure}{r}{0.45\textwidth}
+  \centering
+    \includegraphics[width=0.45\textwidth]{../graphics/tables/Screenshot-DataSources.png}
+  \caption{Sources of Data required for operations}
+\end{wrapfigure}
+
+\newpage
 # CONCLUSIONS
+
+`r lorem::ipsum(paragraphs = 3)`
 
@@
 Two anonymous peer reviewers who increased the quality of this manuscript are thanked.
-Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Hubert Szczygiel, Caitlin Miller.
+Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel.
 
@@
 A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels.
+All data used in this mansucript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript
 
@@
 
+\newpage 
+\small
 # REFERENCES

### 2023-09-09 manuscript/manuscript_APPS.rmd
adding tding text to methods

@@
 --- 
-title: BarnebyLives an R package to create herbarium specimen labels
+title: BarnebyLives an R package to create herbarium specimen labels and digital data sheets
 author:  |
@@
 abstract:  |  
-  **Premise:** Depositing quality specimens to herbaria is a time intensive task.
+  **Premise:** Depositing specimens to herbaria is a time consuming task.
 Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition.
@@
 - \usepackage{wrapfig}
+- \usepackage[export]{adjustbox}
 ---
@@
 These specimens were collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity (@greve2016realising). 
-The rate of accessioning new collections to herbaria diminished in the 20th centry as research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity (@prather2004decline, @pyke2010biological).
+The rate of accessioning new collections to herbaria diminished in the 20th century as research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity (@prather2004decline, @pyke2010biological, @daru2018widespread).
 Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion).
-Fortunately, renewed interest in collections have brought herbaria back to the forefront of plant sciences (@ronsted2020integrative).
+Fortunately, renewed interest in collections have brought herbaria of all sizes back to the forefront of plant sciences (@ronsted2020integrative, @marsico2020small).
 
 Recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have brought about a renaissance in herbarium based studies (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand ....
-of plants ().
-However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, and ...
-(@tosa2021rapid).
+Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
+However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, electronic sensing (@tosa2021rapid).
 
-However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position where the skills of collecting and processing specimens have declined amongst young persons (Daru et al 2017).
+However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position where the skills of collecting and processing specimens, and time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
 The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience in the process. 
@@
 This scenario results in not only the delay in the deposition of many specimens, but undoubtedly the deposition of many collections at all.
-Problems which young collectors face generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of formal education on cartography, natural history, and taxonomy at most universities.
+Problems which young collectors face generally include both the lack of dedicated time awarded to them at a seasons end to process specimens, and a general lack of formal education on cartography, natural history, taxonomy, and plant systematics.
 
@@
   \centering
-    \includegraphics[width=0.35\textwidth]{../graphics/plots/flowchart-trim.png}
+    \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
   \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
@@
 library(tidyverse) # data operations
-# devtools::install_github('sagesteppe/BarnebyLives')
+# devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
 library(BarnebyLives) # for helping accession collections
 ```
+
+## Usage
+BarnebyLives is run entirely from within Rstudio.
+Data may be read in from: Excel, software which can process Comma-separated Values (CSV's) such as LibreOffice, OpenOffice (or Excel), or via the cloud on Googlesheets.
+The latter two options are documented here and in package vignettes.
+BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
+Virtually all of the on-disk memory associated with these data are for storing geospatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation.
+Functions which require the on-disk data require a path to the data as an argument.
+We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, or have the data stored entirely on thumb-drives.
+The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a TeX driver (e.g.
+pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
+While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
 
@@
 data <- read.csv('data/test_data.csv', na.strings = "") %>% 
-  drop_na(c('Longitude', 'Latitude', 'Date_digital'))
+  drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
+  unite(col = 'Scientific_name', Binomial:Infraspecific_authority, na.rm=TRUE, sep = " ", remove = F)
 
@@
   drop_na(Binomial_authority) %>% 
-  count() # 557 groups of authors
+  count() %>% # 557 groups of authors
+  pull()
 
@@
 
+### Herbarium Collections
+The package was finalized using the primary authors collections from 2023.
+The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned.
+Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used.
+This results in a data set of XX records for testing, from XX sites located across Western North America FIGURE XX.
+In total `r n_spp` species (with `r n_sp_authors` authorships), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
+
 ```{r Run pipeline with all steps and benchmarking, eval = F}
+
+time_split_binomials <- system.time({ # split up names into their components
+  data <- split_binomial(data, 'Scientific_name')
+})
 
@@
 
-# time_directions_grabber <- system.time({ # write directions to sites
-#   SoS_gkey = Sys.getenv("Sos_gkey")
-#   data <- directions_grabber(data, api_key = SoS_gkey)
-# })
-
 p2tax <- '/media/steppe/hdd/Barneby_Lives-dev/taxonomic_data'
 
-time_spell_check <- system.time({ # ensure appropriate spellings at taxonomic levels
+time_spell_check <- system.time({ # ensure appropriate spellings of the species
   data <- spell_check(data, path = p2tax)
+})
+
+time_family_spell_check <- system.time({ # ensure appropriate spelling of the family
+  data <- family_spell_check(data, path = p2tax)
 })
@@
 
-# time_powo_searcher <- system.time({ # search for synonyms from plants of the world online
-
-# names <- sf::st_drop_geometry(data) %>% 
-#   pull(Full_name)
-
-# pow_res <- lapply(names,
-#       powo_searcher) %>% 
- #  bind_rows()
-# data <- bind_cols(data, pow_res)
-
-# rm(names, pow_res)
-# })
-
 time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
-  data <- associate_dropper(data)
+  data <- associate_dropper(data, 'Full_name')
 })
@@
 time_date_parser <- system.time({ # parse dates into museum formats
-  data <- date_parser(input, coll_date = 'Date_digital', det_date = 'Determined_date')
+  data <- date_parser(data, coll_date = 'Date_digital')
 })
@@
 time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
-  geodata_writer(data, path = 'someplace', 
+  geodata_writer(data, path = '~/Documents/BL', 
                filename = 'Herbarium_Collections_2023',
@@
 })
-lorem
+
 ```
 
-Here we showcase the reduction in errors which are associated with the implementation of BarnebyLives.
+```{r Run the API services, eval = F}
+
+# we keep these processes in a discrete chunk set not to evaluate so as to not overwhelm
+# the services. Google does charge if the number of queries per month is high.
+
+time_powo_searcher <- system.time({ # search for synonyms from plants of the world online
+
+ names <- sf::st_drop_geometry(data) %>% 
+   pull(Full_name)
+
+ pow_res <- lapply(names,
+       powo_searcher) %>% 
+       bind_rows()
+ data <- bind_cols(data, pow_res)
+
+ rm(names, pow_res)
+}) # has been run
+
+saveRDS(time_powo_searcher, file = 'data/processed/time_powo_searcher')
+saveRDS(data, file = 'data/processed/data_w_POWO_search')
+
+time_directions_grabber <- system.time({ # write directions to sites
+  SoS_gkey = Sys.getenv("Sos_gkey")
+  data <- directions_grabber(data, api_key = SoS_gkey)
+})
+
+saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
+saveRds(data, file = 'data/processed/data_w_Google_Maps')
+```
+
+```{r Total Time of operations}
+
+time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')
+
+time_trials <- data.frame(as.matrix(do.call(rbind, mget(ls(pattern = '^time'))))) %>% 
+  rownames_to_column('Function') %>% 
+  select(-user.child, -sys.child) %>% 
+  mutate(Function = str_remove(Function, 'time_')) %>% 
+  mutate(Module = case_when(
+    Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
+    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 
+                    'political_grabber', 'site_writer') ~ 'Geospatial', 
+    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher') ~ 'Taxonomic',
+    Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
+  ))
+
+# save these as RDS so operations do not need to be run each time document knits
+
+rm(time_trials)
+
+# rm(ls(pattern = '^time'))
+```
+
+It took roughly four (XX exact) minutes to run all local steps of BarnebyLives, and a further XX minutes to search Plants of the World Online, and XX to search Google Maps for site directions.
+Nearly all of this time was attributable to the spatial operations (spatial:, taxonomic:, style:) The POWO online search is likely capable of being carried out much faster, xx, if one decreases the pauses (which are by default relatively long) between each query, a tactic used to avoid adversely impacting the website.
+The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
+
+Even on data which had been manually cleaned and error-checked by a human BarnebyLives was able to reduce transcription errors, identify typos, make nomenclatural suggestions, and reformat text elements for downsteam use.
 
@@
 # CONCLUSIONS
-
-`r lorem::ipsum(paragraphs = 3)`
+BarnebyLives is a tool which is able to rapidly acquire relevant geographic, and taxonomic data.
+It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data.
+The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
 
@@
 Two anonymous peer reviewers who increased the quality of this manuscript are thanked.
-Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel.
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel.
 
@@
 # ORCID
-
 Jeremie Fant https://orcid.org/0000-0001-9276-1111

### 2023-09-09 manuscript/flowchart.R

@@ -25,10 +25,20 @@ digraph boxes_and_circle {
               [label = '     coords2sf', fontsize = 26, penwidth = 6]
           'Geospatial Data' -> 'Political\nBoundaries' 
               [label = '     political_grabber', fontsize = 26, penwidth = 6]
-          'Political\nBoundaries' -> 'Site\nCharacteristics'
+          'Political\nBoundaries' -> 'Site Characteristics' 
               [label = '     physical_grabber', fontsize = 26, penwidth = 6]
-          'Site\nCharacteristics' -> 'Site Name' 
+          'Site Characteristics' -> 'Site Name' 
               [label = '     site_writer', fontsize = 26, penwidth = 6]
+              
+              
+          'Raw Spreadsheet' -> 'Seperate\nname'
+              [label = '     split_binomial', fontsize = 26, penwidth = 6, style = dashed]
+          'Raw Spreadsheet' -> 'Check for\nautofill values' 
+              [label = '     autofill_checker', fontsize = 26, penwidth = 6, style = dashed]
+          'Seperate\nname' -> 'Convert Coordinates\nfrom DMS to DD' 
+              [penwidth = 6, style = dashed]
+          'Check for\nautofill values' -> 'Convert Coordinates\nfrom DMS to DD' 
+              [penwidth = 6, style = dashed]
               
           'Geospatial Data' -> 'Directions'
               [label = '     directions_grabber', fontsize = 26, penwidth =6, style=dashed]
@@ -40,14 +50,16 @@ digraph boxes_and_circle {
       subgraph cluster_taxonomy { style=dashed; color= 'white'; 
         node [color = '#176B87']
   
-          'Site Name' -> 'Spell Check (Collection)' 
+          'Site Name' -> 'Spell Check (Binomial)' 
               [label = '     spell_check', fontsize = 26, penwidth = 6]
-          'Spell Check (Collection)' -> 'Spell Check\n(Authorities)' 
+          'Spell Check (Binomial)' -> 'Spell Check (Family)' 
+              [label = '     spell_check_family', fontsize = 26, penwidth = 6]
+          'Spell Check (Family)' -> 'Spell Check\n(Authorities)' 
               [label = '     author_check', fontsize = 26, penwidth = 6]
           'Spell Check\n(Authorities)' -> 'Spell Check\n(Associated Species)' 
               [label = '    associates_check', fontsize = 26, penwidth = 6]
               
[…]

### 2023-09-17 manuscript/manuscript_APPS.rmd
time trials in, working on results

@@
 --- 
-title: BarnebyLives an R package to create herbarium specimen labels and digital data sheets
+title: BarnebyLives':' an R package to create herbarium specimen labels and digital data sheets
 author:  |
@@
 
-More evidently difficult tasks involve taxonomy and the rapid rate at which taxonomic names have changed since the publication of many Floras.
+Here we provide a description of the BarnebyLives R package. 
+BarnebyLives was named for plant taxonomist extraordinaire Rupert Charles Barneby (1911-200), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the Willian & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
+Select accolades of Rupert include the Asa Gray Award from the American Society of Plant Taxonomists (ASPT) in 1989, the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congresses's (IBC) Millenium Botany Award (1999) (@welsh2001rupert).
 
@@
     \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
-  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
+  \caption{Recommended workflow.}
 \end{wrapfigure}
+
+More evidently difficult tasks involve taxonomy and the rapid rate at which taxonomic names have changed since the publication of many Floras.
 
@@
 library(tidyverse) # data operations
-# devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
+# devtools::install_github('sagesteppe/BarnebyLives')
 library(BarnebyLives) # for helping accession collections
@@
 BarnebyLives is run entirely from within Rstudio. 
-Data may be read in from: Excel, software which can process Comma-separated Values (CSV's) such as LibreOffice, OpenOffice (or Excel), or via the cloud on Googlesheets.
-The latter two options are documented here and in package vignettes.
+Data may be read in from any common spreadsheet management system or database connection such as Excel, LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all of the on-disk memory associated with these data are for storing geospatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation.
+Virtually all of the on-disk memory associated with these data are for storing geo-spatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation.
 Functions which require the on-disk data require a path to the data as an argument.
+Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
 We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, or have the data stored entirely on thumb-drives.
-The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a TeX driver (e.g.
+The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
 pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
 While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
+Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in FIGURE XX.
 
@@
 
+# number of collection sites
+
+n_sites <- data %>% 
+  distinct(Latitude, Longitude) %>% 
+  nrow()
+
 ```
 
+\begin{wrapfigure}{r}{0.35\textwidth}
+  \centering
+    \includegraphics[width=0.35\textwidth]{../graphics/plots/collections_map-trim.png}
+  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.}
+\end{wrapfigure}
+
 ### Herbarium Collections
+
 The package was finalized using the primary authors collections from 2023.
@@
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used.
-This results in a data set of XX records for testing, from XX sites located across Western North America FIGURE XX.
+This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America FIGURE XX.
 In total `r n_spp` species (with `r n_sp_authors` authorships), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
@@
 
+time_spell_check <- system.time({ # ensure appropriate spellings of the species
+  sc_res <- spell_check(data$Binomial, path = p2tax)
+})
+
 time_family_spell_check <- system.time({ # ensure appropriate spelling of the family
@@
 })
+
+data$Family == data$Family
 
@@
 
+rm(p2geo, p2tax)
 ```
@@
 saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
-saveRds(data, file = 'data/processed/data_w_Google_Maps')
+saveRDS(data, file = 'data/processed/data_w_Google_Maps')
 ```
 
-```{r Total Time of operations}
+```{r Total Time of operations, eval = F}
 
 time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')
+time_directions_grabber <- readRDS('data/processed/time_directions_grabber')
 
@@
     Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 
-                    'political_grabber', 'site_writer') ~ 'Geospatial', 
+                    'political_grabber', 'site_writer', 'geodata_writer', 'directions_grabber') ~ 'Geospatial', 
     Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher') ~ 'Taxonomic',
     Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
+    ),
+    Group = if_else(Function %in% c('powo_searcher', 'directions_grabber'), 'Online', 'Local')
+  )
+
+# save these as RDS so operations do not need to be run each time document knits
+saveRDS(time_trials, file = 'data/processed/time_trials')
+
+rm(time_trials)
+rm(list=ls(pattern = 'time'))
+```
+
+```{r load in time trials}
+time_trials <- readRDS('data/processed/time_trials')
+tt_local <- time_trials[time_trials$Group=='Local',]
+```
+
+It took roughly three minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) minutes to run all local steps of BarnebyLives, and a further 15 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
+Nearly all of the local run time is attributable to the spatial operations (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s, style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`, taxonomic: `r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`).
+The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
+
+```{r}
+rm(time_trials, tt_local)
+```
+
+## Results
+
+```{r import results}
+data_p <- readRDS(file = 'data/processed/data_w_POWO_search')
+
+drop_na(data_p, Binomial_authority) %>% 
+  nrow() # nearly all relating to the absence/presence of a period
+
+drop_na(data_p, Infra_auth_issues) %>% 
+  nrow() # nearly all relating to the absence/presence of a period
+
+drop_na(data_p, Long_AutoFill_Flag) %>% 
+  nrow() # 4 auto fill cases
+
+drop_na(data_p, Lat_AutoFill_Flag) %>% 
+  nrow() # 2 autofill cases
+
+## Family spell check; this is a comparision of my input spelling, and the results
+## of the family spell check function
+
+Family_in_positions <- data %>% 
+  drop_na(Family)
+Family_in <- Family_in_positions %>% 
+  filter(Collection_number %in% data_p$Collection_number) %>% 
+  arrange(Collection_number) %>% 
+  pull(Family)
+
+Family_out <- data_p %>% 
+  filter(Collection_number %in% Family_in_positions$Collection_number) %>% 
+  pull(Family)
+Family_in[ which( Family_in != Family_out ) ]
+
+## Family Level Results
+# these species were found by POW, and according to POW, the species are in the f
+# following families. 
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(POW_Family != 'NOT FOUND') %>% 
+  filter(POW_Family != Family) %>% 
+  select(Collection_number, Full_name, Family, POW_Family) %>% 
+  mutate(Situation = 
+    case_when(
+      Family == 'Hydrophyllaceae' ~ 'Preferred', 
+      Family %in% c('Plantaginaceae', 'Athyriaceae') ~ 'Outdated',
+      Family %in% c('Saxifragaceae', 'Solanaceae') ~ 'Submitter_Incorrect', 
+      Family == 'Bataceae' ~ 'Internal Error', 
+      .default = 'Typo'
     ))
 
-# save these as RDS so operations do not need to be run each time document knits
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(POW_Genus == 'NOT FOUND') %>% 
+  select(Collection_number, Full_name, Genus, POW_Genus)
 
-rm(time_trials)
+# how many of my genera were mis-spelled??
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(POW_Genus != 'NOT FOUND') %>% 
+  filter(POW_Genus != Genus)
 
-# rm(ls(pattern = '^time'))
+# the mismatch between POW genus and not
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(POW_Genus != 'NOT FOUND') %>% 
+  filter(POW_Genus != Genus) %>% 
+  select(Collection_number, Full_name, Genus, POW_Genus) %>% 
+  mutate(Situation = 
+    case_when(
+      Genus %in% c('Pleuraphis', 'Glandularia') ~ 'Preferred', # check Pleuraphis, i am confused on this still....
+      Genus %in% c('Oenothera', 'Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia',
+                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
+                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
+                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
+                    ) ~ 'Outdated',
+      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal Error',
+      .default = 'Typo'
+    ),
+    Situation = if_else(Full_name == 'Oenothera caespitosa Nutt.', 'Internal Error', Situation))
+
+data_p %>% 
+  filter(Binomial_authority != POW_Name_authority)
+
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(Epithet != POW_Epithet ) %>% 
+  select(Collection_number, Full_name, Epithet, POW_Epithet)
+
+# NEED TO WORK ON INPUT DATA, INCONSISTENT USE OF AUTHORSHIP IN THE FIRST COLUMN
+data_p %>% 
+  sf::st_drop_geometry() %>% 
+  filter(Full_name != POW_Full_name) %>% 
+  select(Collection_number, Full_name, Binomial, POW_Full_name) 
 ```
 
-It took roughly four (XX exact) minutes to run all local steps of BarnebyLives, and a further XX minutes to search Plants of the World Online, and XX to search Google Maps for site directions.
-Nearly all of this time was attributable to the spatial operations (spatial:, taxonomic:, style:) The POWO online search is likely capable of being carried out much faster, xx, if one decreases the pauses (which are by default relatively long) between each query, a tactic used to avoid adversely impacting the website.
-The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
-
-Even on data which had been manually cleaned and error-checked by a human BarnebyLives was able to reduce transcription errors, identify typos, make nomenclatural suggestions, and reformat text elements for downsteam use.
+Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
+The number of family misspellings were XX (% percent), the number of misspelled genera were XX (% percent), the number of misspelled binomials were XX (% percent).
+The number of author abbreviations which were not in the appropriate format were XX (% percent), generally the presence or absence of a period were the issue.
+Plants of the World Online was able to identify XX new names for the submitted taxa, XX of which the author adopted.
+XX records were appropriately flagged for issues with auto fill incrementation of the longitude value, and three of these records were also auto-flagged for increases in latitude values (% of records).
 
@@
 The project was conceptualized by R.C.B.
-The program was written by R.C.B Data collection and analysis were performed by R.C.B.
+The program was written by R.C.B.
+Data collection and analysis were performed by R.C.B.
 R.C.B.
@@
 
-\newpage 
 \small
 # REFERENCES
+
+<div id="refs"></div>
+
+# SUPPORTING INFORMATION
+Additional supporting information can be found online in the Supporting Information section at the end of this article.
+
+**Appendix S1.** A table of all time trials for each function.

### 2023-09-17 manuscript/GeographicExtentPlot.R

@@ -4,6 +4,7 @@ library(sf)
 library(tidyverse)
 library(tigris)
 library(rnaturalearth)
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
 
 polybb <- st_as_sfc(
   st_bbox(
@@ -23,21 +24,39 @@ lakes <- ne_download(category = 'physical', type='lakes', scale = 'small') %>%
 
 countries <- ne_countries() %>% 
   st_as_sf() %>% 
-  filter(name %in% c('United States of America', 'Canada', 'Mexico'))
+  filter(name %in% c('United States of America', 'Canada', 'Mexico')) %>% 
+  mutate(Cntry_grp = if_else(sov_a3 %in% c('MEX', 'CAN'), 'International', 'Domestic'))
 
 states <- states() %>% 
   st_transform(4326) 
 
-test_data <- read.csv('data/test_data.csv') %>% 
-  drop_na(longitude) %>% 
-  drop_na(latitude) %>% 
-  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)
+test_data <- read.csv('./data/test_data.csv') %>% 
+  drop_na(Longitude) %>% 
+  drop_na(Latitude) %>% 
+  st_as_sf(coords = c('Longitude', 'Latitude'), crs = 4326)
 
+sites <- distinct(test_data, geometry)
+
+cols <- c("International" = "gray55", "Domestic" = "#e6f4f1")
+
+
+png('../graphics/plots/collections_map.png')
 ggplot() +
-  geom_sf(data = countries) + 
-  geom_sf(data = states) + 
-  geom_sf(data = lakes, fill = 'blue') + 
-  geom_sf(data = test_data, shape = 1) +
-  geom_sf(data = polybb, fill = NA, linewidth = 2, color = '#2F0A28') + 
-  coord_sf(xlim = c(-85, -125), ylim = c(30, 50))
+  geom_sf(data = countries, aes(fill = Cntry_grp)) + 
+  scale_fill_manual(values = cols, guide = 'none') +
+  geom_sf(data = states, fill = NA, color = 'grey5') + 
+  geom_sf(data = lakes, fill = '#176B87') + 
+  
+  geom_sf(data = sites, shape = 21, color = 'gold', fill = '#2F0A28', size = 2) +
+  
+  
+  geom_sf(data = polybb, fill = NA, linewidth = 2, color = '#94524A') + 
[…]

### 2023-09-17 scripts/grab-elevs.R

@@ -1,0 +1,21 @@
+setwd('/media/steppe/hdd/Barneby_Lives-dev/scripts')
+
+library(sf)
+library(terra)
+
+p <- '../geodata/elevation'
+el <- mosaic(sprc(file.path( p, list.files(p))))
+
+pts <- data.frame(
+  Col  = c(25, 26),
+  x = c(-117.7172, -117.57132), 
+  y = c(39.26367, 38.08626)
+) |>
+  st_as_sf(coords = c('x', 'y'), crs = 4326) |>
+  vect()
+
+pts_el <- extract(el, pts)
+colnames(pts_el) <- c('site', 'm')
+pts_el$ft <- pts_el$m * 3.28084
+
+pts_el

### 2023-09-19 manuscript/manuscript_APPS.rmd
processing usda plants as lookup table for symbols

@@
   drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
-  unite(col = 'Scientific_name', Binomial:Infraspecific_authority, na.rm=TRUE, sep = " ", remove = F)
+  unite(col = 'Scientific_name', c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F)
 
@@
 time_spell_check <- system.time({ # ensure appropriate spellings of the species
-  data <- spell_check(data, path = p2tax)
-})
-
-time_spell_check <- system.time({ # ensure appropriate spellings of the species
-  sc_res <- spell_check(data$Binomial, path = p2tax)
+  sct <- spell_check(data, column = 'Full_name', path = p2tax)
 })
@@
 })
-
-data$Family == data$Family
 
@@
 
-It took roughly three minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) minutes to run all local steps of BarnebyLives, and a further 15 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
+It took roughly three minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) minutes to run all local steps of BarnebyLives, and a further 16 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
 Nearly all of the local run time is attributable to the spatial operations (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s, style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`, taxonomic: `r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`).
@@
 
-drop_na(data_p, Long_AutoFill_Flag) %>% 
+long_flag <- drop_na(data_p, Long_AutoFill_Flag) %>% 
   nrow() # 4 auto fill cases
 
-drop_na(data_p, Lat_AutoFill_Flag) %>% 
+lat_flag <- drop_na(data_p, Lat_AutoFill_Flag) %>% 
   nrow() # 2 autofill cases
@@
 The number of family misspellings were XX (% percent), the number of misspelled genera were XX (% percent), the number of misspelled binomials were XX (% percent).
-The number of author abbreviations which were not in the appropriate format were XX (% percent), generally the presence or absence of a period were the issue.
+The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.
 Plants of the World Online was able to identify XX new names for the submitted taxa, XX of which the author adopted.
-XX records were appropriately flagged for issues with auto fill incrementation of the longitude value, and three of these records were also auto-flagged for increases in latitude values (% of records).
+`r long_flag` records were appropriately flagged for issues with auto fill incrementation of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
 

### 2023-09-19 scripts/process_USDA_plants.R

@@ -1,0 +1,46 @@
+setwd('/media/steppe/hdd/Barneby_Lives-dev/scripts')
+
+usda <- read.csv('../taxonomic_data_raw/USDA_plants.csv', na.strings = "") %>% 
+  select(Symbol, Synonym.Symbol, Scientific.Name = Scientific.Name.with.Author)
+
+synonyms <- drop_na(usda, Synonym.Symbol) %>% 
+  select(-Symbol, Symbol = Synonym.Symbol)
+
+clean <- filter(usda, is.na(Synonym.Symbol)) %>% 
+  select(-Synonym.Symbol)
+
+gen_clean <- clean %>% 
+  filter(str_detect(Symbol, '[A-Z]{5}') & 
+           str_detect(Scientific.Name, ' var. | ssp. ', negate = TRUE)) %>% 
+  mutate(Scientific.Name = str_remove(Scientific.Name, " .*$"))
+
+gen_syn <- synonyms %>% 
+  filter(str_detect(Symbol, '[A-Z]{5}') & 
+           str_detect(Scientific.Name, ' var. | ssp. ', negate = TRUE)) %>% 
+  mutate(Scientific.Name = str_remove(Scientific.Name, " .*$"))
+
+genera_codes <- bind_rows(gen_clean, gen_syn)
+
+species <- bind_rows(clean, synonyms) %>% 
+  filter(! Symbol %in% genera_codes$Symbol) %>% 
+  mutate(
+    Genus = str_split_i(Scientific.Name, pattern = " ", 1),
+    Species = str_split_i(Scientific.Name, pattern = " ", 2),
+    Infrarank = str_extract(Scientific.Name, 'var[.]|ssp[.]'),
+    Infra = str_split_i(Scientific.Name, pattern = " var[.] | ssp[.] ", 2),
+    Infra = str_split_i(Infra, pattern = " ", 1),
+    Infrarank = if_else(is.na(Infra), NA, Infrarank) # orthographic variants abbreviated as 'var.', drop it
+    ) %>% 
+  unite(Scientific.Name, Genus:Infra, sep = " ", na.rm = T)
+
+out <- bind_rows(genera_codes, species) %>% 
+  distinct(Symbol, Scientific.Name) %>% 
+  arrange(Symbol)
+
+out %>% 
+  group_by(Scientific.Name) %>% 
+  filter(n() > 1)
+
[…]

### 2023-09-29 manuscript/labels/skeleton.Rmd
making labels

---
output: pdf_document
params:
  Collection_number: Collection_number
  UNIQUEID: UNIQUEID
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r load library}
library(BarnebyLives)
```

```{r load data}
data <- readRDS('../data/processed/data_w_Google_Maps') |>
  sf::st_drop_geometry() |>
  dplyr::filter(Collection_number == params$Collection_number) |>  
  sf::st_drop_geometry()
```

\begin{center}
\textbf{\LARGE Flora of Western North America}
\end{center}
\begin{flushright}
\uppercase{`r data$Family`}
\end{flushright}
\begingroup
\large
`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r data$Binomial_authority` `r writer(data$Infraspecific_rank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`  
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r data$latitude_dd` `r data$longitude_dd` (`r data$Datum`). `r data$Gen`. `r writer(data$Site)`

`r writer(data$Habitat)` `r data$physical_environ`  
Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
`r writer(data$Notes)`  
\endgroup
\begingroup
\begin{spacing}{0.1}
{\scriptsize `r data$Directions_BL`}
\end{spacing}
\endgroup

\vspace*{\fill}
\begingroup
`r collection_writer(data)`. `r writer_fide(data)`
\endgroup

\begin{center}
{\footnotesize Collected under the auspices of the Bureau of Land Management}
\end{center}

### 2023-09-29 manuscript/manuscript_APPS.rmd
making labels

@@
 
+```{r install package}
+devtools::install_github('sagesteppe/BarnebyLives')
+```
+
 ```{r load libraries}
 library(tidyverse) # data operations
-# devtools::install_github('sagesteppe/BarnebyLives')
 library(BarnebyLives) # for helping accession collections
@@
 ## Usage
-BarnebyLives is run entirely from within Rstudio. 
+All steps of BarnebyLives except for label generation are run from within Rstudio. 
 Data may be read in from any common spreadsheet management system or database connection such as Excel, LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
@@
 time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
-  geodata_writer(data, path = '~/Documents/BL', 
+  geodata_writer(data, path = 'data/processed', 
                filename = 'Herbarium_Collections_2023',
@@
 
-It took roughly three minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) minutes to run all local steps of BarnebyLives, and a further 16 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
-Nearly all of the local run time is attributable to the spatial operations (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s, style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`, taxonomic: `r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`).
-The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
+It took roughly *six* minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps of BarnebyLives, and roughly 20 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
+Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`.
+The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`) spent performing local taxonomic operations.
+The generation of labels consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
 
@@
 ```{r import results}
-data_p <- readRDS(file = 'data/processed/data_w_POWO_search')
+data_p <- readRDS(file = 'data/processed/data_w_POWO_search') %>% 
+  sf::st_drop_geometry()
 
@@
 
-Family_in_positions <- data %>% 
+Family_in_positions <- data_p %>% 
   drop_na(Family)
@@
 
-# NEED TO WORK ON INPUT DATA, INCONSISTENT USE OF AUTHORSHIP IN THE FIRST COLUMN
-data_p %>% 
-  sf::st_drop_geometry() %>% 
-  filter(Full_name != POW_Full_name) %>% 
-  select(Collection_number, Full_name, Binomial, POW_Full_name) 
 ```

### 2023-09-29 manuscript/labels/making_labels.R

@@ -1,0 +1,31 @@
+library(tidyverse)
+library(BarnebyLives)
+library(googlesheets4)
+
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels')
+dir.create('raw')
+dir.create('processed')
+dir.create('final')
+
+p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
+
+#p2libs <- .libPaths()[ # find the skeleton here and copy to local directory 
+#  grepl(paste0(version$major, '.', sub('\\..*', "", version$minor)), 
+#        .libPaths())]
+
+folds <- c('skeleton.Rmd')
+
+data <- readRDS('../data/processed/data_w_Google_Maps') |>
+  sf::st_drop_geometry() 
+time_label_gen <- system.time({
+  purrr::walk(
+    .x = collections$Collection_number,
+    ~ rmarkdown::render(
+      input = folds,
+      output_file = file.path(p, glue::glue("{.x}.pdf")),
+      params = list(Collection_number = {.x})
+    )
+  )
+})
+
+saveRDS(time_label_gen, file = '../data/processed/time_powo_searcher')

### 2023-11-03 manuscript/labels/making_labels.R
path forward for naming valleys

@@ -28,4 +28,4 @@ time_label_gen <- system.time({
   )
 })
 
-saveRDS(time_label_gen, file = '../data/processed/time_powo_searcher')
+saveRDS(time_label_gen, file = '../data/processed/time_label_gen')

### 2023-11-03 scripts/valleys_create.Rmd
forgot to ignore .zip

@@
 title: "Creating Named Valleys Dataset"
-author: "Payton Lott, Gus Rytting, Reed Benkendorf"
+author: "sagesteppe"
 date: "`r Sys.Date()`"
@@
 ```{r mountains}
-mtns <- st_read('../geodata/mountains/mountains.shp', quiet = T)
+mtns <- st_read('../geodata/mountains/mountains.shp', quiet = T) %>% 
+  st_transform(4269)
 mtn_bbox <- mtns %>% 
-  st_transform(4269) %>% 
   st_bbox()
@@
 
-hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2', quiet = T)  %>% 
-  select(HU2_NAME = name) %>% 
+#hu8 <- st_read(dsn = pbasin, layer = 'WBDHU8', quiet = T)  %>% 
+#  select(HU8_NAME = name) %>% 
+#  st_crop(., mtn_bbox)
+
+hu8 <- st_read(dsn = pbasin, layer = 'WBDHU10', quiet = T) %>% 
+  select(HU10_NAME = name) %>% 
+  st_simplify() %>% 
   st_crop(., mtn_bbox)
 
-hu8 <- st_read(dsn = pbasin, layer = 'WBDHU8', quiet = T)  %>% 
-  select(HU8_NAME = name) %>% 
-  st_crop(., mtn_bbox)
-
-hu8 <- st_join(hu2, hu8) 
+plot(hu8)
 ```
@@
 ```{r}
+p <- '/media/steppe/ExternalHD/SoS_GB/geodata/GNIS/NationalFile'
+setwd(p)
 
-ggplot() + 
-  geom_sf(data = hu8)
+gnis <- read.delim(file.path(p, 'NationalFile_20210825.txt'), sep = "|") %>% 
+  filter(FEATURE_CLASS %in% c('Valley', 'Plain')) %>% 
+  select(MAP_NAME, FEATURE_NAME, FEATURE_CLASS, PRIM_LAT_DEC, PRIM_LONG_DEC) %>% 
+  st_as_sf(coords = c('PRIM_LONG_DEC', 'PRIM_LAT_DEC'), crs = 4269) %>% 
+  st_crop(., mtn_bbox)
 
-gb <- filter(hu2, name == 'Great Basin Region')
+names_maybe <- st_join(hu8, gnis)
+plot(gnis)
 ```
+
+```{r subset valleys}
+
+st_erase = function(x, y) st_difference(x, st_union(y)) 
+output <- st_erase(hu8, mtns)
+
+output1 <- output %>% 
+  st_make_valid() %>% 
+  st_cast('MULTIPOLYGON') %>% 
+  st_cast('POLYGON') %>% 
+  mutate(Valley = "", .before = 2)
+
+## subset to west
+
+states <- tigris::states() %>% 
+  filter(NAME %in% c('California', 'Nevada', 'Utah', 
+                     'Arizona', 'New Mexico', 'Colorado', 'Idaho', 'Washington', 'Oregon')) %>% 
+  select(NAME)
+
+out <- st_intersection(output1, states)
+out <- select(out, -NAME)
+
+```
+
+```{r are gnis valleys useful}
+
+#they do not appear to be useful.
+
+gnis_sub <- gnis[ st_intersects(gnis, output) %>% lengths() > 0, ]
+
+ggplot()+
+  geom_sf(data = output) + 
+  geom_sf(data = gnis_sub)
+
+```
+
+```{r write out features, eval = F}
+
+st_write(gnis_sub, '../ValleysProject/data/places/places.shp')
+st_write(out, '../ValleysProject/data/valleys/valleys.shp', append = F)
+```

### 2023-11-03 manuscript/labels/making_labels.R

@@ -1,0 +1,31 @@
+library(tidyverse)
+library(BarnebyLives)
+library(googlesheets4)
+
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels')
+dir.create('raw')
+dir.create('processed')
+dir.create('final')
+
+p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
+
+#p2libs <- .libPaths()[ # find the skeleton here and copy to local directory 
+#  grepl(paste0(version$major, '.', sub('\\..*', "", version$minor)), 
+#        .libPaths())]
+
+folds <- c('skeleton.Rmd')
+
+data <- readRDS('../data/processed/data_w_Google_Maps') |>
+  sf::st_drop_geometry() 
+time_label_gen <- system.time({
+  purrr::walk(
+    .x = collections$Collection_number,
+    ~ rmarkdown::render(
+      input = folds,
+      output_file = file.path(p, glue::glue("{.x}.pdf")),
+      params = list(Collection_number = {.x})
+    )
+  )
+})
+
+saveRDS(time_label_gen, file = '../data/processed/time_label_gen')

### 2023-11-15 manuscript/manuscript_APPS.rmd
forgot to ignore .zip
working on

@@
 tt_local <- time_trials[time_trials$Group=='Local',]
+time_label_gen <- readRDS('../data/processed/time_label_gen')
+```
+
+```{sh, count number of files in directory}
+cd labels/processed
+ls -1 | wc -l
 ```
@@
 The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`) spent performing local taxonomic operations.
-The generation of labels consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
+The generation of labels consumed around `r time_label_gen` minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
 
@@
 
-data_p %>% 
-  filter(Binomial_authority != POW_Name_authority)
+#data_p %>% 
+#  filter(Binomial_authority != POW_Name_authority)
 

### 2023-11-15 scripts/valleys_create.Rmd
working on

@@
 
-states <- tigris::states() %>% 
-  filter(NAME %in% c('California', 'Nevada', 'Utah', 
-                     'Arizona', 'New Mexico', 'Colorado', 'Idaho', 'Washington', 'Oregon')) %>% 
+# states1 <- tigris::states() %>% 
+#  filter(NAME %in% c('California', 'Nevada', 'Utah', 
+#                     'Arizona', 'New Mexico', 'Colorado', 'Idaho', 'Washington', 'Oregon')) %>% 
+#  select(NAME)
+
+#out <- st_intersection(output1, states)
+# out <- select(out, -NAME)
+
+states1 <- tigris::states() %>% 
+  filter(NAME %in% c('Montana', 'Wyoming', 'Texas', 'Oklahoma', 'South Dakota', 'North Dakota', 'Kansas')) %>% 
   select(NAME)
 
-out <- st_intersection(output1, states)
+out <- st_intersection(output1, states1)
 out <- select(out, -NAME)
@@
 st_write(gnis_sub, '../ValleysProject/data/places/places.shp')
-st_write(out, '../ValleysProject/data/valleys/valleys.shp', append = F)
+
+# st_write(out, '../ValleysProject/data/valleys/valleys.shp', append = F)
+st_write(out, '../ValleysProject/data/valleys/valleys-heartland.shp', append = F)
 ```

### 2023-11-15 database_data/darwincore_names.R

@@ -1,0 +1,14 @@
+setwd('/media/steppe/hdd/Barneby_Lives-dev/database_data')
+
+library(tidyverse)
+
+all_names <- read.csv('all_darwincore_colnames.csv')
+pop_names <- read.csv('darwinCore_crosswalk-draft.csv')
+
+final_names <- filter(all_names, !COLNAMES %in% pop_names$DarwinCore ) %>% 
+  mutate(BL = NA) %>% 
+  rename(DarwinCore = COLNAMES) %>% 
+  bind_rows(., pop_names) %>% 
+  slice(match(all_names$COLNAMES, DarwinCore))
+
+write.csv(final_names, 'DarwinCore_names.csv')

### 2023-11-15 manuscript/labels/making_labels.R

@@ -15,7 +15,7 @@ p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
 
 folds <- c('skeleton.Rmd')
 
-data <- readRDS('../data/processed/data_w_Google_Maps') |>
+collections <- readRDS('../data/processed/data_w_Google_Maps') |>
   sf::st_drop_geometry() 
 time_label_gen <- system.time({
   purrr::walk(

### 2023-11-19 manuscript/manuscript_APPS.rmd
added time trials for pdf render and adding spelling values

--- 2023-09-29 manuscript/manuscript_APPS.rmd
@@
 Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition.
-Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.
+Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology.
+In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data.
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
@@
 keywords: |
-  Herbarium, collections, natural history museum, geospatial, automation, R, software
+  Herbarium, natural history museum, collections, geospatial, automation, R, software
 output:
@@
 
-Here we provide a description of the BarnebyLives R package. 
-BarnebyLives was named for plant taxonomist extraordinaire Rupert Charles Barneby (1911-200), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the Willian & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
-Select accolades of Rupert include the Asa Gray Award from the American Society of Plant Taxonomists (ASPT) in 1989, the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congresses's (IBC) Millenium Botany Award (1999) (@welsh2001rupert).
-
 \begin{wrapfigure}{l}{0.35\textwidth}
@@
 \end{wrapfigure}
+
+Here we provide a description of the BarnebyLives R package. 
+BarnebyLives was named for plant taxonomist extraordinaire Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
+Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
 
@@
 ```{r install package}
-devtools::install_github('sagesteppe/BarnebyLives')
+## devtools::install_github('sagesteppe/BarnebyLives')
 ```
@@
 This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America FIGURE XX.
-In total `r n_spp` species (with `r n_sp_authors` authorships), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
+In total `r n_spp` species (with `r n_sp_authors` sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authors) in `r n_families` families were used for testing.
 
@@
 tt_local <- time_trials[time_trials$Group=='Local',]
+
+time_label_gen <- readRDS('data/processed/time_label_gen')[['elapsed']]
+time_4perpage <- gsub('user.*$', '', readLines('labels/4labsperpage.txt'))[1]
+time_final <- gsub('user.*$', '', readLines('labels/process2final.txt'))[1]
+
+minutemen <- function(x){
+  lkp <- 1:60
+  names(lkp) <-  c('one', 'two', 'three', 'four', 'five', 'six', 
+         'seven', 'eight', 'nine', 'ten', 'eleven', 'twelve', 13:60)
+
+  names(lkp) [ round(x/60, digits = 0) ]
+}
 ```
 
-It took roughly *six* minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps of BarnebyLives, and roughly 20 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
-Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`.
-The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`) spent performing local taxonomic operations.
-The generation of labels consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).
+It took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps of BarnebyLives, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites.
+Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s.
+The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
+The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
 
@@
 
-drop_na(data_p, Binomial_authority) %>% 
+no_binomial_auth_issues <- drop_na(data_p, Binomial_authority) %>% 
   nrow() # nearly all relating to the absence/presence of a period
 
-drop_na(data_p, Infra_auth_issues) %>% 
+no_infra_auth_issues <- drop_na(data_p, Infra_auth_issues) %>% 
   nrow() # nearly all relating to the absence/presence of a period
@@
 # following families. 
-data_p %>% 
+family_incongruence <- data_p %>% 
   sf::st_drop_geometry() %>% 
@@
     case_when(
-      Family == 'Hydrophyllaceae' ~ 'Preferred', 
-      Family %in% c('Plantaginaceae', 'Athyriaceae') ~ 'Outdated',
-      Family %in% c('Saxifragaceae', 'Solanaceae') ~ 'Submitter_Incorrect', 
-      Family == 'Bataceae' ~ 'Internal Error', 
+      Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
+      Family %in% c('Athyriaceae') ~ 'Outdated',
+      Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
+      Family %in% c('Bataceae', 'Ericaceae') ~ 'Internal_Error', 
       .default = 'Typo'
@@
 
-data_p %>% 
+# how many of my genera were mis-spelled??
+genera_mispelled <- data_p %>% 
   sf::st_drop_geometry() %>% 
@@
 
-# how many of my genera were mis-spelled??
-data_p %>% 
+# how many of my genera were outdated?
+genus_incongruence <- data_p %>% 
   sf::st_drop_geometry() %>% 
   filter(POW_Genus != 'NOT FOUND') %>% 
-  filter(POW_Genus != Genus)
+  filter(POW_Genus != Genus)  %>% 
+  select(Collection_number, Full_name, Genus, POW_Genus)
 
 # the mismatch between POW genus and not
-data_p %>% 
+genus_mismatch <- data_p %>% 
   sf::st_drop_geometry() %>% 
@@
                     ) ~ 'Outdated',
-      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal Error',
+      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal_Error',
       .default = 'Typo'
     ),
-    Situation = if_else(Full_name == 'Oenothera caespitosa Nutt.', 'Internal Error', Situation))
+    Situation = if_else(Full_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))
 
-data_p %>% 
+binom_auth_difference <- data_p %>% 
   filter(Binomial_authority != POW_Name_authority)
 
-data_p %>% 
+epithet_diff <- data_p %>% 
   sf::st_drop_geometry() %>% 
-  filter(Epithet != POW_Epithet ) %>% 
+  filter(Epithet != POW_Epithet, !Full_name %in% unique(genus_mismatch$Full_name )) %>% 
   select(Collection_number, Full_name, Epithet, POW_Epithet)
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-The number of family misspellings were XX (% percent), the number of misspelled genera were XX (% percent), the number of misspelled binomials were XX (% percent).
+While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance of an outdated circumscription applied.
+BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` record in error.
+The number of flagged genera were XX (% percent), XX of these were misspelled, XX were outdated, XX
+
+... XX were preferred by author.
+
+the number of misspelled binomials were XX (% percent).
+
 The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.
 Plants of the World Online was able to identify XX new names for the submitted taxa, XX of which the author adopted.
-`r long_flag` records were appropriately flagged for issues with auto fill incrementation of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
+`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
 

### 2023-11-20 manuscript/manuscript_APPS.rmd
addressing epithets

@@
 
-# how many of my genera were outdated?
-genus_incongruence <- data_p %>% 
-  sf::st_drop_geometry() %>% 
-  filter(POW_Genus != 'NOT FOUND') %>% 
-  filter(POW_Genus != Genus)  %>% 
-  select(Collection_number, Full_name, Genus, POW_Genus)
-
 # the mismatch between POW genus and not
@@
     case_when(
-      Genus %in% c('Pleuraphis', 'Glandularia') ~ 'Preferred', # check Pleuraphis, i am confused on this still....
-      Genus %in% c('Oenothera', 'Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia',
+      Genus %in% c('Glandularia', 'Oenothera', 'Chamerion') ~ 'Preferred',
+      # check Pleuraphis, i am confused on this still....
+      Genus %in% c('Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia', 'Pleuraphis',
                     'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
@@
   sf::st_drop_geometry() %>% 
-  filter(Epithet != POW_Epithet, !Full_name %in% unique(genus_mismatch$Full_name )) %>% 
-  select(Collection_number, Full_name, Epithet, POW_Epithet)
+  filter(Epithet != POW_Epithet, !Full_name %in% 
+           unique(c( genera_mispelled$Full_name,  genus_mismatch$Full_name))) %>% 
+  select(Collection_number, Full_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
+  distinct(Full_name, .keep_all = T) %>%  # do we record repeat collections ?
+  mutate(Situation = 
+           case_when(
+             Collection_number %in% c('1217') ~ 'Spelling',
+             Collection_number %in% c('1588', '2445', '2777', '1337', '2412', '2748', 
+                                      '2652', '2805', '2194', '2040', '2039', '2403', '2328') ~ 'Outdated',
+             Collection_number %in% c('2326', '2038') ~ 'Internal_Error' ,
+             Collection_number %in% c('2110') ~ 'Preferred'
 
+           )
+  )
+
+srm(Family_in, Family_out)
 ```
@@
 
-... XX were preferred by author.
+BL was able to identify 41 misspellings ...
+BL made `r nrow(family_incongruence)` suggestions on naming, 
+where an incorrect family was entered, and `r sum(genus_mismatch$Situation=='Outdated', na.rm = T)` instance of an outdated circumscription applied.
+BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
 
-the number of misspelled binomials were XX (% percent).
+Excluding the previous records at the level of genus, BL flagged 75 binomials.
+Not all genera which where flagged in the previous step were identified in the last check, as BL may search for an epithet with a genus which was more similar to the non-target genus, and for which a binomial does not exist.
 
@@
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
+
+```{r}
+rm(genus_mismatch, genera_mispelled, family_incongruence)
+```
 

### 2023-11-20 manuscript/manuscript_APPS.rmd
going to transition local taxonomic look up source to kews! redo parsing for better spell check

@@
 Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
-However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, electronic sensing (@tosa2021rapid).
+Even within systematics, the ability to collect and analyze micro-morphological data has fostered revisionary studies in many groups providing synapomorphies which can be correlated to molecular results to improve the interpretation of monophyletic clades.
+However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, and electronic sensing (@tosa2021rapid).
 
@@
 
+filter(sct, Match == 'fuzzy') %>% 
+  select(Scientific_name, 'SpellCk.scientificName')
+
 time_family_spell_check <- system.time({ # ensure appropriate spelling of the family
@@
 
- pow_res <- lapply(names,
+ pow_res1 <- lapply(names[1],
        powo_searcher) %>% 
@@
 }) # has been run
+
+names
 
@@
              Collection_number %in% c('1588', '2445', '2777', '1337', '2412', '2748', 
-                                      '2652', '2805', '2194', '2040', '2039', '2403', '2328') ~ 'Outdated',
-             Collection_number %in% c('2326', '2038') ~ 'Internal_Error' ,
+                                      '2652', '2805', '2194', '2040', '2039', '2403',
+                                      '2328', '2076') ~ 'Outdated',
+             Collection_number %in% c('2326', '2038', '1424', '2618', '2665', '2297') ~ 'Internal_Error',
              Collection_number %in% c('2110') ~ 'Preferred'
@@
 
-srm(Family_in, Family_out)
+rm(Family_in, Family_out)
 ```

### 2023-11-20 scripts/sheet_writer.R

@@ -1,0 +1,112 @@
+#' write out digital data for mass upload in an database
+#'
+#' This function will write out the data from BarnebyLives into many formats commonly used
+#' by herbaria to database their records. Currently supports: Symbiota, CPNWH, Darwin Core ... Adding additional
+#' support is easy, and the developers are happy to ensure that your museums format is supported, reach out
+#' to coordinate this effort with us.
+#' @param x a data frame generated from the full steps of BarnebyLives pipeline
+#' @param type a string of a supported output format, one of: 'darwincore', 'symbiota', 'cpnwh'
+#' @param BL boolean, whether to also write out a copy of data in BarnebyLives format, which can be used to generate labels
+#' defaults to the name 'BarnebyLives' with the current data appended.
+#' @param basename optional string specifying a filename prefix, e.g. a collector if need be.
+#' @param filename a string specifying the filename, and type, best supported are CSV.
+#' Defaults to the type of data being written out, with the current date.
+#' @param directory a string specifying the path to the directory to write the data to.
+#' Defualts to the current working directory
+#' @example path.R
+#' @export
+#'
+sheet_writer <- function(x, type, filename, directory){
+
+  if(missing(type)){type <- 'symbiota'}
+  if(missing(BL)){BL <- TRUE}
+  if(missing(basename)){basename <- ""}
+  if(missing(directory)){directory <- '.'}
+
+  if(missing(filename)){ if(grepl("", basename){ # if filename not provided create defaults
+    filename <- paste0(type, '_', Sys.Date(), '.csv')
+  }) else {
+    filename <- paste0(basename, '_', type, '_', Sys.Date(), '.csv')}
+  } else{ # if filename provided create the default or user specified input
+    if(grepl("", basename){
+      filename <- paste0(filename, '_', Sys.Date(), '.csv')
+    }) else {
+      filename <- paste0(basename, '_', filename)
+      }
+  }
+
+  x_sub <- dplyr::select(x, -tidyselect::starts_with('POW'), -tidyselect::ends_with('issues')) |>
+    sf::st_drop_geometry()
+
[…]

### 2024-07-09 manuscript/manuscript_APPS.rmd
rerun pipeline with fresh spell check and use of output

@@
 --- 
-title: BarnebyLives':' an R package to create herbarium specimen labels and digital data sheets
+title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
 author:  |
-    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
+    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
@@
 Nearly 400 million specimens are housed in herbaria around the world (@thiers2021herbaria).
-These specimens were collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity (@greve2016realising). 
-The rate of accessioning new collections to herbaria diminished in the 20th century as research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity (@prather2004decline, @pyke2010biological, @daru2018widespread).
+These specimens, collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity, have found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
+However, The rate of accessioning new collections to herbaria diminished in the 20th century as priorities in biology shifted away from describing, documenting, and understanding earths diversity towards understanding cellular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
 Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion).
@@
 
-Recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have brought about a renaissance in herbarium based studies (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
+In fact recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
+These current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
+Further we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, and electronic sensing (@tosa2021rapid).
 Even within systematics, the ability to collect and analyze micro-morphological data has fostered revisionary studies in many groups providing synapomorphies which can be correlated to molecular results to improve the interpretation of monophyletic clades.
-However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, and electronic sensing (@tosa2021rapid).
 
-However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position where the skills of collecting and processing specimens, and time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
-The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience in the process. 
-While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection.
-This scenario results in not only the delay in the deposition of many specimens, but undoubtedly the deposition of many collections at all.
+However, while there is recognized need for more specimens the skills of collecting and processing specimens, and time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
+The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience.  
+While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection and preparing specimens for herbaria.
+This scenario results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for some specimens to ever be accessioned at all. 
 Problems which young collectors face generally include both the lack of dedicated time awarded to them at a seasons end to process specimens, and a general lack of formal education on cartography, natural history, taxonomy, and plant systematics.
@@
 watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g.
-ArcMap, which is relatively expensive at 100$ year), or more likely Google Maps by individual site.
+ArcMap, which is relatively expensive at 100$ year), or more likely Google Maps by individual site a time consuming process.
 This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.
+
+Further issues relate to the typical problems of nomenclature, to wit the pace at which taxonomic innovations are now made and which many persons of all skill levels find difficult to keep up with...
 
@@
 Here we provide a description of the BarnebyLives R package. 
-BarnebyLives was named for plant taxonomist extraordinaire Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
+BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
 
-More evidently difficult tasks involve taxonomy and the rapid rate at which taxonomic names have changed since the publication of many Floras.
-
 # METHODS AND RESULTS
-
-`r lorem::ipsum(paragraphs = 5)`
 
@@
 Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
-We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, or have the data stored entirely on thumb-drives.
+We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, and have the data stored entirely on thumb-drives.
 The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
@@
 While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
-Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in FIGURE XX.
+Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
 
@@
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used.
-This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America FIGURE XX.
+This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2).
 In total `r n_spp` species (with `r n_sp_authors` sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authors) in `r n_families` families were used for testing.
@@
 time_spell_check <- system.time({ # ensure appropriate spellings of the species
-  sct <- spell_check(data, column = 'Full_name', path = p2tax)
+  data <- spell_check(data, column = 'Full_name', path = p2tax)
 })
-
-filter(sct, Match == 'fuzzy') %>% 
-  select(Scientific_name, 'SpellCk.scientificName')
 
@@
 # we keep these processes in a discrete chunk set not to evaluate so as to not overwhelm
-# the services. Google does charge if the number of queries per month is high.
+# the services. Google does charge if the number of queries per month isnt high.
 
@@
  names <- sf::st_drop_geometry(data) %>% 
-   pull(Full_name)
+   pull(SpellCk.taxon_name)
 
- pow_res1 <- lapply(names[1],
+ pow_res <- lapply(names,
        powo_searcher) %>% 
        bind_rows()
- data <- bind_cols(data, pow_res)
+ data <- bind_cols(data, pow_res) |>
+  dplyr::relocate(geometry, .after = dplyr::last_col()) |>
+  sf::st_as_sf(crs = 4326)
 
@@
 }) # has been run
-
-names
 
@@
   SoS_gkey = Sys.getenv("Sos_gkey")
-  data <- directions_grabber(data, api_key = SoS_gkey)
+  data1 <- directions_grabber(data, api_key = SoS_gkey)
 })
@@
 
-It took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps of BarnebyLives, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites.
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites.
 Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s.
@@
   sf::st_drop_geometry() %>% 
-  filter(Epithet != POW_Epithet, !Full_name %in% 
-           unique(c( genera_mispelled$Full_name,  genus_mismatch$Full_name))) %>% 
+  filter(Epithet != POW_Epithet, !Full_name %in% genus_mismatch$Full_name) %>% 
   select(Collection_number, Full_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
@@
            case_when(
-             Collection_number %in% c('1217') ~ 'Spelling',
+             Collection_number %in% c(
+               '1217', '1251', '1283', '1311', '1337', '1391', '1450', '1576', 
+               '1594', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130',
+               '2153', '2199', '2227', '2242', '2256', '2289', '2370', '2379', 
+               '2393', '2415', '2511', '2573', '2648', '2660', '2687', '2690',
+               '2710', '2715', '2750', '2767', '2818')
+             ~ 'Spelling',
+             Collection_number %in% c('1207') ~ 'Binomial_Reversed', 
              Collection_number %in% c('1588', '2445', '2777', '1337', '2412', '2748', 
@@
                                       '2328', '2076') ~ 'Outdated',
-             Collection_number %in% c('2326', '2038', '1424', '2618', '2665', '2297') ~ 'Internal_Error',
-             Collection_number %in% c('2110') ~ 'Preferred'
+             Collection_number %in% c('2644') ~ 'Incorrect',
+             Collection_number %in% c('2326', '2038', '1424', '2618',
+                                      '2665', '2297', '2343') ~ 'Internal_Error',
+             Collection_number %in% c('2110') ~ 'Preferred',
+             Collection_number %in% c('2480', '2561', '2727') ~ 'Not Found'
 
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance of an outdated circumscription applied.
+While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied.
 BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` record in error.
-The number of flagged genera were XX (% percent), XX of these were misspelled, XX were outdated, XX
 
-BL was able to identify 41 misspellings ...
-BL made `r nrow(family_incongruence)` suggestions on naming, 
-where an incorrect family was entered, and `r sum(genus_mismatch$Situation=='Outdated', na.rm = T)` instance of an outdated circumscription applied.
-BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
+BL identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name instance of an outdated circumscription applied (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` genera total).
+BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
 
-Excluding the previous records at the level of genus, BL flagged 75 binomials.
-Not all genera which where flagged in the previous step were identified in the last check, as BL may search for an epithet with a genus which was more similar to the non-target genus, and for which a binomial does not exist.
+BL flagged 75 species
 
@@
 # ORCID
+Reed Benkendorf https://orcid.org/0000-0003-3110-6687  
 Jeremie Fant https://orcid.org/0000-0001-9276-1111

### 2024-07-09 manuscript/manuscript_APPS.rmd
focusing intro

@@
 
-Nearly 400 million specimens are housed in herbaria around the world (@thiers2021herbaria).
-These specimens, collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity, have found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-However, The rate of accessioning new collections to herbaria diminished in the 20th century as priorities in biology shifted away from describing, documenting, and understanding earths diversity towards understanding cellular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
-Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion).
-Fortunately, renewed interest in collections have brought herbaria of all sizes back to the forefront of plant sciences (@ronsted2020integrative, @marsico2020small).
+Nearly 400 million specimens are housed in herbaria around the globe (@thiers2021herbaria).
+These specimens, collected to describe the taxonomic diversity of plants and document the worlds floristic diversity, have recently found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
+However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity towards understanding cellular and molecular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
+This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices (@funk2014erosion).
+Fortunately, renewed interest approaches in collections generated by 'big data approaches' have brought herbarium collections back to the forefront of the natural sciences (@ronsted2020integrative, @marsico2020small).
 
-In fact recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-These current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
-Further we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, community science, and electronic sensing (@tosa2021rapid).
-Even within systematics, the ability to collect and analyze micro-morphological data has fostered revisionary studies in many groups providing synapomorphies which can be correlated to molecular results to improve the interpretation of monophyletic clades.
+In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
+The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
+Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid).
+While image or purely observational (rather than collections based) citizen science initializes (e.g.
+iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images.
+Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes, and seeds or pollen; forever ensuing specimens as the ultimate data source to center most efforts around.
 
-However, while there is recognized need for more specimens the skills of collecting and processing specimens, and time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
+However, while there is recognized need for more specimens the skills of collecting and processing specimens and the time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
 The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience.  
@@
 For example, while the acquisition of political information for a collection site appears simple, it is only so if the collector has the adequate resources at their disposal.
-Given the association of boundaries with topographically complex areas (e.g.
-watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g.
-ArcMap, which is relatively expensive at 100$ year), or more likely Google Maps by individual site a time consuming process.
-This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.
-
-Further issues relate to the typical problems of nomenclature, to wit the pace at which taxonomic innovations are now made and which many persons of all skill levels find difficult to keep up with...
 
@@
 \end{wrapfigure}
+
+Given the association of boundaries with topographically complex areas (e.g.
+watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g.
+ArcMap, which is relatively expensive at $100 year), or more likely Google Maps by individual site a time consuming process.
+This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.
+
+Further issues relate to the typical problems of nomenclature, to wit the pace at which taxonomic innovations are now made and which many persons of all skill levels find difficult to keep up with...
 
@@
   SoS_gkey = Sys.getenv("Sos_gkey")
-  data1 <- directions_grabber(data, api_key = SoS_gkey)
+  data <- directions_grabber(data, api_key = SoS_gkey)
 })
@@
 While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied.
-BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` record in error.
+BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.
 
@@
   \centering
-    \includegraphics[width=0.45\textwidth]{../graphics/tables/Screenshot-DataSources.png}
+    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
   \caption{Sources of Data required for operations}
@@
 
-\newpage
 # CONCLUSIONS

### 2024-07-09 manuscript/manuscript_APPS.rmd
more writing

@@
 Optionally the package writes driving directions to sites using Google Maps.
-Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,
+The package outputs data in a tabular format for review by the user to accept or confirm changes, before dynamically rendering labels.
   **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
 keywords: |
-  Herbarium, natural history museum, collections, geospatial, automation, R, software
+  herbariaum, natural history museum, collections, geospatial, automation, R, software 
 output:
@@
 iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images.
-Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes, and seeds or pollen; forever ensuing specimens as the ultimate data source to center most efforts around.
+Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes (@borges2020schrodinger), and seeds or pollen; eternally ensuing specimens as the ultimate data source to center most efforts around.
 
-However, while there is recognized need for more specimens the skills of collecting and processing specimens and the time allocated for collecting, have declined among young persons (@daru2018widespread, @mishler2020spatial).
-The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience.  
-While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection and preparing specimens for herbaria.
-This scenario results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for some specimens to ever be accessioned at all. 
-Problems which young collectors face generally include both the lack of dedicated time awarded to them at a seasons end to process specimens, and a general lack of formal education on cartography, natural history, taxonomy, and plant systematics.
-
-The successful generation of an herbarium specimen includes many steps which are easy to take for granted. 
-For example, while the acquisition of political information for a collection site appears simple, it is only so if the collector has the adequate resources at their disposal.
+However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow. 
+We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial).
+While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
+Apparent problems include the lack of dedicated time at a field seasons end to process specimens, and a general lack of education on cartography and orienteering, natural history (e.g.
+geology, geomorphology), nomenclature and Latin, various computer programs (e.g.
+Microsoft Office suite), and increasingly - plant systematics.
+In the absence of suitable mentors this assuredly results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for many specimens to be accessioned at all, and increasingly ever collected.
 
@@
 
-Given the association of boundaries with topographically complex areas (e.g.
-watersheds) it often requires topographic maps, which are no longer widespread - resulting in many having difficulties interpreting them, or transcription of coordinates into a Geographic Information System (e.g.
-ArcMap, which is relatively expensive at $100 year), or more likely Google Maps by individual site a time consuming process.
-This lack of topographic maps compounds the issues of young collectors being unable to come up with appropriate site names.
+The generation of an herbarium specimen includes many steps which are easy to take for granted. 
+For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate resources (printed topographic maps, or GIS software) at their disposal.
+In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their position. 
+Finding appropriate sites names is another problem which can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS.
+The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names.
+Upon finding a name they may find it frustrating to hear that while published, the proposal has been accepted by few practitioners and they have unwittingly offended certain curators. 
+Formatting a label correctly (e.g. abbreviations), if successful upon even setting up a mail merge, is a time consuming process and likely to introduce several errors in formatting. 
+Even if a collector navigates all of these hurdles successfully, the time allocated to each step is quite large. 
+Further each step of interfacing with different resources increases the opportunity for transcription errors.
 
-Further issues relate to the typical problems of nomenclature, to wit the pace at which taxonomic innovations are now made and which many persons of all skill levels find difficult to keep up with...
+Here we provide a description of the BarnebyLives R package. 
+BarnebyLives aims to increase both the data quality of labels, and to speed up the process of producing them. 
+It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
+Census Bureau (@walker2022tigris), the Public Land Survey System, and ownership details of public lands via the Protected-Areas Database (PAD-US) from (@usgs2024padus).
+Site names are suggested via finding the closest unambiguously named place feature via the Geographic Name Information System (GNIS), and by precise calculation of the distance and azimuth from these localities to the collection site (@gnis2024).
+Using the GMBA Mountain Inventory V.
+2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region without any ambiguity (@snethlage2022hierarchical).
+Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms (@govaerts2021world, @powo2024).
+Author abbreviations are verified using IPNI's Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviation of authorities (@ipni2024, @powo2024).
+Checks are performed to search for common issues associated with spreadsheets, or transcription, such as the auto-filling of coordinate and date columns. 
+After final review of the data generated by the package, it allows for the option to export spreadsheets which are usable for mass upload of data to multiple common herbarium databases, as well as the generation of herbarium labels.
 
@@
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
+Most importantly, Rupert was remembered as an individual entirely generous with there time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).
 
@@
 Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
-We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, and have the data stored entirely on thumb-drives.
+We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, while having the data stored entirely on thumb-drives.
 The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
@@
 # AUTHOR CONTRIBUTIONS
-The project was conceptualized by R.C.B.
-The program was written by R.C.B.
+The project was conceptualized by R.C.B. The program was written by R.C.B.
 Data collection and analysis were performed by R.C.B.
-R.C.B.
-wrote the manuscript with input from all other authors.
+R.C.B. wrote the manuscript with input from all other authors. 
 All authors approved the final version of the manuscript.
@@
 Two anonymous peer reviewers who increased the quality of this manuscript are thanked. 
+Sofia Garcia is acknowledged for creating the 'Valleys' data set which place naming in the package relies on. 
 Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel.

### 2024-07-10 manuscript/manuscript_APPS.rmd
more citations

@@
 In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative).
+The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative, @davis2023herbarium).
 Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid).
@@
 
-However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow. 
+However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow (@prather2004decline). 
 We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial).
 While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
-Apparent problems include the lack of dedicated time at a field seasons end to process specimens, and a general lack of education on cartography and orienteering, natural history (e.g.
+Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
 geology, geomorphology), nomenclature and Latin, various computer programs (e.g.
-Microsoft Office suite), and increasingly - plant systematics.
+Microsoft Office suite), and increasingly - plant systematics (@nanglu2023nature, @woodland2007botanists, @barrows2016crossroads).
 In the absence of suitable mentors this assuredly results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for many specimens to be accessioned at all, and increasingly ever collected.
@@
 
-The generation of an herbarium specimen includes many steps which are easy to take for granted. 
+The generation of an herbarium specimen includes many steps which are easy to take for granted (@forman1989herbarium). 
 For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate resources (printed topographic maps, or GIS software) at their disposal.
@@
 Here we provide a description of the BarnebyLives R package. 
-BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert).
+BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert).
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
-Most importantly, Rupert was remembered as an individual entirely generous with there time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).
+Most importantly, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).
 
@@
 All steps of BarnebyLives except for label generation are run from within Rstudio. 
-Data may be read in from any common spreadsheet management system or database connection such as Excel, LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
+Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
 The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all of the on-disk memory associated with these data are for storing geo-spatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation.
+Virtually all of the on-disk memory associated with these data are for storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
 Functions which require the on-disk data require a path to the data as an argument.  
-Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
-We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, while having the data stored entirely on thumb-drives.
+Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.  
+We anticipate most personal BarnebyLives instances will be less than several gigabytes (ours covering all of the conterminous Western U.S.
+is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
 The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
 pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
-While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
+While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
@@
 The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
+The total computer run time for processing these XXX specimens was XX minutes.
 
@@
 A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. 
-All data used in this mansucript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript
+All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript
 

### 2024-07-11 manuscript/manuscript_APPS.rmd
adding sresults

@@
   nrow() # 74 families
+
+n_genera <- data %>% 
+  separate(Scientific_name, into = c('Genus', 'Epithet') , remove = FALSE, extra = 'drop') %>% 
+  group_by(Genus) %>% 
+  count() %>% 
+  nrow() # 292 Genera
 
@@
 The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
-The total computer run time for processing these XXX specimens was XX minutes.
+The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.
 
@@
                '1594', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130',
-               '2153', '2199', '2227', '2242', '2256', '2289', '2370', '2379', 
-               '2393', '2415', '2511', '2573', '2648', '2660', '2687', '2690',
-               '2710', '2715', '2750', '2767', '2818')
+               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2343', '2370', '2379', 
+               '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
+               '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
              ~ 'Spelling',
-             Collection_number %in% c('1207') ~ 'Binomial_Reversed', 
-             Collection_number %in% c('1588', '2445', '2777', '1337', '2412', '2748', 
-                                      '2652', '2805', '2194', '2040', '2039', '2403',
-                                      '2328', '2076') ~ 'Outdated',
+             Collection_number %in% c(
+               '1588', '1619', '2445', '2777', '1337', '2412', '2487', '2748', '2652',
+               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
+             ~ 'Outdated', 
              Collection_number %in% c('2644') ~ 'Incorrect',
-             Collection_number %in% c('2326', '2038', '1424', '2618',
-                                      '2665', '2297', '2343') ~ 'Internal_Error',
-             Collection_number %in% c('2110') ~ 'Preferred',
-             Collection_number %in% c('2480', '2561', '2727') ~ 'Not Found'
-
+             Collection_number %in% c(
+               '1207', '2326', '2038', '1424', '2618', '2622', '2665', '2711', '2713') 
+             ~ 'Internal_Error',
+             Collection_number %in% c('2110', '2540') ~ 'Preferred'
            )
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While no families were misspelled, BL made `r nrow(family_incongruence)` suggestions on naming, `r sum(family_incongruence$Situation=='Typo', na.rm = T)` manually entered typos were found, it identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied.
-BL flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied.
+At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.
 
-BL identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name instance of an outdated circumscription applied (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` genera total).
-BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
+In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera). 
+BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
-BL flagged 75 species
+Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records. 
+BL `r sum(epithet_diff$Situation=='Spelling')` detected instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
+In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Full_name']))` unique species). 
+BL flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Full_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
+The final record was an egregious error where the order of the specific epithet and the genus name.
 

### 2024-08-09 manuscript/manuscript_APPS.rmd
added sppell checks to associated spp and geodata writer to time trials

@@
 However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow (@prather2004decline). 
-We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial).
+We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial, @manzano2021fair).
 While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
@@
 Finding appropriate sites names is another problem which can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS.
-The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names.
+The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names @hitchcock2018flora).
 Upon finding a name they may find it frustrating to hear that while published, the proposal has been accepted by few practitioners and they have unwittingly offended certain curators. 
@@
 
-Here we provide a description of the BarnebyLives R package. 
-BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert).
+BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert).
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
@@
 is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
-The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
+The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
 pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
@@
   nrow() # 22 distinct infra species author groups
+
+n_vegetation_fields <- data %>% 
+  drop_na(Vegetation) %>% 
+  distinct(Vegetation) |>
+  nrow()
+
+n_associates_fields <- data %>% 
+  drop_na(Associates) %>% 
+  distinct(Associates) |>
+  nrow()
 
@@
 time_spell_check <- system.time({ # ensure appropriate spellings of the species
-  data <- spell_check(data, column = 'Full_name', path = p2tax)
+  data <- spell_check(data, column = 'Scientific_name', path = p2tax)
 })
@@
   data <- author_check(data, path = p2tax)
+})
+
+associates_spell_check_veg <- system.time({ # ensure associated species are spelled correctly. 
+  data <- associates_spell_check(data, 'Vegetation', p2tax)
+})
+
+associates_spell_check_ass <- system.time({ # ensure associated species are spelled correctly. 
+  data <- associates_spell_check(data, 'Associates', p2tax)
 })
@@
     Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
-    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 
+    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 'time_geodata_writer',
                     'political_grabber', 'site_writer', 'geodata_writer', 'directions_grabber') ~ 'Geospatial', 
-    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher') ~ 'Taxonomic',
+    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher', 
+                    'associates_spell_check_veg', 'associates_spell_check_ass') ~ 'Taxonomic',
     Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
@@
   pull(Family)
-Family_in[ which( Family_in != Family_out ) ]
 
@@
 The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.
-Plants of the World Online was able to identify XX new names for the submitted taxa, XX of which the author adopted.
+
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).

### 2024-08-09 database_data/darwincore_names.R

@@ -2,8 +2,7 @@ setwd('/media/steppe/hdd/Barneby_Lives-dev/database_data')
 
 library(tidyverse)
 
-all_names <- read.csv('all_darwincore_colnames.csv')
-pop_names <- read.csv('darwinCore_crosswalk-draft.csv')
+all_names <- read.csv('Fields.csv')
 
 final_names <- filter(all_names, !COLNAMES %in% pop_names$DarwinCore ) %>% 
   mutate(BL = NA) %>% 

### 2024-08-11 manuscript/flowchart.R
workkflows

@@ -1,5 +1,6 @@
 library(tidyverse)
 library(DiagrammeR)
+library(ggpubr)
 setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
 
 grViz("
@@ -18,30 +19,30 @@ digraph boxes_and_circle {
         node [color = '#ffa600']
   
           'Raw Spreadsheet' -> 'Convert Coordinates\nfrom DMS to DD' 
-              [label = '     dms2dd', fontsize = 26, penwidth = 6]
+              [label = '     dms2dd', fontsize = 40, penwidth = 6]
           'Convert Coordinates\nfrom DMS to DD' -> 'Clean Coordinates' 
-              [label = '     autofill_checker', fontsize = 26, penwidth = 6]
+              [label = '     autofill_checker', fontsize = 40, penwidth = 6]
           'Clean Coordinates' -> 'Geospatial Data' 
-              [label = '     coords2sf', fontsize = 26, penwidth = 6]
+              [label = '     coords2sf', fontsize = 40, penwidth = 6]
           'Geospatial Data' -> 'Political\nBoundaries' 
-              [label = '     political_grabber', fontsize = 26, penwidth = 6]
+              [label = '     political_grabber', fontsize = 40, penwidth = 6]
           'Political\nBoundaries' -> 'Site Characteristics' 
-              [label = '     physical_grabber', fontsize = 26, penwidth = 6]
+              [label = '     physical_grabber', fontsize = 40, penwidth = 6]
           'Site Characteristics' -> 'Site Name' 
-              [label = '     site_writer', fontsize = 26, penwidth = 6]
+              [label = '     site_writer', fontsize = 40, penwidth = 6]
               
               
           'Raw Spreadsheet' -> 'Seperate\nname'
-              [label = '     split_binomial', fontsize = 26, penwidth = 6, style = dashed]
+              [label = '     split_binomial', fontsize = 40, penwidth = 6, style = dashed]
           'Raw Spreadsheet' -> 'Check for\nautofill values' 
-              [label = '     autofill_checker', fontsize = 26, penwidth = 6, style = dashed]
+              [label = '     autofill_checker', fontsize = 40, penwidth = 6, style = dashed]
           'Seperate\nname' -> 'Convert Coordinates\nfrom DMS to DD' 
               [penwidth = 6, style = dashed]
[…]

### 2024-08-16 manuscript/manuscript_APPS.rmd
writing about other software

@@
 - \usepackage[export]{adjustbox}
+- \usepackage{rotating}
 ---
@@
 
-\begin{wrapfigure}{l}{0.35\textwidth}
-  \centering
-    \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
-  \caption{Recommended workflow.}
-\end{wrapfigure}
-
 The generation of an herbarium specimen includes many steps which are easy to take for granted (@forman1989herbarium). 
@@
 
+The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
+The office suite costs money, and in our experience is notoriously finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use.
+PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
+The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections.
+However, not all herbaria use Symbiota and many have original database systems which they maintain. 
+Furthermore, many collectors like to process their own labels, especially as they are likely to be sending sets to different institutions. 
+Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems.
+
 BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert).
@@
 
+\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}
+
 ```{r install package}
-## devtools::install_github('sagesteppe/BarnebyLives')
+## devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
 ```
@@
 time_split_binomials <- system.time({ # split up names into their components
-  data <- split_binomial(data, 'Scientific_name')
+  data1 <- split_scientificName(data, 'Scientific_name')
 })
@@
 BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites.
-Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s.
+Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s.
 The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
@@
 
-## Results
+## RESULTS
 
@@
 
+# DISCUSSION
+
+While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic),
+
+Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly softwares, such as ArcPro, or Adobe Acrobat. 
+While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats
+
+LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions... 
+While the program
+
 # CONCLUSIONS
@@
 Data collection and analysis were performed by R.C.B. 
-R.C.B. wrote the manuscript with input from all other authors. 
-All authors approved the final version of the manuscript.
+R.C.B. & J.B.F wrote the manuscript, and both authors approved the final version of the manuscript.
 

### 2024-08-16 graphics/plots/prep_image.sh

@@ -1,0 +1,6 @@
+#!/bin/sh
+
+mogrify -trim *;
+convert -gravity center -append Spatmod.png Taxmod.png Formatmod.png Reviewmod.png Resultmod.png caption.png workflow.png; 
+convert workflow.png -fuzz 2% -transparent white workflow.png
+exit

### 2024-08-16 manuscript/flowchart.R

@@ -3,112 +3,6 @@ library(DiagrammeR)
 library(ggpubr)
 setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
 
-grViz("
-
-digraph boxes_and_circle {
-  
-  node [shape = oval, fontsize = 28, fontname = arial, style = filled, fontcolor = White]
-
-      subgraph cluster_import{color= 'white'
-        node [color = '#2F0A28']
-        'Raw Spreadsheet'
-      }
-      
-        # Geospatial Operations
-      subgraph cluster_geospatial { style=dashed; color= 'white'; 
-        node [color = '#ffa600']
-  
-          'Raw Spreadsheet' -> 'Convert Coordinates\nfrom DMS to DD' 
-              [label = '     dms2dd', fontsize = 40, penwidth = 6]
-          'Convert Coordinates\nfrom DMS to DD' -> 'Clean Coordinates' 
-              [label = '     autofill_checker', fontsize = 40, penwidth = 6]
-          'Clean Coordinates' -> 'Geospatial Data' 
-              [label = '     coords2sf', fontsize = 40, penwidth = 6]
-          'Geospatial Data' -> 'Political\nBoundaries' 
-              [label = '     political_grabber', fontsize = 40, penwidth = 6]
-          'Political\nBoundaries' -> 'Site Characteristics' 
-              [label = '     physical_grabber', fontsize = 40, penwidth = 6]
-          'Site Characteristics' -> 'Site Name' 
-              [label = '     site_writer', fontsize = 40, penwidth = 6]
-              
-              
-          'Raw Spreadsheet' -> 'Seperate\nname'
-              [label = '     split_binomial', fontsize = 40, penwidth = 6, style = dashed]
-          'Raw Spreadsheet' -> 'Check for\nautofill values' 
-              [label = '     autofill_checker', fontsize = 40, penwidth = 6, style = dashed]
-          'Seperate\nname' -> 'Convert Coordinates\nfrom DMS to DD' 
-              [penwidth = 6, style = dashed]
-          'Check for\nautofill values' -> 'Convert Coordinates\nfrom DMS to DD' 
-              [penwidth = 6, style = dashed]
-              
-          'Geospatial Data' -> 'Directions'
-              [label = '     directions_grabber', fontsize = 40, penwidth =6, style=dashed]
-          'Directions' -> 'Site Name'  [penwidth = 6, style=dashed]
-      }
-              
-              
-          # Taxonomic Operations
[…]

### 2024-08-16 manuscript/manuscript_APPS.rmd
data

@@
 The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
-The office suite costs money, and in our experience is notoriously finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use.
+The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use.
 PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
 The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections.
+Symbiota offers functionality similar to our entire 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
 However, not all herbaria use Symbiota and many have original database systems which they maintain. 
-Furthermore, many collectors like to process their own labels, especially as they are likely to be sending sets to different institutions. 
+Furthermore, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
 Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
+In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats
+after utilizing BLs data cleaning utilities.
 
@@
 The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
-pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
+pdflatex, lualatex, xelatex), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk]().
 While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances of an outdated circumscription applied.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied.
 At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.
@@
 
-While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic),
+While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
+By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
 
-Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly softwares, such as ArcPro, or Adobe Acrobat. 
-While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats
+Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
+While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats.
 
-LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions... 
+LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions. 
+Very good documentation of LaTeX capabilities are offered in multiple areas for instance via the [Overleaf](https://www.overleaf.com/learn)) project. 
 While the program

### 2024-09-12 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
JBF edits and descriptions of fns

--- 2024-08-16 manuscript/manuscript_APPS.rmd
@@
 author:  |
-    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
     |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
+    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
 abstract: |
+  \noindent
   **Premise:** Depositing specimens to herbaria is a time consuming task.
@@
 Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data.
+In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data for fast accessioning of specimens
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
@@
 Optionally the package writes driving directions to sites using Google Maps.
-The package outputs data in a tabular format for review by the user to accept or confirm changes, before dynamically rendering labels.
+The package outputs data in a tabular format, as well as a spatial format, for review by the user to accept or confirm changes, before dynamically rendering labels using LaTeX.
   **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
 keywords: |
-  herbariaum, natural history museum, collections, geospatial, automation, R, software 
+  herbarium, natural history museum, collections, geospatial, R, LaTeX   
 output:
@@
 
-Nearly 400 million specimens are housed in herbaria around the globe (@thiers2021herbaria).
-These specimens, collected to describe the taxonomic diversity of plants and document the worlds floristic diversity, have recently found myriad new applications in several adjacent fields such as conservation biology and ecology (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity towards understanding cellular and molecular processes (@prather2004decline, @pyke2010biological, @daru2018widespread).
-This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices (@funk2014erosion).
-Fortunately, renewed interest approaches in collections generated by 'big data approaches' have brought herbarium collections back to the forefront of the natural sciences (@ronsted2020integrative, @marsico2020small).
+Nearly 400 million specimens are housed in herbaria around the globe [@thiers2021herbaria].
+However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes [@prather2004decline; @pyke2010biological; @daru2018widespread].
+This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion].
+istorically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+A renewed interest in herbarium data through 'big data approaches', such as museuomics, has brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
-In fact innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative).
-The current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants (@ronsted2020integrative, @davis2023herbarium).
-Further we anticipate that collections are yet to gain their most widespread utilization as a revitalization of natural history appears underway in ecology, fostered via novel approaches such as remote and electronic sensing, meta-barcoding, and community science  (@tosa2021rapid).
-While image or purely observational (rather than collections based) citizen science initializes (e.g.
-iNaturalist) have dovetailed to meet many needs of these studies specimens contain rich data which are not accessible via images.
-Namely specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes (@borges2020schrodinger), and seeds or pollen; eternally ensuing specimens as the ultimate data source to center most efforts around.
+Innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+The current uses of specimens and derived data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
+We anticipate that collections have yet to gain their full, fostered by novel approaches become more accessible, such as remote and electronic sensing, meta-barcoding, and community science  [@tosa2021rapid].
+While image based or purely observational (rather than collections based) citizen science approaches (e.g.
+iNaturalist, BudBurst) have dovetailed to meet many of these research needs specimens contain rich data which are not accessible via images.
+Specimens have the additional potential to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
+These factors will ensure that specimens will remain the ultimate botanical data source into perpetuity.
 
-However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow (@prather2004decline). 
-We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, than simply taking photographs via well-developed apps (@daru2018widespread, @mishler2020spatial, @manzano2021fair).
-While many young botanists, capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
+However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow [@prather2004decline]. 
+We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, when compared to taking photographs via professionally developed apps which can be run on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
+While many young botanists are capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
 Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
-geology, geomorphology), nomenclature and Latin, various computer programs (e.g.
-Microsoft Office suite), and increasingly - plant systematics (@nanglu2023nature, @woodland2007botanists, @barrows2016crossroads).
-In the absence of suitable mentors this assuredly results in not only the delay in the deposition of many specimens, but undoubtedly in a failure for many specimens to be accessioned at all, and increasingly ever collected.
+geology, geomorphology), nomenclature, familiarity with various computer programs (e.g.
+Microsoft Office suite), and increasingly - foundational knowledge of plant systematics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+In the absence of suitable mentors this results in not only the delay in the deposition of many specimens, but in a failure for many specimens to be accessioned at all, and increasingly ever collected.
 
-The generation of an herbarium specimen includes many steps which are easy to take for granted (@forman1989herbarium). 
-For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate resources (printed topographic maps, or GIS software) at their disposal.
-In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their position. 
-Finding appropriate sites names is another problem which can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS.
-The rate at which taxonomic innovations are occurring has left many Floras difficult for young users to interpret and has made it difficult for them to find more recently applied names @hitchcock2018flora).
-Upon finding a name they may find it frustrating to hear that while published, the proposal has been accepted by few practitioners and they have unwittingly offended certain curators. 
-Formatting a label correctly (e.g. abbreviations), if successful upon even setting up a mail merge, is a time consuming process and likely to introduce several errors in formatting. 
-Even if a collector navigates all of these hurdles successfully, the time allocated to each step is quite large. 
-Further each step of interfacing with different resources increases the opportunity for transcription errors.
+The generation of an herbarium specimen includes many steps which are easy to take for granted [@forman1989herbarium]. 
+For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate cartographic resources (printed topographic maps, or GIS software) at their disposal.
+In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position.
+Even finding appropriate sites names can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS. 
+Similarly, the rate at which taxonomic innovations are occurring has made it difficult to find more recently applied names [@hitchcock2018flora]. 
+Furthermore formatting a label correctly (e.g. abbreviations) is a time consuming process and likely to introduce several errors in formatting. 
+Anecdotally, many mail merge templates still require collectors to modify many variables by hand, e.g. applying italicization. 
+Even if a collector navigates all of these hurdles, the time allocated to each step is quite large.
 
-Here we provide a description of the BarnebyLives R package. 
-BarnebyLives aims to increase both the data quality of labels, and to speed up the process of producing them. 
+As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the data quality of labels, and to speed up the process of producing them. 
 It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
@@
 Using the GMBA Mountain Inventory V.
-2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region without any ambiguity (@snethlage2022hierarchical).
+2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region with less ambiguity (@snethlage2022hierarchical).
 Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms (@govaerts2021world, @powo2024).
@@
 
-The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
-The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use.
-PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
-The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections.
-Symbiota offers functionality similar to our entire 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
-However, not all herbaria use Symbiota and many have original database systems which they maintain. 
-Furthermore, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
+Currently the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
+The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation.  
+PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
+The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to work on all operating systems.
+Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
+However, not all herbaria use Symbiota and many have original database systems which they maintain (e.g.
+Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html, Missouri Botanical Garden https://tropicos.org/specimen/Search, and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/).
+However, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
 Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
@@
 
+BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
+Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.
+
 ## Usage
-All steps of BarnebyLives except for label generation are run from within Rstudio. 
+All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
 Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all of the on-disk memory associated with these data are for storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
+Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
 Functions which require the on-disk data require a path to the data as an argument. 
 Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.
-We anticipate most personal BarnebyLives instances will be less than several gigabytes (ours covering all of the conterminous Western U.S.
-is XX GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+
+We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
 The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
-pdflatex, lualatex, xelatex), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk]().
-While these steps are run through bash, we have wrapped them in R functions which bypass the need to enter the commands to a terminal.
+[pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
+While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
+
+## Functionality
+BarnebyLives can be thought of as consisting of five main modules (Figure 1): spatial, taxonomic, formatting, manual review, and data exporting.
+
+The spatial module has five required functions, and two optional functions.
+*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs, and will emit a warning if they are encountered. 
+*coords2sf* will create a spatially explicit simple feature (sf) geometry data set of the input data. 
+*political_grabber* will determine many levels of administrative ownership including land management, and public land survey system sections. 
+*physical_grabber* provides various geographic data such as elevation, landform position, and aspect using 90m resolution spatial data. 
+*site_writer* will write directions from an officially named and located placename to the collection site. 
+*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (e.g. 42°08'39.9"N 87°47'08.3"W) to decimal degree format (e.g. 42.14439, -87.78569).
+*directions_grabber* is an optional function which writes driving directions from a reasonably sized town to the closet drivable area to the site using the Google Maps API; this will require a valid Google account, which is free per month for most personal and smaller academic usages.
+
+The taxonomic module has four required functions, and one optional function. 
+*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World Database filtered to the local continents, or a user specific backbone. 
+*spell_check_family* performs a spell check on the family entered for each scientific name. 
+*author_check* ensures authors are entered in a valid format, e.g. the correct standard abbreviations are used. 
+*associates_check* will perform a spell check on all of the associated species using the local taxonomic database.
+*powo_searcher* can be used tandem to the functions *spell_check_family*, and *author-check*, but we use it in lieu of them; this function will search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species.
+No API key or any registration is required to use *powo_searcher*.
+
+The formatting modules has three functions, the first two we will detail are technically optional, but they are run locally and so quickly there is no reason to skip them. 
+*associate_dropper* if present this function will silently remove the collected species from the list of associated species, however it will search for the species to remove using scientific name entered initially by the user rather than returned via spell checks etc.
+*field_lengths* will emit messages for any fields which we suspect will create an 'overflow' on the physical label and should be truncated for clarity. 
+*date_parser* is mandatory and will parse an input date into various formats for notating collection and determination dates on labels.
+
+The manual review process technically only has one function which is optional, and may be ran during the spatial process (after *coords2sf*), but the importance of manual review is important enough that it warrants explicit mention.
+*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/).
+Notably many of the flags which BarnebyLives generates will be placed into columns (e.g.
+NAMES OF COLUMNS FROM FLAGS), and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.
+
+The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
+*labels_skeleton* is an R script which will require a few modification steps to tailor to each instiution, these r scripts will put data into a TEMPLATE...
 
@@
 
-\begin{wrapfigure}{r}{0.35\textwidth}
-  \centering
-    \includegraphics[width=0.35\textwidth]{../graphics/plots/collections_map-trim.png}
+### Herbarium Collections
+
+\begin{wrapfigure}{R}{0.35\textwidth}  
+ \begin{center}
+    \includegraphics[width=.35\textwidth]{../graphics/plots/collections_map-trim.png}  
   \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.} 
+\end{center}
 \end{wrapfigure}
 
-### Herbarium Collections
-
-The package was finalized using the primary authors collections from 2023. 
+The package was released into beta testing using the primary authors collections from 2023. 
 The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned. 
@@
 
-BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites.
-Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s.
-The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
-The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec.
+The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
+The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, `r time_4perpage`sec to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
 The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera). 
-BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
-Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records. 
-BL `r sum(epithet_diff$Situation=='Spelling')` detected instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species). 
+Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
 In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Full_name']))` unique species). 
-BL flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Full_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
+It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Full_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
 The final record was an egregious error where the order of the specific epithet and the genus name.
 
-The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.
-
+*The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.*
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
@@
 
-\begin{wrapfigure}{r}{0.45\textwidth}
-  \centering
+\begin{wrapfigure}{R}{0.45\textwidth}
+  \begin{center}
     \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
   \caption{Sources of Data required for operations}
+  \end{center}
 \end{wrapfigure}
@@
 While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
+We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format. 
 By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
@@
 Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
-While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats.
+While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats.
 
 LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions.
-Very good documentation of LaTeX capabilities are offered in multiple areas for instance via the [Overleaf](https://www.overleaf.com/learn)) project. 
-While the program
+Anecdotally, using its default settings it is able to produce more aesthetically pleasing results than the typical word processors.
+Very good documentation of LaTeX capabilities is offered in multiple areas for instance via the [Overleaf](https://www.overleaf.com/learn)) project.
 
 # CONCLUSIONS
-BarnebyLives is a tool which is able to rapidly acquire relevant geographic, and taxonomic data. 
+BarnebyLives is an R package which can rapidly acquire relevant geographic, and taxonomic data. 
 It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data. 
@@
 # ACKNOWLEDGEMENTS 
-The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for the majority of his specimen collection activities. 
+The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for most of his specimen collection activities. 
 Two anonymous peer reviewers who increased the quality of this manuscript are thanked. 
@@
 Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel. 
+Rosalind Rowe is thanked for providing useful feedback during the later development stages of the program.
 
@@
 A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. 
-All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript
+All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.
 
 # ORCID
-Reed Benkendorf https://orcid.org/0000-0003-3110-6687  
+Reed Clark Benkendorf https://orcid.org/0000-0003-3110-6687  
 Jeremie Fant https://orcid.org/0000-0001-9276-1111

### 2024-09-12 manuscript/flowchart.R

@@ -12,7 +12,7 @@ digraph boxes_and_circle {
   node [shape = circle, fontsize = 36, fontname = arial, style = filled, fontcolor = White]
 
       subgraph cluster_import{color= 'white'
-        node [color = '#CDF7F6', shape = egg]
+        node [color = 'black', shape = egg]
         'Raw\nSpreadsheet'
       }
       
@@ -21,11 +21,13 @@ digraph boxes_and_circle {
         node [color = '#94524A']
   
           'Raw\nSpreadsheet' -> 'Convert\nCoordinates\nfrom\nDMS to DD' 
-              [label = '     dms2dd', fontsize = 50, penwidth = 10]
-          'Convert\nCoordinates\nfrom\nDMS to DD' -> 'Clean\nCoordinates' 
-              [label = '     autofill_checker', fontsize = 50, penwidth = 6]
-          'Clean\nCoordinates' -> 'Geospatial\nData' 
+              [label = '     dms2dd', fontsize = 50, penwidth = 6, style = dashed]
+          'Convert\nCoordinates\nfrom\nDMS to DD' -> 'Check for\nautofill\nvalues' 
+              [label = '     autofill_checker', fontsize = 50, penwidth = 6, style = dashed]
+          'Check for\nautofill\nvalues' -> 'Geospatial\nData' 
               [label = '     coords2sf', fontsize = 50, penwidth = 6]
+          'Raw\nSpreadsheet' -> 'Check for\nautofill\nvalues' 
+              [label = '     autofill_checker', fontsize = 50, penwidth = 6]
           'Geospatial\nData' -> 'Political\nBoundaries' 
               [label = '     political_grabber', fontsize = 50, penwidth = 6]
           'Political\nBoundaries' -> 'Site\nCharacteristics' 
@@ -33,10 +35,6 @@ digraph boxes_and_circle {
           'Site\nCharacteristics' -> 'Site\nName' 
               [label = '     site_writer', fontsize = 50, penwidth = 6]
               
-          'Raw\nSpreadsheet' -> 'Check for\nautofill\nvalues' 
-              [label = '     autofill_checker', fontsize = 50, penwidth = 6, style = dashed]
-          'Check for\nautofill\nvalues' -> 'Convert\nCoordinates\nfrom\nDMS to DD' 
-              [penwidth = 6, style = dashed]
           'Geospatial\nData' -> 'Directions'
[…]

### 2024-12-22 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
edits to manuscript

@@
   \noindent
-  **Premise:** Depositing specimens to herbaria is a time consuming task.
-Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition.
-Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data for fast accessioning of specimens
+  **Premise:** Depositing specimens to herbaria is a time consuming process.
+Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated tomaking collections and curatorial tasks.
+Despite this, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology.
+In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data to increase the rate of accessioning specimens
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
+Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
 It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations.
 Optionally the package writes driving directions to sites using Google Maps.
-The package outputs data in a tabular format, as well as a spatial format, for review by the user to accept or confirm changes, before dynamically rendering labels using LaTeX.
+The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.
   **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
@@
 output:
+  word_document: default
+  toc: no
   pdf_document: default
-  toc: no
-  word_document: default
 csl: "../citations/american-journal-of-botany.csl"
@@
 This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion].
-istorically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-A renewed interest in herbarium data through 'big data approaches', such as museuomics, has brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
+Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+However a renewed interest in herbarium data through 'big data approaches', such as museuomics, have brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
-Innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have likely brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-The current uses of specimens and derived data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
-We anticipate that collections have yet to gain their full, fostered by novel approaches become more accessible, such as remote and electronic sensing, meta-barcoding, and community science  [@tosa2021rapid].
+Innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+The current uses of specimens and their ancillary data extend beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
+We anticipate that collections have yet to gain their full potential, fostered by novel approaches become more accessible, such as remote and electronic sensing, meta-barcoding, and community science  [@tosa2021rapid].
 While image based or purely observational (rather than collections based) citizen science approaches (e.g.
-iNaturalist, BudBurst) have dovetailed to meet many of these research needs specimens contain rich data which are not accessible via images.
-Specimens have the additional potential to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
-These factors will ensure that specimens will remain the ultimate botanical data source into perpetuity.
+iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many of these research needs specimens contain rich data which are not accessible via images.
+Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
+These factors will ensure that specimens remain the premier botanical data source into perpetuity.
 
-However, despite this renewed recognition of the utility of collections, efforts to continually grow them appear slow [@prather2004decline]. 
-We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, when compared to taking photographs via professionally developed apps which can be run on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys to reliably identify - and able to collect satisfactory - material exist, we have observed that they face difficulties navigating several aspects of data collection and preparation of labels for submission to herbaria.
+However, despite the renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline]. 
+We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
+While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria.
 Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
 geology, geomorphology), nomenclature, familiarity with various computer programs (e.g.
-Microsoft Office suite), and increasingly - foundational knowledge of plant systematics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 In the absence of suitable mentors this results in not only the delay in the deposition of many specimens, but in a failure for many specimens to be accessioned at all, and increasingly ever collected.
@@
 For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate cartographic resources (printed topographic maps, or GIS software) at their disposal.
-In topographically complex areas, where borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position.
-Even finding appropriate sites names can rarely be solved without a printed map, as many software maps now consider many features which would serve as site names extraneous in the era of GPS. 
-Similarly, the rate at which taxonomic innovations are occurring has made it difficult to find more recently applied names [@hitchcock2018flora]. 
-Furthermore formatting a label correctly (e.g. abbreviations) is a time consuming process and likely to introduce several errors in formatting. 
-Anecdotally, many mail merge templates still require collectors to modify many variables by hand, e.g. applying italicization. 
+In topographically complex areas, where administrative borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
+Even finding appropriate sites names can rarely be solved without a printed map, as many navigation related softwares, now consider many features which would serve as site names extraneous.  
+Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these are unanimously accepted by group experts [@hitchcock2018flora].
+Furthermore formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
+Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, e.g. applying italicization. 
 Even if a collector navigates all of these hurdles, the time allocated to each step is quite large.
 
-As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the data quality of labels, and to speed up the process of producing them. 
+As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases, and to speed up the process of producing labels.
 It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
-Census Bureau (@walker2022tigris), the Public Land Survey System, and ownership details of public lands via the Protected-Areas Database (PAD-US) from (@usgs2024padus).
-Site names are suggested via finding the closest unambiguously named place feature via the Geographic Name Information System (GNIS), and by precise calculation of the distance and azimuth from these localities to the collection site (@gnis2024).
-Using the GMBA Mountain Inventory V.
-2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region with less ambiguity (@snethlage2022hierarchical).
-Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms (@govaerts2021world, @powo2024).
-Author abbreviations are verified using IPNI's Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviation of authorities (@ipni2024, @powo2024).
-Checks are performed to search for common issues associated with spreadsheets, or transcription, such as the auto-filling of coordinate and date columns. 
-After final review of the data generated by the package, it allows for the option to export spreadsheets which are usable for mass upload of data to multiple common herbarium databases, as well as the generation of herbarium labels.
+Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) from [@usgs2024padus].
+Site names are suggested via finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS), and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
+Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V.
+2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
+Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024].
+Author abbreviations are verified using International Plant Names Index (IPNI's) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviation of authorities [@ipni2024; @powo2024].
+Checks to search for and flag common issues associated with spreadsheet softwares, or data transcription, such as the auto-filling of coordinate and date columns. 
+After final review of the data generated by the package, it allows for the option to export spreadsheets which are suitable for mass upload of data to multiple common herbarium databases, as well as the generation of herbarium labels.
 
-Currently the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
+Currently herbarium label generation functionality is provided explicitly by two programs PLabel, and Symbiota, and is perhaps most commonly implemented by the Microsoft Word tool and functionality Mail Merge [@perkins2020Plabel; @gries2014symbiota].
 The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation.  
-PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
+PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection.
+While some sources indicate it can only be used on Microsoft we speculate it can be accessed on Linux and Mac using Windows emulators like Wine (**CITE**). 
 The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to work on all operating systems.
@@
 
-BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing in 2000 (@welsh2001rupert).
-Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) (@welsh2001rupert).
-Most importantly, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy (@holmgren2017).
+BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until his passing in 2000 [@welsh2001rupert].
+Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert].
+Most importantly, Rupert was remembered as being generous with his time and in assisting younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].
 
@@
 
-BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
-Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.
+BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botany teams over two years. 
+Essentially, continual updates have been made as the developers became aware of more idiosyncrasies of collection notes and data entry.
 
@@
 All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
-Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
+Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 300 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
+Virtually all the on-disk memory associated with the package are used in storing spatial data. 
+The amount of spatial data will vary by the domain which the user decides to support (Figure XX). 
 Functions which require the on-disk data require a path to the data as an argument. 
@@
 
-We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
-at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
-The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
+We anticipate that for the typical user BarnebyLives will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+The final steps of BarnebyLives, generating the labels, requires working installations of Rmarkdown, a LaTeX installation (e.g.
 [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
 While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
+Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
@@
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
-This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
-In total `r n_spp` species (with `r n_sp_authors` sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authors) in `r n_families` families were used for testing.
+Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
+In total `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
 
@@
 
-BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
 Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera). 
+BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
-Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
+Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
 In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Full_name']))` unique species). 
@@
 
-While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
+While numerous tools have been developed for cleaning of existing herbarium and museum records, few tools help with ensuring that the data entered are accurate (@patten2024geographic). 
 We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format. 
-By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
+By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
 
-Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
-While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats.
+Accessioning often relies on the use of the Microsoft Office suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
+While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats, making them more accessible to many users.
+Whereas the other software often cost money, these are also free, and we recommend users install an open source PDF viewer such as Okular to review their rendered documents.
 
@@
 It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data. 
-The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
+The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
 
@@
 Sofia Garcia is acknowledged for creating the 'Valleys' data set which place naming in the package relies on. 
-Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel. 
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz & Dakota Becerra, Hannah Lovell and , Caitlin Miller & Hubert Szczygiel. 
 Rosalind Rowe is thanked for providing useful feedback during the later development stages of the program.
@@
 # DATA AVAILABILITY STATEMENT
-The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives), and the stable version is available on CRAN. 
-The package includes three real use-case vignettes (tutorials) on usage. 
+The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives).  
+The package includes three real use-case vignettes (tutorials) on usage available at a Github pages site (https://sagesteppe.github.io/BarnebyLives/). 
 One vignette "setting_up_files" explores setting up a instance for a certain geographic area. 

### 2024-12-22 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd

@@
 The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.
+  **Premise:** Depositing specimens to herbaria is a time consuming task.
+Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition.
+Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology.
+In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data for fast accessioning of specimens
+  **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
+Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
+It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations.
+Optionally the package writes driving directions to sites using Google Maps.
+The package outputs data in a tabular format, as well as a spatial format, for review by the user to accept or confirm changes, before rendering labels using LaTeX.
   **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
@@
 Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
-In the absence of suitable mentors this results in not only the delay in the deposition of many specimens, but in a failure for many specimens to be accessioned at all, and increasingly ever collected.
+Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+A renewed interest in herbarium data through 'big data approaches', such as museuomics, has brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
@@
 
-Currently herbarium label generation functionality is provided explicitly by two programs PLabel, and Symbiota, and is perhaps most commonly implemented by the Microsoft Word tool and functionality Mail Merge [@perkins2020Plabel; @gries2014symbiota].
+Currently the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
 The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation.  
-PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection.
-While some sources indicate it can only be used on Microsoft we speculate it can be accessed on Linux and Mac using Windows emulators like Wine (**CITE**). 
+PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
 The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to work on all operating systems.
@@
 Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
+BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
+Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.
+
+## Usage
+All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
+Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
+BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
+Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
+Functions which require the on-disk data require a path to the data as an argument. 
+Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.
+
+We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
+[pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
+While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
@@
 BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+
 Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera). 
-BL flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
 Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
+Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
 In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Full_name']))` unique species). 
@@
 
+While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
+We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format. 
+By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
+
+Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
+While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats.
+
 LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions.
@@
 The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
+
+The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
 
@@
 Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz & Dakota Becerra, Hannah Lovell and , Caitlin Miller & Hubert Szczygiel. 
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel. 
 Rosalind Rowe is thanked for providing useful feedback during the later development stages of the program.

### 2025-01-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
updated

--- 2024-09-12 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
     |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
-    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
+    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
 abstract: |

### 2025-01-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
use rmarkdown not latex to insert imgs

--- 2024-12-22 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
 output:
+  pdf_document: default
+  toc: no
   word_document: default
-  toc: no
-  pdf_document: default
 csl: "../citations/american-journal-of-botany.csl"
@@
 
-\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}
+![Recommended workflow](../graphics/plots/workflow.png){width=100%}
 
@@
 Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry. 
-
-## Usage
-All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
-Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
-BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
-Functions which require the on-disk data require a path to the data as an argument. 
-Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.
-
-We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
-at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
-The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
-[pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
-While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
@@
 
-\begin{wrapfigure}{R}{0.35\textwidth}  
- \begin{center}
-    \includegraphics[width=.35\textwidth]{../graphics/plots/collections_map-trim.png}  
-  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.} 
-\end{center}
-\end{wrapfigure}
+![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}
 
@@
 
-*The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.*
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
@@
 
-\begin{wrapfigure}{R}{0.45\textwidth}
-  \begin{center}
-    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
-  \caption{Sources of Data required for operations}
-  \end{center}
-\end{wrapfigure}
+![Data Sources](../graphics/tables/Screenshot-DataSources.png){width=50%}
 

### 2025-01-26 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
typo

@@
 The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
-*labels_skeleton* is an R script which will require a few modification steps to tailor to each instiution, these r scripts will put data into a TEMPLATE...
+*labels_skeleton* is an R script which will require a few modification steps to tailor to each institution, these r scripts will put data into a TEMPLATE...
 

### 2025-03-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
lets make some templates

--- 2025-01-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
     |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
-    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
+    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
 abstract: |
@@
   **Premise:** Depositing specimens to herbaria is a time consuming process.
-Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated tomaking collections and curatorial tasks.
-Despite this, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data to increase the rate of accessioning specimens
+Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated towards collections and curatorial tasks.
+However, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology.
+In order to faciliate the continued growth of herbaria BarnebyLives was developed as a tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
@@
 The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.
-  **Premise:** Depositing specimens to herbaria is a time consuming task.
-Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition.
-Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data for fast accessioning of specimens
-  **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
-It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations.
-Optionally the package writes driving directions to sites using Google Maps.
-The package outputs data in a tabular format, as well as a spatial format, for review by the user to accept or confirm changes, before rendering labels using LaTeX.
   **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
@@
 
-Nearly 400 million specimens are housed in herbaria around the globe [@thiers2021herbaria].
-However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes [@prather2004decline; @pyke2010biological; @daru2018widespread].
+Nearly 400 million specimens are housed in herbaria around the world [@thiers2021herbaria].
+However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread].
 This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion].
 Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-However a renewed interest in herbarium data through 'big data approaches', such as museuomics, have brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
+However, a renewed interest in herbarium collections utilizing 'big data approaches', such as museuomics, have brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
-Innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-The current uses of specimens and their ancillary data extend beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
-We anticipate that collections have yet to gain their full potential, fostered by novel approaches become more accessible, such as remote and electronic sensing, meta-barcoding, and community science  [@tosa2021rapid].
+In fact, innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+The current uses of specimens and their ancillary data extend well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
+We suspect that collections have yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
 While image based or purely observational (rather than collections based) citizen science approaches (e.g.
-iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many of these research needs specimens contain rich data which are not accessible via images.
-Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, notes on the status and composition of the biotic and abiotic settings at time of collection, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
+iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs specimens contain rich data which are not accessible via images.
+Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
 These factors will ensure that specimens remain the premier botanical data source into perpetuity.
@@
 We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria.
-Apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
+While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and are able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria.
+Evidently, apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
 geology, geomorphology), nomenclature, familiarity with various computer programs (e.g.
 Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
-Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-A renewed interest in herbarium data through 'big data approaches', such as museuomics, has brought herbarium collections back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
@@
 Even finding appropriate sites names can rarely be solved without a printed map, as many navigation related softwares, now consider many features which would serve as site names extraneous.  
-Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these are unanimously accepted by group experts [@hitchcock2018flora].
-Furthermore formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
+Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these names are unanimously accepted by group experts [@hitchcock2018flora].
+Furthermore, formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
 Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, e.g. applying italicization. 
@@
 
-Currently the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
+Currently, the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
 The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation. 
-PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine.
-The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to work on all operating systems.
-Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see FIG). 
+PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection.
+While some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows 'emulators' like Wine. 
+The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
+Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see Figure 1). 
 However, not all herbaria use Symbiota and many have original database systems which they maintain (e.g.
@@
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert].
-Most importantly, Rupert was remembered as being generous with his time and in assisting younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].
+Most germanely, Rupert was remembered as being generous with his time to assist younger botanists with the more arcane aspects of field botany and taxonomy [@holmgren2017].
 
@@
 
-![Recommended workflow](../graphics/plots/workflow.png){width=100%}
+\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}
 
@@
 All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
-Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets.
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 300 real-world examples are on a Google Sheets accessible from the page.
+Data may be read in from any common spreadsheet management system or database connection such as Excel, or free alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and 96 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
 Virtually all the on-disk memory associated with the package are used in storing spatial data. 
-The amount of spatial data will vary by the domain which the user decides to support (Figure XX). 
+The amount of spatial data will vary by the domain which the user decides to support (Figure 3). 
 Functions which require the on-disk data require a path to the data as an argument. 
@@
 
-We anticipate that for the typical user BarnebyLives will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+We anticipate that for the typical user BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
 at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
-The final steps of BarnebyLives, generating the labels, requires working installations of Rmarkdown, a LaTeX installation (e.g.
+The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g.
 [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
 While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
-Unfortunately we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
+Unfortunately, we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
 BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
 Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry.
+
+## Usage
+All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
+Data may be read in from any common spreadsheet management system or database connection such as Excel, or free of charge alternatives such as: LibreOffice, OpenOffice, or via the cloud on Googlesheets.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over *3*00 real-world examples are on a Google Sheets accessible from the page.
+BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
+Virtually all the on-disk memory associated with these data are used in storing spatial data, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is fully documented in the package documentation.
+Functions which require the on-disk data require a path to the data as an argument. 
+Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.
+
+We anticipate BarnebyLives usage will requite less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+at 90m resolution is ~16 GiB), and the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+The final steps of BarnebyLives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g.
+[pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
+While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal.
 Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
@@
 The taxonomic module has four required functions, and one optional function. 
-*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World Database filtered to the local continents, or a user specific backbone. 
+*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents, or a user specified backbone. 
 *spell_check_family* performs a spell check on the family entered for each scientific name. 
@@
 *geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/).
-Notably many of the flags which BarnebyLives generates will be placed into columns (e.g.
-NAMES OF COLUMNS FROM FLAGS), and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.
+Notably many of the flags which BarnebyLives generates will be placed into columns with obviously flagged names, and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.
 
 The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
-*labels_skeleton* is an R script which will require a few modification steps to tailor to each instiution, these r scripts will put data into a TEMPLATE...
+*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
 
@@
 
-![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}
+\begin{wrapfigure}{R}{0.35\textwidth}  
+ \begin{center}
+    \includegraphics[width=.35\textwidth]{../graphics/plots/collections_map-trim.png}  
+  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.} 
+\end{center}
+\end{wrapfigure}
 
 The package was released into beta testing using the primary authors collections from 2023. 
-The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned. 
+The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, all of which are un-accessioned. 
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
 Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
-In total `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
+In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
 
@@
 ```{r load in time trials}
-time_trials <- readRDS('data/processed/time_trials')
+time_trials <- readRDS('./data/processed/time_trials')
 tt_local <- time_trials[time_trials$Group=='Local',]
 
-time_label_gen <- readRDS('data/processed/time_label_gen')[['elapsed']]
-time_4perpage <- gsub('user.*$', '', readLines('labels/4labsperpage.txt'))[1]
-time_final <- gsub('user.*$', '', readLines('labels/process2final.txt'))[1]
+time_label_gen <- readRDS('./data/processed/time_label_gen')[['elapsed']]
+time_4perpage <- gsub('user.*$', '', readLines('./labels/4labsperpage.txt'))[1]
+time_final <- gsub('user.*$', '', readLines('./labels/process2final.txt'))[1]
 
@@
 Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`sec), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`sec), while formatting data for labels took `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`sec.
-The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
+The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
 The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, `r time_4perpage`sec to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
@@
 
+*The number of author abbreviations which were not in the appropriate format were XX (% percent), in nearly all cases the presence or absence of a period were the issue.*
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
@@
 
-![Data Sources](../graphics/tables/Screenshot-DataSources.png){width=50%}
+\begin{wrapfigure}{R}{0.45\textwidth}
+  \begin{center}
+    \includegraphics[width=0.40\textwidth]{../graphics/tables/Screenshot-DataSources.png}
+  \caption{Sources of Data required for operations}
+  \end{center}
+\end{wrapfigure}
 

### 2025-03-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
need troubleshoot label making

@@
 time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
-  data <- associate_dropper(data, 'Full_name')
+  data <- associate_dropper(data, 'Full_name', col  = 'Associates')
+})
+
+time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
+  data <- associate_dropper(data, 'Full_name', col  = 'Vegetation')
 })
@@
 saveRDS(data, file = 'data/processed/data_w_Google_Maps')
+
+```
+
+```{r accept or reject changes, eval = F}
+
+# for this example we will just ACCEPT all changes blindly. 
+# basically we just want some data for running timing exercises on the label making.
+
+# just overwrite the original Associates and Vegetation columns
+data <- data %>% 
+  select(-Associates, -Vegetation) |>
+  rename(Associates = 'SpellCk.Associates', Vegetation = 'SpellCk.Vegetation')
+
+# honestly just overwrite these too. 
+data <- data %>% 
+  select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 
+        -Infrarank, -Infraspecies) |>
+  rename(Full_name = SpellCk.taxon_name, Genus = SpellCk.genus, Epithet = SpellCk.species, 
+         Infrarank = SpellCk.infraspecific_rank, Infraspecies = SpellCk.infraspecies) %>% 
+  mutate(
+    Slope = slope, 
+    Aspect = aspect
+  ) |>
+  unite(Name_authority, 
+        Genus, Epithet, Binomial_authority, Infrarank, Infraspecies, Infraspecific_authority,
+        remove = F, na.rm = T, sep = ' ') |>
+  sf::st_drop_geometry()
+
+# also save a spreadsheet of the cleaned data, we'll use this to create the labels. 
+write.csv(data, file = 'data/processed/cleaned_data.csv', row.names = FALSE)
 ```
@@
 rm(list=ls(pattern = 'time'))
+```
+
+```{r Create Labels}
+processed <- read.csv('data/processed/cleaned_data.csv') %>% 
+  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
+         Coordinate_Uncertainty = '+/- 5m') %>% 
+  data.frame()
+
+processed <- data.frame( apply(processed, 2, as.character) )
+processed <- mutate(processed, Collection_number = as.numeric(Collection_number))
+
+dir.create('./labels/raw')
+p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
+
+purrr::walk(
+  .x = processed$Collection_number,
+  ~ rmarkdown::render(
+    input = './labels/skeleton.Rmd',
+    output_file = file.path(p, glue::glue("{.x}.pdf")),
+    params = list(Collection_number = {.x})
+  )
+)
 ```

### 2025-03-01 manuscript/labels/skeleton.Rmd
need troubleshoot label making

@@
   Collection_number: Collection_number
-  UNIQUEID: UNIQUEID
 geometry: paperheight=4in,paperwidth=4in,margin=0.1in
@@
 ```{r load data}
-data <- readRDS('../data/processed/data_w_Google_Maps') |>
-  sf::st_drop_geometry() |>
-  dplyr::filter(Collection_number == params$Collection_number) |>  
-  sf::st_drop_geometry()
+data <- read.csv('../data/test_data.csv') |>
+  dplyr::mutate(
+    Collection_number = as.numeric(Collection_number), 
+    Coordinate_Uncertainty = '5m') |>
+  dplyr::filter(Collection_number == params$Collection_number)
 ```
@@
 \begin{center}
-\textbf{\LARGE Flora of Western North America}
+\textbf{\LARGE}
 \end{center}
@@
 \large
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r data$Binomial_authority` `r writer(data$Infraspecific_rank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
+`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
 \normalsize
@@
 \begingroup
-`r data$latitude_dd` `r data$longitude_dd` (`r data$Datum`). `r data$Gen`. `r writer(data$Site)`
+`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).
 
-`r writer(data$Habitat)` `r data$physical_environ`  
+`r writer(data$Habitat)`. `r data$physical_environ`  
 Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
@@
 \endgroup
-
-\begin{center}
-{\footnotesize Collected under the auspices of the Bureau of Land Management}
-\end{center}

### 2025-03-01 manuscript/labels/making_labels.R

@@ -1,27 +1,23 @@
 library(tidyverse)
 library(BarnebyLives)
-library(googlesheets4)
 
 setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels')
 dir.create('raw')
 dir.create('processed')
 dir.create('final')
 
+collections <- read.csv('../data/processed/cleaned_data.csv') %>% 
+  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
+         Coordinate_Uncertainty = '+/- 5m') %>% 
+  data.frame() 
+
 p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
 
-#p2libs <- .libPaths()[ # find the skeleton here and copy to local directory 
-#  grepl(paste0(version$major, '.', sub('\\..*', "", version$minor)), 
-#        .libPaths())]
-
-folds <- c('skeleton.Rmd')
-
-collections <- readRDS('../data/processed/data_w_Google_Maps') |>
-  sf::st_drop_geometry() 
 time_label_gen <- system.time({
   purrr::walk(
     .x = collections$Collection_number,
     ~ rmarkdown::render(
-      input = folds,
+      input = 'skeleton.Rmd',
       output_file = file.path(p, glue::glue("{.x}.pdf")),
       params = list(Collection_number = {.x})
     )

### 2025-03-02 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
workflow

@@
 ```{r install package}
-## devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
+## devtools::install_github('sagesteppe/BarnebyLives')
 ```
@@
 
-```{r Create Labels}
+```{r Create Labels, eval = F}
 processed <- read.csv('data/processed/cleaned_data.csv') %>% 
@@
          Coordinate_Uncertainty = '+/- 5m') %>% 
+  filter(Collection_number != 'sine non') %>% 
   data.frame()
@@
 dir.create('./labels/raw')
-p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw'
+dir.create('./labels/raw/Reed-raw')
+p <- '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw/Reed-raw'
 
+raw_lab_time <- system.time({
   purrr::walk(
@@
   )
+})
+
+saveRDS(raw_lab_time, file = 'data/processed/time_label_gen')
+
+# now copy bash script to location # note we are one dir up from the lowest level. 
+p2libs <- system.file(package = 'BarnebyLives')
+folds <- file.path( 'rmarkdown', 'render_labels.sh')
+
+file.copy(from = file.path(p2libs, folds), 
+          to = '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw')
+```
+
+```{r, eval = T}
+## this SHOULD be a sh chunk, so launch in a shell! I just want to be sure you can see the code. 
+## and eval = TRUE will be respected. 
+user: pwd
+/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw
+# this will have both your script and a dir full of raw labels with a users name
+# now launch the bash script with: 
+bash render_labels.sh collector='Reed'
+# or chmod it etc!
+# et voila! data will be saved up on dir in a 'final' directory. 
 ```

### 2025-03-02 manuscript/labels/skeleton.Rmd
workflow

@@
 ```{r load data}
-data <- read.csv('../data/test_data.csv') |>
+data <- read.csv('../data/processed/cleaned_data.csv') |>
   dplyr::mutate(
@@
 \begin{center}
-\textbf{\LARGE}
+\textbf{\LARGE `r data$Project_name`}
 \end{center}

### 2025-03-26 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
ordered coll nos

@@
 
-```{r install package}
+```{r install package, eval = F}
 ## devtools::install_github('sagesteppe/BarnebyLives')
@@
 
-```{r, eval = T}
+```{r combine labels, eval = F}
 ## this SHOULD be a sh chunk, so launch in a shell! I just want to be sure you can see the code. 
@@
 # et voila! data will be saved up on dir in a 'final' directory. 
+# we timed it like this:
+\time -o ../process2final.txt  bash render_labels.sh collector='Reed'
 ```
@@
 time_label_gen <- readRDS('./data/processed/time_label_gen')[['elapsed']]
-time_4perpage <- gsub('user.*$', '', readLines('./labels/4labsperpage.txt'))[1]
 time_final <- gsub('user.*$', '', readLines('./labels/process2final.txt'))[1]
@@
 The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
-The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, `r time_4perpage`sec to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
+The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
 The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.
@@
   filter(POW_Family != Family) %>% 
-  select(Collection_number, Full_name, Family, POW_Family) %>% 
+  select(Collection_number, Scientific_name, Family, POW_Family) %>% 
   mutate(Situation = 
@@
   filter(POW_Genus == 'NOT FOUND') %>% 
-  select(Collection_number, Full_name, Genus, POW_Genus)
+  select(Collection_number, Scientific_name, Genus, POW_Genus)
 

### 2025-03-29 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
det date last day of year

@@
 time_date_parser <- system.time({ # parse dates into museum formats
-  data <- date_parser(data, coll_date = 'Date_digital')
+  data <- date_parser(data, coll_date = 'Date_digital', det_date = 'Determined_date')
 })
@@
 
-time_directions_grabber <- system.time({ # write directions to sites
-  SoS_gkey = Sys.getenv("Sos_gkey")
-  data <- directions_grabber(data, api_key = SoS_gkey)
-})
+#time_directions_grabber <- system.time({ # write directions to sites
+#  SoS_gkey = Sys.getenv("Sos_gkey")
+#  data <- directions_grabber(data, api_key = SoS_gkey)
+#})
 
-saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
+#saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
 saveRDS(data, file = 'data/processed/data_w_Google_Maps')
@@
   select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 
-        -Infrarank, -Infraspecies) |>
+        -Infrarank, -Infraspecies, -Binomial_authority, -Infraspecific_authority) |>
   rename(Full_name = SpellCk.taxon_name, Genus = SpellCk.genus, Epithet = SpellCk.species, 
+         Binomial_authority = POW_Binom_authority, Infraspecific_authority = POW_Infra_authority,
          Infrarank = SpellCk.infraspecific_rank, Infraspecies = SpellCk.infraspecies) %>% 

### 2025-03-30 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
notes

@@
 
+time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
+  geodata_writer(data, path = 'data/processed', 
+               filename = 'Herbarium_Collections',
+               filetype = 'kml')
+})
+
 p2geo <- '/media/steppe/hdd/Barneby_Lives-dev/geodata'
@@
   data <- date_parser(data, coll_date = 'Date_digital', det_date = 'Determined_date')
-})
-
-time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
-  geodata_writer(data, path = 'data/processed', 
-               filename = 'Herbarium_Collections_2023',
-               filetype = 'kml')
 })

### 2025-03-30 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
more data

@@
 # now copy bash script to location # note we are one dir up from the lowest level. 
-p2libs <- system.file(package = 'BarnebyLives')
-folds <- file.path( 'rmarkdown', 'render_labels.sh')
+# note that I customized the label template after import, so we will not run this again. 
+# p2libs <- system.file(package = 'BarnebyLives')
+# folds <- file.path( 'rmarkdown', 'render_labels.sh')
 
-file.copy(from = file.path(p2libs, folds), 
-          to = '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw')
+# file.copy(from = file.path(p2libs, folds), 
+#           to = '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw')
 ```

### 2025-03-30 manuscript/labels/skeleton.Rmd
more data

@@
     Collection_number = as.numeric(Collection_number), 
-    Coordinate_Uncertainty = '5m') |>
+    Coordinate_Uncertainty = '+/- 5m') |>
   dplyr::filter(Collection_number == params$Collection_number)
@@
 \endgroup
-\begingroup
-\begin{spacing}{0.1}
-{\scriptsize `r data$Directions_BL`}
-\end{spacing}
-\endgroup
 

### 2025-04-07 contacts_grabber.R
data

@@ -1,0 +1,86 @@
+#' Given an index herbariorum code find a shipping address and a curator's contact information
+#'
+#' @description Assuming you know an IH code this function will acquire some basic information from IH for creating a letterhead on transmittal notices.
+#' It's utility is somewhat restricted by idiosyncrasies within the IH database itself, and some data which have somewhat free forms of formatting.
+#' @param x Character. an IH code. If you don't know it, you can find it at the following link https://sweetgum.nybg.org/science/ih/.
+
+
+museums
+
+museums$address[, grep('physical|postal', colnames(museums$address)) ]
+
+contacts_grabber <- function(x){
+
+  # CAT THE CODE WE ARE SEARCHING FOR HERE!!! NO NEED TO DL THE WHOLE COUNTRY TO GET
+  # THE RELEVANT RESULTS, START OUT SUBSET...
+
+  url <- "http://sweetgum.nybg.org/science/api/v1/institutions/search?country=U.S.A."
+  museums <- jsonlite::fromJSON(url)$data
+
+  museums <- cbind(
+
+    museums[,c('organization', 'code')], # these two fields from 1st table
+
+    { # here we grab the shipping addresses.
+      museum_addy <- museums$address[, grep('physical|postal', colnames(museums$address)) ]
+      museum_addy[museum_addy==""] <- NA # it's easy to test for NA's
+
+      # overwrite original values that are NA, with the physical address information.
+      # really long winded- but want to be able to deal with any IH name changes
+      # without having to refactor.
+      post <- cnames[grep('postal', cnames)]
+      for (i in seq_along(post)){
+        museum_addy[,museum_addy[i]] <- ifelse(
+          is.na(
+            museum_addy[,museum_addy[i]]),
+          museum_addy[, gsub('postal', 'physical', museum_addy[i])], museum_addy[,museum_addy[i]]
+        ) # this only catches pretty egregious errors in the DB.
+      }
+      # no need for the physical 'walk in' addresses for the institutes.
+      museumsTEST2 <- museumsTEST2[, grep('postal', colnames(museumsTEST2))]
+    },
+
+    museums$contact
+  )
+
+  # now we will just drop many institute with missing street addresses. The odds that
+  # a user of this package would be shipping to those locations is very low.
[…]

### 2025-04-08 LabelStyles/SoS-skeleton-Det.Rmd
working on vignette example

--- 2025-03-30 manuscript/labels/skeleton.Rmd
@@
 ```{r load data}
-data <- read.csv('../data/processed/cleaned_data.csv') |>
+data <- read.csv('SoS-ExampleCollection.csv') |>
   dplyr::mutate(
@@
 \begin{center}
-\textbf{\LARGE `r data$Project_name`}
+\textbf{\LARGE Seeds of Success}
 \end{center}
@@
 \normalsize
+`r writer_fide(data)`
 \endgroup
@@
 `r writer(data$Habitat)`. `r data$physical_environ`
+
 Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
@@
 \begingroup
-`r collection_writer(data)`. `r writer_fide(data)`
+`r collection_writer(data)`.
 \endgroup
+
+\begin{center}
+{\footnotesize Collected under the auspices of the Bureau of Land Management}
+\end{center}

### 2025-04-08 LabelStyles/SoS-skeleton-familyLeft.Rmd
working on vignette example

--- 2025-04-08 LabelStyles/SoS-skeleton-Det.Rmd
@@
 \begin{center}
-\textbf{\LARGE Seeds of Success}
+\textbf{\LARGE `r data$Project_name`}
 \end{center}
-\begin{flushright}
+\large
 \uppercase{`r data$Family`}  
-\end{flushright}
 \begingroup
-\large
 `r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
+\endgroup
 \normalsize
-`r writer_fide(data)`
-\endgroup
 
@@
 \begingroup
-`r collection_writer(data)`.
+`r collection_writer(data)`. `r writer_fide(data)`
 \endgroup

### 2025-04-16 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
data
order of operations

@@
 
-associates_spell_check_veg <- system.time({ # ensure associated species are spelled correctly. 
-  data <- associates_spell_check(data, 'Vegetation', p2tax)
+time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
+  data <- associate_dropper(data, 'Binomial', col  = 'Associates')
 })
@@
 time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
-  data <- associate_dropper(data, 'Full_name', col  = 'Associates')
+  data <- associate_dropper(data, 'Binomial', col  = 'Vegetation')
 })
 
-time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
-  data <- associate_dropper(data, 'Full_name', col  = 'Vegetation')
+associates_spell_check_ass <- system.time({ # ensure associated species are spelled correctly. 
+  data <- associates_spell_check(data, 'Vegetation', p2tax)
 })
@@
   select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 
-        -Infrarank, -Infraspecies, -Binomial_authority, -Infraspecific_authority) |>
-  rename(Full_name = SpellCk.taxon_name, Genus = SpellCk.genus, Epithet = SpellCk.species, 
-         Binomial_authority = POW_Binom_authority, Infraspecific_authority = POW_Infra_authority,
-         Infrarank = SpellCk.infraspecific_rank, Infraspecies = SpellCk.infraspecies) %>% 
+        -Infrarank, -Infraspecies, -Binomial_authority, -Infraspecific_authority, -Family) |>
+  rename(Full_name = SpellCk.taxon_name,
+         Genus = SpellCk.genus, 
+         Epithet = SpellCk.species, 
+         Binomial_authority = POW_Binom_authority, 
+         Infraspecific_authority = POW_Infra_authority,
+         Infrarank = SpellCk.infraspecific_rank, 
+         Infraspecies = SpellCk.infraspecies,
+         Family = POW_Family) %>% 
   mutate(
@@
   purrr::walk(
-  .x = processed$Collection_number,
+  .x = na.omit(processed$Collection_number),
   ~ rmarkdown::render(
@@
 # now launch the bash script with:
-bash render_labels.sh collector='Reed'
+
 # or chmod it etc!

### 2025-04-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
copy edits

@@
   \noindent
-  **Premise:** Depositing specimens to herbaria is a time consuming process.
-Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated towards collections and curatorial tasks.
-However, the continual generation of herbarium specimens are essential for current and future research in evolution and ecology.
-In order to faciliate the continued growth of herbaria BarnebyLives was developed as a tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.
+  **Premise:** Depositing specimens in herbaria is a time-consuming process.
+Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks.
+However, the continual generation of herbarium specimens is essential for current and future research on evolution and ecology.
+To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-Census Bureau to provide administrative jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
-It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations.
-Optionally the package writes driving directions to sites using Google Maps.
-The package outputs data in a tabular format, as well as a spatial format, for review by the user before rendering labels using LaTeX.
-  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
+The Census Bureau provides administrative jurisdictional information and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
+It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations.
+Optionally, the package writes driving directions to the sites using Google Maps.
+The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.
+  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
 keywords: |
@@
 
-Nearly 400 million specimens are housed in herbaria around the world [@thiers2021herbaria].
+Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
 However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread].
-This shift, among other factors, lead to a decline in the funding allocated to collections based research, the number of staff maintaining and accessioning new collections, and educating students in these practices [@funk2014erosion].
-Historically specimens have been used to describe the taxonomic diversity of plants and document the worlds floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-However, a renewed interest in herbarium collections utilizing 'big data approaches', such as museuomics, have brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
+This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
+Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
 
-In fact, innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-The current uses of specimens and their ancillary data extend well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
-We suspect that collections have yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
-While image based or purely observational (rather than collections based) citizen science approaches (e.g.
-iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs specimens contain rich data which are not accessible via images.
+Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
+The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
+We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
+While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images.
 Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
-These factors will ensure that specimens remain the premier botanical data source into perpetuity.
+These factors ensure that the specimens remain the premier botanical data source.
 
-However, despite the renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline]. 
-We conjecture this is in part because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys, and other resources, to reliably identify - and are able to collect satisfactory  material, we have observed that they face difficulties navigating several aspects of data acquisition and processing and preparation of labels for submission to herbaria.
-Evidently, apparent problems include the lack of dedicated time at a field seasons end to process specimens, a general lack of education on cartography and orienteering, natural history (e.g.
-geology, geomorphology), nomenclature, familiarity with various computer programs (e.g.
-Microsoft Office suite), and increasingly - foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
+We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
+While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observed that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
+Evidently, apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
-The generation of an herbarium specimen includes many steps which are easy to take for granted [@forman1989herbarium]. 
-For example while acquiring appropriate political information for a collection site appears simple, young collectors rarely have the adequate cartographic resources (printed topographic maps, or GIS software) at their disposal.
-In topographically complex areas, where administrative borders are often associated with hydrologic basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
-Even finding appropriate sites names can rarely be solved without a printed map, as many navigation related softwares, now consider many features which would serve as site names extraneous.  
-Similarly, the rate at which taxonomic innovations are occurring, the volume of the literature and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification, has made it difficult to find more recently applied names - even when these names are unanimously accepted by group experts [@hitchcock2018flora].
-Furthermore, formatting a label correctly (e.g. author abbreviations) is a time consuming process with many opportunities to introduce several errors in formatting. 
-Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, e.g. applying italicization. 
-Even if a collector navigates all of these hurdles, the time allocated to each step is quite large.
+The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
+For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
+In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
+Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as extraneous site names. 
+Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by group experts [@hitchcock2018flora].
+Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
+Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization.
+Even if a collector navigates all these hurdles, the time allocated to each step is quite large.
 
-As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases, and to speed up the process of producing labels.
+As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
 It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
-Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) from [@usgs2024padus].
-Site names are suggested via finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS), and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
+Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
+Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
 Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V.
-2, a standardized named mountain data set with global coverage, which we have supplemented with over XXXX valleys allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
+2, a standardized named mountain data set with global coverage, which we have supplemented with over *XXXX* valleys allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
 Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024].
-Author abbreviations are verified using International Plant Names Index (IPNI's) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviation of authorities [@ipni2024; @powo2024].
-Checks to search for and flag common issues associated with spreadsheet softwares, or data transcription, such as the auto-filling of coordinate and date columns. 
-After final review of the data generated by the package, it allows for the option to export spreadsheets which are suitable for mass upload of data to multiple common herbarium databases, as well as the generation of herbarium labels.
+Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
+Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
+After a final review of the data generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.
 
-Currently, the label generation functionality is provided explicitly by two programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota).
-The office suite costs money, and in our experience is finicky, further it's functionality ends with label creation. 
-PLabel is a standalone program which has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality, unfortunately it does not include data cleaning functionalities beyond verifying nations of collection.
-While some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows 'emulators' like Wine. 
-The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections, and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
-Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries' as well (see Figure 1). 
-However, not all herbaria use Symbiota and many have original database systems which they maintain (e.g.
-Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html, Missouri Botanical Garden https://tropicos.org/specimen/Search, and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/).
-However, many collectors like to generate their own labels, especially as they are likely to be sending different sets of collections to different institutions. 
-Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems. 
-In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats
-after utilizing BLs data cleaning utilities.
+Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
+The office suite costs money, and in our experience, it is finicky; further, its functionality ends with label creation. 
+PLabel is a standalone program that has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
+However, it does not include data cleaning functionalities beyond verifying nations of collection. 
+While some sources indicate that it can only be used in Microsoft, we expect it to be accessed on Linux and Mac using Windows 'emulators' like Wine. 
+The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
+Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
+However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/).
+However, many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
+Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.
+In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.
 
-BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until his passing in 2000 [@welsh2001rupert].
+BarnebyLives was named by plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William and Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until he passed in 2000 [@welsh2001rupert].
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert].
@@
 
-BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botany teams over two years. 
-Essentially, continual updates have been made as the developers became aware of more idiosyncrasies of collection notes and data entry.
+BarnebyLives was iteratively developed based on data submitted by approximately 20 seasonal field botany teams over two years. 
+Essentially, continual updates were made as the developers became aware of the idiosyncrasies of collection notes and data entry.
 
 ## Usage
-All steps of BarnebyLives except for label generation are run within the freely available Rstudio. 
-Data may be read in from any common spreadsheet management system or database connection such as Excel, or free alternatives such as: LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and 96 real-world examples are on a Google Sheets accessible from the page.
-BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
-Virtually all the on-disk memory associated with the package are used in storing spatial data. 
-The amount of spatial data will vary by the domain which the user decides to support (Figure 3). 
-Functions which require the on-disk data require a path to the data as an argument. 
-Manually supplying the path argument allows for users to judiciously decide a storage location suitable for their needs.
+All steps of BarnebyLives, except for label generation are run within the freely available Rstudio. 
+Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page.
+BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
+Virtually all on-disk memory associated with the package are used to store spatial data.
+The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
+Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to judiciously determine a storage location suitable for their needs.
 
-We anticipate that for the typical user BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
-at 90m resolution is ~16 GiB), while the processing takes relatively little RAM, hence we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
+at 90m resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
 The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g.
 [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
-While these steps are run through a shell scripting language like bash, we have wrapped them in R functions which bypass the need to enter the commands directly into a terminal. 
-Unfortunately, we have not found Windows alternatives to pdfam and pdftek so we are unable to offer the final label generating functionality on that operating system. 
-BarnebyLives was iteratively developed based on data submitted by roughly 20 seasonal field botanists. 
-Essentially, additions to the code have been continually made as the developers were exposed to more idiosyncrasies of collection notes and data entry. 
-Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in Figure 1.
+While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a terminal.
+Unfortunately, we did not find Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system. 
+BarnebyLives was iteratively developed based on the data submitted by approximately 20 seasonal field botanists. 
+Essentially, additions to the code have been continually made, as developers were exposed to more idiosyncrasies of collection notes and data entry.
+Several commands in BarnebyLives require output from previous functions, and a workflow that satisfies these requirements is presented in Figure 1.
 
@@
 
-The spatial module has five required functions, and two optional functions.
-*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs, and will emit a warning if they are encountered. 
-*coords2sf* will create a spatially explicit simple feature (sf) geometry data set of the input data. 
-*political_grabber* will determine many levels of administrative ownership including land management, and public land survey system sections. 
-*physical_grabber* provides various geographic data such as elevation, landform position, and aspect using 90m resolution spatial data. 
-*site_writer* will write directions from an officially named and located placename to the collection site. 
-*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (e.g. 42°08'39.9"N 87°47'08.3"W) to decimal degree format (e.g. 42.14439, -87.78569).
-*directions_grabber* is an optional function which writes driving directions from a reasonably sized town to the closet drivable area to the site using the Google Maps API; this will require a valid Google account, which is free per month for most personal and smaller academic usages.
+The spatial module has five required functions and two optional functions.
 
-The taxonomic module has four required functions, and one optional function. 
-*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents, or a user specified backbone. 
+*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered. 
+*coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data.
+political_grabber determines many levels of administrative ownership, including land management and public land survey system sections.
+*physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.
+site_writer will write directions from an officially named placename to the collection site.
+dms2dd is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).
+*directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages.
+
+The taxonomic module has four required functions and one optional function. 
+*spell_check* will perform a spell check on the entered scientific name based on a local copy of Kew Plants of the World database filtered to the local continents or a user-specified backbone. 
 *spell_check_family* performs a spell check on the family entered for each scientific name.
-*author_check* ensures authors are entered in a valid format, e.g. the correct standard abbreviations are used. 
-*associates_check* will perform a spell check on all of the associated species using the local taxonomic database.
-*powo_searcher* can be used tandem to the functions *spell_check_family*, and *author-check*, but we use it in lieu of them; this function will search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species.
-No API key or any registration is required to use *powo_searcher*.
+author_check ensures that the authors are entered in a valid format, for example, the correct standard abbreviations are used.
+*associates_check* performs a spell check on all associated species using the local taxonomic database.
+*powo_searcher* can be used in tandem with the functions spell_check_family and author-check, but we use it in lieu of them to search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species.
+No API key or registration is required to use powo_searcher.
 
-The formatting modules has three functions, the first two we will detail are technically optional, but they are run locally and so quickly there is no reason to skip them. 
-*associate_dropper* if present this function will silently remove the collected species from the list of associated species, however it will search for the species to remove using scientific name entered initially by the user rather than returned via spell checks etc.
-*field_lengths* will emit messages for any fields which we suspect will create an 'overflow' on the physical label and should be truncated for clarity. 
-*date_parser* is mandatory and will parse an input date into various formats for notating collection and determination dates on labels.
+The formatting module has three functions. The first two we will detail are technically optional; however, they are run locally and so quickly that there is no reason to skip them. 
+*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks.
+*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
+*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.
 
-The manual review process technically only has one function which is optional, and may be ran during the spatial process (after *coords2sf*), but the importance of manual review is important enough that it warrants explicit mention.
+The manual review process technically only has one function that is optional and may be executed during the spatial process (after coords2sf), but the importance of manual review is important enough to warrant explicit mention.
 *geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/).
-Notably many of the flags which BarnebyLives generates will be placed into columns with obviously flagged names, and can be manually reviewed by the analyst, many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.
+Notably, many of the flags that BarnebyLives generates will be placed into columns with obviously flagged names and can be manually reviewed by the analyst, and many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.
 
-The data exporting module contains three functions, which interact with LaTeX templates and require slightly more advanced R user interactivity such as setting up mapping functions using the tidyverses purrr package.
+The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
 *labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
+"
 
@@
 
-While numerous tools have been developed for cleaning of existing herbarium and museum records, few tools help with ensuring that the data entered are accurate (@patten2024geographic). 
+While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate (@patten2024geographic).
+We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
+By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
+
+Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
+While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
+
+While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate (@patten2024geographic).
 We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format.
-By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
+By utilizing both R and LaTeX and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications to project branches.
 
-Accessioning often relies on the use of the Microsoft Office suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
-While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats, making them more accessible to many users.
-Whereas the other software often cost money, these are also free, and we recommend users install an open source PDF viewer such as Okular to review their rendered documents.
+Accessioning often relies on the use of the Office Suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat.  
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats.
 
-While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic). 
-We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format. 
-By utilizing both R and LaTeX, and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues, and implementing upgrades and modifications on project branches.
-
-Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly software, such as ArcPro, or Adobe Acrobat. 
-While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet like formats.
-
-LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions.
-Anecdotally, using its default settings it is able to produce more aesthetically pleasing results than the typical word processors.
-Very good documentation of LaTeX capabilities is offered in multiple areas for instance via the [Overleaf](https://www.overleaf.com/learn)) project.
+LaTeX offers well-documented and detailed functionality to customize labels for individuals and institutions. 
+Anecdotally, using its default settings, it can produce more aesthetically pleasing results than typical word processors. 
+Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn)) project.
 
 # CONCLUSIONS
-BarnebyLives is an R package which can rapidly acquire relevant geographic, and taxonomic data. 
-It is also capable of performing specialized spell checks, and assorted curatorial tasks to produce both digital and analog data. 
+BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
+It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
 The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
-
-The package relies on no licensed Software, such as the Microsoft suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
 
@@
 # ACKNOWLEDGEMENTS 
-The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for most of his specimen collection activities. 
-Two anonymous peer reviewers who increased the quality of this manuscript are thanked. 
+The Bureau of Land Management is gratefully acknowledged as a provider of funding to R. C. B. for most of his specimen collection activities. 
+We thank the two anonymous peer reviewers who have increased the quality of this manuscript.
 Sofia Garcia is acknowledged for creating the 'Valleys' data set which place naming in the package relies on. 
-Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz & Dakota Becerra, Hannah Lovell and , Caitlin Miller & Hubert Szczygiel. 
-Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Caitlin Miller & Hubert Szczygiel. 
-Rosalind Rowe is thanked for providing useful feedback during the later development stages of the program.
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz & Dakota Becerra, Hannah Lovell and, Caitlin Miller & Hubert Szczygiel. 
+We thank Rosalind Rowe for providing useful feedback during the later developmental stages of the program.
 
@@
 The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
-The package includes three real use-case vignettes (tutorials) on usage available at a Github pages site (https://sagesteppe.github.io/BarnebyLives/). 
-One vignette "setting_up_files" explores setting up a instance for a certain geographic area. 
-Another vignette "running_pipeline" showcases the usage of the package for processing data entered on a spreadsheet. 
-A final vignette "creating_labels" shows the usage of an R, and Bash script launched from RStudio to produce print-ready labels. 
+The package includes three real use-case vignettes (tutorials) available on a Github page site (https://sagesteppe.github.io/BarnebyLives/).
+One vignette *"setting_up_files"* explores setting up a instance for a certain geographic area.  
+Another vignette *"running_pipeline"* showcases the usage of the package for processing data entered on a spreadsheet.  
+The final vignette *"creating_labels"* shows the usage of an R and Bash script launched from RStudio to produce print-ready labels. 
 All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.

### 2025-04-23 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
back to manuscript!

@@
 
+#subs_batch <- c(1384, 1388, 1394, 1396, 1397, 1400, 1435, 1438, 2161, 2187, 2268, 
+#                2301, 2302, 2577, 2746, 2785, 1460, 1461, 1379, 1381)
+
 data <- read.csv('data/test_data.csv', na.strings = "") %>% 
   drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
-  unite(col = 'Scientific_name', c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F)
+  unite(col = 'Scientific_name', 
+        c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
+  filter(Scientific_name != '')#, Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
 
@@
 
-#time_directions_grabber <- system.time({ # write directions to sites
-#  SoS_gkey = Sys.getenv("Sos_gkey")
-#  data <- directions_grabber(data, api_key = SoS_gkey)
-#})
+time_directions_grabber <- system.time({ # write directions to sites
+  SoS_gkey = Sys.getenv("Sos_gkey")
+  data <- directions_grabber(data, api_key = SoS_gkey)
+})
 
-#saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
+saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
 saveRDS(data, file = 'data/processed/data_w_Google_Maps')
@@
 # how many of my genera were mis-spelled??
+
+## HEY REED SPLIT OUT GENERA FOR THIS !!!!! AGAIN .... 
+data_p <- split_scientificName(data_p, sciName_col = 'Scientific_name')
+
 genera_mispelled <- data_p %>% 
@@
   filter(POW_Genus != Genus) %>% 
-  select(Collection_number, Full_name, Genus, POW_Genus) %>% 
+  select(Collection_number, Scientific_name, Genus, POW_Genus) %>% 
   mutate(Situation = 
@@
     ),
-    Situation = if_else(Full_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))
+    Situation = if_else(Scientific_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))
 
@@
   sf::st_drop_geometry() %>% 
-  filter(Epithet != POW_Epithet, !Full_name %in% genus_mismatch$Full_name) %>% 
-  select(Collection_number, Full_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
-  distinct(Full_name, .keep_all = T) %>%  # do we record repeat collections ?
+  filter(Epithet != POW_Epithet, !Scientific_name %in% genus_mismatch$Scientific_name) %>% 
+  select(Collection_number, Scientific_name, Epithet, POW_Genus, POW_Epithet, POW_Infraspecies) %>% 
+  distinct(Scientific_name, .keep_all = T) %>%  # do we record repeat collections ?
   mutate(Situation = 
@@
              Collection_number %in% c(
-               '1217', '1251', '1283', '1311', '1337', '1391', '1450', '1576', 
-               '1594', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130',
-               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2343', '2370', '2379', 
+               '1217', '1251', '1283', '1311', '1337', '1391', '1415', '1421', '1435', '1450', '1576', 
+               '1594', '1624', '1627', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130', '2135', 
+               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2343', '2370', '2379', 
                '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
@@
              Collection_number %in% c(
-               '1588', '1619', '2445', '2777', '1337', '2412', '2487', '2748', '2652',
+               '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
                '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
@@
 Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
-In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Full_name']))` unique species). 
-It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Full_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
+In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
+It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
 The final record was an egregious error where the order of the specific epithet and the genus name.

### 2025-04-23 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
scored veg spell check

@@
   - \linenumbers
-- \usepackage[width=\textwidth]{caption}
-- \usepackage{wrapfig}
-- \usepackage[export]{adjustbox}
   - \usepackage{rotating}
@@
 The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
-Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to judiciously determine a storage location suitable for their needs.
+Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.
 
@@
 The package was released into beta testing using the primary authors collections from 2023. 
-The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, all of which are un-accessioned. 
+The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
 Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
-In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.
+In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.
 
@@
 The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
-The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).
+The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` **sheets** to a single Portable Document Format (PDF).
 The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.
@@
       Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
-      Family %in% c('Bataceae', 'Ericaceae') ~ 'Internal_Error', 
+      Family %in% c('Ranunculaceae', 'Ericaceae') ~ 'Internal_Error', 
       .default = 'Typo'
@@
                     ) ~ 'Outdated',
-      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal_Error',
+      Genus %in% c('Allium', 'Vaccinium', 'Eleocharis') ~ 'Internal_Error',
       .default = 'Typo'
@@
            case_when(
+             Collection_number %in% c('1207') ~ 'User Error', 
              Collection_number %in% c(
-               '1217', '1251', '1283', '1311', '1337', '1391', '1415', '1421', '1435', '1450', '1576', 
-               '1594', '1624', '1627', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130', '2135', 
-               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2343', '2370', '2379', 
-               '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
+               '1217', '1251', '1283', '1311', '1391', '1415', '1435', '1450', '1556','1576', 
+               '1594', '1624', '1627', '1745', '2043', '2108', '2109', '2120', '2127', '2130', '2135', 
+               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2370', '2379', 
+               '2393', '2480', '2511', '2561', '2573', '2648', '2660', 
                '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
@@
              Collection_number %in% c(
-               '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
-               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
+               '1337', '1421', '1588', '1619','2415', '2445', '2777', '1337', '2004', '2086', '2343', '2412', '2487', '2748', '2652',
+               '2805', '2194', '2040', '2039', '2403',  '2076') 
              ~ 'Outdated', 
@@
              ~ 'Internal_Error',
-             Collection_number %in% c('2110', '2540') ~ 'Preferred'
+             Collection_number %in% c('2110', '2252', '2328', '2540') ~ 'Preferred'
            )
@@
 rm(Family_in, Family_out)
+
+## check spell checks on both vegetation and associates columns
+
+veg_diff <- data_p %>% 
+  sf::st_drop_geometry() %>% 
+  select(Collection_number, Vegetation, SpellCk.Vegetation) %>% 
+  filter(Vegetation != SpellCk.Vegetation) %>% 
+  distinct(Vegetation, SpellCk.Vegetation, .keep_all = TRUE) %>%
+
+  # grouped sites - these sites are essentially replicates of each other, I have just
+  # remove a focal taxon from them. I feel they are pseudoreplicates so they will be discarded
+
+  filter(!Collection_number %in%
+   c('2058', '2107', '2110', '2112', '2117', '2220', '2221', '2407', '2428', '2429', 
+     '2430', '2432', '2540', '2541', '2560', '1407') 
+  ) %>% 
+  mutate(Situation = 
+           case_when(
+             Collection_number %in% c(  
+               '1207', '1286', '1596', '1598', '2004', '2033', '2034', '2133', 
+               '2175', '2402', '2407', '2427', '2760' 
+               ) ~ 'User Error',
+             Collection_number %in% c(
+               '1214', '1218', '1220', '1227', '1228', '1250', '1257', '1258', '1270', 
+               '1273', '1287', '1292', '1293', '1302', '1341', '1343', '1383', 
+               '1384', '1390', '1401', '1455', '1557', '1558', '1565', '1596', 
+               '2000', '2010', '2012', '2014', '2032', '2049', '2054', '2067', '2105', 
+               '2116', '2163', '2219', '2224', '2230', '2395', '2406', '2454', '2528', 
+               '2539', '2562', '2566', '2571', '2572', '2622', '2625', '2656', 
+               '2664', '2668', '2745', '2759', '2787'
+             )
+             ~ 'Spelling',
+             Collection_number %in% c(
+                '1278', '1339', '1405'
+               ) ~ 'Incorrect', # that is the program returns the wrong taxon
+             Collection_number %in% c(
+               '2527' 
+             ) 
+             ~ 'Internal_Error',
+           )
+  )
 ```
@@
 In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
-It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
+It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` record in error.
 The final record was an egregious error where the order of the specific epithet and the genus name.
@@
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
+Vegetation data were entered for `r nrow(data) - sum(is.na(data$Vegetation))` records, and after removal of duplicate site information BL flagged  `r nrow(veg_diff)` records. 
+BL was able to correct the spelling in `r nrow(veg_diff[veg_diff$Situation == 'Spelling',])` instances, and in  `r nrow(veg_diff[veg_diff$Situation == 'User Error',])` or instances the user entered data in a format beyond the capabilities of BL to fix.
+BL returned an incorrect spell check match for  `r nrow(veg_diff[veg_diff$Situation == 'Incorrect',])` records, two of which had the same misspelled species, in `r nrow(veg_diff[veg_diff$Situation == 'Internal_Error',])` instance BL had an internal errors where it returned an incorrect match due to abbreviations.
 
@@
 
+\singlespacing
 # ORCID

### 2025-04-24 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
edits from printed

@@
 Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks.
-However, the continual generation of herbarium specimens is essential for current and future research on evolution and ecology.
-To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the rate of accessioning specimens.
+However, the continual generation of herbarium specimens is essential for current and future research on ecology, evolution, and conservation.
+To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens.
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-The Census Bureau provides administrative jurisdictional information and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions.
+The Census Bureau for administrative jurisdictional data, and other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions and locality information.
 It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations.
@@
 The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.
-  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.
+  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.
 keywords: |
@@
 In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
-Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as extraneous site names. 
-Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by group experts [@hitchcock2018flora].
+Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
+Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora].
 Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
@@
 Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V.
-2, a standardized named mountain data set with global coverage, which we have supplemented with over *XXXX* valleys allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
-Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the collected species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024].
+2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
+Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024].
 Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
 Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
-After a final review of the data generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.
+After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.
 
 Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
-The office suite costs money, and in our experience, it is finicky; further, its functionality ends with label creation. 
-PLabel is a standalone program that has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
+The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
+PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
 However, it does not include data cleaning functionalities beyond verifying nations of collection. 
-While some sources indicate that it can only be used in Microsoft, we expect it to be accessed on Linux and Mac using Windows 'emulators' like Wine. 
+While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
 The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
@@
 However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/).
-However, many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
+However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
 Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.
@@
 
-BarnebyLives was named by plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William and Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until he passed in 2000 [@welsh2001rupert].
+BarnebyLives was named for plant taxonomist Rupert Charles Barneby (1911-2000), who published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the William and Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970 and sporadically until he passed in 2000 [@welsh2001rupert].
 Select accolades of Rupert include the 1989 Asa Gray Award from the American Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congress's (IBC) Millennium Botany Award (1999) [@welsh2001rupert].
@@
 Essentially, continual updates were made as the developers became aware of the idiosyncrasies of collection notes and data entry.  
+Several commands in BarnebyLives require output from previous functions, and a workflow that satisfies these requirements is presented in Figure 1.
 
 ## Usage
-All steps of BarnebyLives, except for label generation are run within the freely available Rstudio. 
+All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
 Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
@@
 We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S.
-at 90m resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
 The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g.
 [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
-While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a terminal.
-Unfortunately, we did not find Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system. 
-BarnebyLives was iteratively developed based on the data submitted by approximately 20 seasonal field botanists. 
-Essentially, additions to the code have been continually made, as developers were exposed to more idiosyncrasies of collection notes and data entry.
-Several commands in BarnebyLives require output from previous functions, and a workflow that satisfies these requirements is presented in Figure 1.
+While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
+Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.
 
@@
 *spell_check_family* performs a spell check on the family entered for each scientific name.  
-author_check ensures that the authors are entered in a valid format, for example, the correct standard abbreviations are used.
+*author_check* ensures that the authors are entered in a valid format, for example, the correct standard abbreviations are used.  
 *associates_check* performs a spell check on all associated species using the local taxonomic database. 
-*powo_searcher* can be used in tandem with the functions spell_check_family and author-check, but we use it in lieu of them to search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species.
-No API key or registration is required to use powo_searcher.
+*powo_searcher* can be used in tandem with the functions *spell_check_family* and *author_check*, but we use it in lieu of them to search the current Plants of the World Online to determine relevant synonyms and alternative higher taxonomy for the focal species.
+No API key or registration is required to use *powo_searcher*.
 
-The formatting module has three functions. The first two we will detail are technically optional; however, they are run locally and so quickly that there is no reason to skip them. 
+The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
 *associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks.
@@
 
-The manual review process technically only has one function that is optional and may be executed during the spatial process (after coords2sf), but the importance of manual review is important enough to warrant explicit mention.
+The manual review process technically only has one function that is optional and may be executed during the spatial process (after *coords2sf*), but the importance of manual review is important enough to warrant explicit mention.
 *geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/).
@@
 *labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
-"
 
@@
 BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
-BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
 
@@
 The spell check of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`sec) spent performing local taxonomic operations.
-The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the `r length(list.files('labels/processed'))` **sheets** to a single Portable Document Format (PDF).
-The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.
+The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`sec) for the rendering, and an additional `r time_final`sec to combine the 182 sheets to a single Portable Document Format (PDF).
+The total label generation run time for processing these `r n_spp` collections was `r minutemen(sum(time_trials$elapsed))` minutes. 
+In total the `r n_spp` collections, which underwent all processing steps, took `minutemen(sum(time_trials$elapsed, time_label_gen, as.numeric(time_final)))` minutes to process.
 
@@
       Family %in% c('Hydrophyllaceae', 'Namaceae') ~ 'Preferred', 
-      Family %in% c('Athyriaceae') ~ 'Outdated',
-      Family %in% c('Saxifragaceae', 'Plantaginaceae') ~ 'Submitter_Incorrect', 
-      Family %in% c('Ranunculaceae', 'Ericaceae') ~ 'Internal_Error', 
+      Collection_number %in% c('1594', '2375', '2547', '2563', '2729') ~'Autofilled', 
+      Collection_number %in% c('1576', '2229', '2376', '2401', '2603', '2688') ~ 'Submitter_Incorrect', 
+      Collection_number %in% c('2377') ~ 'Outdated', 
+      Collection_number %in% c('2284', '2435', '2517', '2622', '2705', '2725', '2726') ~ 'Internal_Error', 
       .default = 'Typo'
@@
              Collection_number %in% c(
-               '1217', '1251', '1283', '1311', '1391', '1415', '1435', '1450', '1556','1576', 
-               '1594', '1624', '1627', '1745', '2043', '2108', '2109', '2120', '2127', '2130', '2135', 
-               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2370', '2379', 
-               '2393', '2480', '2511', '2561', '2573', '2648', '2660', 
+               '1217', '1251', '1283', '1311', '1337', '1391', '1415', '1421', '1435', '1450', '1576', 
+               '1594', '1624', '1627', '1745', '2043', '2086', '2108', '2109', '2120', '2127', '2130', '2135', 
+               '2153', '2199', '2227', '2242', '2256', '2289', '2297', '2315', '2343', '2370', '2379', 
+               '2393', '2415', '2480', '2511', '2561', '2573', '2648', '2660', 
                '2687', '2690', '2710', '2715', '2727', '2750', '2767', '2818')
@@
              Collection_number %in% c(
-               '1337', '1421', '1588', '1619','2415', '2445', '2777', '1337', '2004', '2086', '2343', '2412', '2487', '2748', '2652',
-               '2805', '2194', '2040', '2039', '2403',  '2076') 
+               '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
+               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
              ~ 'Outdated', 
@@
              ~ 'Internal_Error',
-             Collection_number %in% c('2110', '2252', '2328', '2540') ~ 'Preferred'
+             Collection_number %in% c('2110', '2540') ~ 'Preferred'
            )
@@
 rm(Family_in, Family_out)
-
-## check spell checks on both vegetation and associates columns
-
-veg_diff <- data_p %>% 
-  sf::st_drop_geometry() %>% 
-  select(Collection_number, Vegetation, SpellCk.Vegetation) %>% 
-  filter(Vegetation != SpellCk.Vegetation) %>% 
-  distinct(Vegetation, SpellCk.Vegetation, .keep_all = TRUE) %>%
-
-  # grouped sites - these sites are essentially replicates of each other, I have just
-  # remove a focal taxon from them. I feel they are pseudoreplicates so they will be discarded
-
-  filter(!Collection_number %in%
-   c('2058', '2107', '2110', '2112', '2117', '2220', '2221', '2407', '2428', '2429', 
-     '2430', '2432', '2540', '2541', '2560', '1407') 
-  ) %>% 
-  mutate(Situation = 
-           case_when(
-             Collection_number %in% c(  
-               '1207', '1286', '1596', '1598', '2004', '2033', '2034', '2133', 
-               '2175', '2402', '2407', '2427', '2760' 
-               ) ~ 'User Error',
-             Collection_number %in% c(
-               '1214', '1218', '1220', '1227', '1228', '1250', '1257', '1258', '1270', 
-               '1273', '1287', '1292', '1293', '1302', '1341', '1343', '1383', 
-               '1384', '1390', '1401', '1455', '1557', '1558', '1565', '1596', 
-               '2000', '2010', '2012', '2014', '2032', '2049', '2054', '2067', '2105', 
-               '2116', '2163', '2219', '2224', '2230', '2395', '2406', '2454', '2528', 
-               '2539', '2562', '2566', '2571', '2572', '2622', '2625', '2656', 
-               '2664', '2668', '2745', '2759', '2787'
-             )
-             ~ 'Spelling',
-             Collection_number %in% c(
-                '1278', '1339', '1405'
-               ) ~ 'Incorrect', # that is the program returns the wrong taxon
-             Collection_number %in% c(
-               '2527' 
-             ) 
-             ~ 'Internal_Error',
-           )
-  )
 ```
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Typo', na.rm = T)` typos, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where an incorrect family was entered, and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied.
-At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)`records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied.
+At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.
 
@@
 In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
-It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` record in error.
+It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
 The final record was an egregious error where the order of the specific epithet and the genus name.
 
-`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values (% of records).
-Vegetation data were entered for `r nrow(data) - sum(is.na(data$Vegetation))` records, and after removal of duplicate site information BL flagged  `r nrow(veg_diff)` records. 
-BL was able to correct the spelling in `r nrow(veg_diff[veg_diff$Situation == 'Spelling',])` instances, and in  `r nrow(veg_diff[veg_diff$Situation == 'User Error',])` or instances the user entered data in a format beyond the capabilities of BL to fix.
-BL returned an incorrect spell check match for  `r nrow(veg_diff[veg_diff$Situation == 'Incorrect',])` records, two of which had the same misspelled species, in `r nrow(veg_diff[veg_diff$Situation == 'Internal_Error',])` instance BL had an internal errors where it returned an incorrect match due to abbreviations.
+`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values.
 
@@
 
-While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate (@patten2024geographic).
+While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate [@patten2024geographic].
 We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format.
@@
 Anecdotally, using its default settings, it can produce more aesthetically pleasing results than typical word processors. 
-Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn)) project.
+Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project.
 
@@
 We thank the two anonymous peer reviewers who have increased the quality of this manuscript.
-Sofia Garcia is acknowledged for creating the 'Valleys' data set which place naming in the package relies on. 
-Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz & Dakota Becerra, Hannah Lovell and, Caitlin Miller & Hubert Szczygiel. 
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota Becerra, Caitlin Miller, Hubert Szczygiel. 
 We thank Rosalind Rowe for providing useful feedback during the later developmental stages of the program.
@@
 The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
-The package includes three real use-case vignettes (tutorials) available on a Github page site (https://sagesteppe.github.io/BarnebyLives/).
-One vignette *"setting_up_files"* explores setting up a instance for a certain geographic area.  
-Another vignette *"running_pipeline"* showcases the usage of the package for processing data entered on a spreadsheet.  
-The final vignette *"creating_labels"* shows the usage of an R and Bash script launched from RStudio to produce print-ready labels. 
+The package includes three real use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
+The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain). 
+The next two vignettes *"BarnebyLives!
+Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* how to export data in both digital and analog formats.
+*"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium.
 All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.
@@
 
+\clearpage
 <div id="refs"></div>

### 2025-04-24 manuscript/CoverLetter.rmd
edits from printed

---
output:
  pdf_document: default
  word_document: default
---

\begin{flushleft}
Brianna Gross \newline
Editor in Chief \newline
Applications in Plant Sciences \newline
\end{flushleft}

\hfill April 24^th^, 2025

Dear Dr. Gross, 
\vspace{3em}

I am writing to propose an original software article titled *"BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets"*

I have no conflicts of interest to disclose.

### 2025-04-24 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
function description cleaned up again

@@
 author:  |
+    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
     |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
-    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
 abstract: |
@@
 
-*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.  
 *coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data.
@@
 *physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
-site_writer will write directions from an officially named placename to the collection site.
-dms2dd is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).
+*political_grabber*  grabs a wide range of administrative information for a collection site. 
+*site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
 *directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages.
+*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).
+*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.
 
@@
 
-#subs_batch <- c(1384, 1388, 1394, 1396, 1397, 1400, 1435, 1438, 2161, 2187, 2268, 
-#                2301, 2302, 2577, 2746, 2785, 1460, 1461, 1379, 1381)
+subs_batch <- c(11207,  1263, 1326, 1384, 1388, 1394, 1396, 1397, 1400, 1435, 1437, 1438,
+                2161, 2301, 2517, 2746, 2785)
 
@@
 
-## HEY REED SPLIT OUT GENERA FOR THIS !!!!! AGAIN .... 
 data_p <- split_scientificName(data_p, sciName_col = 'Scientific_name')
-
 genera_mispelled <- data_p %>% 

### 2025-04-24 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
working on cover letter; but too real, will need to make old jerk curators the heros

@@
 The total label generation run time for processing these `r n_spp` collections was `r minutemen(sum(time_trials$elapsed))` minutes. 
-In total the `r n_spp` collections, which underwent all processing steps, took `minutemen(sum(time_trials$elapsed, time_label_gen, as.numeric(time_final)))` minutes to process.
+In total the `r n_spp` collections, which underwent all processing steps, took `r minutemen(sum(time_trials$elapsed, time_label_gen, as.numeric(time_final)))` minutes to process.
 

### 2025-04-24 manuscript/CoverLetter.rmd
working on cover letter; but too real, will need to make old jerk curators the heros

@@
 
+While a great variety of software tools and packages exist for interacting with accessioned herbarium data, far fewer are dedicated towards accessioning material. 
+And even fewer of these tools are intended for an actual collector to interact with. 
+This isn't terribly surprising, as the number of collectors who are not curators has decreased over the last few years, and the barrier to entry for both activities is generally a degree of eccentricity and particularity.
+Unfortunately, the particularity of curators often makes depositing specimens to herbaria a challenging task for collectors, especially those who work on large projects implemented by federal agencies, such as germplasm collection (e.g.
+Seeds of Success), ecological monitoring (e.g.
+National Wetland Condition Assessment, and Assess, Inventory, and Monitor), and have to satisfy the particularities of multiple curators simultaneously.
+
+Fortunately, I have been scolded (which on occasion has included some constructive criticism), by about 10 curators over the last decade and built up a wide knowledge of quirks to watch our for. 
+As a manager of young botanists for the last handful of years, I have insisted on maintaining a degree of separation between my employees and curators, in hopes that the small glimmer of hope which the youths have isn't extinguished.
+
 I have no conflicts of interest to disclose.

### 2025-04-25 manuscript/CoverLetter.rmd
part of letter to editor; v difficult to not be cheeky given why this actually exists...

--- 2025-04-24 manuscript/CoverLetter.rmd
@@
 
+While a variety of software tools exist for interacting with accessioned herbarium data, fewer are dedicated towards accessioning material. 
+And even fewer of these tools are intended for an actual collector to interact with. 
+This isn't surprising, as the number of collectors who are not curators has decreased, and the barrier to entry for both activities is a degree of eccentricity and particularity. 
+Unfortunately, the particularity of curators often makes depositing specimens to herbaria a challenging task for collectors. 
+A common challenge for collectors, especially those who work on large projects implemented by federal agencies, such as germplasm collection (e.g.
+Seeds of Success), ecological monitoring (e.g.
+National Wetland Condition Assessment, and Assess, Inventory, and Monitor), are the needs to satisfy the particularities of multiple curators simultaneously.
+
+Fortunately, I have been scolded by many curators over the last decade (and on occasion even received constructive criticism), and built up a wide knowledge of quirks to cater to. 
+As a manager of young botanists for the last handful of years, I have insisted on maintaining a degree of separation between my employees and curators, in hopes that the small glimmer of hope which the youths have for the profession isn't extinguished.
+To assist myself in running quality control checks, for both mine and their collections, I developed a wind range of functions which help meet a broad set of interests; in fact the package with all of these tools have been rolled into - BarnebyLives - has 30 user facing functions, at least another 10 internals, and 8 data sets.
+The package is used to both access commonly mistaken data, perform an array of spell checks, offer botanists more recently applied scientific names, and export data and write labels for them. 
+Curators across roughly 15 institutes have been very impressed by the results, while one or two may just dislike me as individual.
+
 I have no conflicts of interest to disclose.
+
+\newpage
+
+Reviewers:
+
+Shelly Gaynor mlgaynor@umich.edu 
+R package author extraordinaire, multiple thorough publications with APPS for software notes, and (former?) member of iDigBio team.
+
+Ben Legler blegler@uwyo.edu 
+Database administrator and creator of two large herbaria consortia, field botanist extraordinaire.
+
+Jason Alexander jason_alexander@berkeley.edu 
+I don't actually know Jason, word from our mutuals is he knows his stuff and this should be within his wheelhouse.
+
+Charles Davies cdavis@oeb.harvard.edu 
+Yeah, this one is a bit of a stretch, but I need to suggest four reviewers. 
+I doubt he has time for this, but maybe one of his graduate students would be interested. 
+Reasoning: general using herbaria specimens in ecology at global scale perspectives, as well as active taxonomist.

### 2025-04-25 manuscript/CoverLetter.rmd
letter to editor

--- 2025-04-24 manuscript/CoverLetter.rmd
@@
 
-I am writing to propose an original software article titled *"BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets"*
+I am writing to propose an original software article titled *"BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets"*. 
+Please bear with me when I cheekily explain it's impetus for existence, otherwise it's utility may seem opaque.
+
+While a variety of software tools exist for interacting with accessioned herbarium data, few are dedicated to accessioning new material. 
+And even fewer of these tools are intended for an actual collector to use.  
+This isn't surprising, as the number of collectors who are not curators has decreased, and the barrier to entry for both activities is a degree of particularity and eccentricity. 
+Unfortunately, the particularity of curators often makes contributing to herbaria a challenging task for collectors. 
+A common challenge for collectors, especially those who work on large projects implemented by federal agencies, such as germplasm collection (e.g.
+SOS) and ecological monitoring (e.g.
+NWCA, and AIM), is the need to simultaneously satisfy the particularities of multiple curators under tight deadlines.
+
+Fortunately, I have been scolded by many curators over the last decade, on occasion even receiving constructive criticism, and have built up a wide knowledge of quirks to cater to. 
+To assist in running quality control checks, for both my - and the young botanists whom I manage - collections, I developed a wide range of functions which help meet a broad set of interests.
+In fact 'BarnebyLives' the package which integrates these tools, has 30 user facing functions,  10+ internal functions, and 8 built in data sets.
+
+The package is used to both evaluate commonly mistaken data, perform an array of spell checks, offer botanists more recently applied scientific names, and export data and write labels for them. 
+Curators across roughly 15 institutes have been satisfied by the results while one curator just seems to just dislike me as individual. 
+While I note it as a an R package, it also creeps into LaTeX and some bash utilities, which are 99% restricted to the RStudio environment, making it slightly outside the realm of a typical R Package; we could probably render labels through html, but why bother to go through all this work to create labels which look terrible?
+
+I think that APPS is a wonderful place for this paper to be published, and APPS would be happy to have this paper. 
+Several editors of APPS have been unrelenting in their support of herbaria, and that herbaria maintain their role as the center of academic and field botany. 
+We believe that on occasion herbaria are not seen as resources for the future, but rather catalogues of the past, and would like to make our small contribution to ensure that does not become true.
 
 I have no conflicts of interest to disclose.
+
+\newpage
+
+Reviewers:
+
+Shelly Gaynor mlgaynor@umich.edu 
+R package author extraordinaire, including multiple documented in APPS software notes, and (former?) member of iDigBio team.
+
+Ben Legler blegler@uwyo.edu 
+Database administrator and creator of two large herbaria consortia websites, and a *bona fide* field botanist.
+
+Jason Alexander jason_alexander@berkeley.edu 
+I don't actually know Jason, word from our mutuals is he knows his stuff and this should be within his wheelhouse.
+
+Charles Davies cdavis@oeb.harvard.edu 
+A bit of a stretch, but I need to suggest four reviewers. 
+I doubt he has time for this, but maybe one of his graduate students would be interested. 
+Reasoning: Charles is a profuse user of herbarium specimens for both global scale ecology studies, localized natural history, and systematics.

### 2025-04-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
part of letter to editor; v difficult to not be cheeky given why this actually exists...
letter to editor
formatting

@@
 
-subs_batch <- c(11207,  1263, 1326, 1384, 1388, 1394, 1396, 1397, 1400, 1435, 1437, 1438,
-                2161, 2301, 2517, 2746, 2785)
+subs_batch <- c(1586, 1880, 2187, 2159)
 
@@
 
+![A label generated from a default template](../graphics/plots/RCB2759.png){width=50%}
+
 LaTeX offers well-documented and detailed functionality to customize labels for individuals and institutions. 
-Anecdotally, using its default settings, it can produce more aesthetically pleasing results than typical word processors. 
+Anecdotally, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
 Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project.
@@
 The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
+Thematically, it's area of interest is Western North America.
 
@@
 # ACKNOWLEDGEMENTS 
-The Bureau of Land Management is gratefully acknowledged as a provider of funding to R. C. B. for most of his specimen collection activities. 
+The Bureau of Land Management is gratefully acknowledged as a provider of funding to R.C.B. for most of his specimen collection activities. 
 We thank the two anonymous peer reviewers who have increased the quality of this manuscript.
 Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota Becerra, Caitlin Miller, Hubert Szczygiel.
-We thank Rosalind Rowe for providing useful feedback during the later developmental stages of the program.
 

### 2025-04-25 manuscript/CoverLetter.rmd
formatting

@@
 And even fewer of these tools are intended for an actual collector to use. 
-This isn't surprising, as the number of collectors who are not curators has decreased, and the barrier to entry for both activities is a degree of particularity and eccentricity. 
+Which isn't surprising, as the number of collectors who are not curators has decreased, and the barrier to entry for both activities is a high degree of particularity and eccentricity. 
 Unfortunately, the particularity of curators often makes contributing to herbaria a challenging task for collectors. 
@@
 
-Fortunately, I have been scolded by many curators over the last decade, on occasion even receiving constructive criticism, and have built up a wide knowledge of quirks to cater to. 
+Fortunately, I have been scolded by many curators over the last decade, and on occasion received actionable criticism, and have built up a wide knowledge of quirks to cater to. 
 To assist in running quality control checks, for both my - and the young botanists whom I manage - collections, I developed a wide range of functions which help meet a broad set of interests.
@@
 
-\newpage
-
-Reviewers:
-
-Shelly Gaynor mlgaynor@umich.edu 
-R package author extraordinaire, including multiple documented in APPS software notes, and (former?) member of iDigBio team.
-
-Ben Legler blegler@uwyo.edu 
-Database administrator and creator of two large herbaria consortia websites, and a *bona fide* field botanist.
-
-Jason Alexander jason_alexander@berkeley.edu 
-I don't actually know Jason, word from our mutuals is he knows his stuff and this should be within his wheelhouse.
-
-Charles Davies cdavis@oeb.harvard.edu 
-A bit of a stretch, but I need to suggest four reviewers. 
-I doubt he has time for this, but maybe one of his graduate students would be interested. 
-Reasoning: Charles is a profuse user of herbarium specimens for both global scale ecology studies, localized natural history, and systematics.
+Sincerely,  
+Reed

### 2025-04-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
cleaning up manuscript

@@
 Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
-These factors ensure that the specimens remain the premier botanical data source.
+These factors ensure that the specimens remain the premier botanical data source into perpetuity.
 
@@
 We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observed that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
-Evidently, apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
+Evidently apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
@@
 Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
-Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora].
-Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
+Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora].
+Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting. 
 Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization.
@@
 
-As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
-It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
+As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
+BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S.
 Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
@@
 
-Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
+Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota].
 The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
 PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
-However, it does not include data cleaning functionalities beyond verifying nations of collection. 
-While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
+However, beyond verifying the nations of collection it does not include data cleaning functionalities.  
+While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
 The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
@@
 Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github Pages (https://sagesteppe.github.io/BarnebyLives/) and around 100 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
@@
 at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
+Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
 The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g.
@@
 While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
-Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.
+Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.
 
@@
 The spatial module has five required functions and two optional functions.  
-
+*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.  
 *coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data. 
-political_grabber determines many levels of administrative ownership, including land management and public land survey system sections.
+*political_grabber* determines many levels of administrative ownership, including land management and public land survey system sections.  
 *physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
-*political_grabber*  grabs a wide range of administrative information for a collection site. 
 *site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
@@
 *dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).
-*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.
+
+Please note that the function *physical_grabber* is the one portion of the package where a decoupling may exist between the collection site, and the resolution of the spatial data. 
+While we expect the mismatch to be negligible for all effective purposes relating to: elevation, major geology type, and in general aspect, estimates of slope at this resolution may be biased - generally to lower angles.
+For these reasons collectors must always make notes on the truly local environment which taxa are found in, and consider that the notes from BL reflect the greater landscape which a microfeature may be present in.
+While this mis-match will seldom effect landscape ecologists, it may have implications for other data users.
 
@@
 
-The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
+The formatting module has three functions.Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
+*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
 *associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks.
 *field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
-*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.
 
@@
 *labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
+*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
+*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.
 
@@
 
-The package was released into beta testing using the primary authors collections from 2023. 
-The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
-Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
-Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
-In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.
+The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
+Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
+In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
+In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.
 
@@
 
-BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online for preferred synonyms, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
 
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)`records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)` records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance where an outdated circumscription was applied.
 At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
 
-Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
-Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
-In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
+Of `r n_spp` distinct species analysed BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling', na.rm = TRUE)`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling',  na.rm = TRUE)` unique species).
+In `r sum(epithet_diff$Situation=='Outdated', na.rm = TRUE)` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
 It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
@@
 
-While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate (@patten2024geographic).
+While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
 We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
@@
 Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
-While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
 While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
-
-While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate [@patten2024geographic].
-We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format.
-By utilizing both R and LaTeX and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications to project branches.
-
-Accessioning often relies on the use of the Office Suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
-While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats.
 

### 2025-04-28 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
workflow image slightly more legible

@@
 
-The formatting module has three functions.Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
+The formatting module has three functions. Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
 *date_parser* parses an input date into various formats for notating collection and determination dates on labels.
@@
 
-![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}
+![The spatial extent-or domain- (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map.png){width=50%}
 
@@
     case_when(
-      Genus %in% c('Glandularia', 'Oenothera', 'Chamerion') ~ 'Preferred',
       # check Pleuraphis, i am confused on this still....
-      Genus %in% c('Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia', 'Pleuraphis',
-                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
-                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
-                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
+      Collection_number %in% c(
+        '1231', '1243', '1245', '1248', '1251', '1263', '1299', '1363', '1373',
+        '1392', '1400', '1438', '1458', '1570', '1599', '1603', '1628', '2060',
+        '2094', '2185', '2233', '2249', '2249', '2258', '2264', '2291', '2421', 
+        '2424', '2426', '2473', '2538', '2545', '2569', '2581', '2598', '2620', 
+        '2638', '2684', '2686', '2697', '2715', '2806', '2810', '2812'
                     ) ~ 'Outdated',
-      Genus %in% c('Allium', 'Vaccinium', 'Eleocharis') ~ 'Internal_Error',
+      Collection_number %in% c('1236', '207', '2115', '2212', '2803') ~ 'Preferred', 
+      Collection_number %in% c('2435', '2517' ) ~ 'Internal_Error',
       .default = 'Typo'
@@
 While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
-We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
+We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume that responsibility in a relatively fast and streamlined format.
 By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
+
+LaTeX, a software system used for typesetting, allows users to focus on the content rather than the style of the documents rendered from it. 
+However, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
+Additionally LaTeX offers users a wide variety of ways which they can modify labels which are under-explored in the package. 
+Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project. 
+While the templates in the package are quite simple, LaTeX also offers the ability to use custom fonts, to alter font weights and colors, alter line spacing, to include images (e.g.
+dot maps) and customize labels beyond what the default templates support.
+
+Thematically, BarnebyLives is set up to cover Western North America. 
+However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
+Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e.
+the administrative information on Township Section and Range (or 'TRS'), but are considered less value in other geographic regions.
+Further several of the abiotic variables which it acquires information on: slope, aspect, and geology have long been considered prominent drivers of plant distributions in semi-arid and montane systems and warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous.
+Finally, it is plausible people in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the west we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.
 
@@
 
-LaTeX offers well-documented and detailed functionality to customize labels for individuals and institutions. 
-Anecdotally, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
-Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project.
-
 # CONCLUSIONS
@@
 The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
-Thematically, it's area of interest is Western North America.
 

### 2025-04-28 graphics/plots/prep_image.sh

@@ -3,4 +3,5 @@
 mogrify -trim *;
 convert -gravity center -append Spatmod.png Taxmod.png Formatmod.png Reviewmod.png Resultmod.png caption.png workflow.png; 
 convert workflow.png -fuzz 2% -transparent white workflow.png
+find . -name '*.png~' -delete
 exit

### 2025-04-28 manuscript/flowchart.R

@@ -1,15 +1,15 @@
 library(tidyverse)
 library(DiagrammeR)
 library(ggpubr)
-setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
-
+#setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
+setwd('/home/sagesteppe/Documents/assoRted/Barneby_Lives_dev/manuscript')
 
 grViz("
 
 digraph boxes_and_circle {
-  graph[rankdir = LR, ratio = 0.1]
+  graph[rankdir = LR, ratio = 0.2, nodesep = 0.01, ranksep = 0.01]
   
-  node [shape = circle, fontsize = 36, fontname = arial, style = filled, fontcolor = White]
+  node [shape = circle, fontsize = 72, fontname = arial, style = filled, fontcolor = White]
 
       subgraph cluster_import{color= 'white'
         node [color = 'black', shape = egg]
@@ -20,24 +20,25 @@ digraph boxes_and_circle {
       subgraph cluster_geospatial {style=dashed; color= 'white'; 
         node [color = '#94524A']
   
-          'Raw\nSpreadsheet' -> 'Convert\nCoordinates\nfrom\nDMS to DD' 
-              [label = '     dms2dd', fontsize = 50, penwidth = 6, style = dashed]
-          'Convert\nCoordinates\nfrom\nDMS to DD' -> 'Check for\nautofill\nvalues' 
-              [label = '     autofill_checker', fontsize = 50, penwidth = 6, style = dashed]
+          'Raw\nSpreadsheet' -> 'Convert\nfrom\nDMS to DD' 
+              [label = '     dms2dd', fontsize = 72, penwidth = 6, style = dashed]
+          'Convert\nfrom\nDMS to DD' -> 'Check for\nautofill\nvalues' 
+              [label = '     autofill_checker', fontsize = 72, penwidth = 6, style = dashed]
           'Check for\nautofill\nvalues' -> 'Geospatial\nData' 
-              [label = '     coords2sf', fontsize = 50, penwidth = 6]
+              [label = '     coords2sf', fontsize = 72, penwidth = 6]
           'Raw\nSpreadsheet' -> 'Check for\nautofill\nvalues' 
-              [label = '     autofill_checker', fontsize = 50, penwidth = 6]
+              [label = '     autofill_checker', fontsize = 72, penwidth = 6]
           'Geospatial\nData' -> 'Political\nBoundaries' 
[…]

### 2025-04-28 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
format things

@@
 Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks.
-However, the continual generation of herbarium specimens is essential for current and future research on ecology, evolution, and conservation.
-To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens.
+However, the continual accessioning of herbarium specimens is essential for current and future research on ecology, evolution, and conservation.
+To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens for collectors.
   **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-The Census Bureau for administrative jurisdictional data, and other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions and locality information.
+The Census Bureau, the United States Geological Survey, and other sources to supplement collection notes by providing information on administrative jurisdictions, abiotic site conditions and locality information.
 It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations.
@@
 Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].
+However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].
 
@@
 The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium].
-We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
+We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as electronic and remote sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid].
 While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images.
 Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
-These factors ensure that the specimens remain the premier botanical data source into perpetuity.
+These factors will ensure that the specimens remain the premier botanical data source into perpetuity.
 
 However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
-We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
+We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
 While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
-Evidently apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
@@
 For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
-In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error.
-Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
-Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora].
-Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting. 
+In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report administrative details in error.
+Even finding appropriate site names can rarely be resolved without a printed map, as many navigation-related software now consider most features that would serve as site names extraneous. 
+Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied scientific names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora].
+Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting which reduce the apparent credibility of a collector.
 Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. 
-Even if a collector navigates all these hurdles, the time allocated to each step is quite large.
+Even if a collector successfully navigates all these hurdles, the time allocated to each step is quite large, and may discourage them from further collecting.
 

### 2025-04-29 manuscript/BarnebyLives-An-R-package-to-create-and-qc-herbarium-labels.rmd
abstract was way too long, now under 150 words

--- 2025-04-28 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
   \noindent
-  **Premise:** Depositing specimens in herbaria is a time-consuming process.
-Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks.
-However, the continual accessioning of herbarium specimens is essential for current and future research on ecology, evolution, and conservation.
-To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens for collectors.
-  **Methods and Results:** BarnebyLives uses geospatial data from the U.S.
-The Census Bureau, the United States Geological Survey, and other sources to supplement collection notes by providing information on administrative jurisdictions, abiotic site conditions and locality information.
-It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations.
-Optionally, the package writes driving directions to the sites using Google Maps.
-The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.
-  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.
+  **Premise:** Accessioning herbarium specimens is labor-intensive, yet remains vital for research in ecology, evolution, and conservation.
+As institutional support for herbaria declines, efficient tools are needed to streamline this process.
+BarnebyLives was developed to assist collectors by supplementing collection notes, verifying taxonomic data, conducting quality checks, generating labels, and submitting digital records.
+  **Methods and Results:** It integrates geospatial data from U.S.
+government sources to provide jurisdictional and site information, and checks taxonomic names using in-house spell checkers, IPNI author standards, and Kew’s Plants of the World Online.
+Optional features include generating Google Maps driving directions.
+The tool outputs data in tabular and spatial formats for review before producing LaTeX-based labels and shipping manifests.
+  **Conclusions:** BarnebyLives improves data accuracy, ensures up-to-date taxonomy, and significantly reduces the time and effort required to accession herbarium specimens.        
 keywords: |

### 2025-04-29 manuscript/BarnebyLives-An-R-package-to-create-and-qc-herbarium-labels.tex
abstract was way too long, now under 150 words

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{endfloat}
\usepackage{setspace}\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage{rotating}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets},
  pdfkeywords={herbarium, natural history museum, collections,
geospatial, R, LaTeX},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{BarnebyLives: an R package to create herbarium specimen labels
and clean spreadsheets}
\author{Reed Clark Benkendorf\(^1\)\(^,\)\(^2\)\footnote{Author for
  Correspondence:
  \href{mailto:rbenkendorf@chicagobotanic.org}{\nolinkurl{rbenkendorf@chicagobotanic.org}}},
Jeremie B. Fant\(^1\)\(^,\)\(^2\)\\
\strut ~\(^1\)Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe,
Illinois 60022, USA\\
\strut ~\(^2\)Plant Biology and Conservation, Northwestern University,
Evanston, Illinois 60208, USA}
\date{}

\begin{document}
\maketitle
\begin{abstract}
\noindent \textbf{Premise:} Accessioning herbarium specimens is
labor-intensive, yet remains vital for research in ecology, evolution,
and conservation. As institutional support for herbaria declines,
efficient tools are needed to streamline this process. BarnebyLives was
developed to assist collectors by supplementing collection notes,
verifying taxonomic data, conducting quality checks, generating labels,
and submitting digital records. \textbf{Methods and Results:} It
integrates geospatial data from U.S. government sources to provide
jurisdictional and site information, and checks taxonomic names using
in-house spell checkers, IPNI author standards, and Kew's Plants of the
World Online. Optional features include generating Google Maps driving
directions. The tool outputs data in tabular and spatial formats for
review before producing LaTeX-based labels and shipping manifests.\\
\textbf{Conclusions:} BarnebyLives improves data accuracy, ensures
up-to-date taxonomy, and significantly reduces the time and effort
required to accession herbarium specimens.
\end{abstract}

Nearly 400 million specimens are housed worldwide in herbaria
(\citeproc{ref-thiers2021herbaria}{Thiers, 2021}). However, The rate of
accessioning new collections to herbaria diminished in the
20\textsuperscript{th} century as priorities in biology shifted away
from describing and documenting earths biodiversity and towards
understanding cellular and molecular processes underpinning life
(\citeproc{ref-prather2004decline}{Prather et al., 2004};
\citeproc{ref-pyke2010biological}{Pyke and Ehrlich, 2010};
\citeproc{ref-daru2018widespread}{Daru et al., 2018}). This shift, among
other factors, led to a decline in the funding allocated to
collection-based research, the number of staff maintaining and accessing
new collections, and educating students in these practices
(\citeproc{ref-funk2014erosion}{Funk, 2014}). Historically, specimens
have been used to describe the taxonomic diversity of plants and
document global floristic diversity
(\citeproc{ref-greve2016realising}{Greve et al., 2016};
\citeproc{ref-james2018herbarium}{James et al., 2018};
\citeproc{ref-brewer2019factors}{Brewer et al., 2019};
\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020}). However,
renewed interest in herbarium collections utilizing `big data
approaches,' such as museuomics, has brought herbaria back to the
forefront of the natural sciences and grearly expanded their roles in
science (\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020};
\citeproc{ref-marsico2020small}{Marsico et al., 2020}).

Innovations in specimen digitization, data sharing, computing, DNA
sequencing, and statistics have perhaps brought about greater use of
herbarium specimens than ever before
(\citeproc{ref-greve2016realising}{Greve et al., 2016};
\citeproc{ref-james2018herbarium}{James et al., 2018};
\citeproc{ref-brewer2019factors}{Brewer et al., 2019};
\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020}). The
current use of specimens and their ancillary data extends well beyond
their traditional roles in systematics and floristics, and studies
utilizing collections are regularly carried out to better understand the
ecological niches, phenological processes, and interactions of plants
(\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020};
\citeproc{ref-davis2023herbarium}{Davis, 2023}). We suspect that
collections are yet to realize their full potential, and as currently
novel approaches, such as electronic and remote sensing and
meta-barcoding, become more accessible the use of collections will
increase (\citeproc{ref-tosa2021rapid}{Tosa et al., 2021}). While
image-based or purely observational (rather than collection-based)
citizen science approaches (e.g., iNaturalist, BudBurst) have recently
dovetailed with herbarium specimens to meet many current research needs,
specimens contain rich data that are not accessible via images. Only
specimens have the ability to: provide samples of DNA, secondary
metabolites, or proteins, material for measuring (micro-)morphological
attributes (\citeproc{ref-borges2020schrodinger}{Borges et al., 2020}),
and seeds or pollen. These factors will ensure that the specimens remain
the premier botanical data source into perpetuity.

However, despite renewed recognition of the utility of collections,
efforts to grow them appear slow
(\citeproc{ref-prather2004decline}{Prather et al., 2004}). We conjecture
that this is partly because collecting and depositing specimens is a
fundamentally slower process, especially for novice collectors, relative
to taking photographs via commercially developed apps on smartphones
(\citeproc{ref-daru2018widespread}{Daru et al., 2018};
\citeproc{ref-mishler2020spatial}{Mishler et al., 2020};
\citeproc{ref-manzano2021fair}{Manzano and Julier, 2021}). While many
young botanists are capable of using dichotomous keys and other
resources to reliably identify and collect satisfactory material, we
observe that they face difficulties navigating several aspects of data
acquisition, processing, and preparation of labels for submission to
herbaria. Some of the apparent problems include the lack of dedicated
time at the end of a field season to process specimens, a general lack
of education on cartography and orienteering, natural history (e.g.,
geology, geomorphology), nomenclature, and familiarity with various
computer programs (for example, Microsoft Office suite), and increasing
foundational knowledge of plant systematics and phylogenetics
(\citeproc{ref-woodland2007botanists}{Woodland, 2007};
\citeproc{ref-barrows2016crossroads}{Barrows et al., 2016};
\citeproc{ref-nanglu2023nature}{Nanglu et al., 2023}).

The generation of an herbarium specimen involves many steps that are
easy to take for granted (\citeproc{ref-forman1989herbarium}{Forman and
Bridson, 1989}). For example, while acquiring appropriate political
information for a collection site appears simple, young collectors
rarely have adequate cartographic resources (printed topographic maps or
GIS software) at their disposal. In topographically complex areas, where
administrative borders are often associated with hydrological basins and
the ridges defining them, collectors are liable to misinterpret their
true geographic position and report administrative details in error.
Even finding appropriate site names can rarely be resolved without a
printed map, as many navigation-related software now consider most
features that would serve as site names extraneous. Similarly, the rate
at which taxonomic innovations occur, the volume of the literature, and
the reluctance of some regional curators to embrace a phylogenetic
approach to plant classification have made it difficult to find more
recently applied scientific names, even when these names are unanimously
accepted by taxonomic specialists in the group and other regional
curators (\citeproc{ref-hitchcock2018flora}{Hitchcock and Cronquist,
2018}). Furthermore, formatting a label correctly (e.g., author
abbreviations, italicization, etc.) is a time-consuming process with
many opportunities to introduce errors in formatting which reduce the
apparent credibility of a collector. Anecdotally, many mail merge
templates offered by herbaria still require collectors to modify many
variables by hand, for example, applying italicization. Even if a
collector successfully navigates all these hurdles, the time allocated
to each step is quite large, and may discourage them from further
collecting.

As a result of these concerns, we have developed an R package,
BarnebyLives, that aims to increase both the quality of data rendered to
labels and recorded in databases and to speed up the generation of
labels. BarnebyLives rapidly provides political and administrative
boundary information for a collection site using data from the U.S.
Census Bureau (\citeproc{ref-walker2022tigris}{Walker, 2024}), the
Public Land Survey System (PLSS), and ownership details of public lands
via the Protected-Areas Database (PAD-US)
(\citeproc{ref-usgs2024padus}{Gap Analysis Project (GAP), 2024}). Site
names are suggested by finding the closest unambiguously named place
feature in the Geographic Name Information System (GNIS) and the precise
calculation of distance and azimuth from this feature to the collection
site (\citeproc{ref-gnis2024}{Survey, 2023}). Using the Global Mountain
Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized
named mountain data set with global coverage allows for a relevant
descriptor of the general region with less ambiguity
(\citeproc{ref-snethlage2022hierarchical}{Snethlage et al., 2022}).
Spell checks on all scientific names (including associated species) are
performed using a copy of the World Checklist of Vascular Plants, and
the resolved species may be searched via Kew's Plant of the World Online
for relevant synonyms (\citeproc{ref-govaerts2021world}{Govaerts et al.,
2021}; \citeproc{ref-powo2024}{POWO, 2024}). Author abbreviations are
verified using the International Plant Names Index (IPNI) Standard
Author Abbreviation Checklist and also returned by Kew's Plants of the
World Online to ensure proper abbreviations of authorities
(\citeproc{ref-ipni2024}{The Royal Botanic Gardens and Herbarium, 2024};
\citeproc{ref-powo2024}{POWO, 2024}). Checks to search for and flag
common issues associated with spreadsheet software or data
transcription, such as the auto-filling of coordinate and date columns.
After a final review of the data, flagged or generated by the package,
it allows for the option to export spreadsheets that are suitable for
mass uploading of data to multiple common herbarium databases as well as
the generation of herbarium labels.

Currently, to our knowledge label generation functionality is provided
explicitly by two programs, PLabel and Symbiota, and by the Microsoft
Word tool Mail Merge (\citeproc{ref-gries2014symbiota}{Gries et al.,
2014}; \citeproc{ref-perkins2020Plabel}{Perkins, 2020}). The office
suite costs money, and in our experience, is finicky; further, its
functionality ends with label creation. PLabel is a standalone program
that has greatly enhanced functionality relative to a mail merge,
allowing users to specify the layout and formatting of label components
using an intuitive and local graphical user interface (GUI)
functionality. However, beyond verifying the nations of collection it
does not include data cleaning functionalities.\\
While some sources indicate that it can only be used on Microsoft, we
expect it to be usable on Linux and Mac using Windows `emulators' like
Wine. The increasingly popular Symbiota biodiversity data management
software not only provides label generation capabilities but also
provides data cleaning functionality in an attractive GUI web portal
allowing for live management of collections and bypassing the need for a
local installation, allowing it to be accessed on all operating systems.
Symbiota offers functionality similar to the first four of our five
stages of our `Taxonomic' module and to our knowledge a check of the
`Political Boundaries' (see Figure 1). However, not all herbaria use
Symbiota and many have original database systems that they maintain (for
example, Harvard University Herbarium,
\url{https://kiki.huh.harvard.edu/databases/specimen_index.html};
Missouri Botanical Garden \url{https://tropicos.org/specimen/Search};
and The Consortium of Pacific Northwest Herbaria
\url{https://www.pnwherbaria.org/}). However, and most importantly many
collectors prefer to generate their own labels, especially as they are
likely to send different sets of collections to different institutions.
Accordingly, the functionality of Symbiota should exist in an ecosystem
with alternative systems. In scenarios where users want to keep
rendering labels in either of the three existing alternatives, they can
easily export data in the appropriate formats after utilizing BLs data
cleaning utilities.

BarnebyLives was named for plant taxonomist Rupert Charles Barneby
(1911-2000), who published over 6,500 pages of text, described over 750
taxa, and is notable for balancing his studies at the William and Lynda
Steere Herbarium at the New York Botanical Garden with annual collection
trips in Western North America from 1937-1970 and sporadically until he
passed in 2000 (\citeproc{ref-welsh2001rupert}{Welsh, 2001}). Select
accolades of Rupert include the 1989 Asa Gray Award from the American
Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from
the International Association of Plant Taxonomists (IAPT), as well as
being one of eight recipients of the International Botanical Congress's
(IBC) Millennium Botany Award (1999)
(\citeproc{ref-welsh2001rupert}{Welsh, 2001}). Most germanely, Rupert
was remembered as being generous with his time to assist younger
botanists with the more arcane aspects of field botany and taxonomy
(\citeproc{ref-holmgren2017}{Holmgren and Holmgren, 1988}).

\section{METHODS AND RESULTS}\label{methods-and-results}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{../graphics/plots/workflow.png}
\caption{Recommended workflow}
\end{figure}

BarnebyLives was iteratively developed based on data submitted by
approximately 20 seasonal field botany teams over two years.
Essentially, continual updates were made as the developers became aware
of the idiosyncrasies of collection notes and data entry. Several
commands in BarnebyLives require output from previous functions, and a
workflow that satisfies these requirements is presented in Figure 1.

\subsection{Usage}\label{usage}

All steps of BarnebyLives, except for label generation are run within
the freely available RStudio. Data may be read from any common
spreadsheet management system or database connection such as Excel, or
free alternatives such as LibreOffice, OpenOffice, or via the cloud on
Google Sheets. The latter two options are documented here and in package
vignettes, detailed descriptions of the required and suggested input
columns are located on a Github Pages
(\url{https://sagesteppe.github.io/BarnebyLives/}) and around 100
real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical for R packages in that it requires a
considerable amount of data to operate (Table 1). Virtually all on-disk
memory associated with the package are used to store spatial data. The
amount of spatial data varies according to the domain that the user
decides to support (Figure 3). Functions that require on-disk data
require a path to data as an argument. Manually supplying the path
argument allows users to determine an appropriate storage location
suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less
than a couple gigabytes of memory (ours covering all of the conterminous
Western U.S. at 3-arc second (\textasciitilde90m) resolution is
\textasciitilde16 GiB), while the processing requires relatively little
RAM; hence, we believe installations can work on hardware as limited as
Chromebooks, while having the data stored entirely on thumb-drives.
Given that the attributes which the package collects data on are
tailored to the Western U.S. region, we do not expect local installs to
exceed the size of ours. The final steps of BarnebyLives, generating the
labels, requires working installations of R Markdown, a LaTeX
installation
(e.g.~\href{https://www.tug.org/applications/pdftex/}{pdfTeX},
\href{https://www.luatex.org/}{LuaTeX},
\href{https://www.overleaf.com/learn/latex/XeLaTeX}{XeLaTeX}), and the
open source command line tools
\href{https://github.com/rrthomas/pdfjam}{pdfjam} and
\href{https://linux.die.net/man/1/pdftk}{pdftk}. While these steps are
run through a shell scripting language such as bash, we have wrapped
them in R functions that bypass the need to enter the commands directly
into a shell terminal outside of RStudio. Unfortunately, we have not
found Windows alternatives to pdfam and pdftek, so we are unable to
offer the final label-generating functionality on that operating system,
but suspect Ubuntu subsystem for Windows may allow for integration of
these tools.

\subsection{Functionality}\label{functionality}

BarnebyLives can be thought of as consisting of five main modules
(Figure 1): spatial, taxonomic, formatting, manual review, and data
exporting.

The spatial module has five required functions and two optional
functions.\\
\emph{autofill\_checker} searches for patterns in the input latitude and
longitude data associated with autofilling from various spreadsheet
programs and will emit a warning if they are encountered.\\
\emph{coords2sf} creates a spatially explicit simple feature (sf)
geometry dataset for the input data. \emph{political\_grabber}
determines many levels of administrative ownership, including land
management and public land survey system sections.\\
\emph{physical\_grabber} provides various geographic data, such as
elevation, landform position, and aspect using 90m resolution spatial
data.\\
\emph{site\_writer} write distance and azimuth to collection site from
the nearest official named place from the GNIS database.
\emph{directions\_grabber} is an optional function that writes driving
directions from a reasonably sized town to the closest drivable area to
the site using the Google Maps API, which will require a valid Google
account that is free per month for most personal and smaller academic
usages.\\
\emph{dms2dd} is an optional function used to convert from coordinates
denoted in the degrees minutes and second format (for example,
42°08'39.9"N 87°47'08.3"W) to decimal degree format (for example
42.14439, -87.78569).

Please note that the function \emph{physical\_grabber} is the one
portion of the package where a decoupling may exist between the
collection site, and the resolution of the spatial data. While we expect
the mismatch to be negligible for all effective purposes relating to:
elevation, major geology type, and in general aspect, estimates of slope
at this resolution may be biased - generally to lower angles. For these
reasons collectors must always make notes on the truly local environment
which taxa are found in, and consider that the notes from BL reflect the
greater landscape which a microfeature may be present in. While this
mis-match will seldom effect landscape ecologists, it may have
implications for other data users.

The taxonomic module has four required functions and one optional
function.\\
\emph{spell\_check} will perform a spell check on the entered scientific
name based on a local copy of Kew Plants of the World database filtered
to the local continents or a user-specified backbone.\\
\emph{spell\_check\_family} performs a spell check on the family entered
for each scientific name.\\
\emph{author\_check} ensures that the authors are entered in a valid
format, for example, the correct standard abbreviations are used.\\
\emph{associates\_check} performs a spell check on all associated
species using the local taxonomic database. \emph{powo\_searcher} can be
used in tandem with the functions \emph{spell\_check\_family} and
\emph{author\_check}, but we use it in lieu of them to search the
current Plants of the World Online to determine relevant synonyms and
alternative higher taxonomy for the focal species. No API key or
registration is required to use \emph{powo\_searcher}.

The formatting module has three functions. Two are optional; however,
they are run locally and so quickly that there is no reason to skip
them. \emph{date\_parser} parses an input date into various formats for
notating collection and determination dates on labels.
\emph{associate\_dropper} silently removes the collected species from
the list of associated species; however, it searches for the species to
be removed using the scientific name entered initially by the user
rather than returned via spell checks. \emph{field\_lengths} will emit
messages for any fields that we suspect will create an `overflow' on the
physical label and should be truncated for clarity.

The manual review process technically only has one function that is
optional and may be executed during the spatial process (after
\emph{coords2sf}), but the importance of manual review is important
enough to warrant explicit mention.\\
\emph{geodata\_writer} will write out a spatial copy of the data set to
any geospatial format supported by the sf package, but defaults to
writing out `kmls' which are readily used with
\href{https://earth.google.com/web/}{Google Earth}, and can also be
opened in several other free geographic information system (GIS)
softwares such as \href{https://www.qgis.org/}{QGIS}. Notably, many of
the flags that BarnebyLives generates will be placed into columns with
obviously flagged names and can be manually reviewed by the analyst, and
many of these issues can be resolved by simply addressing the relevant
issues in the original data input spreadsheet.

The data exporting module contains three functions that interact with
LaTeX templates and require slightly more advanced R user interactivity,
such as setting up mapping functions using the tidyverses purrr
package.\\
\emph{labels\_skeleton} is an R `script' which will require a few
modification steps to tailor to each institution, these R scripts will
put data into a user specified template, and serve as the interface to
LaTeX.\\
\emph{label\_writer} write from a flatfile or spreadsheet to small 4x4
inch herbarium labels (users can modify these dimensions as they see
fit). \emph{format\_database\_import} will write out a spreadsheet of
cleaned data in a variety of formats, currently: Jepson, Symbiota, and
Consortium of Pacific Northwest herbaria are supported.

\subsubsection{Herbarium Collections}\label{herbarium-collections}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/plots/collections_map.png}
\caption{The spatial extent-or domain- (orange), and herbarium
collection sites (burgundy) tested in this manuscript.}
\end{figure}

The testing of the package within this manuscript was performed using a
subset of the authors collections from 2018-2022, while most development
was performed on their 2023 and 2024 collections. Only collections which
had identifications to the level of species or lower, and transcribed
collection dates and coordinates were used for most functionality. In
total 981 records were used for testing various functions, these records
were from 234 sites located across Western North America (Figure 2). In
total this data set had 729 species (with 558 distinct sets of authors),
with 83 infraspecies (22 authorships) in 74 families.

BarnebyLives took roughly four minutes (222.886sec) to run all local
steps, and roughly ten minutes (584.05sec) to search Plants of the World
Online for preferred synonyms, and a minute 63.945sec to search Google
Maps and write directions to sites.

Most of the local run time is attributable to the spatial (205.254sec),
and taxonomic operations (17.253sec), while formatting data for labels
took 0.379sec. The spell check of the scientific name accounted for
nearly all of the time (17.006sec) spent performing local taxonomic
operations. The generation of labels consumed around eight minutes
(509.931sec) for the rendering, and an additional 58.60sec to combine
the 182 sheets to a single Portable Document Format (PDF). The total
label generation run time for processing these 729 collections was 15
minutes. In total the 729 collections, which underwent all processing
steps, took 24 minutes to process.

\subsection{RESULTS}\label{results}

Even on data which had been manually cleaned and error-checked by a
human several times BarnebyLives was able to reduce transcription
errors, identify typos, make nomenclature suggestions, and reformat text
elements for downstream use. While none of the 74 families were
misspelled, BarnebyLives made 25 suggestions on naming, identified 6
instances where the user entered an unequivocally incorrect family (or
taxonomic entity), identified 5 records where families were autofilled,
and 1 instance where an outdated circumscription was applied. At the
level of family BarnebyLives flagged 6 records where the author follows
an alternative taxonomy, and flagged 7 records in error, it appears most
of these errors are due to issues in the backbone used by the earlier
spell check function.

In the 326 genera analysed BarnebyLives identified 75 discrepancies at
the level of genus between user submitted and processed data. In 42 of
these instances the user supplied an outdated name (21 unique genera)
flagged 4 records where the author follows an alternative taxonomy (2
genera total), and flagged 2 record in error.

Of 729 distinct species analysed BarnebyLives flagged 61 records, and
detected 33 instances of misspelled epithets (33 unique species). In 15
of these instances the user supplied an outdated name (15 unique
species). It also flagged 2 records where the author follows an
alternative taxonomy (2 unique species), and flagged 8 records in error.
The final record was an egregious error where the order of the specific
epithet and the genus name.

5 records were appropriately flagged for issues with auto fill increment
of the longitude value, and 3 records were also auto-flagged for
increases in latitude values.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/tables/Screenshot-DataSources.png}
\caption{Data Sources}
\end{figure}

\section{DISCUSSION}\label{discussion}

While numerous tools have been developed for cleaning existing herbarium
and museum records, few tools help to ensure that the data entered are
accurate (\citeproc{ref-patten2024geographic}{Patten et al., 2024}). We
argue that the original collectors are the most qualified individuals to
perform quality control checks and that BarnebyLives allows them to
assume that responsibility in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code
on Github, this program allows users immediate familiarity with the
system for troubleshooting issues and implementing upgrades and
modifications in project branches.

LaTeX, a software system used for typesetting, allows users to focus on
the content rather than the style of the documents rendered from it.
However, using its default settings, it can produce aesthetically
pleasing results (Figure 4). Additionally LaTeX offers users a wide
variety of ways which they can modify labels which are under-explored in
the package. Very good documentation of LaTeX capabilities is offered in
multiple areas; for instance, via the
\href{https://www.overleaf.com/learn}{Overleaf} project. While the
templates in the package are quite simple, LaTeX also offers the ability
to use custom fonts, to alter font weights and colors, alter line
spacing, to include images (e.g.~dot maps) and customize labels beyond
what the default templates support.

Thematically, BarnebyLives is set up to cover Western North America.
However, the package supports the use of a `domain' being drawn over any
of the conterminous United States. Several of the attributes which it
collects and displays on labels, relate to topics which more senior
curators are interested in, i.e.~the administrative information on
Township Section and Range (or `TRS'), but are considered less value in
other geographic regions.\\
Further several of the abiotic variables which it acquires information
on: slope, aspect, and geology have long been considered prominent
drivers of plant distributions in semi-arid and montane systems and
warranted on a label in these types of systems, whereas curators in
other regions may find this information superfluous. Finally, it is
plausible people in other geographic areas are less interested in
displaying which land management agency has jurisdiction over a
collection; however in the west we believe this is useful information
which may help a collector interested in revisiting a site to determine
if they will require permits for access or to make new collections.

Accessioning often relies on the use of the Microsoft Office suite of
programs and may utilize other costly software such as ArcPro or Adobe
Acrobat. While BarnebyLives does not have its own graphic user
interface, the functionality of commonly used Interactive Development
Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now
offer functionality to readily view and filter datasets using familiar
spreadsheet-like formats, making them more accessible to many users.
While other software often cost money, these are also free, and we
recommend that users install an open-source PDF viewer such as Okular to
review their rendered documents.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/plots/RCB2759.png}
\caption{A label generated from a default template}
\end{figure}

\section{CONCLUSIONS}\label{conclusions}

BarnebyLives is an R package that can be used to rapidly acquire
relevant geographic and taxonomic data. It can also perform specialized
spell checks and assorted curatorial tasks to produce both digital and
analog data. The package relies on no licensed software, such as the
Microsoft Office suite, and is suitable for install on all major
operating systems (Windows, Mac, Linux), however currently label
generation support is only offered on Linux and Mac, with a small amount
of use of the command line, which may be called from the Rstudio rather
than a `traditional' terminal.

\section{AUTHOR CONTRIBUTIONS}\label{author-contributions}

The project was conceptualized by R.C.B. The program was written by
R.C.B. Data collection and analysis were performed by R.C.B. R.C.B. \&
J.B.F wrote the manuscript, and both authors approved the final version
of the manuscript.

\section{ACKNOWLEDGEMENTS}\label{acknowledgements}

The Bureau of Land Management is gratefully acknowledged as a provider
of funding to R.C.B. for most of his specimen collection activities. We
thank the two anonymous peer reviewers who have increased the quality of
this manuscript. Several prominent associated collectors of specimens
used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota
Becerra, Caitlin Miller, Hubert Szczygiel.

\section{DATA AVAILABILITY STATEMENT}\label{data-availability-statement}

The BarnebyLives R package is open source, the development version is
available on GitHub (\url{https://github.com/sagesteppe/BarnebyLives}).
The package includes three real use-case vignettes (tutorials) available
on a Github Pages site
(\url{https://sagesteppe.github.io/BarnebyLives/}). The first vignette
\emph{``Preparing to use BarnebyLives!''} shows how to set up an
instance for a certain geographic area (domain). The next two vignettes
\emph{``BarnebyLives! Running pipeline''} showcases the usage of the
package for processing data entered on a spreadsheet, and
\emph{``Printing herbarium labels and exporting a digital copy of
data''} how to export data in both digital and analog formats.
\emph{``Custom label templates''} shows how to customize labels in
LaTeX, and \emph{``Rendering a shipping manifest''} details how to
produce a shipping manifest for gifting or transferring material to an
herbarium. All data used in this manuscript are available at:
\url{https://github.com/sagesteppe/Barneby_Lives_dev/manuscript}.

\singlespacing

\section{ORCID}\label{orcid}

Reed Clark Benkendorf \url{https://orcid.org/0000-0003-3110-6687}\\
Jeremie Fant \url{https://orcid.org/0000-0001-9276-1111}

\small

\section{REFERENCES}\label{references}

\clearpage

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-barrows2016crossroads}
Barrows, C. W., M. L. Murphy-Mariscal, and R. R. Hernandez. 2016. At a
crossroads: The nature of natural history in the twenty-first century.
\emph{BioScience} 66: 592--599.

\bibitem[\citeproctext]{ref-borges2020schrodinger}
Borges, L. M., V. C. Reis, and R. Izbicki. 2020. Schrodinger's
phenotypes: Herbarium specimens show two-dimensional images are both
good and (not so) bad sources of morphological data. \emph{Methods in
Ecology and Evolution} 11: 1296--1308.

\bibitem[\citeproctext]{ref-brewer2019factors}
Brewer, G. E., J. J. Clarkson, O. Maurin, A. R. Zuntini, V. Barber, S.
Bellot, N. Biggs, et al. 2019. Factors affecting targeted sequencing of
353 nuclear genes from herbarium specimens spanning the diversity of
angiosperms. \emph{Frontiers in plant science} 10: 1102.

\bibitem[\citeproctext]{ref-daru2018widespread}
Daru, B. H., D. S. Park, R. B. Primack, C. G. Willis, D. S. Barrington,
T. J. Whitfeld, T. G. Seidler, et al. 2018. Widespread sampling biases
in herbaria revealed from large-scale digitization. \emph{New
Phytologist} 217: 939--955.

\bibitem[\citeproctext]{ref-davis2023herbarium}
Davis, C. C. 2023. The herbarium of the future. \emph{Trends in Ecology
\& Evolution} 38: 412--423.

\bibitem[\citeproctext]{ref-forman1989herbarium}
Forman, L., and D. Bridson. 1989. The herbarium handbook. Royal Botanic
Gardens Kew.

\bibitem[\citeproctext]{ref-funk2014erosion}
Funk, V. A. 2014. The erosion of collections-based science: Alarming
trend or coincidence. \emph{The Plant Press} 17: 1--13.

\bibitem[\citeproctext]{ref-usgs2024padus}
Gap Analysis Project (GAP), U. S. G. S. (USGS). 2024. Protected areas
database of the united states (PAD-US) 4.0.

\bibitem[\citeproctext]{ref-govaerts2021world}
Govaerts, R., E. Nic Lughadha, N. Black, R. Turner, and A. Paton. 2021.
The world checklist of vascular plants, a continuously updated resource
for exploring global plant diversity. \emph{Scientific data} 8: 215.

\bibitem[\citeproctext]{ref-greve2016realising}
Greve, M., A. M. Lykke, C. W. Fagg, R. E. Gereau, G. P. Lewis, R.
Marchant, A. R. Marshall, et al. 2016. Realising the potential of
herbarium records for conservation biology. \emph{South African Journal
of Botany} 105: 317--323.

\bibitem[\citeproctext]{ref-gries2014symbiota}
Gries, C., M. E. E. Gilbert, and N. M. Franz. 2014. Symbiota--a virtual
platform for creating voucher-based biodiversity information
communities. \emph{Biodiversity data journal}.

\bibitem[\citeproctext]{ref-hitchcock2018flora}
Hitchcock, C. L., and A. Cronquist. 2018. Flora of the pacific
northwest: An illustrated manual. University of Washington Press.

\bibitem[\citeproctext]{ref-holmgren2017}
Holmgren, N., and P. Holmgren. 1988. Intermountain flora v. 7. The New
York Botanical Garden Press, New York.

\bibitem[\citeproctext]{ref-james2018herbarium}
James, S. A., P. S. Soltis, L. Belbin, A. D. Chapman, G. Nelson, D. L.
Paul, and M. Collins. 2018. Herbarium data: Global biodiversity and
societal botanical needs for novel research. \emph{Applications in plant
sciences} 6: e1024.

\bibitem[\citeproctext]{ref-manzano2021fair}
Manzano, S., and A. C. Julier. 2021. How FAIR are plant sciences in the
twenty-first century? The pressing need for reproducibility in plant
ecology and evolution. \emph{Proceedings of the Royal Society B} 288:
20202597.

\bibitem[\citeproctext]{ref-marsico2020small}
Marsico, T. D., E. R. Krimmel, J. R. Carter, E. L. Gillespie, P. D.
Lowe, R. McCauley, A. B. Morris, et al. 2020. Small herbaria contribute
unique biogeographic records to county, locality, and temporal scales.
\emph{American journal of botany} 107: 1577--1587.

\bibitem[\citeproctext]{ref-mishler2020spatial}
Mishler, B. D., R. Guralnick, P. S. Soltis, S. A. Smith, D. E. Soltis,
N. Barve, J. M. Allen, and S. W. Laffan. 2020. Spatial phylogenetics of
the north american flora. \emph{Journal of Systematics and Evolution}
58: 393--405.

\bibitem[\citeproctext]{ref-nanglu2023nature}
Nanglu, K., D. de Carle, T. M. Cullen, E. B. Anderson, S. Arif, R. A.
Castañeda, L. M. Chang, et al. 2023. The nature of science: The
fundamental role of natural history in ecology, evolution, conservation,
and education. \emph{Ecology and Evolution} 13: e10621.

\bibitem[\citeproctext]{ref-patten2024geographic}
Patten, N. N., M. L. Gaynor, D. E. Soltis, and P. S. Soltis. 2024.
Geographic and taxonomic occurrence r-based scrubbing (gatoRs): An r
package and workflow for processing biodiversity data.
\emph{Applications in Plant Sciences} 12: e11575.

\bibitem[\citeproctext]{ref-perkins2020Plabel}
Perkins, K. 2020. Plabel.

\bibitem[\citeproctext]{ref-powo2024}
POWO. 2024. Geographic names information system (GNIS) - USGS national
map downloadable data collection: U.s. Geological survey.

\bibitem[\citeproctext]{ref-prather2004decline}
Prather, L. A., O. Alvarez-Fuentes, M. H. Mayfield, and C. J. Ferguson.
2004. The decline of plant collecting in the united states: A threat to
the infrastructure of biodiversity studies. \emph{Systematic Botany} 29:
15--28.

\bibitem[\citeproctext]{ref-pyke2010biological}
Pyke, G. H., and P. R. Ehrlich. 2010. Biological collections and
ecological/environmental research: A review, some observations and a
look to the future. \emph{Biological reviews} 85: 247--266.

\bibitem[\citeproctext]{ref-ronsted2020integrative}
Rønsted, N., O. M. Grace, and M. A. Carine. 2020. Integrative and
translational uses of herbarium collections across time, space, and
species. \emph{Frontiers in Plant Science} 11: 1319.

\bibitem[\citeproctext]{ref-snethlage2022hierarchical}
Snethlage, M. A., J. Geschke, A. Ranipeta, W. Jetz, N. G. Yoccoz, C.
Körner, E. M. Spehn, et al. 2022. A hierarchical inventory of the
world's mountains for global comparative mountain science.
\emph{Scientific data} 9: 149.

\bibitem[\citeproctext]{ref-gnis2024}
Survey, U. S. G. 2023. Geographic names information system (GNIS) - USGS
national map downloadable data collection: U.s. Geological survey.

\bibitem[\citeproctext]{ref-ipni2024}
The Royal Botanic Gardens, H. U. H. \&. L., Kew, and A. N. Herbarium.
2024. International plant names index.

\bibitem[\citeproctext]{ref-thiers2021herbaria}
Thiers, B. M. 2021. The world's herbaria 2021: A summary report based on
data from index herbarium.

\bibitem[\citeproctext]{ref-tosa2021rapid}
Tosa, M. I., E. H. Dziedzic, C. L. Appel, J. Urbina, A. Massey, J.
Ruprecht, C. E. Eriksson, et al. 2021. The rapid rise of next-generation
natural history. \emph{Frontiers in Ecology and Evolution} 9: 698131.

\bibitem[\citeproctext]{ref-walker2022tigris}
Walker, K. 2024.
\href{https://CRAN.R-project.org/package=tigris}{Tigris: Load census
TIGER/line shapefiles}.

\bibitem[\citeproctext]{ref-welsh2001rupert}
Welsh, S. L. 2001. Rupert c. Barneby (1911-2000). \emph{Taxon}.

\bibitem[\citeproctext]{ref-woodland2007botanists}
Woodland, D. W. 2007. Are botanists becoming the dinosaurs of biology in
the 21st century? \emph{South African Journal of Botany} 73: 343--346.

\end{CSLReferences}

\section{SUPPORTING INFORMATION}\label{supporting-information}

Additional supporting information can be found online in the Supporting
Information section at the end of this article.

\textbf{Appendix S1.} A table of all time trials for each function.

\end{document}

### 2025-05-01 manuscript/BarnebyLives -COPY2 ACTUALLY CLEAN.rmd
script to clean up the problem records identified in manuscript to accession

--- 2025-04-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
 Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.
-These factors ensure that the specimens remain the premier botanical data source into perpetuity.
+These factors ensure that the specimens remain the premier botanical data source.
 
@@
 We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
-Evidently apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observed that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
+Evidently, apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
@@
 Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
-Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora].
-Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting. 
+Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora].
+Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
 Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization.
@@
 
-As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
-BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S.
+As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
+It rapidly provides political and administrative boundary information for a collection site using data from the U.S.
 Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
@@
 
-Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota].
+Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
 The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
 PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
-However, beyond verifying the nations of collection it does not include data cleaning functionalities.  
-While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
+However, it does not include data cleaning functionalities beyond verifying nations of collection. 
+While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
 The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
@@
 Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
-The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github Pages (https://sagesteppe.github.io/BarnebyLives/) and around 100 real-world examples are on a Google Sheets accessible from the page.
+The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page.
 BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
@@
 at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives.
-Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
 The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g.
@@
 While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
-Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.
+Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.
 
@@
 The spatial module has five required functions and two optional functions.
-*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.  
+
 *coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data.
-*political_grabber* determines many levels of administrative ownership, including land management and public land survey system sections.  
+political_grabber determines many levels of administrative ownership, including land management and public land survey system sections.
 *physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
+*political_grabber*  grabs a wide range of administrative information for a collection site. 
 *site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
@@
 *dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).
-
-Please note that the function *physical_grabber* is the one portion of the package where a decoupling may exist between the collection site, and the resolution of the spatial data. 
-While we expect the mismatch to be negligible for all effective purposes relating to: elevation, major geology type, and in general aspect, estimates of slope at this resolution may be biased - generally to lower angles.
-For these reasons collectors must always make notes on the truly local environment which taxa are found in, and consider that the notes from BL reflect the greater landscape which a microfeature may be present in.
-While this mis-match will seldom effect landscape ecologists, it may have implications for other data users.
+*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.
 
@@
 
-The formatting module has three functions.Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
-*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
+The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
 *associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks.
 *field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
+*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.
 
@@
 *labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
-*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
-*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.
 
@@
 
-subs_batch <- c(1586, 1880, 2187, 2159)
+subs_batch <- c(1207, 2818, 2747, 2748, 2760:2770, 2782:2784, 2435, 2517, 1424, 
+                2644, 2038, 2326, 2618)
 
@@
         c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
-  filter(Scientific_name != '')#, Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
-
-n_families <- data %>% 
-  group_by(Family) %>% 
-  count() %>% 
-  nrow() # 74 families
-
-n_genera <- data %>% 
-  separate(Scientific_name, into = c('Genus', 'Epithet') , remove = FALSE, extra = 'drop') %>% 
-  group_by(Genus) %>% 
-  count() %>% 
-  nrow() # 292 Genera
-
-n_spp <- data %>% 
-  group_by(Binomial) %>% 
-  count() %>% 
-  nrow() # 616 species
-
-n_infraspecies <- data %>% 
-  drop_na(Infraspecies) %>% 
-  group_by(Binomial, Infraspecies) %>% 
-  count() %>% 
-  nrow() # 66 distinct infraspecies
-
-n_sp_authors <- data %>% 
-  drop_na(Binomial_authority) %>% 
-  count() %>% # 557 groups of authors
-  pull()
-
-n_infra_sp_auths <- data %>% 
-  drop_na(Infraspecific_authority) %>% 
-  group_by(Binomial, Infraspecies) %>% 
-  count() %>% 
-  nrow() # 22 distinct infra species author groups
-
-n_vegetation_fields <- data %>% 
-  drop_na(Vegetation) %>% 
-  distinct(Vegetation) |>
-  nrow()
-
-n_associates_fields <- data %>% 
-  drop_na(Associates) %>% 
-  distinct(Associates) |>
-  nrow()
-
-# number of collection sites
-
-n_sites <- data %>% 
-  distinct(Latitude, Longitude) %>% 
-  nrow()
+  filter(Scientific_name != '', Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
 
@@
 
-The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
-Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
-In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
-In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.
+The package was released into beta testing using the primary authors collections from 2023. 
+The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
+Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
+Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
+In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.
 
@@
 
-saveRDS(time_powo_searcher, file = 'data/processed/time_powo_searcher')
-saveRDS(data, file = 'data/processed/data_w_POWO_search')
-
-time_directions_grabber <- system.time({ # write directions to sites
-  SoS_gkey = Sys.getenv("Sos_gkey")
-  data <- directions_grabber(data, api_key = SoS_gkey)
-})
-
-saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
-saveRDS(data, file = 'data/processed/data_w_Google_Maps')
-
 ```
@@
 # also save a spreadsheet of the cleaned data, we'll use this to create the labels. 
-write.csv(data, file = 'data/processed/cleaned_data.csv', row.names = FALSE)
+write.csv(data, file = 'data/processed/cleaned_data-problemchildren.csv', row.names = FALSE)
 ```
@@
 ```{r Total Time of operations, eval = F}
-
-time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')
-time_directions_grabber <- readRDS('data/processed/time_directions_grabber')
-
-time_trials <- data.frame(as.matrix(do.call(rbind, mget(ls(pattern = '^time'))))) %>% 
-  rownames_to_column('Function') %>% 
-  select(-user.child, -sys.child) %>% 
-  mutate(Function = str_remove(Function, 'time_')) %>% 
-  mutate(Module = case_when(
-    Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
-    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 'time_geodata_writer',
-                    'political_grabber', 'site_writer', 'geodata_writer', 'directions_grabber') ~ 'Geospatial', 
-    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher', 
-                    'associates_spell_check_veg', 'associates_spell_check_ass') ~ 'Taxonomic',
-    Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
-    ),
-    Group = if_else(Function %in% c('powo_searcher', 'directions_grabber'), 'Online', 'Local')
-  )
-
-# save these as RDS so operations do not need to be run each time document knits
-saveRDS(time_trials, file = 'data/processed/time_trials')
-
-rm(time_trials)
 rm(list=ls(pattern = 'time'))
@@
 ```{r Create Labels, eval = F}
-processed <- read.csv('data/processed/cleaned_data.csv') %>% 
+processed <- read.csv('data/processed/cleaned_data-problemchildren.csv.csv') %>% 
   mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
@@
 
-# now copy bash script to location # note we are one dir up from the lowest level. 
-# note that I customized the label template after import, so we will not run this again. 
-# p2libs <- system.file(package = 'BarnebyLives')
-# folds <- file.path( 'rmarkdown', 'render_labels.sh')
-
-# file.copy(from = file.path(p2libs, folds), 
-#           to = '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw')
 ```
@@
 ```{r combine labels, eval = F}
-## this SHOULD be a sh chunk, so launch in a shell! I just want to be sure you can see the code. 
-## and eval = TRUE will be respected. 
-user: pwd
-/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw
-# this will have both your script and a dir full of raw labels with a users name
-# now launch the bash script with:
-
 # or chmod it etc!
 # et voila! data will be saved up on dir in a 'final' directory. 
-# we timed it like this:
-\time -o ../process2final.txt  bash render_labels.sh collector='Reed'
+../process2final.txt  bash render_labels.sh collector='Reed'
 ```
@@
 
-BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online for preferred synonyms, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
+BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`sec) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`sec) to search Plants of the World Online, and a minute `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`sec to search Google Maps and write directions to sites.
 
@@
                '1556', '1588', '1619', '2445', '2777', '1337', '2004', '2412', '2487', '2748', '2652',
-               '2805', '2194', '2040', '2039', '2403', '2328', '2076') 
+               '2805', '2194', '2040', '2039', '2252', '2403', '2328', '2076') 
              ~ 'Outdated', 
@@
 Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)` records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance where an outdated circumscription was applied.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)`records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instances where an outdated circumscription was applied.
 At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
-Of `r n_spp` distinct species analysed BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling', na.rm = TRUE)`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling',  na.rm = TRUE)` unique species).
-In `r sum(epithet_diff$Situation=='Outdated', na.rm = TRUE)` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
+Of the `r nrow(data)` records analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
+Of the `r nrow(data)` species analysed (`r n_spp` distinct species) BarnebyLives flagged `r nrow(epithet_diff)` records, and detected `r sum(epithet_diff$Situation=='Spelling')`  instances of misspelled epithets (`r sum(epithet_diff$Situation=='Spelling')` unique species).
+In `r sum(epithet_diff$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Outdated', 'Scientific_name']))` unique species).
 It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
@@
 
-While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
+While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate (@patten2024geographic).
 We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
@@
 Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
-While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
 While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
 
-![A label generated from a default template](../graphics/plots/RCB2759.png){width=50%}
+While numerous tools have been developed for cleaning existing herbarium and museum records, few help ensure that the entered data are accurate [@patten2024geographic].
+We argue that the original collectors are the most qualified individuals to perform quality control checks, and BarnebyLives allows them to do so in a relatively fast and streamlined format.
+By utilizing both R and LaTeX and having publicly available source code on Github, this program allows most users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications to project branches.
+
+Accessioning often relies on the use of the Office Suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offers functionality to readily view and filter datasets using familiar spreadsheet-like formats.
 
 LaTeX offers well-documented and detailed functionality to customize labels for individuals and institutions. 
-Anecdotally, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
+Anecdotally, using its default settings, it can produce more aesthetically pleasing results than typical word processors. 
 Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project.
@@
 # ACKNOWLEDGEMENTS 
-The Bureau of Land Management is gratefully acknowledged as a provider of funding to R.C.B. for most of his specimen collection activities. 
+The Bureau of Land Management is gratefully acknowledged as a provider of funding to R. C. B. for most of his specimen collection activities. 
 We thank the two anonymous peer reviewers who have increased the quality of this manuscript.
 Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota Becerra, Caitlin Miller, Hubert Szczygiel. 
+We thank Rosalind Rowe for providing useful feedback during the later developmental stages of the program.
 

### 2025-05-01 manuscript/CoverLetter.rmd

@@
 \end{flushleft} 
+\vspace{1em}
 
 \hfill April 24^th^, 2025  
+\vspace{1em}
 
@@
 
-I am writing to propose an original software article titled *"BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets"*. 
-Please bear with me when I cheekily explain it's impetus for existence, otherwise it's utility may seem opaque.
+I am pleased to submit a software note titled *"BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets"* for consideration in Applications in Plant Sciences.  
+\vspace{2em}
 
-While a variety of software tools exist for interacting with accessioned herbarium data, few are dedicated to accessioning new material. 
-And even fewer of these tools are intended for an actual collector to use. 
-Which isn't surprising, as the number of collectors who are not curators has decreased, and the barrier to entry for both activities is a high degree of particularity and eccentricity. 
-Unfortunately, the particularity of curators often makes contributing to herbaria a challenging task for collectors. 
-A common challenge for collectors, especially those who work on large projects implemented by federal agencies, such as germplasm collection (e.g.
-SOS) and ecological monitoring (e.g.
-NWCA, and AIM), is the need to simultaneously satisfy the particularities of multiple curators under tight deadlines.
+This package was developed to assist field botanists, particularly those engaged in large-scale federal efforts such as germplasm collection (e.g., Seeds of Success) or ecological monitoring programs (e.g., National Wetland Condition Assessment, or  Assess, Inventory and Monitor), in generating herbarium specimen labels and ensuring the quality and veracity of their data.
+While several tools exist for working with already accessioned herbarium data, few are tailored to the needs of collectors preparing specimens for deposit.
+Fewer yet are designed with the realities of tight timelines and varying herbarium requirements in mind.
+\vspace{2em}
 
-Fortunately, I have been scolded by many curators over the last decade, and on occasion received actionable criticism, and have built up a wide knowledge of quirks to cater to. 
-To assist in running quality control checks, for both my - and the young botanists whom I manage - collections, I developed a wide range of functions which help meet a broad set of interests.
-In fact 'BarnebyLives' the package which integrates these tools, has 30 user facing functions,  10+ internal functions, and 8 built in data sets.
-
-The package is used to both evaluate commonly mistaken data, perform an array of spell checks, offer botanists more recently applied scientific names, and export data and write labels for them. 
-Curators across roughly 15 institutes have been satisfied by the results while one curator just seems to just dislike me as individual. 
-While I note it as a an R package, it also creeps into LaTeX and some bash utilities, which are 99% restricted to the RStudio environment, making it slightly outside the realm of a typical R Package; we could probably render labels through html, but why bother to go through all this work to create labels which look terrible?
+The software has already been used to prepare data for collections, from a variety of teams, deposited at approximately 15 herbaria, and the output from it has been received quiet warmly. 
+Feedback from curators has helped shape its features and scope, and we hope that making it widely available will help empower field botanists to contribute high-quality collections with greater efficiency and confidence.
+\vspace{2em}
 
 I think that APPS is a wonderful place for this paper to be published, and APPS would be happy to have this paper. 
-Several editors of APPS have been unrelenting in their support of herbaria, and that herbaria maintain their role as the center of academic and field botany. 
+Several editors of APPS have been unrelenting in their support of herbaria and that herbaria maintain their role as the center of academic and field botany. 
 We believe that on occasion herbaria are not seen as resources for the future, but rather catalogues of the past, and would like to make our small contribution to ensure that does not become true. 
+\vspace{2em}
 
-I have no conflicts of interest to disclose.
+We affirm that this submission is original, has not been published elsewhere, and is not under consideration by another journal. There are no conflicts of interest to disclose.
+\vspace{2em}
+
+Thank you for considering this submission. I look forward to your response.
+\vspace{1em}
 

### 2025-05-01 SubsetRecords2CBG.rmd

---
author: "Reed Benkendorf"
date: "2025-04-29"
output: html_document
---

```{r}
library(tidyverse)
```

```{r}
p <- './manuscript/data'

dat_sub <- read.csv(file.path(p, 'test_data.csv')) |>
  select(Collection_number, CBG_ACC_BOOL) |>
  mutate(CBG_ACC_BOOL = if_else(is.na(CBG_ACC_BOOL), TRUE, FALSE)) |>
  filter(CBG_ACC_BOOL == TRUE)

read.csv( file.path(p, 'processed', 'cleaned_data.csv') ) |>
  select(-CBG_ACC_BOOL) |>
  right_join(dat_sub)
```

### 2025-05-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd

@@
         c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
-  filter(Scientific_name != '')#, Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
+  filter(Scientific_name != '', Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
 

### 2025-05-01 manuscript/BarnebyLives -COPY2 ACTUALLY CLEAN.rmd
ensured results from POW are compared, not results to internal spellcheck... results are the same - but matched!

@@
 
-data <- read.csv('data/test_data.csv', na.strings = "") %>% 
+data <- read.csv('data/test_data-2accession.csv', na.strings = "") %>% 
   drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
@@
 
-# for this example we will just ACCEPT all changes blindly. 
-# basically we just want some data for running timing exercises on the label making.
-
 # just overwrite the original Associates and Vegetation columns
@@
 
-# honestly just overwrite these too. 
-data <- data %>% 
+original_names <- c(1424, 2038, 2326, 2435, 2517, 2618, 2644, 2782:2784)
+
+data1 <- data |>
+  sf::st_drop_geometry() |>
+  split_scientificName(sciName_col = 'Scientific_name')
+
+  # for these records, we want to retain the originally entered information 
+  mutate(
+    POW_Family = if_else(Collection_number %in% c(2435, 2517), 'Ranunculaceae', POW_Family)), 
+    POW_Genus
+
+  ) |>
+
   select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 

### 2025-05-01 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
ensured results from POW are compared, not results to internal spellcheck... results are the same - but matched!

@@
 We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
-While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
+While many novice botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
 Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
@@
 The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
-For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
+For example, while acquiring appropriate political information for a collection site appears simple, novice collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
 In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report administrative details in error.
@@
 
-subs_batch <- c(1586, 1880, 2187, 2159)
+#subs_batch <- c(1586, 1880, 2187, 2159)
 
@@
         c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
-  filter(Scientific_name != '', Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
+  filter(Scientific_name != '')#, Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)
 
@@
 saveRDS(data, file = 'data/processed/data_w_Google_Maps')
-
 ```
@@
 # basically we just want some data for running timing exercises on the label making. 
-
+data <- readRDS( file = 'data/processed/data_w_Google_Maps')
 # just overwrite the original Associates and Vegetation columns
@@
         -Infrarank, -Infraspecies, -Binomial_authority, -Infraspecific_authority, -Family) |>
-  rename(Full_name = SpellCk.taxon_name,
-         Genus = SpellCk.genus, 
-         Epithet = SpellCk.species, 
+  rename(Full_name = POW_Full_name,
+         Genus = POW_Genus, 
+         Epithet = POW_Epithet, 
          Binomial_authority = POW_Binom_authority, 
@@
 data_p <- readRDS(file = 'data/processed/data_w_POWO_search') %>% 
-  sf::st_drop_geometry()
+  sf::st_drop_geometry() %>% 
+  split_scientificName('Scientific_name')
 

### 2025-05-01 manuscript/flowchart.R

@@ -10,18 +10,11 @@ digraph boxes_and_circle {
   graph[rankdir = LR, ratio = 0.2, nodesep = 0.01, ranksep = 0.01]
   
   node [shape = circle, fontsize = 72, fontname = arial, style = filled, fontcolor = White]
-
-      subgraph cluster_import{color= 'white'
-        node [color = 'black', shape = egg]
-        'Raw\nSpreadsheet'
-      }
       
         # Geospatial Operations
       subgraph cluster_geospatial {style=dashed; color= 'white'; 
         node [color = '#94524A']
-  
-          'Raw\nSpreadsheet' -> 'Convert\nfrom\nDMS to DD' 
-              [label = '     dms2dd', fontsize = 72, penwidth = 6, style = dashed]
+
           'Convert\nfrom\nDMS to DD' -> 'Check for\nautofill\nvalues' 
               [label = '     autofill_checker', fontsize = 72, penwidth = 6, style = dashed]
           'Check for\nautofill\nvalues' -> 'Geospatial\nData' 

### 2025-05-01 manuscript/flowchart.R
writing

@@ -1,16 +1,16 @@
 library(tidyverse)
 library(DiagrammeR)
 library(ggpubr)
-#setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
-setwd('/home/sagesteppe/Documents/assoRted/Barneby_Lives_dev/manuscript')
+setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
+#setwd('/home/sagesteppe/Documents/assoRted/Barneby_Lives_dev/manuscript')
 
 grViz("
 
 digraph boxes_and_circle {
   graph[rankdir = LR, ratio = 0.2, nodesep = 0.01, ranksep = 0.01]
   
-  node [shape = circle, fontsize = 72, fontname = arial, style = filled, fontcolor = White]
-      
+  node [shape = circle, fontsize = 72, fontname = 'arial, Bold', style = filled, fontcolor = White]
+    
         # Geospatial Operations
       subgraph cluster_geospatial {style=dashed; color= 'white'; 
         node [color = '#94524A']
@@ -19,8 +19,6 @@ digraph boxes_and_circle {
               [label = '     autofill_checker', fontsize = 72, penwidth = 6, style = dashed]
           'Check for\nautofill\nvalues' -> 'Geospatial\nData' 
               [label = '     coords2sf', fontsize = 72, penwidth = 6]
-          'Raw\nSpreadsheet' -> 'Check for\nautofill\nvalues' 
-              [label = '     autofill_checker', fontsize = 72, penwidth = 6]
           'Geospatial\nData' -> 'Political\nBoundaries' 
               [label = '     political_grabber', fontsize = 72, penwidth = 6]
           'Political\nBoundaries' -> 'Site\nCharacteristics' 
@@ -171,9 +169,7 @@ digraph boxes_and_circle {
   charToRaw %>% 
   rsvg::rsvg_png("../graphics/plots/Resultmod.png")
 
-
 text <- paste(
-  "Sample workflow for using BarnebyLives on data which has been manually entered into a spreadsheet.",
   "The top two rows indicate the main data cleaning functionality and are best run in the order outlined",
   "above although taxonomic steps may be ran before spatial steps. The third row can be interspersed",
   "with the above two, includes creation of labels, which allows for detection of formatting or other",

### 2025-05-01 SubsetRecords2CBG.rmd
data writeout for CHIC - ready, do a final manual review

@@
 ```{r}
-p <- './manuscript/data'
+p <- './manuscript/data/processed'
 
-dat_sub <- read.csv(file.path(p, 'test_data.csv')) |>
-  select(Collection_number, CBG_ACC_BOOL) |>
-  mutate(CBG_ACC_BOOL = if_else(is.na(CBG_ACC_BOOL), TRUE, FALSE)) |>
-  filter(CBG_ACC_BOOL == TRUE)
+dat_sub <- read.csv(file.path(p, 'cleaned_data.csv')) |>
+  mutate(CBG_ACC_BOOL = if_else(is.na(CBG_ACC_BOOL), TRUE, FALSE)) 
+dat_sub <- data.frame( apply(dat_sub, 2, as.character) )
 
-read.csv( file.path(p, 'processed', 'cleaned_data.csv') ) |>
-  select(-CBG_ACC_BOOL) |>
-  right_join(dat_sub)
+dat_manual <- read.csv(file.path(p, 'cleaned_data-problemchildren.csv')) |>
+  mutate(CBG_ACC_BOOL = if_else(is.na(CBG_ACC_BOOL), TRUE, FALSE)) 
+dat_manual <- data.frame( apply(dat_manual, 2, as.character) )
+
+dat_out <- filter(dat_sub, ! Collection_number %in% dat_manual$Collection_number) |>
+  bind_rows( dat_manual) |>
+  filter(CBG_ACC_BOOL == 'TRUE') |>
+  arrange(Collection_number)
+
+write.csv(dat_out, file.path(p, 'CHIC-RCB-2025.csv'))
 ```

### 2025-05-01 manuscript/labels/skeleton.Rmd
data
data writeout for CHIC - ready, do a final manual review

@@
 ```{r load data}
-data <- read.csv('../data/processed/cleaned_data.csv') |>
+data <- #read.csv('../data/processed/cleaned_data.csv') |>
+  read.csv('../data/processed/cleaned_data-problemchildren.csv') |>
   dplyr::mutate(

### 2025-07-19 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
writing
floating ?function
level

@@
 ```
+
+# INTRODUCTION
 
@@
 `r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values. 
+All flags were correct, and in several instances more errors were found in the rows following the flagged values.
 
@@
 
-# ACKNOWLEDGEMENTS 
+# ACKNOWLEDGMENTS 
 The Bureau of Land Management is gratefully acknowledged as a provider of funding to R.C.B. for most of his specimen collection activities. 
@@
 <div id="refs"></div>
-
-# SUPPORTING INFORMATION
-Additional supporting information can be found online in the Supporting Information section at the end of this article.
-
-**Appendix S1.** A table of all time trials for each function.

### 2025-07-19 manuscript/BarnebyLives -COPY2 ACTUALLY CLEAN.rmd

@@
 
-data1 <- data |>
+data <- data |>
   sf::st_drop_geometry() |>
-  split_scientificName(sciName_col = 'Scientific_name')
+  split_scientificName(sciName_col = 'Scientific_name') |>
 
@@
   mutate(
-    POW_Family = if_else(Collection_number %in% c(2435, 2517), 'Ranunculaceae', POW_Family)), 
-    POW_Genus
-
+    POW_Family = if_else(Collection_number %in% c(2435, 2517), 'Ranunculaceae', POW_Family),
+    POW_Genus = if_else(Collection_number %in% original_names, Genus, POW_Genus),
+    POW_Epithet = if_else(Collection_number %in% original_names, Epithet, POW_Epithet),
   ) |> 
-
-  select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, 
+  select(-SpellCk.gGRP, -Match, -SpellCk.taxon_rank, -Genus, -Epithet,
         -Infrarank, -Infraspecies, -Binomial_authority, -Infraspecific_authority, -Family) |>
-  rename(Full_name = SpellCk.taxon_name,
-         Genus = SpellCk.genus, 
-         Epithet = SpellCk.species, 
+  rename(Full_name = POW_Full_name,
+         Genus = POW_Genus, 
+         Epithet = POW_Epithet, 
          Binomial_authority = POW_Binom_authority, 
@@
 ```{r Create Labels, eval = F}
-processed <- read.csv('data/processed/cleaned_data-problemchildren.csv.csv') %>% 
+processed <- read.csv('data/processed/cleaned_data-problemchildren.csv') %>% 
   mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
@@
 
-processed <- data.frame( apply(processed, 2, as.character) )
+processed <- data.frame(apply(processed, 2, as.character))
 processed <- mutate(processed, Collection_number = as.numeric(Collection_number))
@@
 })
-
-saveRDS(raw_lab_time, file = 'data/processed/time_label_gen')
 

### 2025-07-19 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
reviewer 1 edits

@@
 Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
-However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread].
+However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting Earth's biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread].
 This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
@@
 While many novice botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
-Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
@@
 Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota].
-The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
+The office suite costs money, and in our experience, can be problematic to set up; further, its functionality ends with label creation. 
 PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
@@
 While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
-Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.
+Unfortunately, we have not found Windows alternatives to pdfjam and pdftk, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.
 
@@
 *associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks.
-*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
+*field_lengths* will supply a warning messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
 
-The manual review process technically only has one function that is optional and may be executed during the spatial process (after *coords2sf*), but the importance of manual review is important enough to warrant explicit mention.
+The manual review process technically only has one optional function (after *coords2sf*), however given the importance of ensuring that coordinates are appropriately located for many of the modules in the software, we suggest all users perform it.
 *geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/).
@@
 
-The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
-*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.
-*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
+The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the purrr package from the tidyverse.
+*labels_skeleton* is an R 'script' which will require modifications to customize for institutions, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.  
+*label_writer* writes from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
 *format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.
@@
 
-The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
+The testing of the package within this manuscript was performed using a subset of the first author's collections from 2018-2022, while most development was performed on their 2023 and 2024 collections.
 Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
@@
 The next two vignettes *"BarnebyLives!
-Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* how to export data in both digital and analog formats.
+Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats.
 *"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium.

### 2025-07-19 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
citations

@@
 Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
-However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].
+However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].
 
@@
 While many novice botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
-Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
+Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, familiarity with various computer programs (for example, Microsoft Office suite), and foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].
 
@@
 
-Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
-While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on naming, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)` records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance where an outdated circumscription was applied.
+Even on our test data set which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use.
+While none of the `r n_families` families were misspelled, BarnebyLives made `r nrow(family_incongruence)` suggestions on family level nomenclature, identified `r sum(family_incongruence$Situation=='Submitter_Incorrect', na.rm = T)` instances where the user entered an unequivocally incorrect family (or taxonomic entity), identified `r sum(family_incongruence$Situation=='Autofilled', na.rm = T)` records where families were autofilled,  and `r sum(family_incongruence$Situation=='Outdated', na.rm = T)` instance where an outdated circumscription was applied.
 At the level of family BarnebyLives flagged `r sum(family_incongruence$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy, and flagged `r sum(family_incongruence$Situation=='Internal_Error', na.rm = T)` records in error, it appears most of these errors are due to issues in the backbone used by the earlier spell check function.
@@
 In the `r n_genera` genera analysed  BarnebyLives identified `r nrow(genus_mismatch)` discrepancies at the level of genus between user submitted and processed data. 
-In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` record in error.
+In `r sum(genus_mismatch$Situation=='Outdated')` of these instances the user supplied an outdated name (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Outdated', 'Genus']))` unique genera) flagged `r sum(genus_mismatch$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(genus_mismatch[genus_mismatch$Situation=='Preferred', 'Genus']))` genera total), and flagged `r sum(genus_mismatch$Situation=='Internal_Error', na.rm = T)` records in error.
 
@@
 It also flagged `r sum(epithet_diff$Situation=='Preferred', na.rm = T)` records where the author follows an alternative taxonomy (`r nrow(unique(epithet_diff[epithet_diff$Situation=='Preferred', 'Scientific_name']))` unique species), and flagged `r sum(epithet_diff$Situation=='Internal_Error', na.rm = T)` records in error.
-The final record was an egregious error where the order of the specific epithet and the genus name.
+The final record was an egregious error where the order of the specific epithet and the genus name were swapped.
 
-`r long_flag` records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values. 
+Five records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values. 
 All flags were correct, and in several instances more errors were found in the rows following the flagged values.
@@
 Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e.
-the administrative information on Township Section and Range (or 'TRS'), but are considered less value in other geographic regions.
-Further several of the abiotic variables which it acquires information on: slope, aspect, and geology have long been considered prominent drivers of plant distributions in semi-arid and montane systems and warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous.
-Finally, it is plausible people in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the west we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.
+the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.
+Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane systems and is warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous.
+Finally, it is plausible botanists in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the Western United States we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.
 
@@
 While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
-While other software often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
+While other software packages often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
 

### 2025-07-19 manuscript/ComputerSpecsAppendix.Rmd

# Appendix A: System Information for Benchmarking

The following system details are provided to support reproducibility of the performance measurements reported in this work.
All benchmarks were conducted on a local machine with the specifications shown below. 
Information was collected via standard system commands at time of receiving reviewer comments.

## Operating System

```{sh, eval = F}
lsb_release -a
```

Distributor ID:	Ubuntu  
Description:	Ubuntu 22.04.5 LTS  
Release:	22.04  
Codename:	jammy

## CPU

```{sh, eval = F}
lscpu
```

Architecture:             x86_64  
  CPU op-mode(s):         32-bit, 64-bit  
  Address sizes:          39 bits physical, 48 bits virtual  
  Byte Order:             Little Endian  
CPU(s):                   16  
  On-line CPU(s) list:    0-15  
Vendor ID:                GenuineIntel  
  Model name:             11th Gen Intel(R) Core(TM) i7-11700K @ 3.60GHz  
    CPU family:           6  
    Model:                167  
    Thread(s) per core:   2  
    Core(s) per socket:   8  
    Socket(s):            1  
    Stepping:             1  
    CPU max MHz:          5100.0000  
    CPU min MHz:          800.0000  
    BogoMIPS:             7200.00

## Memory

```{sh, eval = F}
free -h
```
               total        used        free      shared  buff/cache   available  
Mem:           125Gi       3.1Gi       112Gi       569Mi       9.7Gi       120Gi  
Swap:          2.0Gi          0B       2.0Gi

## Internet Speed

```{sh, eval = F}
speedtest-cli
```

Retrieving speedtest.net configuration...  
Testing from Astound Broadband (168.91.196.6)...  
Retrieving speedtest.net server list...  
Selecting best server based on ping...  
Hosted by Sangoma (Chicago, IL) [17.63 km]: 18.388 ms  
Testing download speed..........................................................  
Download: 21.58 Mbit/s  
Testing upload speed............................................................  
Upload: 27.43 Mbit/s

## Filesystem Usage

```{sh, eval = F}
df -h
```

Filesystem      Size  Used Avail Use% Mounted on  
tmpfs            13G  2.6M   13G   1% /run  
/dev/nvme0n1p2  468G  242G  203G  55% /  
tmpfs            63G   14M   63G   1% /dev/shm  
tmpfs           5.0M  4.0K  5.0M   1% /run/lock  
efivarfs        192K   88K  100K  47% /sys/firmware/efi/efivars  
/dev/nvme0n1p1  511M  6.1M  505M   2% /boot/efi  
tmpfs            13G  132K   13G   1% /run/user/1000  
/dev/sda1       3.6T  3.2T  295G  92% /media/steppe/hdd

## I/O Performance

```{sh, eval = F}
sudo hdparm -Tt /dev/nvme0n1p2
```

Timing cached reads:   44096 MB in  2.00 seconds = 22095.91 MB/sec  
Timing buffered disk reads: 4470 MB in  3.00 seconds = 1489.83 MB/sec

All files were stored on the following drive.

```{sh, eval = F}
sudo hdparm -Tt /dev/sda1
```

Timing cached reads:   45870 MB in  2.00 seconds = 22986.94 MB/sec  
Timing buffered disk reads: 514 MB in  3.01 seconds = 170.57 MB/sec

### 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
edits

@@
 Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
-Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
+Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the autofilling of coordinate and date columns. 
 After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.
@@
 
-Five records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values. 
+Five records were appropriately flagged for issues with autofills of the longitude value, and `r lat_flag` records were also flagged for autofilling increases in latitude values. 
 All flags were correct, and in several instances more errors were found in the rows following the flagged values.
@@
 
-Thematically, BarnebyLives is set up to cover Western North America. 
+Thematically, many of the variables which BarnebyLives collects information on are tailored to Western North America. 
 However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
@@
 the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.
-Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane systems and is warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous.
+Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane environments and are warranted on an herbarium label for specimens collected from these environments, whereas curators in other regions may find this information superfluous.
 Finally, it is plausible botanists in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the Western United States we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.

### 2025-08-09 LabelStyles/skeleton-customfont-twofonts.Rmd
working on scope of users sentence/s

---
output: 
  pdf_document:
    latex_engine: xelatex # or lualatex
    keep_tex: true
params:
  Collection_number: Collection_number
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
  - \usepackage{fontspec}
  - \setmainfont{Georgia}
  - \newfontfamily\couriernew{Courier New}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r load library}
library(BarnebyLives)
```

```{r load data}
local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')

record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
  sf::st_drop_geometry() |>
  data.frame() |>
  dplyr::filter(Collection_number == params$Collection_number) 
```

\begin{center}
\textbf{\Large `r writer(record[['Project_name']])`}
\end{center}
\begin{flushright}
\uppercase{`r record[['Family']]`}
\end{flushright}
\begingroup
\large
`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`  
`r writer_fide(record)`
\normalsize
\endgroup

\vspace*{\fill}
\begingroup
`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).

\couriernew{`r writer(record[['Habitat']])`. `r record[['physical_environ']]`}
Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
`r writer(record[['Notes']])`  
\endgroup

\vspace*{\fill}
\begingroup
`r collection_writer(record)`
\endgroup

### 2025-08-09 LabelStyles/skeleton-customfont.Rmd
working on scope of users sentence/s

--- 2025-08-09 LabelStyles/skeleton-customfont-twofonts.Rmd
@@
   - \setmainfont{Georgia}
-  - \newfontfamily\couriernew{Courier New}
 ---
@@
 
-\couriernew{`r writer(record[['Habitat']])`. `r record[['physical_environ']]`}
+`r writer(record[['Habitat']])`. `r record[['physical_environ']]`  
 Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`

### 2025-08-09 LabelStyles/skeleton-maps.Rmd
working on scope of users sentence/s

--- 2025-08-09 LabelStyles/skeleton-customfont.Rmd
@@
 ---
-output: 
-  pdf_document:
-    latex_engine: xelatex # or lualatex
-    keep_tex: true
+output: pdf_document
 params:
   Collection_number: Collection_number
+  imagepath: imagepath
 geometry: paperheight=4in,paperwidth=4in,margin=0.1in
@@
   - \usepackage{setspace}
-  - \usepackage{fontspec}
-  - \setmainfont{Georgia}
 ---
@@
 ```{r load data}
-local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
+record <- collection_examples |>
+  dplyr::filter(Collection_number == params$Collection_number) |>  
+  sf::st_drop_geometry()
 
-record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
-  sf::st_drop_geometry() |>
-  data.frame() |>
-  dplyr::filter(Collection_number == params$Collection_number) 
+imagepath <- params$imagepath
 ```
@@
 \begin{center}
-\textbf{\Large `r writer(record[['Project_name']])`}
+\textbf{\Large `r writer(record[['Project_Name']])`}
 \end{center}
@@
 \end{flushright}
+
+\noindent
+\begin{minipage}[t]{0.78\textwidth}\vspace*{0pt}
+
 \begingroup
@@
 \vspace*{\fill}
+
 \begingroup
-`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_Uncertainty']])`).
 
@@
 \endgroup
+\end{minipage}%
+\hfill
+\begin{minipage}[c]{0.18\textwidth}\vspace*{0pt}
+\hspace*{4mm}% left margin
+\centering
+\includegraphics[width=\linewidth,height=0.66\textheight,keepaspectratio]{`r imagepath`}
+\end{minipage}
+
+convert -density 300 Molecular_nets-crop.pdf -quality 100 \
+        -gravity West -background white -splice 100x0 \
+        \( -background transparent -fill black -pointsize 12 \
+           label:"Time within Season" -rotate -90 \) \
+        -geometry +0+250 -composite Molecular_nets-crop-labelled.png

### 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
working on scope of users sentence/s

@@
 
+- User groups 
+  - Academia
+    - Botanists (floristics, inventory work)
+    - Ecology (community ecologists)
+  - herbaria personnel
+
+We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume that responsibility in a relatively fast and streamlined format
+While we largely highlight the usage of BarnebyLives in the context of herbaria, it's original use case was for natural resource management workers (e.g.
+US Forest Service, heritage programs, natural history surveys), and graduate students in ecology and conservation (e.g.
+inventory, and community ecology work).
+
+We believe that these 
+additional uses cases exist in domains where users seldom interface with curatorial staff, such as
+
 LaTeX, a software system used for typesetting, allows users to focus on the content rather than the style of the documents rendered from it. 

### 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
minor edits in discussion

--- 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
 While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
-We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume that responsibility in a relatively fast and streamlined format.
+We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume this responsibility in a relatively fast and streamlined format.
+While BarnebyLives may be adopted by herbarium curators, its original use was for natural resource management employees (for example, US Forest Service, heritage programs, natural history surveys) and graduate students in ecology and conservation (e.g., performing inventory and community ecology work), and we expect it to be most widely used in this realm.
+Beyond always having the data collected by BarnebyLives terminating in a label for a physical specimen, its utility may be extended to providing information for various biotic databases (e.g., heritage program databases) or modelling applications (e.g., modelling in community ecology) as a user sees fit.
 By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
-
-LaTeX, a software system used for typesetting, allows users to focus on the content rather than the style of the documents rendered from it. 
-However, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
-Additionally LaTeX offers users a wide variety of ways which they can modify labels which are under-explored in the package. 
+LaTeX, a software system used for typesetting, was developed to allow users to focus on the content rather than the style of the rendered documents. 
+Indeed, using the default LaTeX settings along with standard templates allows users to produce aesthetically pleasing results (Figure 4). 
+However, LaTeX also offers a rich suite of customizable parameters that offer users a wide variety of ways to modify labels that are under-explored in the package.
 Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project. 
@@
 However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
-Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e.
+Several of the attributes which it collects and displays on labels, relate to topics in which more senior curators are interested, i.e.
 the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.
@@
 
-![A label generated from a default template](../graphics/plots/RCB2759.png){width=50%}
+![A label generated from a default template included with the package](../graphics/plots/RCB2759.png){width=50%}
 
@@
 We thank the two anonymous peer reviewers who have increased the quality of this manuscript.
-Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota Becerra, Caitlin Miller, Hubert Szczygiel.
+Several prominent associated collectors of specimens used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota Becerra, Caitlin Miller, and Hubert Szczygiel.
 

### 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
revamp paragraph on data themes

@@
 
-Thematically, many of the variables which BarnebyLives collects information on are tailored to Western North America. 
-However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
-Several of the attributes which it collects and displays on labels, relate to topics in which more senior curators are interested, i.e.
-the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.
-Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane environments and are warranted on an herbarium label for specimens collected from these environments, whereas curators in other regions may find this information superfluous.
-Finally, it is plausible botanists in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the Western United States we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.
+BarnebyLives allows for the usage of 'domains' (i.e., custom geographic extents), which analysis is restricted to, allowing users to specify any portion of the United States for setting up their instance.
+However, many of the environmental and geographic variables that BarnebyLives collects information on are tailored to Western North America. 
+Several of the variables that BarnebyLives collects and displays on labels relate to topics in which more senior Western curators are interested, that is, the administrative information on Township Section and Range (or 'TRS') or the agency administering public land, but are considered less valuable–or even unavailable or inapplicable–in other geographic regions.
+Environmental variables such as slope, aspect, and geology are prominent drivers of plant distributions in semi-arid environments and warrant inclusion on herbarium labels for specimens from Western environments, but curators may consider them superfluous for other environments.
+Accordingly, users may consider developing new functionality that either removes or supplements the default variables to better reflect regional needs.
 

### 2025-09-01 LabelStyles/custom-font2823.tex
map example

--- 2025-04-29 manuscript/BarnebyLives-An-R-package-to-create-and-qc-herbarium-labels.tex
@@
 \usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
+\usepackage[paperheight=4in,paperwidth=4in,margin=0.1in]{geometry}
 \usepackage{graphicx}
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{endfloat}
-\usepackage{setspace}\doublespacing
-\usepackage{lineno}
-\linenumbers
-\usepackage{rotating}
+\pagenumbering{gobble}
+\usepackage{setspace}
 \ifLuaTeX
@@
 \hypersetup{
-  pdftitle={BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets},
-  pdfkeywords={herbarium, natural history museum, collections,
-geospatial, R, LaTeX},
   hidelinks,
@@
 
-\title{BarnebyLives: an R package to create herbarium specimen labels
-and clean spreadsheets}
-\author{Reed Clark Benkendorf\(^1\)\(^,\)\(^2\)\footnote{Author for
-  Correspondence:
-  \href{mailto:rbenkendorf@chicagobotanic.org}{\nolinkurl{rbenkendorf@chicagobotanic.org}}},
-Jeremie B. Fant\(^1\)\(^,\)\(^2\)\\
-\strut ~\(^1\)Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe,
-Illinois 60022, USA\\
-\strut ~\(^2\)Plant Biology and Conservation, Northwestern University,
-Evanston, Illinois 60208, USA}
-\date{}
+\author{}
+\date{\vspace{-2.5em}}
 
 \begin{document}
-\maketitle
-\begin{abstract}
-\noindent \textbf{Premise:} Accessioning herbarium specimens is
-labor-intensive, yet remains vital for research in ecology, evolution,
-and conservation. As institutional support for herbaria declines,
-efficient tools are needed to streamline this process. BarnebyLives was
-developed to assist collectors by supplementing collection notes,
-verifying taxonomic data, conducting quality checks, generating labels,
-and submitting digital records. \textbf{Methods and Results:} It
-integrates geospatial data from U.S. government sources to provide
-jurisdictional and site information, and checks taxonomic names using
-in-house spell checkers, IPNI author standards, and Kew's Plants of the
-World Online. Optional features include generating Google Maps driving
-directions. The tool outputs data in tabular and spatial formats for
-review before producing LaTeX-based labels and shipping manifests.\\
-\textbf{Conclusions:} BarnebyLives improves data accuracy, ensures
-up-to-date taxonomy, and significantly reduces the time and effort
-required to accession herbarium specimens.
-\end{abstract}
 
-Nearly 400 million specimens are housed worldwide in herbaria
-(\citeproc{ref-thiers2021herbaria}{Thiers, 2021}). However, The rate of
-accessioning new collections to herbaria diminished in the
-20\textsuperscript{th} century as priorities in biology shifted away
-from describing and documenting earths biodiversity and towards
-understanding cellular and molecular processes underpinning life
-(\citeproc{ref-prather2004decline}{Prather et al., 2004};
-\citeproc{ref-pyke2010biological}{Pyke and Ehrlich, 2010};
-\citeproc{ref-daru2018widespread}{Daru et al., 2018}). This shift, among
-other factors, led to a decline in the funding allocated to
-collection-based research, the number of staff maintaining and accessing
-new collections, and educating students in these practices
-(\citeproc{ref-funk2014erosion}{Funk, 2014}). Historically, specimens
-have been used to describe the taxonomic diversity of plants and
-document global floristic diversity
-(\citeproc{ref-greve2016realising}{Greve et al., 2016};
-\citeproc{ref-james2018herbarium}{James et al., 2018};
-\citeproc{ref-brewer2019factors}{Brewer et al., 2019};
-\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020}). However,
-renewed interest in herbarium collections utilizing `big data
-approaches,' such as museuomics, has brought herbaria back to the
-forefront of the natural sciences and grearly expanded their roles in
-science (\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020};
-\citeproc{ref-marsico2020small}{Marsico et al., 2020}).
+\begin{center}
+\textbf{\Large Custom Fonts}
+\end{center}
+\begin{flushright}
+\uppercase{Apiaceae}
+\end{flushright}
 
-Innovations in specimen digitization, data sharing, computing, DNA
-sequencing, and statistics have perhaps brought about greater use of
-herbarium specimens than ever before
-(\citeproc{ref-greve2016realising}{Greve et al., 2016};
-\citeproc{ref-james2018herbarium}{James et al., 2018};
-\citeproc{ref-brewer2019factors}{Brewer et al., 2019};
-\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020}). The
-current use of specimens and their ancillary data extends well beyond
-their traditional roles in systematics and floristics, and studies
-utilizing collections are regularly carried out to better understand the
-ecological niches, phenological processes, and interactions of plants
-(\citeproc{ref-ronsted2020integrative}{Rønsted et al., 2020};
-\citeproc{ref-davis2023herbarium}{Davis, 2023}). We suspect that
-collections are yet to realize their full potential, and as currently
-novel approaches, such as electronic and remote sensing and
-meta-barcoding, become more accessible the use of collections will
-increase (\citeproc{ref-tosa2021rapid}{Tosa et al., 2021}). While
-image-based or purely observational (rather than collection-based)
-citizen science approaches (e.g., iNaturalist, BudBurst) have recently
-dovetailed with herbarium specimens to meet many current research needs,
-specimens contain rich data that are not accessible via images. Only
-specimens have the ability to: provide samples of DNA, secondary
-metabolites, or proteins, material for measuring (micro-)morphological
-attributes (\citeproc{ref-borges2020schrodinger}{Borges et al., 2020}),
-and seeds or pollen. These factors will ensure that the specimens remain
-the premier botanical data source into perpetuity.
+\noindent
 
-However, despite renewed recognition of the utility of collections,
-efforts to grow them appear slow
-(\citeproc{ref-prather2004decline}{Prather et al., 2004}). We conjecture
-that this is partly because collecting and depositing specimens is a
-fundamentally slower process, especially for novice collectors, relative
-to taking photographs via commercially developed apps on smartphones
-(\citeproc{ref-daru2018widespread}{Daru et al., 2018};
-\citeproc{ref-mishler2020spatial}{Mishler et al., 2020};
-\citeproc{ref-manzano2021fair}{Manzano and Julier, 2021}). While many
-young botanists are capable of using dichotomous keys and other
-resources to reliably identify and collect satisfactory material, we
-observe that they face difficulties navigating several aspects of data
-acquisition, processing, and preparation of labels for submission to
-herbaria. Some of the apparent problems include the lack of dedicated
-time at the end of a field season to process specimens, a general lack
-of education on cartography and orienteering, natural history (e.g.,
-geology, geomorphology), nomenclature, and familiarity with various
-computer programs (for example, Microsoft Office suite), and increasing
-foundational knowledge of plant systematics and phylogenetics
-(\citeproc{ref-woodland2007botanists}{Woodland, 2007};
-\citeproc{ref-barrows2016crossroads}{Barrows et al., 2016};
-\citeproc{ref-nanglu2023nature}{Nanglu et al., 2023}).
+\begin{minipage}[t]{0.78\textwidth}\vspace*{0pt}
 
-The generation of an herbarium specimen involves many steps that are
-easy to take for granted (\citeproc{ref-forman1989herbarium}{Forman and
-Bridson, 1989}). For example, while acquiring appropriate political
-information for a collection site appears simple, young collectors
-rarely have adequate cartographic resources (printed topographic maps or
-GIS software) at their disposal. In topographically complex areas, where
-administrative borders are often associated with hydrological basins and
-the ridges defining them, collectors are liable to misinterpret their
-true geographic position and report administrative details in error.
-Even finding appropriate site names can rarely be resolved without a
-printed map, as many navigation-related software now consider most
-features that would serve as site names extraneous. Similarly, the rate
-at which taxonomic innovations occur, the volume of the literature, and
-the reluctance of some regional curators to embrace a phylogenetic
-approach to plant classification have made it difficult to find more
-recently applied scientific names, even when these names are unanimously
-accepted by taxonomic specialists in the group and other regional
-curators (\citeproc{ref-hitchcock2018flora}{Hitchcock and Cronquist,
-2018}). Furthermore, formatting a label correctly (e.g., author
-abbreviations, italicization, etc.) is a time-consuming process with
-many opportunities to introduce errors in formatting which reduce the
-apparent credibility of a collector. Anecdotally, many mail merge
-templates offered by herbaria still require collectors to modify many
-variables by hand, for example, applying italicization. Even if a
-collector successfully navigates all these hurdles, the time allocated
-to each step is quite large, and may discourage them from further
-collecting.
+\begingroup
+\large
+*Lomatium nevadense* (S. Watson) J.M. Coult. & Rose     
+Fide: *Intermountain Flora*, det.: RCB, 29 Aug, 2023.
+\normalsize
+\endgroup
 
-As a result of these concerns, we have developed an R package,
-BarnebyLives, that aims to increase both the quality of data rendered to
-labels and recorded in databases and to speed up the generation of
-labels. BarnebyLives rapidly provides political and administrative
-boundary information for a collection site using data from the U.S.
-Census Bureau (\citeproc{ref-walker2022tigris}{Walker, 2024}), the
-Public Land Survey System (PLSS), and ownership details of public lands
-via the Protected-Areas Database (PAD-US)
-(\citeproc{ref-usgs2024padus}{Gap Analysis Project (GAP), 2024}). Site
-names are suggested by finding the closest unambiguously named place
-feature in the Geographic Name Information System (GNIS) and the precise
-calculation of distance and azimuth from this feature to the collection
-site (\citeproc{ref-gnis2024}{Survey, 2023}). Using the Global Mountain
-Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized
-named mountain data set with global coverage allows for a relevant
-descriptor of the general region with less ambiguity
-(\citeproc{ref-snethlage2022hierarchical}{Snethlage et al., 2022}).
-Spell checks on all scientific names (including associated species) are
-performed using a copy of the World Checklist of Vascular Plants, and
-the resolved species may be searched via Kew's Plant of the World Online
-for relevant synonyms (\citeproc{ref-govaerts2021world}{Govaerts et al.,
-2021}; \citeproc{ref-powo2024}{POWO, 2024}). Author abbreviations are
-verified using the International Plant Names Index (IPNI) Standard
-Author Abbreviation Checklist and also returned by Kew's Plants of the
-World Online to ensure proper abbreviations of authorities
-(\citeproc{ref-ipni2024}{The Royal Botanic Gardens and Herbarium, 2024};
-\citeproc{ref-powo2024}{POWO, 2024}). Checks to search for and flag
-common issues associated with spreadsheet software or data
-transcription, such as the auto-filling of coordinate and date columns.
-After a final review of the data, flagged or generated by the package,
-it allows for the option to export spreadsheets that are suitable for
-mass uploading of data to multiple common herbarium databases as well as
-the generation of herbarium labels.
+\vspace*{\fill}
 
-Currently, to our knowledge label generation functionality is provided
-explicitly by two programs, PLabel and Symbiota, and by the Microsoft
-Word tool Mail Merge (\citeproc{ref-gries2014symbiota}{Gries et al.,
-2014}; \citeproc{ref-perkins2020Plabel}{Perkins, 2020}). The office
-suite costs money, and in our experience, is finicky; further, its
-functionality ends with label creation. PLabel is a standalone program
-that has greatly enhanced functionality relative to a mail merge,
-allowing users to specify the layout and formatting of label components
-using an intuitive and local graphical user interface (GUI)
-functionality. However, beyond verifying the nations of collection it
-does not include data cleaning functionalities.\\
-While some sources indicate that it can only be used on Microsoft, we
-expect it to be usable on Linux and Mac using Windows `emulators' like
-Wine. The increasingly popular Symbiota biodiversity data management
-software not only provides label generation capabilities but also
-provides data cleaning functionality in an attractive GUI web portal
-allowing for live management of collections and bypassing the need for a
-local installation, allowing it to be accessed on all operating systems.
-Symbiota offers functionality similar to the first four of our five
-stages of our `Taxonomic' module and to our knowledge a check of the
-`Political Boundaries' (see Figure 1). However, not all herbaria use
-Symbiota and many have original database systems that they maintain (for
-example, Harvard University Herbarium,
-\url{https://kiki.huh.harvard.edu/databases/specimen_index.html};
-Missouri Botanical Garden \url{https://tropicos.org/specimen/Search};
-and The Consortium of Pacific Northwest Herbaria
-\url{https://www.pnwherbaria.org/}). However, and most importantly many
-collectors prefer to generate their own labels, especially as they are
-likely to send different sets of collections to different institutions.
-Accordingly, the functionality of Symbiota should exist in an ecosystem
-with alternative systems. In scenarios where users want to keep
-rendering labels in either of the three existing alternatives, they can
-easily export data in the appropriate formats after utilizing BLs data
-cleaning utilities.
+\begingroup
+U.S.A., Nevada, Washoe Co., 17N 20E 19. Washoe Summit. 39.32677 -119.78537 (WGS84 5m).
 
-BarnebyLives was named for plant taxonomist Rupert Charles Barneby
-(1911-2000), who published over 6,500 pages of text, described over 750
-taxa, and is notable for balancing his studies at the William and Lynda
-Steere Herbarium at the New York Botanical Garden with annual collection
-trips in Western North America from 1937-1970 and sporadically until he
-passed in 2000 (\citeproc{ref-welsh2001rupert}{Welsh, 2001}). Select
-accolades of Rupert include the 1989 Asa Gray Award from the American
-Society of Plant Taxonomists (ASPT), the 1991 Engler Silver Medal from
-the International Association of Plant Taxonomists (IAPT), as well as
-being one of eight recipients of the International Botanical Congress's
-(IBC) Millennium Botany Award (1999)
-(\citeproc{ref-welsh2001rupert}{Welsh, 2001}). Most germanely, Rupert
-was remembered as being generous with his time to assist younger
-botanists with the more arcane aspects of field botany and taxonomy
-(\citeproc{ref-holmgren2017}{Holmgren and Holmgren, 1988}).
+Granitoid derived aeolian loamy sand soils. At 5,200 ft, on a slope, 4° slo. 57° asp.; geology: Igneous and Sedimentary, undifferentiated.  
+Veg.: *Artemisia tridentata* var. *wyomimgensis, Poa secunda, Agropyron cristatum, Achnatherum thurberianum, Purshia tridenata*. Ass.: *Bromus tectorum, Phlox longifolia*.
 
-\section{METHODS AND RESULTS}\label{methods-and-results}
+\endgroup
 
-\begin{figure}
+\vspace*{\fill}
+\begingroup
+Reed Clark Benkendorf 2823; 6 May, 2023
+\endgroup
+\end{minipage}
+
+\% \hfill
+
+\begin{minipage}[c]{0.18\textwidth}\vspace*{0pt}
+\hspace*{4mm}% left margin
 \centering
-\includegraphics[width=1\textwidth,height=\textheight]{../graphics/plots/workflow.png}
-\caption{Recommended workflow}
-\end{figure}
-
-BarnebyLives was iteratively developed based on data submitted by
-approximately 20 seasonal field botany teams over two years.
-Essentially, continual updates were made as the developers became aware
-of the idiosyncrasies of collection notes and data entry. Several
-commands in BarnebyLives require output from previous functions, and a
-workflow that satisfies these requirements is presented in Figure 1.
-
-\subsection{Usage}\label{usage}
-
-All steps of BarnebyLives, except for label generation are run within
-the freely available RStudio. Data may be read from any common
-spreadsheet management system or database connection such as Excel, or
-free alternatives such as LibreOffice, OpenOffice, or via the cloud on
-Google Sheets. The latter two options are documented here and in package
-vignettes, detailed descriptions of the required and suggested input
-columns are located on a Github Pages
-(\url{https://sagesteppe.github.io/BarnebyLives/}) and around 100
-real-world examples are on a Google Sheets accessible from the page.
-BarnebyLives is atypical for R packages in that it requires a
-considerable amount of data to operate (Table 1). Virtually all on-disk
-memory associated with the package are used to store spatial data. The
-amount of spatial data varies according to the domain that the user
-decides to support (Figure 3). Functions that require on-disk data
-require a path to data as an argument. Manually supplying the path
-argument allows users to determine an appropriate storage location
-suitable for their needs.
-
-We anticipate that for a typical user, BarnebyLives will require less
-than a couple gigabytes of memory (ours covering all of the conterminous
-Western U.S. at 3-arc second (\textasciitilde90m) resolution is
-\textasciitilde16 GiB), while the processing requires relatively little
-RAM; hence, we believe installations can work on hardware as limited as
-Chromebooks, while having the data stored entirely on thumb-drives.
-Given that the attributes which the package collects data on are
-tailored to the Western U.S. region, we do not expect local installs to
-exceed the size of ours. The final steps of BarnebyLives, generating the
-labels, requires working installations of R Markdown, a LaTeX
-installation
-(e.g.~\href{https://www.tug.org/applications/pdftex/}{pdfTeX},
-\href{https://www.luatex.org/}{LuaTeX},
-\href{https://www.overleaf.com/learn/latex/XeLaTeX}{XeLaTeX}), and the
-open source command line tools
-\href{https://github.com/rrthomas/pdfjam}{pdfjam} and
-\href{https://linux.die.net/man/1/pdftk}{pdftk}. While these steps are
-run through a shell scripting language such as bash, we have wrapped
-them in R functions that bypass the need to enter the commands directly
-into a shell terminal outside of RStudio. Unfortunately, we have not
-found Windows alternatives to pdfam and pdftek, so we are unable to
-offer the final label-generating functionality on that operating system,
-but suspect Ubuntu subsystem for Windows may allow for integration of
-these tools.
-
-\subsection{Functionality}\label{functionality}
-
-BarnebyLives can be thought of as consisting of five main modules
-(Figure 1): spatial, taxonomic, formatting, manual review, and data
-exporting.
-
-The spatial module has five required functions and two optional
-functions.\\
-\emph{autofill\_checker} searches for patterns in the input latitude and
-longitude data associated with autofilling from various spreadsheet
-programs and will emit a warning if they are encountered.\\
-\emph{coords2sf} creates a spatially explicit simple feature (sf)
-geometry dataset for the input data. \emph{political\_grabber}
-determines many levels of administrative ownership, including land
-management and public land survey system sections.\\
-\emph{physical\_grabber} provides various geographic data, such as
-elevation, landform position, and aspect using 90m resolution spatial
-data.\\
-\emph{site\_writer} write distance and azimuth to collection site from
-the nearest official named place from the GNIS database.
-\emph{directions\_grabber} is an optional function that writes driving
-directions from a reasonably sized town to the closest drivable area to
-the site using the Google Maps API, which will require a valid Google
-account that is free per month for most personal and smaller academic
-usages.\\
-\emph{dms2dd} is an optional function used to convert from coordinates
-denoted in the degrees minutes and second format (for example,
-42°08'39.9"N 87°47'08.3"W) to decimal degree format (for example
-42.14439, -87.78569).
-
-Please note that the function \emph{physical\_grabber} is the one
-portion of the package where a decoupling may exist between the
-collection site, and the resolution of the spatial data. While we expect
-the mismatch to be negligible for all effective purposes relating to:
-elevation, major geology type, and in general aspect, estimates of slope
-at this resolution may be biased - generally to lower angles. For these
-reasons collectors must always make notes on the truly local environment
-which taxa are found in, and consider that the notes from BL reflect the
-greater landscape which a microfeature may be present in. While this
-mis-match will seldom effect landscape ecologists, it may have
-implications for other data users.
-
-The taxonomic module has four required functions and one optional
-function.\\
-\emph{spell\_check} will perform a spell check on the entered scientific
-name based on a local copy of Kew Plants of the World database filtered
-to the local continents or a user-specified backbone.\\
-\emph{spell\_check\_family} performs a spell check on the family entered
-for each scientific name.\\
-\emph{author\_check} ensures that the authors are entered in a valid
-format, for example, the correct standard abbreviations are used.\\
-\emph{associates\_check} performs a spell check on all associated
-species using the local taxonomic database. \emph{powo\_searcher} can be
-used in tandem with the functions \emph{spell\_check\_family} and
-\emph{author\_check}, but we use it in lieu of them to search the
-current Plants of the World Online to determine relevant synonyms and
-alternative higher taxonomy for the focal species. No API key or
-registration is required to use \emph{powo\_searcher}.
-
-The formatting module has three functions. Two are optional; however,
-they are run locally and so quickly that there is no reason to skip
-them. \emph{date\_parser} parses an input date into various formats for
-notating collection and determination dates on labels.
-\emph{associate\_dropper} silently removes the collected species from
-the list of associated species; however, it searches for the species to
-be removed using the scientific name entered initially by the user
-rather than returned via spell checks. \emph{field\_lengths} will emit
-messages for any fields that we suspect will create an `overflow' on the
-physical label and should be truncated for clarity.
-
-The manual review process technically only has one function that is
-optional and may be executed during the spatial process (after
-\emph{coords2sf}), but the importance of manual review is important
-enough to warrant explicit mention.\\
-\emph{geodata\_writer} will write out a spatial copy of the data set to
-any geospatial format supported by the sf package, but defaults to
-writing out `kmls' which are readily used with
-\href{https://earth.google.com/web/}{Google Earth}, and can also be
-opened in several other free geographic information system (GIS)
-softwares such as \href{https://www.qgis.org/}{QGIS}. Notably, many of
-the flags that BarnebyLives generates will be placed into columns with
-obviously flagged names and can be manually reviewed by the analyst, and
-many of these issues can be resolved by simply addressing the relevant
-issues in the original data input spreadsheet.
-
-The data exporting module contains three functions that interact with
-LaTeX templates and require slightly more advanced R user interactivity,
-such as setting up mapping functions using the tidyverses purrr
-package.\\
-\emph{labels\_skeleton} is an R `script' which will require a few
-modification steps to tailor to each institution, these R scripts will
-put data into a user specified template, and serve as the interface to
-LaTeX.\\
-\emph{label\_writer} write from a flatfile or spreadsheet to small 4x4
-inch herbarium labels (users can modify these dimensions as they see
-fit). \emph{format\_database\_import} will write out a spreadsheet of
-cleaned data in a variety of formats, currently: Jepson, Symbiota, and
-Consortium of Pacific Northwest herbaria are supported.
-
-\subsubsection{Herbarium Collections}\label{herbarium-collections}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/plots/collections_map.png}
-\caption{The spatial extent-or domain- (orange), and herbarium
-collection sites (burgundy) tested in this manuscript.}
-\end{figure}
-
-The testing of the package within this manuscript was performed using a
-subset of the authors collections from 2018-2022, while most development
-was performed on their 2023 and 2024 collections. Only collections which
-had identifications to the level of species or lower, and transcribed
-collection dates and coordinates were used for most functionality. In
-total 981 records were used for testing various functions, these records
-were from 234 sites located across Western North America (Figure 2). In
-total this data set had 729 species (with 558 distinct sets of authors),
-with 83 infraspecies (22 authorships) in 74 families.
-
-BarnebyLives took roughly four minutes (222.886sec) to run all local
-steps, and roughly ten minutes (584.05sec) to search Plants of the World
-Online for preferred synonyms, and a minute 63.945sec to search Google
-Maps and write directions to sites.
-
-Most of the local run time is attributable to the spatial (205.254sec),
-and taxonomic operations (17.253sec), while formatting data for labels
-took 0.379sec. The spell check of the scientific name accounted for
-nearly all of the time (17.006sec) spent performing local taxonomic
-operations. The generation of labels consumed around eight minutes
-(509.931sec) for the rendering, and an additional 58.60sec to combine
-the 182 sheets to a single Portable Document Format (PDF). The total
-label generation run time for processing these 729 collections was 15
-minutes. In total the 729 collections, which underwent all processing
-steps, took 24 minutes to process.
-
-\subsection{RESULTS}\label{results}
-
-Even on data which had been manually cleaned and error-checked by a
-human several times BarnebyLives was able to reduce transcription
-errors, identify typos, make nomenclature suggestions, and reformat text
-elements for downstream use. While none of the 74 families were
-misspelled, BarnebyLives made 25 suggestions on naming, identified 6
-instances where the user entered an unequivocally incorrect family (or
-taxonomic entity), identified 5 records where families were autofilled,
-and 1 instance where an outdated circumscription was applied. At the
-level of family BarnebyLives flagged 6 records where the author follows
-an alternative taxonomy, and flagged 7 records in error, it appears most
-of these errors are due to issues in the backbone used by the earlier
-spell check function.
-
-In the 326 genera analysed BarnebyLives identified 75 discrepancies at
-the level of genus between user submitted and processed data. In 42 of
-these instances the user supplied an outdated name (21 unique genera)
-flagged 4 records where the author follows an alternative taxonomy (2
-genera total), and flagged 2 record in error.
-
-Of 729 distinct species analysed BarnebyLives flagged 61 records, and
-detected 33 instances of misspelled epithets (33 unique species). In 15
-of these instances the user supplied an outdated name (15 unique
-species). It also flagged 2 records where the author follows an
-alternative taxonomy (2 unique species), and flagged 8 records in error.
-The final record was an egregious error where the order of the specific
-epithet and the genus name.
-
-5 records were appropriately flagged for issues with auto fill increment
-of the longitude value, and 3 records were also auto-flagged for
-increases in latitude values.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/tables/Screenshot-DataSources.png}
-\caption{Data Sources}
-\end{figure}
-
-\section{DISCUSSION}\label{discussion}
-
-While numerous tools have been developed for cleaning existing herbarium
-and museum records, few tools help to ensure that the data entered are
-accurate (\citeproc{ref-patten2024geographic}{Patten et al., 2024}). We
-argue that the original collectors are the most qualified individuals to
-perform quality control checks and that BarnebyLives allows them to
-assume that responsibility in a relatively fast and streamlined format.
-By utilizing both R and LaTeX and having publicly available source code
-on Github, this program allows users immediate familiarity with the
-system for troubleshooting issues and implementing upgrades and
-modifications in project branches.
-
-LaTeX, a software system used for typesetting, allows users to focus on
-the content rather than the style of the documents rendered from it.
-However, using its default settings, it can produce aesthetically
-pleasing results (Figure 4). Additionally LaTeX offers users a wide
-variety of ways which they can modify labels which are under-explored in
-the package. Very good documentation of LaTeX capabilities is offered in
-multiple areas; for instance, via the
-\href{https://www.overleaf.com/learn}{Overleaf} project. While the
-templates in the package are quite simple, LaTeX also offers the ability
-to use custom fonts, to alter font weights and colors, alter line
-spacing, to include images (e.g.~dot maps) and customize labels beyond
-what the default templates support.
-
-Thematically, BarnebyLives is set up to cover Western North America.
-However, the package supports the use of a `domain' being drawn over any
-of the conterminous United States. Several of the attributes which it
-collects and displays on labels, relate to topics which more senior
-curators are interested in, i.e.~the administrative information on
-Township Section and Range (or `TRS'), but are considered less value in
-other geographic regions.\\
-Further several of the abiotic variables which it acquires information
-on: slope, aspect, and geology have long been considered prominent
-drivers of plant distributions in semi-arid and montane systems and
-warranted on a label in these types of systems, whereas curators in
-other regions may find this information superfluous. Finally, it is
-plausible people in other geographic areas are less interested in
-displaying which land management agency has jurisdiction over a
-collection; however in the west we believe this is useful information
-which may help a collector interested in revisiting a site to determine
-if they will require permits for access or to make new collections.
-
-Accessioning often relies on the use of the Microsoft Office suite of
-programs and may utilize other costly software such as ArcPro or Adobe
-Acrobat. While BarnebyLives does not have its own graphic user
-interface, the functionality of commonly used Interactive Development
-Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now
-offer functionality to readily view and filter datasets using familiar
-spreadsheet-like formats, making them more accessible to many users.
-While other software often cost money, these are also free, and we
-recommend that users install an open-source PDF viewer such as Okular to
-review their rendered documents.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth,height=\textheight]{../graphics/plots/RCB2759.png}
-\caption{A label generated from a default template}
-\end{figure}
-
-\section{CONCLUSIONS}\label{conclusions}
-
-BarnebyLives is an R package that can be used to rapidly acquire
-relevant geographic and taxonomic data. It can also perform specialized
-spell checks and assorted curatorial tasks to produce both digital and
-analog data. The package relies on no licensed software, such as the
-Microsoft Office suite, and is suitable for install on all major
-operating systems (Windows, Mac, Linux), however currently label
-generation support is only offered on Linux and Mac, with a small amount
-of use of the command line, which may be called from the Rstudio rather
-than a `traditional' terminal.
-
-\section{AUTHOR CONTRIBUTIONS}\label{author-contributions}
-
-The project was conceptualized by R.C.B. The program was written by
-R.C.B. Data collection and analysis were performed by R.C.B. R.C.B. \&
-J.B.F wrote the manuscript, and both authors approved the final version
-of the manuscript.
-
-\section{ACKNOWLEDGEMENTS}\label{acknowledgements}
-
-The Bureau of Land Management is gratefully acknowledged as a provider
-of funding to R.C.B. for most of his specimen collection activities. We
-thank the two anonymous peer reviewers who have increased the quality of
-this manuscript. Several prominent associated collectors of specimens
-used in this study are thanked: Dani Yashinovitz, Hannah Lovell, Dakota
-Becerra, Caitlin Miller, Hubert Szczygiel.
-
-\section{DATA AVAILABILITY STATEMENT}\label{data-availability-statement}
-
-The BarnebyLives R package is open source, the development version is
-available on GitHub (\url{https://github.com/sagesteppe/BarnebyLives}).
-The package includes three real use-case vignettes (tutorials) available
-on a Github Pages site
-(\url{https://sagesteppe.github.io/BarnebyLives/}). The first vignette
-\emph{``Preparing to use BarnebyLives!''} shows how to set up an
-instance for a certain geographic area (domain). The next two vignettes
-\emph{``BarnebyLives! Running pipeline''} showcases the usage of the
-package for processing data entered on a spreadsheet, and
-\emph{``Printing herbarium labels and exporting a digital copy of
-data''} how to export data in both digital and analog formats.
-\emph{``Custom label templates''} shows how to customize labels in
-LaTeX, and \emph{``Rendering a shipping manifest''} details how to
-produce a shipping manifest for gifting or transferring material to an
-herbarium. All data used in this manuscript are available at:
-\url{https://github.com/sagesteppe/Barneby_Lives_dev/manuscript}.
-
-\singlespacing
-
-\section{ORCID}\label{orcid}
-
-Reed Clark Benkendorf \url{https://orcid.org/0000-0003-3110-6687}\\
-Jeremie Fant \url{https://orcid.org/0000-0001-9276-1111}
-
-\small
-
-\section{REFERENCES}\label{references}
-
-\clearpage
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{1}
-\bibitem[\citeproctext]{ref-barrows2016crossroads}
-Barrows, C. W., M. L. Murphy-Mariscal, and R. R. Hernandez. 2016. At a
-crossroads: The nature of natural history in the twenty-first century.
-\emph{BioScience} 66: 592--599.
-
-\bibitem[\citeproctext]{ref-borges2020schrodinger}
-Borges, L. M., V. C. Reis, and R. Izbicki. 2020. Schrodinger's
-phenotypes: Herbarium specimens show two-dimensional images are both
-good and (not so) bad sources of morphological data. \emph{Methods in
-Ecology and Evolution} 11: 1296--1308.
-
-\bibitem[\citeproctext]{ref-brewer2019factors}
-Brewer, G. E., J. J. Clarkson, O. Maurin, A. R. Zuntini, V. Barber, S.
-Bellot, N. Biggs, et al. 2019. Factors affecting targeted sequencing of
-353 nuclear genes from herbarium specimens spanning the diversity of
-angiosperms. \emph{Frontiers in plant science} 10: 1102.
-
-\bibitem[\citeproctext]{ref-daru2018widespread}
-Daru, B. H., D. S. Park, R. B. Primack, C. G. Willis, D. S. Barrington,
-T. J. Whitfeld, T. G. Seidler, et al. 2018. Widespread sampling biases
-in herbaria revealed from large-scale digitization. \emph{New
-Phytologist} 217: 939--955.
-
-\bibitem[\citeproctext]{ref-davis2023herbarium}
-Davis, C. C. 2023. The herbarium of the future. \emph{Trends in Ecology
-\& Evolution} 38: 412--423.
-
-\bibitem[\citeproctext]{ref-forman1989herbarium}
-Forman, L., and D. Bridson. 1989. The herbarium handbook. Royal Botanic
-Gardens Kew.
-
-\bibitem[\citeproctext]{ref-funk2014erosion}
-Funk, V. A. 2014. The erosion of collections-based science: Alarming
-trend or coincidence. \emph{The Plant Press} 17: 1--13.
-
-\bibitem[\citeproctext]{ref-usgs2024padus}
-Gap Analysis Project (GAP), U. S. G. S. (USGS). 2024. Protected areas
-database of the united states (PAD-US) 4.0.
-
-\bibitem[\citeproctext]{ref-govaerts2021world}
-Govaerts, R., E. Nic Lughadha, N. Black, R. Turner, and A. Paton. 2021.
-The world checklist of vascular plants, a continuously updated resource
-for exploring global plant diversity. \emph{Scientific data} 8: 215.
-
-\bibitem[\citeproctext]{ref-greve2016realising}
-Greve, M., A. M. Lykke, C. W. Fagg, R. E. Gereau, G. P. Lewis, R.
-Marchant, A. R. Marshall, et al. 2016. Realising the potential of
-herbarium records for conservation biology. \emph{South African Journal
-of Botany} 105: 317--323.
-
-\bibitem[\citeproctext]{ref-gries2014symbiota}
-Gries, C., M. E. E. Gilbert, and N. M. Franz. 2014. Symbiota--a virtual
-platform for creating voucher-based biodiversity information
-communities. \emph{Biodiversity data journal}.
-
-\bibitem[\citeproctext]{ref-hitchcock2018flora}
-Hitchcock, C. L., and A. Cronquist. 2018. Flora of the pacific
-northwest: An illustrated manual. University of Washington Press.
-
-\bibitem[\citeproctext]{ref-holmgren2017}
-Holmgren, N., and P. Holmgren. 1988. Intermountain flora v. 7. The New
-York Botanical Garden Press, New York.
-
-\bibitem[\citeproctext]{ref-james2018herbarium}
-James, S. A., P. S. Soltis, L. Belbin, A. D. Chapman, G. Nelson, D. L.
-Paul, and M. Collins. 2018. Herbarium data: Global biodiversity and
-societal botanical needs for novel research. \emph{Applications in plant
-sciences} 6: e1024.
-
-\bibitem[\citeproctext]{ref-manzano2021fair}
-Manzano, S., and A. C. Julier. 2021. How FAIR are plant sciences in the
-twenty-first century? The pressing need for reproducibility in plant
-ecology and evolution. \emph{Proceedings of the Royal Society B} 288:
-20202597.
-
-\bibitem[\citeproctext]{ref-marsico2020small}
-Marsico, T. D., E. R. Krimmel, J. R. Carter, E. L. Gillespie, P. D.
-Lowe, R. McCauley, A. B. Morris, et al. 2020. Small herbaria contribute
-unique biogeographic records to county, locality, and temporal scales.
-\emph{American journal of botany} 107: 1577--1587.
-
-\bibitem[\citeproctext]{ref-mishler2020spatial}
-Mishler, B. D., R. Guralnick, P. S. Soltis, S. A. Smith, D. E. Soltis,
-N. Barve, J. M. Allen, and S. W. Laffan. 2020. Spatial phylogenetics of
-the north american flora. \emph{Journal of Systematics and Evolution}
-58: 393--405.
-
-\bibitem[\citeproctext]{ref-nanglu2023nature}
-Nanglu, K., D. de Carle, T. M. Cullen, E. B. Anderson, S. Arif, R. A.
-Castañeda, L. M. Chang, et al. 2023. The nature of science: The
-fundamental role of natural history in ecology, evolution, conservation,
-and education. \emph{Ecology and Evolution} 13: e10621.
-
-\bibitem[\citeproctext]{ref-patten2024geographic}
-Patten, N. N., M. L. Gaynor, D. E. Soltis, and P. S. Soltis. 2024.
-Geographic and taxonomic occurrence r-based scrubbing (gatoRs): An r
-package and workflow for processing biodiversity data.
-\emph{Applications in Plant Sciences} 12: e11575.
-
-\bibitem[\citeproctext]{ref-perkins2020Plabel}
-Perkins, K. 2020. Plabel.
-
-\bibitem[\citeproctext]{ref-powo2024}
-POWO. 2024. Geographic names information system (GNIS) - USGS national
-map downloadable data collection: U.s. Geological survey.
-
-\bibitem[\citeproctext]{ref-prather2004decline}
-Prather, L. A., O. Alvarez-Fuentes, M. H. Mayfield, and C. J. Ferguson.
-2004. The decline of plant collecting in the united states: A threat to
-the infrastructure of biodiversity studies. \emph{Systematic Botany} 29:
-15--28.
-
-\bibitem[\citeproctext]{ref-pyke2010biological}
-Pyke, G. H., and P. R. Ehrlich. 2010. Biological collections and
-ecological/environmental research: A review, some observations and a
-look to the future. \emph{Biological reviews} 85: 247--266.
-
-\bibitem[\citeproctext]{ref-ronsted2020integrative}
-Rønsted, N., O. M. Grace, and M. A. Carine. 2020. Integrative and
-translational uses of herbarium collections across time, space, and
-species. \emph{Frontiers in Plant Science} 11: 1319.
-
-\bibitem[\citeproctext]{ref-snethlage2022hierarchical}
-Snethlage, M. A., J. Geschke, A. Ranipeta, W. Jetz, N. G. Yoccoz, C.
-Körner, E. M. Spehn, et al. 2022. A hierarchical inventory of the
-world's mountains for global comparative mountain science.
-\emph{Scientific data} 9: 149.
-
-\bibitem[\citeproctext]{ref-gnis2024}
-Survey, U. S. G. 2023. Geographic names information system (GNIS) - USGS
-national map downloadable data collection: U.s. Geological survey.
-
-\bibitem[\citeproctext]{ref-ipni2024}
-The Royal Botanic Gardens, H. U. H. \&. L., Kew, and A. N. Herbarium.
-2024. International plant names index.
-
-\bibitem[\citeproctext]{ref-thiers2021herbaria}
-Thiers, B. M. 2021. The world's herbaria 2021: A summary report based on
-data from index herbarium.
-
-\bibitem[\citeproctext]{ref-tosa2021rapid}
-Tosa, M. I., E. H. Dziedzic, C. L. Appel, J. Urbina, A. Massey, J.
-Ruprecht, C. E. Eriksson, et al. 2021. The rapid rise of next-generation
-natural history. \emph{Frontiers in Ecology and Evolution} 9: 698131.
-
-\bibitem[\citeproctext]{ref-walker2022tigris}
-Walker, K. 2024.
-\href{https://CRAN.R-project.org/package=tigris}{Tigris: Load census
-TIGER/line shapefiles}.
-
-\bibitem[\citeproctext]{ref-welsh2001rupert}
-Welsh, S. L. 2001. Rupert c. Barneby (1911-2000). \emph{Taxon}.
-
-\bibitem[\citeproctext]{ref-woodland2007botanists}
-Woodland, D. W. 2007. Are botanists becoming the dinosaurs of biology in
-the 21st century? \emph{South African Journal of Botany} 73: 343--346.
-
-\end{CSLReferences}
-
-\section{SUPPORTING INFORMATION}\label{supporting-information}
-
-Additional supporting information can be found online in the Supporting
-Information section at the end of this article.
-
-\textbf{Appendix S1.} A table of all time trials for each function.
+\includegraphics[width=\linewidth,height=0.66\textheight,keepaspectratio]{~/Documents/assoRted/Barneby_Lives_dev/LabelStyles/map_2823.png}
+\end{minipage}
 

### 2025-09-01 LabelStyles/skeleton-maps.tex
map example

--- 2025-09-01 LabelStyles/custom-font2823.tex
@@
 \usepackage{setspace}
+\usepackage{wrapfig}
 \ifLuaTeX
@@
 \begin{center}
-\textbf{\Large Custom Fonts}
+\textbf{\Large Label with Map}
 \end{center}
@@
 
-\noindent
-
-\begin{minipage}[t]{0.78\textwidth}\vspace*{0pt}
-
 \begingroup
 \large
-*Lomatium nevadense* (S. Watson) J.M. Coult. & Rose     
-Fide: *Intermountain Flora*, det.: RCB, 29 Aug, 2023.
-\normalsize
+
+\emph{Lomatium nevadense} (S. Watson) J.M. Coult. \& Rose\\
+\normalsize Fide: \emph{Intermountain Flora}, det.: RCB, 29 Aug, 2023.
 \endgroup
+
+\begin{wrapfigure}[5]{r}[0pt]{1in} % wrap text around fig, [5] lines of text,
+% {r} align, [0pt] no additional spacing, {1in} image
+\vspace{-5.0ex} % push the image upwards above it's starting line. 
+\includegraphics[width=1in,height=1in,keepaspectratio]{~/Documents/assoRted/Barneby_Lives_dev/LabelStyles/maps/map_2823.png} % path to image
+\end{wrapfigure}
 
@@
 \begingroup
-U.S.A., Nevada, Washoe Co., 17N 20E 19. Washoe Summit. 39.32677 -119.78537 (WGS84 5m).
 
-Granitoid derived aeolian loamy sand soils. At 5,200 ft, on a slope, 4° slo. 57° asp.; geology: Igneous and Sedimentary, undifferentiated.  
-Veg.: *Artemisia tridentata* var. *wyomimgensis, Poa secunda, Agropyron cristatum, Achnatherum thurberianum, Purshia tridenata*. Ass.: *Bromus tectorum, Phlox longifolia*.
+U.S.A., Nevada, Washoe Co., 17N 20E 19. Washoe Summit. 39.32677
+-119.78537 (WGS84 5m).
+
+Granitoid derived aeolian loamy sand soils. At 5,200 ft, on a slope, 4°
+slo. 57° asp.; geology: Igneous and Sedimentary, undifferentiated.
+
+\par
+
+Veg.: \emph{Artemisia tridentata} var. \emph{wyomimgensis, Poa secunda,
+Agropyron cristatum, Achnatherum thurberianum, Purshia tridenata}. Ass.:
+\emph{Bromus tectorum, Phlox longifolia}.
 
@@
 \begingroup
-Reed Clark Benkendorf 2823; 6 May, 2023
-\endgroup
-\end{minipage}
 
-\% \hfill
-
-\begin{minipage}[c]{0.18\textwidth}\vspace*{0pt}
-\hspace*{4mm}% left margin
-\centering
-\includegraphics[width=\linewidth,height=0.66\textheight,keepaspectratio]{~/Documents/assoRted/Barneby_Lives_dev/LabelStyles/map_2823.png}
-\end{minipage}
+Reed Clark Benkendorf 2823; 6 May, 2023 \endgroup
 

### 2025-09-01 manuscript/DataSourceTable.R

@@ -1,5 +1,7 @@
 library(gt)
-setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
+library(tidyverse)
+# setwd('/media/steppe/hdd/Barneby_Lives-dev/manuscript')
+setwd('/home/sagesteppe/Documents/assoRted/Barneby_Lives_dev/manuscript')
 
 footnote = 'Counties and States are merged into the same dataset while setting up the package. The value for "County" includes State.'
 
@@ -61,17 +63,67 @@ data_sources <- local_data |>
     Subset_Size = "Size (GiB)") |>  
   tab_footnote(footnote) |> 
   
-  tab_options(heading.background.color = "#94524A",
-              heading.border.lr.color='#2F0A28', 
-              table.background.color='#eff4ec',
-              row.striping.background_color = 'black',
-              footnotes.background.color = '#A27E6F',
-              heading.padding = 0)
+  tab_options(
+    heading.background.color = "#94524A",
+    heading.border.lr.color='#2F0A28', 
+    table.background.color='#eff4ec',
+    row.striping.background_color = 'black',
+    footnotes.background.color = '#A27E6F',
+    heading.padding = 0
+    )
 
 
 gtsave(data_sources, "../graphics/tables/rab_tab_examples.png")
 
-
-
-
 sessioninfo::session_info()
+
+## and create a second table for the alternative data sources page. 
+
+data_dropins <- select(local_data, -Subset_Size) |>
+  mutate(
+    across(.cols = everything(), ~ na_if(.x, ''))
+    ) |> 
+  fill(everything()) |>
+  filter(Source %in% c('US Census Bureau', 'US Geological Survey')) |>
+  mutate(
+    Alternative = case_when(
+      Variable == 'County' ~ 'geoBoundaries', 
+      Variable == 'State' ~ 'geoBoundaries',
+      Variable == 'Surficial Geology' ~ 'Macrostrat',  # NOTE API, requests, not local. 
+      Variable == 'Place Names' ~ 'GeoNames',  # NOTE API, requests, not local. 
+      .default = as.character('?')
+    )
+  ) |>
+  mutate(
+    Status = case_when(
+      Alternative == 'geoBoundaries' ~ 'Downloadable', 
+      Alternative == 'GeoNames' ~ 'Downloadable',
+      Alternative == 'Macrostrat' ~ 'API Calls',
+      .default = as.character('?')
+    ),
[…]

### 2025-09-20 LabelStyles/SoS-skeleton-Spaced.Rmd
working on vignette example
reviewer comments on linux mac

--- 2025-04-08 LabelStyles/SoS-skeleton-Det.Rmd
@@
     Collection_number = as.numeric(Collection_number), 
-    Coordinate_Uncertainty = '+/- 5m') |>
+    Coordinate_uncertainty = '+/- 5m') |>
   dplyr::filter(Collection_number == params$Collection_number)
@@
 \normalsize
-`r writer_fide(data)`
 \endgroup
@@
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).
+`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
 

### 2025-09-20 LabelStyles/SoS-skeleton-large.Rmd
working on vignette example
reviewer comments on linux mac

--- 2025-04-08 LabelStyles/SoS-skeleton-Det.Rmd
@@
 \begin{center}
-\textbf{\LARGE Seeds of Success}
+\textbf{\large Seeds of Success}
 \end{center}

### 2025-09-20 LabelStyles/SoS-skeleton-noDir.Rmd
working on vignette example
reviewer comments on linux mac

--- 2025-03-30 manuscript/labels/skeleton.Rmd
@@
 ```{r load data}
-data <- read.csv('../data/processed/cleaned_data.csv') |>
+data <- read.csv('SoS-ExampleCollection.csv') |>
   dplyr::mutate(
     Collection_number = as.numeric(Collection_number), 
-    Coordinate_Uncertainty = '+/- 5m') |>
+    Coordinate_uncertainty = '+/- 5m') |>
   dplyr::filter(Collection_number == params$Collection_number)
@@
 \endgroup
+
+\begin{center}
+{\footnotesize Collected under the auspices of the Bureau of Land Management}
+\end{center}

### 2025-09-20 LabelStyles/skeleton-research.Rmd
reviewer comments on linux mac

---
output: pdf_document
params:
  Collection_number: NULL
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r load data, eval = T}
library(BarnebyLives)
local <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
  dplyr::filter(Collection_number == params$Collection_number)

# escape character columns to ensure characters which interfere with latex
# rendering are escaped. 
char_cols <- sapply(record, is.character)
char_col_names <- names(record)[char_cols]
record[char_col_names] <- lapply(record[char_col_names], escape_latex)
```

\begin{center}
\textbf{\Large `r writer(record[['Project_name']])`}
\end{center}
\uppercase{`r record[['Family']]`}

\begingroup
\normalsize
`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`  
\endgroup

\vspace{\fill}
\begingroup
\small
`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).

`r writer(record[['Habitat']])`. `r record[['physical_environ']]`  
Veg.: `r species_font(record[['Vegetation']])` 
`r associates_writer(record[['Associates']], full = TRUE)`
\endgroup

\vspace{\fill}
\begingroup
`r collection_writer(record)`
`r writer_fide(record)`
\endgroup

\begingroup
\scriptsize
`r writer(record[['Research_notes']])`  
\endgroup

### 2025-09-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
map example
reviewer comments on linux mac

@@
 While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images.
-Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
+Of the two, specimens alone can provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
 These factors will ensure that the specimens remain the premier botanical data source into perpetuity.
@@
 Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota].
-The office suite costs money, and in our experience, can be problematic to set up; further, its functionality ends with label creation. 
+The office suite costs money, and in our experience, can be difficult to set up; further, its functionality ends with label creation. 
 PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality.
 However, beyond verifying the nations of collection it does not include data cleaning functionalities. 
-While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
 The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems.
@@
 It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
-The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
+The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Mac and Linux, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
 

### 2025-10-12 valleys2/NameValleys.R
refresh on valley naming problem

@@ -1,0 +1,376 @@
+setwd('~/Documents/assoRted/Barneby_Lives_dev/valleys2')
+
+library(sf)
+library(spdep)
+library(tidyverse)
+set.seed(999)
+
+valleys <- st_read('valleys.gpkg') |>
+  st_cast('POLYGON') |>
+  st_transform(5070) |>
+  mutate(ID = 1:n())
+
+sizes <- as.numeric(units::set_units(st_area(valleys), "km^2"))
+table(sizes > 20)
+
+large_valleys <- valleys[sizes > 20,]
+st_write(large_valleys, 'large_valleys.gpkg', append = FALSE)
+
+
+bbv <- st_bbox(valleys)
+
+GeoNames_names <- st_read('valleys_glimpse.gpkg') |>
+  st_transform(5070) |>
+  st_crop(bbv)
+
+total_voron <- st_voronoi( st_union(GeoNames_names))
+total_voron <- st_collection_extract(total_voron, "POLYGON") |>
+  st_as_sf() |>
+  rename(geometry = x) |>
+  st_crop(bbv)
+
+total_voron_attributed <- total_voron |>
+  st_join(GeoNames_names)
+
+rm(total_voron, bbv)
+
+ggplot() +
+  geom_sf(data = total_voron_attributed, fill = NA) +
+  geom_sf(data = valleys, col = 'red')
+
+
+################################################################################
+
+# match one which HU10 watershed has a valley fall in it. 
+
+
+## identify which watersheds have valleys assigned for them 
+ints <- st_intersects(valleys, GeoNames_names)
+matched_single <- st_drop_geometry(valleys)[ lengths(ints) == 1, 'ID']
+
+valleys_single_match <- valleys |>
+  filter(ID %in% matched_single) |>
+  st_join(GeoNames_names, join =  st_contains)
+
+## identify the watersheds with > 1 valley name
+valleys_sub <- valleys[ lengths(ints) >= 2, ]
+
+################################################################################
+
+# match two, multiple valley labels per watersheed simpply split by closest label. 
+
+## these valleys get split by the voronoi, and names assigned based on voronoi intersections
+matched_multi <- st_drop_geometry(valleys)[ lengths(ints) >= 2, 'ID']
+
+valleys_multi_match <- valleys |>
+  filter(ID %in% matched_multi) |>
+  st_join(GeoNames_names, join =  st_contains)
+
+
[…]

### 2025-10-12 valleys2/Valleys_simplifcation.R

@@ -1,0 +1,77 @@
+setwd('~/Documents/assoRted/Barneby_Lives_dev/valleys2')
+
+library(sf)
+library(tidyverse)
+library(rmapshaper)
+library(tigris)
+
+v1 <- st_read(file.path('valleys', 'valleys.shp')) |>
+  st_cast('POLYGON') |>
+  mutate(TYPE = 'VALLEY') |>
+  st_transform(4326)
+
+st <- tigris::states()
+st_s <- filter(st, STUSPS %in% c('CA', 'OR', 'WA', 'NV', 'ID', 'MT', 'CO', 'AZ', 'NM', 'WY', 'UT')) |>
+  st_combine() |>
+  st_make_valid() |>
+  st_transform(4326)
+
+v2 <- st_read(file.path('valleys', 'valleys-heartland.shp')) |>
+  st_cast('POLYGON') |>
+  mutate(TYPE = 'VALLEY') |>
+  st_transform(4326)
+
+v2 <- v2[unlist(st_intersects(st_s, v2)),]
+
+rm(st, st_s)
+# COMBINE
+valleys <- bind_rows(v1, v2) 
+
+# rm(v1, v2)
+
+# NOW READ IN GMBA BEFORE SIMPLIGYING GEOMETRY
+
+mt <- st_read(file.path('mountains', 'mountains.shp')) |>
+  mutate(TYPE = 'MOUNTAIN') |>
+  select(-Mountains) |>
+  st_transform(4326)
+
+data <- bind_rows(v1, mt)
+
+# keep compute low, run the pieces separately
+data <- st_make_valid(data)
+data_0.5 <- ms_simplify(data, keep = 0.025, keep_shapes = TRUE) 
+data_0.5 <- filter(data_0.5, TYPE == 'VALLEY')
+data_0.5 <- data_0.5[st_geometry_type(data_0.5) %in%   c("POLYGON", "MULTIPOLYGON"), ]
+data_0.5 <- st_make_valid(data_0.5)
+data_v1 <- st_cast(data_0.5, 'POLYGON')
+
+data <- bind_rows(v2, mt)
+
+data <- st_make_valid(data)
+data_0.5 <- ms_simplify(data, keep = 0.025, keep_shapes = TRUE) 
+data_0.5 <- filter(data_0.5, TYPE == 'VALLEY')
[…]

### 2025-10-12 LabelStyles/SoS-skeleton.Rmd
working on vignette example

--- 2025-04-08 LabelStyles/SoS-skeleton-Det.Rmd
@@
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_Uncertainty)`).
+`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
 
 `r writer(data$Habitat)`. `r data$physical_environ`  
-
 Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
 `r writer(data$Notes)`  
+\endgroup
+\begingroup
+\begin{spacing}{0.1}
+{\scriptsize `r data$Directions_BL`}
+\end{spacing}
 \endgroup

### 2025-10-12 LabelStyles/skeleton-logo.Rmd

---
output: 
  pdf_document:
    latex_engine: xelatex # or lualatex
    keep_tex: true
params:
  Collection_number: NULL
geometry: paperheight=4in,paperwidth=4in,margin=0.1in
header-includes:
  - \pagenumbering{gobble}
  - \usepackage{setspace}
  - \usepackage{wrapfig}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r load library, eval = T}
library(BarnebyLives)

local <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
logopath <- file.path(local, 'logo.png')

record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |> # default feed in the path to your csv. 
  dplyr::filter(Collection_number == params$Collection_number) |>  
  sf::st_drop_geometry() |>
  dplyr::mutate(Project_name = 'Logos', Coordinate_uncertainty = ' +/- 5m')

# escape character columns to ensure characters which interfere with latex
# rendering are escaped. 
char_cols <- sapply(record, is.character)
char_col_names <- names(record)[char_cols]
record[char_col_names] <- lapply(record[char_col_names], escape_latex)
```

\begin{center}
\textbf{\Large `r writer(record[['Project_name']])`}
\end{center}
\begin{flushright}
\uppercase{`r record[['Family']]`}
\end{flushright}

\begingroup
\large
`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)`  `r writer(record[['Infraspecific_authority']])`
\normalsize
\endgroup

\vspace{\fill}

\begingroup
`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).

`r writer(record[['Habitat']])`. `r record[['physical_environ']]` \par
Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
`r writer(record[['Notes']])`  
\endgroup

\vspace{\fill}

\noindent
\begin{minipage}[t]{0.75\textwidth}
  \begingroup 
  `r collection_writer(record)`
  `r writer_fide(record)`
  \endgroup
\end{minipage}
\hfill
\begin{minipage}[b]{0.2\textwidth}
  \includegraphics[width=\linewidth,keepaspectratio]{`r logopath`}
\end{minipage}

### 2025-10-12 LabelStyles/skeleton-modern.Rmd

--- 2025-10-12 LabelStyles/skeleton-logo.Rmd
@@
   - \usepackage{setspace}
-  - \usepackage{wrapfig}
+  - \usepackage{fontspec}
+  - \setmainfont{Ubuntu Mono}
 ---
@@
 
-```{r load library, eval = T}
+```{r load library}
 library(BarnebyLives)
+```
 
-local <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
-logopath <- file.path(local, 'logo.png')
+```{r load data, eval = T}
+local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
 
-record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |> # default feed in the path to your csv. 
-  dplyr::filter(Collection_number == params$Collection_number) |>  
-  sf::st_drop_geometry() |>
-  dplyr::mutate(Project_name = 'Logos', Coordinate_uncertainty = ' +/- 5m')
+record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
+  dplyr::filter(Collection_number == params$Collection_number)
 
@@
 \end{flushright}
-
 \begingroup
@@
 \vspace{\fill}
+\begingroup
+`r record[['Gen']]`. `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).  `r writer(record[['Site']])`
 
-\begingroup
-`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
-
-`r writer(record[['Habitat']])`. `r record[['physical_environ']]` \par
-Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Habitat']])`. `r record[['physical_environ']]`  
+Vegetation: `r species_font(record[['Vegetation']])` 
+`r associates_writer(record[['Associates']], full = TRUE)` 
 `r writer(record[['Notes']])`  
@@
 \vspace{\fill}
-
-\noindent
-\begin{minipage}[t]{0.75\textwidth}
 \begingroup
-  `r collection_writer(record)`
+`r collection_writer(record)`.  
 `r writer_fide(record)`
 \endgroup
-\end{minipage}
-\hfill
-\begin{minipage}[b]{0.2\textwidth}
-  \includegraphics[width=\linewidth,keepaspectratio]{`r logopath`}
-\end{minipage}

### 2025-10-12 valleys2/pull_hu10_for_valleys_project_oct10_2025.qmd
Reworked Valleys dataset for BL

---
title: "Pull hu10"
format: html
editor: visual
---

```{r}
library(sf)
library(tidyverse)
library(rmapshaper)
getwd()
```

```{r, eval = F}
p <- '/media/steppe/hdd/Geospatial_data/NHDPlusNationalData/NHDPlusV21_National_Seamless_Flattened_Lower48.gdb'
st_layers(p)
hu10 <- st_read(p, layer = 'HUC12') |>
  st_make_valid()

domain <- st_read('/media/steppe/hdd/Barneby_Lives-dev/ValleysProject/data/domain.gpkg') |>
  st_transform(st_crs(hu10))

hu10 <- hu10[ unlist(st_intersects(domain, hu10)),] |>
  st_make_valid()

mountains <- st_read('/media/steppe/hdd/Barneby_Lives-dev/geodata/mountains') |>
  st_make_valid() |>
  st_transform(st_crs(hu10)) |>
  st_union()

mountains <- sf::st_set_precision(mountains, precision = 1e3) |>
  st_make_valid()
hu10 <- sf::st_set_precision(hu10, precision = 1e3)|>
  st_make_valid()

hu10 <- st_difference(hu10, mountains)
hu10 <- st_make_valid(hu10)

polys <- hu10[complete.cases(
  sf::st_coordinates(sf::st_centroid(sf::st_geometry(hu10)))
), ]

polys <- polys %>%
  sf::st_collection_extract("POLYGON")

polys <- polys %>%
    st_make_valid() %>%                     # Step 1: Repair invalid geometries
    st_cast("MULTIPOLYGON", warn = FALSE) %>%  # Step 2: Ensure MULTIPOLYGON type
    st_cast("POLYGON", warn = FALSE) %>%       # Step 3: Flatten to POLYGONs
    filter(!st_is_empty(.))           # Final: Drop empty geometries

st_write(polys, '/media/steppe/hdd/Barneby_Lives-dev/ValleysProject/data/Main_huc12_polygons.gpkg', append = F)
#hu10 <- rmapshaper::ms_simplify(hu10, keep = 0.05, keep_shapes = TRUE)
```

```{r}
polys <- st_read(
  '/media/steppe/hdd/Barneby_Lives-dev/ValleysProject/data/Main_huc12_polygons.gpkg') |>
  sf::st_make_valid()

polys_simp <- rmapshaper::ms_simplify(polys, keep = 0.1, keep_shapes  = TRUE)
polys_simp <- polys_simp[!st_is_empty(polys_simp),]

st_write(
  polys_simp, 
  '/media/steppe/hdd/Barneby_Lives-dev/ValleysProject/data/Simplified_huc12_polygons.gpkg', 
  append = F)
```

### 2025-10-12 valleys2/NameValleys.R

@@ -55,322 +55,105 @@ valleys_single_match <- valleys |>
 ## identify the watersheds with > 1 valley name
 valleys_sub <- valleys[ lengths(ints) >= 2, ]
 
-################################################################################
-
-# match two, multiple valley labels per watersheed simpply split by closest label. 
-
-## these valleys get split by the voronoi, and names assigned based on voronoi intersections
-matched_multi <- st_drop_geometry(valleys)[ lengths(ints) >= 2, 'ID']
-
-valleys_multi_match <- valleys |>
-  filter(ID %in% matched_multi) |>
-  st_join(GeoNames_names, join =  st_contains)
 
 
-valleys_multi_voronoi <- valleys |>
-  filter(ID %in% matched_multi) |>
-  st_intersection(total_voron_attributed) 
+#################################################################################
 
-# now remove these old geometries from the data set and replace them with these. 
+## using tagged data
 
-valleys_partial_label <- valleys |>
-  filter(!ID %in% c(matched_single, matched_multi)) |>
-  bind_rows(valleys_single_match, valleys_multi_voronoi) |>
-  arrange(ID)
+twelve_geometry <- sf::st_read(file.path('.', 'data', 'Simplified_huc12_polygons.gpkg')) |>
+  st_make_valid()
+twelve_geometry <- filter(twelve_geometry, st_is_valid(twelve_geometry))
 
-rm(matched_multi, matched_single, valleys_sub, valleys_multi_voronoi, valleys_single_match,
-   ints, total_voron_attributed)
+tagged <- bind_rows(
+  sf::st_read(file.path('.', 'valleys',  'Valleys_oct11_2025.gpkg'), quiet = T)|>
+    filter(!is.na(Valley)),
+  sf::st_read(file.path('.', 'valleys', 'valleys-heartland.shp'), quiet = T) |>
+    filter(!is.na(Valley)) |>
+    rename(geom = geometry)
+) |>
+  st_cast('POLYGON') |>
+  st_make_valid()
 
-################################################################################
+# now simplify the geometries to make overlapping polygons united
+tagged <- tagged |>
+  group_by(Valley) |>
+  summarize(geometry = st_union(geom)) |>
+  st_make_valid() |>
[…]

### 2025-10-25 LabelStyles/skeleton-default.Rmd
label changes - check diff on vignettes

--- 2025-09-20 LabelStyles/skeleton-research.Rmd
@@
 ```{r setup, include=FALSE}
+knitr::opts_chunk$set(dpi = 300) 
 knitr::opts_chunk$set(echo = FALSE)
-knitr::opts_chunk$set(dpi = 300) 
 knitr::opts_chunk$set(message = FALSE)
@@
 
-```{r load data, eval = T}
+```{r load data, eval = T, echo= F}
 library(BarnebyLives)
-local <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
+
+local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
+
 record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
+  sf::st_drop_geometry() |>
+  data.frame() |>
   dplyr::filter(Collection_number == params$Collection_number)
@@
 \end{center}
+\begin{flushright}
 \uppercase{`r record[['Family']]`}
+\end{flushright}
+\begingroup
+\large
 
-\begingroup
-\normalsize
-`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`\
 \endgroup
@@
 \begingroup
-\small
+\normalsize
+
 `r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
@@
 `r writer(record[['Habitat']])`. `r record[['physical_environ']]`
-Veg.: `r species_font(record[['Vegetation']])` 
-`r associates_writer(record[['Associates']], full = TRUE)`
-\endgroup
+
+Vegetation: `r species_font(record[['Vegetation']])`
+
+`r associates_writer(record[['Associates']], full = TRUE)` \endgroup
 
@@
 \begingroup
-`r collection_writer(record)`
-`r writer_fide(record)`
+
+`r collection_writer(record, date=FALSE)`
+
+\begin{flushright}
+`r record[['Date_digital_text']]`
+\end{flushright}
 \endgroup
-
 \begingroup
-\scriptsize
-`r writer(record[['Research_notes']])`  
+\small
+`r writer(record[['Notes']])`
 \endgroup

### 2025-10-25 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
label changes - check diff on vignettes

@@
 author:  |
-    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
     |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
     |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
+    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
 abstract: |
@@
 output: 
+  word_document: default
+  toc: no
   pdf_document: default
-  toc: no
-  word_document: default
 csl: "../citations/american-journal-of-botany.csl"

### 2025-11-20 LabelStyles/SoS-skeleton-Det.Rmd
resubmitted version

@@
 ```{r load data}
-data <- read.csv('SoS-ExampleCollection.csv') |>
+record <- read.csv('SoS-ExampleCollection.csv') |>
   dplyr::mutate(
@@
     Coordinate_uncertainty = '+/- 5m') |>
-  dplyr::filter(Collection_number == params$Collection_number)
+  dplyr::filter(Collection_number == params$Collection_number) |>
+  dplyr::mutate(Project_name = 'SOS Template')
 ```
@@
 \begin{center}
-\textbf{\LARGE Seeds of Success}
+\textbf{\LARGE `r record[['Project_name']]`}
 \end{center}
-\begin{flushright}
-\uppercase{`r data$Family`}
-\end{flushright}
+\vspace{-\baselineskip} 
+\uppercase{`r record[['Family']]`}  
+\vspace{-\baselineskip} 
 \begingroup
 \large
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
 \normalsize 
-`r writer_fide(data)`
+`r writer_fide(record)`
 \endgroup
 
-\vspace*{\fill}
+\vspace{\fill}
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
+\small
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
 
-`r writer(data$Habitat)`. `r data$physical_environ`
+`r writer(record$Habitat)`. `r record[['physical_environ']]`
 
-Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
-`r writer(data$Notes)`  
+Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Notes']])`  
 \endgroup
 
-\vspace*{\fill}
+\par\vspace{\fill}
+
 \begingroup
-`r collection_writer(data)`.
+\small
+`r collection_writer(record)`
 \endgroup
 
+\vspace{-\baselineskip} 
 \begin{center}

### 2025-11-20 LabelStyles/SoS-skeleton-Spaced.Rmd
resubmitted version

@@
 \begin{center}
-\textbf{\LARGE Seeds of Success}
+\textbf{\LARGE `r record[['Project_name']]`}
 \end{center}
+\vspace{-\baselineskip} 
 \begin{flushright}
-\uppercase{`r data$Family`}
+\uppercase{`r record[['Family']]`}
 \end{flushright}
+\vspace{-\baselineskip} 
 \begingroup
 \large
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
-\normalsize
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
 \endgroup
 
-\vspace*{\fill}
+\vspace{\fill}
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
+\small
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
 
-`r writer(data$Habitat)`. `r data$physical_environ`
+`r writer(record$Habitat)`. `r record[['physical_environ']]`
 
-Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
-`r writer(data$Notes)`  
+Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Notes']])`  
 \endgroup
 
-\vspace*{\fill}
+\par\vspace{\fill}
+
 \begingroup
-`r collection_writer(data)`. `r writer_fide(data)`
+\small
+`r collection_writer(record)`. `r writer_fide(record)`
 \endgroup
 
+\vspace{-\baselineskip} 
 \begin{center}

### 2025-11-20 LabelStyles/SoS-skeleton-familyLeft.Rmd
resubmitted version

@@
     Coordinate_uncertainty = '+/- 5m') |>
-  dplyr::filter(Collection_number == params$Collection_number)
+  dplyr::filter(Collection_number == params$Collection_number)  |>
+  dplyr::mutate(Project_name = 'SOS Template')
 ```
@@
 \begin{center}
-\textbf{\LARGE `r data$Project_name`}
+\textbf{\LARGE `r record[['Project_name']]`}
 \end{center}
+\vspace{-\baselineskip} 
+\uppercase{`r record[['Family']]`}  
+\vspace{-\baselineskip} 
+\begingroup
 \large
-\uppercase{`r data$Family`}  
-\begingroup
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
-\endgroup
-\normalsize
-
-\vspace*{\fill}
-\begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
-
-`r writer(data$Habitat)`. `r data$physical_environ`
-
-Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
-`r writer(data$Notes)`  
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
 \endgroup
 
-\vspace*{\fill}
+\vspace{\fill}
 \begingroup
-`r collection_writer(data)`. `r writer_fide(data)`
+\small
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
+
+`r writer(record$Habitat)`. `r record[['physical_environ']]`
+
+Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Notes']])`  
 \endgroup
 
+\par\vspace{\fill}
+
+\begingroup
+\small
+`r collection_writer(record)`. `r writer_fide(record)`
+\endgroup
+
+\vspace{-\baselineskip} 
 \begin{center}

### 2025-11-20 LabelStyles/SoS-skeleton-large.Rmd
resubmitted version

@@
 \begin{center}
-\textbf{\large Seeds of Success}
+\textbf{\large `r record[['Project_name']]`}
 \end{center}
+\vspace{-\baselineskip} 
 \begin{flushright}
-\uppercase{`r data$Family`}
+\uppercase{`r record[['Family']]`}
 \end{flushright}
+\vspace{-\baselineskip} 
 \begingroup
 \large
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
-\normalsize
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
 \endgroup
 
-\vspace*{\fill}
+\vspace{\fill}
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
+\small
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
 
-`r writer(data$Habitat)`. `r data$physical_environ`
+`r writer(record$Habitat)`. `r record[['physical_environ']]`
 
-Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
-`r writer(data$Notes)`  
+Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Notes']])`  
 \endgroup
 
-\vspace*{\fill}
+\par\vspace{\fill}
+
 \begingroup
-`r collection_writer(data)`. `r writer_fide(data)`
+\small
+`r collection_writer(record)`. `r writer_fide(record)`
 \endgroup
 
+\vspace{-\baselineskip} 
 \begin{center}

### 2025-11-20 LabelStyles/SoS-skeleton-noDir.Rmd
resubmitted version

@@
 \begin{center}
-\textbf{\LARGE Seeds of Success}
+\textbf{\LARGE `r record[['Project_name']]`}
 \end{center}
+\vspace{-\baselineskip} 
 \begin{flushright}
-\uppercase{`r data$Family`}
+\uppercase{`r record[['Family']]`}
 \end{flushright}
+\vspace{-\baselineskip} 
 \begingroup
 \large
-`r writer(paste(data$Genus, data$Epithet), italics = TRUE)` `r writer(data$Binomial_authority)` `r writer(data$Infrarank)` `r  writer(data$Infraspecies, italics = TRUE)` `r writer(data$Infraspecific_authority)`
-\normalsize
+`r writer(paste(record[['Genus']], record[['Epithet']]), italics = TRUE)` `r writer(record[['Binomial_authority']])` `r writer(record[['Infrarank']])` `r writer(record[['Infraspecies']], italics = TRUE)` `r writer(record[['Infraspecific_authority']])`
 \endgroup
 
-\vspace*{\fill}
+\vspace{\fill}
 \begingroup
-`r data$Gen`. `r writer(data$Site)` `r data$latitude_dd` `r data$longitude_dd` (`r data$Datum` `r writer(data$Coordinate_uncertainty)`).
+\small
+`r record[['Gen']]`. `r writer(record[['Site']])` `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).
 
-`r writer(data$Habitat)`. `r data$physical_environ`  
-Veg.: `r species_font(data$Vegetation)` `r associates_writer(data$Associates)`
-`r writer(data$Notes)`  
+`r writer(record$Habitat)`. `r record[['physical_environ']]`  
+Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+`r writer(record[['Notes']])`  
 \endgroup
 
-\vspace*{\fill}
+\par\vspace{\fill}
+
 \begingroup
-`r collection_writer(data)`. `r writer_fide(data)`
+\small
+`r collection_writer(record)`. `r writer_fide(record)`
 \endgroup
 
+\vspace{-\baselineskip} 
 \begin{center}

### 2025-11-20 LabelStyles/creating_example_labels.Rmd
resubmitted version

---
title: "Printing herbarium labels and exporting a digital copy of data"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Printing herbarium labels and exporting a digital copy of data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
description: >
  Rendering herbarium labels
---

```{r load libraries and import data}
# remotes::install_github('sagesteppe/BarnebyLives') 
# OR devtools::install_github('sagesteppe/BarnebyLives')
library(BarnebyLives)
library(dplyr)
library(glue)
data('collection_examples')
```

## creating a driver script to feed the skeleton data

```{r}
p <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
driver <- function(x, name){
  purrr::walk(
    .x = 2823,
  ~ rmarkdown::render(
    input = x,                      # template to populate
    output_file = file.path(p, glue("{name}{.x}.pdf")),# write one PDF per record
    params = list(Collection_number = {.x})      # pass record to skeleton
    )
  )
}
```

### BUILT IN TEMPLATES

```{r, eval = F}
driver("skeleton-default.Rmd", name = 'default')

driver("skeleton-research.Rmd", name = 'research')
driver("skeleton-modern.Rmd", name = 'modern')
driver("skeleton-SOS.Rmd", name = 'SOS')
driver("skeleton-customfont.Rmd", name = 'customfont')
```

### LABELS WITH GRAPHICS

```{r}
#driver("skeleton-maps.Rmd", name = 'map')
#driver("skeleton-default.Rmd", name = 'default-')
```

## customizing a label template

```{r customizing label templates, eval = F}
driver("skeleton-SOS.Rmd", name = 'default-SOS-')
driver("SoS-skeleton-noDir.Rmd", name = 'noDir-SOS-')
driver("SoS-skeleton-Spaced.Rmd", name = 'Spaced-SOS-')
driver("SoS-skeleton-large.Rmd", name = 'large-SOS-')
driver("SoS-skeleton-familyLeft.Rmd", name = 'familyLeft-SOS-')
driver("SoS-skeleton-Det.Rmd", name = 'Det-SOS-')
```

## customizing label fonts

(completed)

```{r, eval = F}
driver("skeleton-customfont.Rmd", name = 'custom-font')
driver("skeleton-customfont-twofonts.Rmd", name = 'custom-twofonts')
driver("skeleton-customfont-downloaded.Rmd", name = 'custom-font-downloaded')
```

### 2025-11-20 LabelStyles/skeleton-SOS.Rmd
resubmitted version

--- 2025-11-20 LabelStyles/SoS-skeleton-noDir.Rmd
@@
     Collection_number = as.numeric(Collection_number), 
-    Coordinate_uncertainty = '+/- 5m') |>
+    Coordinate_Uncertainty = '+/- 5m') |>
   dplyr::filter(Collection_number == params$Collection_number) |>
@@
 \endgroup
+\begingroup
+\begin{spacing}{0.02}
+{\scriptsize `r record[['Directions_BL']]`}
+\end{spacing}
+\endgroup
 

### 2025-11-20 LabelStyles/skeleton-customfont-downloaded.Rmd
resubmitted version

--- 2025-10-12 LabelStyles/skeleton-logo.Rmd
@@
 params:
-  Collection_number: NULL
+  Collection_number: Collection_number
 geometry: paperheight=4in,paperwidth=4in,margin=0.1in
@@
   - \usepackage{setspace}
-  - \usepackage{wrapfig}
+  - \usepackage{fontspec}
+  - |
+    \setmainfont[
+      Path = /home/sagesteppe/Documents/assoRted/Barneby_Lives_dev/LabelStyles/Nunito/static/,
+      UprightFont = *-Regular,
+      ItalicFont = *-Italic,
+      BoldFont = *-Bold,
+      BoldItalicFont = *-BoldItalic
+    ]{Nunito}
 ---
@@
 
-```{r load library, eval = T}
+```{r load library}
 library(BarnebyLives)
+```
 
-local <- file.path(path.expand('~'), 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
-logopath <- file.path(local, 'logo.png')
+```{r load data}
+local <- file.path('~', 'Documents', 'assoRted', 'Barneby_Lives_dev', 'LabelStyles')
 
-record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |> # default feed in the path to your csv. 
+record <- read.csv(file.path(local, 'SoS-ExampleCollection.csv')) |>
   dplyr::filter(Collection_number == params$Collection_number)  |>
-  sf::st_drop_geometry() |>
-  dplyr::mutate(Project_name = 'Logos', Coordinate_uncertainty = ' +/- 5m')
-
-# escape character columns to ensure characters which interfere with latex
-# rendering are escaped. 
-char_cols <- sapply(record, is.character)
-char_col_names <- names(record)[char_cols]
-record[char_col_names] <- lapply(record[char_col_names], escape_latex)
+  dplyr::mutate(Project_name = 'Custom font downloaded')
 ```
@@
 \vspace{\fill}
-
 \begingroup
@@
 `r writer(record[['Habitat']])`. `r record[['physical_environ']]` \par
-Veg.: `r species_font(record[['Vegetation']])` `r associates_writer(record[['Associates']])`
+Vegetation: `r species_font(record[['Vegetation']])` \par
+`r associates_writer(record[['Associates']])`
 `r writer(record[['Notes']])`  
@@
 \vspace{\fill}
-
-\noindent
-\begin{minipage}[t]{0.75\textwidth}
 \begingroup
-  `r collection_writer(record)`
+`r collection_writer(record)` \par
 `r writer_fide(record)`
 \endgroup
-\end{minipage}
-\hfill
-\begin{minipage}[b]{0.2\textwidth}
-  \includegraphics[width=\linewidth,keepaspectratio]{`r logopath`}
-\end{minipage}

### 2025-11-20 LabelStyles/skeleton-customfont-twofonts.Rmd
resubmitted version

@@
   data.frame() |>
-  dplyr::filter(Collection_number == params$Collection_number) 
+  dplyr::filter(Collection_number == params$Collection_number)  |>
+  dplyr::mutate(Project_name = 'Custom font - two of them')
 ```

### 2025-11-20 LabelStyles/skeleton-customfont.Rmd
resubmitted version

@@
   data.frame() |>
-  dplyr::filter(Collection_number == params$Collection_number) 
+  dplyr::filter(Collection_number == params$Collection_number)  |>
+  dplyr::mutate(Project_name = 'Custom font Skeleton')
 ```

### 2025-11-20 LabelStyles/skeleton-default.Rmd
resubmitted version

@@
   data.frame() |>
-  dplyr::filter(Collection_number == params$Collection_number)
+  dplyr::filter(Collection_number == params$Collection_number) |>
+  dplyr::mutate(Project_name = 'Default Template')
 

### 2025-11-20 LabelStyles/skeleton-modern.Rmd
resubmitted version

@@
 \vspace{\fill}
+
 \begingroup
 `r record[['Gen']]`. `r record[['latitude_dd']]` `r record[['longitude_dd']]` (`r record[['Datum']]` `r writer(record[['Coordinate_uncertainty']])`).  `r writer(record[['Site']])` 
+\endgroup
 
+\vspace{\fill}
+
+\begingroup
 `r writer(record[['Habitat']])`. `r record[['physical_environ']]` 
-Vegetation: `r species_font(record[['Vegetation']])` 
-`r associates_writer(record[['Associates']], full = TRUE)` 
+\endgroup
+
+\vspace{\fill}
+
+\begingroup
+Vegetation: `r species_font(record[['Vegetation']])` \newline 
+`r associates_writer(record[['Associates']], full = TRUE)` \newline 
 `r writer(record[['Notes']])`  
@@
 \vspace{\fill}
+
 \begingroup

### 2025-11-20 manuscript/Appendices.Rmd
resubmitted version

--- 2025-07-19 manuscript/ComputerSpecsAppendix.Rmd
@@
-# Appendix A: System Information for Benchmarking
+---
+output:
+  pdf_document: default
+---
+# Appendix 1: System Information for Benchmarking
 
@@
 
-Architecture:             x86_64  
-  CPU op-mode(s):         32-bit, 64-bit  
-  Address sizes:          39 bits physical, 48 bits virtual  
-  Byte Order:             Little Endian  
-CPU(s):                   16  
-  On-line CPU(s) list:    0-15  
-Vendor ID:                GenuineIntel  
-  Model name:             11th Gen Intel(R) Core(TM) i7-11700K @ 3.60GHz  
-    CPU family:           6  
-    Model:                167  
-    Thread(s) per core:   2  
-    Core(s) per socket:   8  
-    Socket(s):            1  
-    Stepping:             1  
-    CPU max MHz:          5100.0000  
-    CPU min MHz:          800.0000  
-    BogoMIPS:             7200.00
+| Field               | Value                                          |
+| ------------------- | ---------------------------------------------- |
+| Architecture        | x86_64                                         |
+| CPU op-mode(s)      | 32-bit, 64-bit                                 |
+| Address sizes       | 39 bits physical, 48 bits virtual              |
+| Byte Order          | Little Endian                                  |
+| CPU(s)              | 16                                             |
+| On-line CPU(s) list | 0-15                                           |
+| Vendor ID           | GenuineIntel                                   |
+| Model name          | 11th Gen Intel(R) Core(TM) i7-11700K @ 3.60GHz |
+| CPU family          | 6                                              |
+| Model               | 167                                            |
+| Thread(s) per core  | 2                                              |
+| Core(s) per socket  | 8                                              |
+| Socket(s)           | 1                                              |
+| Stepping            | 1                                              |
+| CPU max MHz         | 5100.0000                                      |
+| CPU min MHz         | 800.0000                                       |
+| BogoMIPS            | 7200.00                                        |
+
+\newpage
 
@@
 ```
-               total        used        free      shared  buff/cache   available  
-Mem:           125Gi       3.1Gi       112Gi       569Mi       9.7Gi       120Gi  
-Swap:          2.0Gi          0B       2.0Gi
+
+| Type | Total | Used  | Free  | Shared | Buff/Cache | Available |
+| ---- | ----- | ----- | ----- | ------ | ---------- | --------- |
+| Mem  | 125Gi | 3.1Gi | 112Gi | 569Mi  | 9.7Gi      | 120Gi     |
+| Swap | 2.0Gi | 0B    | 2.0Gi | —      | —          | —         |
 
@@
 
-Filesystem      Size  Used Avail Use% Mounted on  
-tmpfs            13G  2.6M   13G   1% /run  
-/dev/nvme0n1p2  468G  242G  203G  55% /  
-tmpfs            63G   14M   63G   1% /dev/shm  
-tmpfs           5.0M  4.0K  5.0M   1% /run/lock  
-efivarfs        192K   88K  100K  47% /sys/firmware/efi/efivars  
-/dev/nvme0n1p1  511M  6.1M  505M   2% /boot/efi  
-tmpfs            13G  132K   13G   1% /run/user/1000  
-/dev/sda1       3.6T  3.2T  295G  92% /media/steppe/hdd
+| Filesystem     | Size | Used | Avail | Use% | Mounted on                |
+| -------------- | ---- | ---- | ----- | ---- | ------------------------- |
+| tmpfs          | 13G  | 2.6M | 13G   | 1%   | /run                      |
+| /dev/nvme0n1p2 | 468G | 242G | 203G  | 55%  | /                         |
+| tmpfs          | 63G  | 14M  | 63G   | 1%   | /dev/shm                  |
+| tmpfs          | 5.0M | 4.0K | 5.0M  | 1%   | /run/lock                 |
+| efivarfs       | 192K | 88K  | 100K  | 47%  | /sys/firmware/efi/efivars |
+| /dev/nvme0n1p1 | 511M | 6.1M | 505M  | 2%   | /boot/efi                 |
+| tmpfs          | 13G  | 132K | 13G   | 1%   | /run/user/1000            |
+| /dev/sda1      | 3.6T | 3.2T | 295G  | 92%  | /media/steppe/hdd         |
 
@@
 Timing buffered disk reads: 514 MB in  3.01 seconds = 170.57 MB/sec
+
+\newpage
+
+# Appendix 2: Label Templates
+
+\begin{center}
+\includegraphics[width=420pt]{./labels/TemplateExamples.png}
+\end{center}

### 2025-11-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
resubmitted version

--- 2025-09-20 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
@@
 While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
-Unfortunately, we have not found Windows alternatives to pdfjam and pdftk, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.
+However, installation of the utilities on the three supported Operating systems (Debian Linux, MacOS, and Windows 10 and 11)  will require some minimal use of the command line to install software which are required for the label-generating functionality as detailed on the README page.
 
@@
 BarnebyLives can be thought of as consisting of five main modules (Figure 1): spatial, taxonomic, formatting, manual review, and data exporting. 
+Core functions are detailed below, while all functions are documented on reference page https://sagesteppe.github.io/BarnebyLives/reference/index.html.
 
@@
 Beyond always having the data collected by BarnebyLives terminating in a label for a physical specimen, its utility may be extended to providing information for various biotic databases (e.g., heritage program databases) or modelling applications (e.g., modelling in community ecology) as a user sees fit.
-By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
+By utilizing both R and LaTeX and having publicly available source code on GitHub, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
 LaTeX, a software system used for typesetting, was developed to allow users to focus on the content rather than the style of the rendered documents. 
@@
 Accessioning often relies on the use of the Microsoft Office suite of programs and may utilize other costly software such as ArcPro or Adobe Acrobat. 
-While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
+While BarnebyLives does not have its own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as RStudio, Positron, VisualStudio (VS) Code, now offer functionality to readily view and filter datasets using familiar spreadsheet-like formats, making them more accessible to many users.
 While other software packages often cost money, these are also free, and we recommend that users install an open-source PDF viewer such as Okular to review their rendered documents.
@@
 It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
-The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Mac and Linux, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
+The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), requiring only a small amount of use of the command line, which may be called from RStudio or Positron rather than a 'traditional' terminal.
+Further while the software is intended for use in the context of producing herbarium labels, it's basic functionality to resolve taxonomic names, and query environmental attributes may make it useful for various land management agencies which require these utilities for various research activities.
 
@@
 The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
-The package includes three real use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
+The package includes seven use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
 The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain).
-The next two vignettes *"BarnebyLives!
-Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats.
-*"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium.
+The next vignette *"BarnebyLives! Running pipeline"* showcases the core functionality of the package for processing data entered on a spreadsheet. 
+*"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats while *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium.
+The remaining vignettes detail different aspects of label usage, such as including maps and logos in *"Labels with graphics"*, and customization for example the position of template elements *"Customizing a label template"*, and fonts in *"Customizing label fonts"*.
+
 All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.
